### 2016-05-03 DCMIP2016-Part1-Models.tex
Initial revision

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package which can be downloaded from the different journal webpages.
%% For further assistance please contact the Copernicus Publications at: publications@copernicus.org
%% http://publications.copernicus.org

%% Please use the following documentclass and Journal Abbreviations for Discussion Papers and Final Revised Papers.

%% 2-Column Papers and Discussion Papers
\documentclass[gmd, manuscript]{copernicus}

%% Journal Abbreviations (Please use the same for Discussion Papers and Final Revised Papers)

% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Geothermal Energy Science (gtes)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

% Custom commands
\newcommand{\vb}{\mathbf}
\newcommand{\vg}{\boldsymbol}
\newcommand{\mat}{\mathsf}
\newcommand{\diff}[2]{\frac{d #1}{d #2}}
\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}

\begin{document}

\title{DCMIP2016, Part 1: Models and Equation Sets}

% \Author[affil]{given_name}{surname}

\Author[1]{Paul A.}{Ullrich}
\Author[2]{Christiane}{Jablonowski}
\Author[3]{James}{Kent}
\Author[4]{Peter}{Lauritzen}
\Author[4]{Ramachandran}{Nair}
\Author[5]{Kevin A.}{Reed}
\Author[4]{Colin}{Zarzycki}

\Author[6]{Thomas}{Dubos}
\Author[7]{Marco}{Giorgetta}
\Author[8]{Elijah}{Goodfriend}
\Author[9]{David A.}{Hall}
\Author[10]{Lucas}{Harris}
\Author[8]{Hans}{Johansen}
\Author[11]{Christian}{Kuehnlein}
\Author[12]{Vivian}{Lee}
\Author[13]{Thomas}{Melvin}
\Author[14]{Hiroaki}{Miura}
\Author[15]{David}{Randall}
\Author[16]{Alex}{Reinecke}
\Author[4]{William}{Skamarock}
\Author[16]{Kevin}{Viner}
\Author[17]{Robert}{Walko}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
\affil[7]{Max Planck Institute for Meteorology}
\affil[8]{Lawrence Berkeley National Laboratory}
\affil[9]{University of Colorado, Boulder}
\affil[10]{Geophysical Fluid Dynamics Laboratory}
\affil[11]{European Center for Medium-Range Weather Forecasting}
\affil[12]{Environment Canada}
\affil[13]{U.K. Met Office}
\affil[14]{University of Tokyo}
\affil[15]{Colorado State University}
\affil[16]{Naval Research Laboratory}
\affil[17]{University of Miami}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

\runningtitle{DCMIP2016: Models and Equation Sets}

\runningauthor{Ullrich, et al.}

\correspondence{Paul A. Ullrich (paullrich@ucdavis.edu)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
This paper provides a comprehensive review of the design of modern non-hydrostatic atmospheric dynamical cores, including relevant equation sets, numerical stabilization techniques and idealized physics routines.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]

{\color{red}INSTRUCTIONS FOR AUTHORS}

{\color{red} Fill in text in section 3, 4, 5, 6, 7 and 8 below.}

\section{Notation}

\subsection{List of Symbols}

Table \ref{tab:symbols} lists the symbols used in this paper.

\begin{table}[h]
\caption{List of symbols used in this manuscript} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$\delta$ & Divergence\\
$\zeta$ & Relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\Theta$ & Potential temperature \\
$\Theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

\subsection{List of Physical Constants}
A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.  Constants which are specific to each test case are similarly tabulated at the beginning of each section.

\begin{table}[h]
\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
%\ \\
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
\hline Constant & Description & Value \\
\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
$c_p$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$c_v$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
$R_\nu$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
$\varepsilon$ & Ratio of $R_d$ to $R_\nu$ & $0.622$ \\
$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
\hline 
\end{tabular*}

\end{table}

\subsection{Great Circle Distance}

The great circle distance is used throughout the document and is given by
\begin{equation}
R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Model Grids} \label{sec:ModelGrids}

{\color{red}[ALL] Add a short description of your model grid here.}

\subsection{Latitude-longitude grid}

\subsection{Cubed-sphere grid}

The equiangular cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $i_p \in \{1,2,3,4,5,6\}$.  By convention, we choose panels 1--4 to be along the equator and panels $5$ and $6$ to be centered on the northern and southern pole, respectively.

\subsection{Icosahedral grid}

\subsection{Centroidal Voronoi tessellation grid}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Equation Sets} \label{sec:EquationSets}

{\color{red}[ALL] Include the continuous equation set that you use for your model here.}

\subsection{Tempest} \label{sec:TempestEquations}

The continuity, momentum and thermodynamic equations can be written as:
\begin{align}
\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
\pdiff{\vb{u}}{t} &= - \nabla (K + \Phi) - \theta \nabla \Pi + \vg{\eta} \times \vb{u}, \\
\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v,
\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, geopotential $\Phi = g_c z$ and absolute vorticity $\vg{\eta} = \vg{\zeta} + \vg{\Omega}$, which consists of relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$ and planetary vorticity $\vg{\Omega}$.  The Exner pressure is related to the prognosed density and potential temperature via
\begin{equation} \label{eq:BaseNonhydrostaticEOS}
\Pi = c_p \left( \frac{p_0}{p} \right)^{R_d/c_p} = c_p \left( \frac{R_d \rho \theta_v}{p_0} \right)^{R_d/c_v}.
\end{equation}

\subsection{Tracer transport}

\noindent Lagrangian form:
\begin{align}
\diff{q}{t} &= 0.
\end{align}

\noindent Non-conservative Eulerian form:
\begin{align}
\pdiff{q}{t} &= - \vb{u} \cdot \nabla q.
\end{align}

\noindent Flux form:
\begin{align}
\pdiff{}{t} (\rho q) &= - \nabla \cdot (\rho q \vb{u}).
\end{align}

\subsection{Height-based coordinates}

{\color{blue}Define Unstaggered, Lorenz and Charney-Phillips staggering here}

\subsection{Mass-based coordinates}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Diffusion and Stabilization} \label{sec:DiffusionStabilization}

{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}

\subsection{Scalar viscosity}

\noindent Scalar viscosity (direct):
\begin{align} \label{eq:scalar_viscosity_direct}
\diff{s}{t} &= \ldots + \nu \nabla \cdot \nabla s.
\end{align}

\noindent Scalar viscosity (conservative):
\begin{align} \label{eq:scalar_viscosity_conservative}
\diff{}{t} (\rho q) &= \ldots + \nu \nabla \cdot ( \rho \nabla q ).
\end{align}

\subsection{Smagorinsky eddy viscosity}

\subsection{Vector viscosity, divergence and vorticity damping}

\noindent Vector viscosity:
\begin{align} \label{eq:vector_viscosity}
\diff{\vb{u}}{t} &= \ldots + \nu \nabla^2 \vb{u}
\end{align}

\noindent Divergence damping:
\begin{align} \label{eq:divergence_damping}
\diff{\vb{u}}{t} &= \ldots + \nu_{div} \nabla (\nabla \cdot \vb{u})
\end{align}

\noindent Vorticity damping:
\begin{align} \label{eq:vorticity_damping}
\diff{\vb{u}}{t} &= \ldots + \nu_{vort} \nabla \times (\nabla \times \vb{u})
\end{align}

\subsection{Hyperviscosity} \label{sec:Diffusion_Hyperviscosity}

{\color{blue}Repeated application of the scalar and vector viscosity operators}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Filters and Fixers} \label{sec:FiltersFixers}

{\color{red}[ALL] Include explicit filters and fixers that you have utilized in the dynamical core here.}

\subsection{Mass borrowing (positive definite preservation)}

\subsection{Mass fixers}

\subsection{Energy fixers}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Temporal Discretizations} \label{sec:TemporalDiscretizations}

{\color{red}[ALL] Describe the time-stepping scheme / temporal discretization employed by your dynamical core here.}

\subsection{Runge-Kutta}

\subsubsection{Ullrich-Kinnmark-Gray 5 step 3rd order scheme} \label{sec:UKG53Scheme}

Explicit terms are evolved using a Runge-Kutta method which supports a large stability bound for spatial discretizations with purely imaginary eigenvalues. This particular scheme is based on \cite{kinnmark1984onestepA, kinnmark1984onestepB} and takes the form
\begin{align}
\psi^{(1)} &= \psi^{(0)} + \tfrac{\Delta t}{5} f(\psi^{(0)}), \nonumber \\
\psi^{(2)} &= \psi^{(0)} + \tfrac{\Delta t}{5} f(\psi^{(1)}), \nonumber \\
\psi^{(3)} &= \psi^{(0)} + \tfrac{\Delta t}{3} f(\psi^{(2)}), \\
\psi^{(4)} &= \psi^{(0)} + \tfrac{2 \Delta t}{3} f(\psi^{(3)}), \nonumber \\
\psi^{(5)} &= -\tfrac{1}{4} \psi^{(0)} + \tfrac{5}{4} \psi^{(1)} + \tfrac{3 \Delta t}{4} f(\psi^{(4)}). \nonumber
\end{align}

\subsection{Semi-Implicit time integration}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Dynamical Cores}

{\color{red}In this section provide a short description (approximately 0.5 pages) of the dynamical core, focusing on unique features or design specifications.  Do not include information on the physical parameterizations used by the modeling system.  Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}

\subsection{Tempest}

{\color{red}[ULLRICH]}

The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.  The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.  These continuous equations are given in section \ref{sec:TempestEquations}.  The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.  Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).  Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.

\subsection{High-Order Method Modeling Environment (HOMME)}

{\color{red}[HALL]}

\subsection{Model for Prediction Across Scales (MPAS)}

{\color{red}[SKAMAROCK]}

\subsection{Colorado State University Model (CSU)}

{\color{red}[RANDALL]}

\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV3)}

{\color{red}[HARRIS]}

\subsection{Chombo}

{\color{red}[JOHANSEN]}

\subsection{Naval Research Laboratory NEPTUNE Model}

{\color{red}[VINER, REINECKE]}

\subsection{Global Environmental Multiscale (GEM) Model}

{\color{red}[LEE]}

\subsection{Ocean-Land-Atmosphere Model (OLAM)}

{\color{red}[WALKO]}

\subsection{DYNAMICO}

{\color{red}[DUBOS]}

\subsection{ECMWF PentaRei Finite Volume Model}

{\color{red}[KUEHNLEIN]}

\subsection{Icosahedral Non-hydrostatic (ICON) Model}

{\color{red}[GIORGETTA]}

\subsection{Nonhydrostatic ICosahedral Atmospheric Model (NICAM) Model}

{\color{red}[MIURA]}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Idealized Physical Parameterizations}

\subsection{Kessler Physics} \label{sec:KesslerPhysics}

The cloud microphysics update according to the following equation set:
\begin{alignat}{5}
\frac{\Delta \theta}{\Delta t} = & - \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big) & \\
\frac{\Delta q_v}{\Delta t} = & & \frac{\Delta q_{vs}}{\Delta t} & + E_r \\
\frac{\Delta q_c}{\Delta t} = & & - \frac{\Delta q_{vs}}{\Delta t} & & - A_r & - C_r \\
\frac{\Delta q_r}{\Delta t} = & & & - E_r & + A_r & + C_r & - V_r \pdiff{q_r}{z},
\end{alignat} where $L$ is the latent heat of condensation, $A_r$ is the autoconversion rate of cloud water to rain water, $C_r$ is the collection rate of rain water, $E_r$ is the rain water evaporation rate, and $V_r$ is the rain water terminal velocity.

The pressure follows from the equation of state
\begin{equation}
p=\rho R_dT(1+0.61q_v)
\end{equation} with $p$ the pressure, $\rho$ the density of moist air, $R_d$ the gas constant for dry air, $T$ the temperature and $q_v$ the mixing ratio of water vapor. The equation is rewritten as a nondimensional pressure $\Pi$ equation.
\begin{equation}
\pi = \left(\frac{p}{p_0}\right)^{\frac{R_dT}{cp}}
\end{equation}

To determine the saturation vapor mixing ratio the Teten's formula is used,
\begin{equation}
q_{vs}(p,T) = \left( \frac{380.0}{p} \right) \exp\left(17.27 \times \frac{T-273.0}{T-36.0}\right)
\end {equation}

The autoconvection rate ($A_r$) and collection rate ($C_r$) follow Kessler parametrization and are defined by:
\begin{align}
A_r &= k_1(q_c-a) \\
C_r &= k_2q_cq_r^{0.875}
\end{align} With $k_1=0.001 \text{s}^{-1}$, $a=0.001 \text{g}.\text{g}^{-1}$ and $k_2=2.2 \text{s}^{-1}$

Deriving from \cite{klemp1978simulation} description of cloud water,rain water and water vapor mixing ratios. they are define as followed:
\begin{equation}
q_c^{n+1}=\mbox{max}(q_c^r-\Delta q_r,0)
\end{equation}
\begin{equation}
q_r^{n+1}=\mbox{max}(q_r^r-\Delta q_r+S,0)
\end{equation} where $S$ is the sedimentation term and $\Delta q_r$ is defined as
\begin{equation}
\Delta q_r=q_c^n-\frac{q_c^n-\Delta \text{t}\ \mbox{max}(A_r,0)}{1+\Delta \text{t} C_r}
\end{equation}

The Rain evaporation equation is defined similarly to \cite{ogura1971numerical} description:
\begin{equation}
E_r=\frac{1}{\rho}\frac{\left(1-\frac{q_v}{q_{vs}}\right)C(\rho q_r)^{0.525}}{5.4\times10^5+\frac{2.55\times10^6}{pq_{vs}}}
\end{equation}  With ventilation factor C define as 
\begin{equation}
C_r=1.6+124.9(\rho q_r)^{0.2046}
\label{venti}
\end{equation}

The liquid water terminal velocity is similar to \cite{soong1973comparison} description with a mean density adjustment as suggested by \cite{kessler1969distribution}:
\begin{equation}
V_r = 36349(\rho q_r)^{0.1346}\left(\frac{\rho}{\rho_0}\right)^{-\frac{1}{2}}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Simplified Mixing in the Planetary Boundary Layer} \label{sec:PlanetaryBoundaryLayer}

The forcing by the planetary boundary layer is described in \cite{reed2012idealized} and is partly reproduced here.  To parameterize the surface fluxes that impact the zonal velocity $u$, the meridional velocity $v$ and moisture $q$ we start with the time rate of change equations
\begin{eqnarray}
\frac{\partial u}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'u'}}{\partial z} \label{eqturb}  \\
\frac{\partial v}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'v'}}{\partial z}  \label{eqturb_v} \\
\frac{\partial q}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'q'}}{\partial z}.  \label{eqturb_last}
\end{eqnarray}  Potential temperature, as opposed to temperature, is used in the boundary layer parameterization because the vertical profile of the potential temperature is a suitable indicator of static stability.  This adds the time rate of change equation
\begin{eqnarray}
\frac{\partial \Theta}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'\Theta'}}{\partial z}.
\end{eqnarray}  Here $u'$, $v'$, $w'$, $\Theta'$ and $q'$ symbolize the deviations of the zonal velocity, meridional velocity, vertical velocity, potential temperature and specific humidity from their averages, respectively. The average is indicated by an overbar.  Note, assuming pressure is held constant (which is a common assumption in physical parameterizations), the potential temperature time tendency can be converted back to a temperature tendency of the following form
\begin{eqnarray}
\frac{\partial T}{\partial t} &=& - \frac{1}{\rho} \left (\frac{p}{p_0} \right )^{\kappa} \frac{\partial \rho \ \overline{w'\Theta'}}{\partial z}.
\end{eqnarray}
with the reference pressure $p_0 = 1000$ hPa.

The turbulent mixing is characterized by a constant vertical eddy diffusivity to represent Ekman-like profiles of boundary layers
\begin{eqnarray}\label{ktheory}
\overline{w'u'} = -K_m \frac{\partial u}{\partial z} \\
\overline{w'v'} = -K_m \frac{\partial v}{\partial z} \\
\overline{w'\Theta'} = -K_E \frac{\partial \Theta}{\partial z} \\
\overline{w'q'} = -K_E \frac{\partial q}{\partial z}. 
\end{eqnarray}
Here, $K_m$ is the eddy diffusivity coefficient for momentum and $K_E$ is the eddy diffusivity coefficient for energy and set equal to that for water vapor. In order to calculate the eddy diffusivity coefficients, the eddy diffusivity is matched to that for the surface flux calculated in Appendix~\ref{sec:OceanSurfaceFluxes} at the lowermost model level.  To allow for a smooth transition above the boundary layer ($p_{top} = 850$ hPa)  the diffusivity coefficients for momentum taper to zero as
\begin{equation} \label{Kmtaper}
\begin{array}{ll}
K_m = C_d \vert \vec{v}_{a} \vert z_a & \mbox{for} \; p > p_{top} \\
K_m = C_d \vert \vec{v}_{a} \vert z_a \exp \left ( -\left [ \frac{p_{top} - p}{p_{strato}} \right ]^2 \right ) & \mbox{for} \; p \leq p_{top}.
\end{array}
\end{equation}
Here the constant $p_{strato}$ determines the rate of decrease and is set to 100 hPa. $K_E$ is defined by
\begin{equation}
\begin{array}{ll}\label{KEtaper}
K_E = C_E \vert \vec{v}_{a} \vert z_a & \mbox{for} \; p > p_{top} \\
K_E = C_E \vert \vec{v}_{a} \vert z_a \exp \left ( -\left [ \frac{p_{top} - p}{p_{strato}} \right ]^2 \right ) & \mbox{for} \; p \leq p_{top}.
\end{array}
\end{equation}
We suggest implementing the boundary layer scheme with an implicit temporal discretization to avoid numerical instabilities. The details of this discretization are somewhat complicated, and so we refer to implementation details in Appendix D of \cite{reed2012idealized}. In addition, we supply the DCMIP modeling groups with the complete ``simple-physics'' package as used in the model CAM which can serve as a template routine.

\conclusions  %% \conclusions[modified heading if necessary]
TEXT

%\appendix
%\section{}    %% Appendix A

%\subsection{}                               %% Appendix A1, A2, etc.

\authorcontribution{TEXT}

\begin{acknowledgements}
{\color{blue}[Include a complete list of DCMIP2016 student participants here along with sponsors]}
\end{acknowledgements}

%% REFERENCES

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

%% FIGURES

%% ONE-COLUMN FIGURES

%%f
%\begin{figure}[t]
%\includegraphics[width=8.3cm]{FILE NAME}
%\caption{TEXT}
%\end{figure}
%
%%% TWO-COLUMN FIGURES
%
%%f
%\begin{figure*}[t]
%\includegraphics[width=12cm]{FILE NAME}
%\caption{TEXT}
%\end{figure*}
%
%
%%% TABLES
%%%
%%% The different columns must be seperated with a & command and should
%%% end with \\ to identify the column brake.
%
%%% ONE-COLUMN TABLE
%
%%t
%\begin{table}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table}
%
%%% TWO-COLUMN TABLE
%
%%t
%\begin{table*}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table*}
%
%
%%% NUMBERING OF FIGURES AND TABLES
%%%
%%% If figures and tables must be numbered 1a, 1b, etc. the following command
%%% should be inserted before the begin{} command.
%
%\addtocounter{figure}{-1}\renewcommand{\thefigure}{\arabic{figure}a}
%
%
%%% MATHEMATICAL EXPRESSIONS
%
%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
%%%
%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
%%% Vectors are identified in bold italic font using \vec{x}
%%% Matrices are identified in bold roman font
%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
%%% The character * should not be applied as mutliplication sign
%
%
%%% EQUATIONS
%
%%% Single-row equation
%
%\begin{equation}
%
%\end{equation}
%
%%% Multiline equation
%
%\begin{align}
%& 3 + 5 = 8\\
%& 3 + 5 = 8\\
%& 3 + 5 = 8
%\end{align}
%
%
%%% MATRICES
%
%\begin{matrix}
%x & y & z\\
%x & y & z\\
%x & y & z\\
%\end{matrix}
%
%
%%% ALGORITHM
%
%\begin{algorithm}
%\caption{…}
%\label{a1}
%\begin{algorithmic}
%…
%\end{algorithmic}
%\end{algorithm}
%
%
%%% CHEMICAL FORMULAS AND REACTIONS
%
%%% For formulas embedded in the text, please use \chem{}
%
%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
%
%\begin{reaction}
%%% \rightarrow should be used for normal (one-way) chemical reactions
%%% \rightleftharpoons should be used for equilibria
%%% \leftrightarrow should be used for resonance structures
%\end{reaction}
%
%
%%% PHYSICAL UNITS
%%%
%%% Please use \unit{} and apply the exponential notation

\end{document}

### 2016-08-18 DCMIP2016-Part1-Models.tex
Corrected authors and acknowledgements

@@
 \Author[3]{James}{Kent}
-\Author[4]{Peter}{Lauritzen}
+\Author[4]{Peter H.}{Lauritzen}
 \Author[4]{Ramachandran}{Nair}
 \Author[5]{Kevin A.}{Reed}
-\Author[4]{Colin}{Zarzycki}
+\Author[4]{Colin M.}{Zarzycki}
 
-\Author[6]{Thomas}{Dubos}
-\Author[7]{Marco}{Giorgetta}
-\Author[8]{Elijah}{Goodfriend}
-\Author[9]{David A.}{Hall}
-\Author[10]{Lucas}{Harris}
-\Author[8]{Hans}{Johansen}
-\Author[11]{Christian}{Kuehnlein}
-\Author[12]{Vivian}{Lee}
-\Author[13]{Thomas}{Melvin}
-\Author[14]{Hiroaki}{Miura}
-\Author[15]{David}{Randall}
-\Author[16]{Alex}{Reinecke}
+\Author[6]{David A.}{Hall}
+
+\Author[7]{Alex}{Reinecke}
+\Author[7]{Kevin}{Viner}
+
+\Author[8]{Don}{Dazlich}
+\Author[8]{Ross}{Heikes}
+\Author[8]{Celal}{Konor}
+\Author[8]{David}{Randall}
+
+\Author[9]{Xi}{Chen}
+\Author[9]{Lucas}{Harris}
+
+\Author[10]{Marco}{Giorgetta}
+\Author[11]{Daniel}{Reinert}
+
+\Author[12]{Christian}{Kuehnlein}
+
+\Author[13]{Robert}{Walko}
+
+\Author[14]{Vivian}{Lee}
+\Author[14]{Abdessamad}{Qaddouri}
+\Author[14]{Claude}{Girard}
+
+\Author[15]{Hiroaki}{Miura}
+\Author[15]{Tomoki}{Ohno}
+\Author[16]{Ryuji}{Yoshida}
+
+\Author[4]{Joseph}{Klemp}
+\Author[4]{Sang-Hun}{Park}
 \Author[4]{William}{Skamarock}
-\Author[16]{Kevin}{Viner}
-\Author[17]{Robert}{Walko}
+
+\Author[17]{Thomas}{Dubos}
+\Author[17]{Yann}{Meurdesoif}
+
+\Author[18]{Elijah}{Goodfriend}
+\Author[18]{Hans}{Johansen}
 
@@
 \affil[5]{Stony Brook University}
-\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
-\affil[7]{Max Planck Institute for Meteorology}
-\affil[8]{Lawrence Berkeley National Laboratory}
-\affil[9]{University of Colorado, Boulder}
-\affil[10]{Geophysical Fluid Dynamics Laboratory}
-\affil[11]{European Center for Medium-Range Weather Forecasting}
-\affil[12]{Environment Canada}
-\affil[13]{U.K. Met Office}
-\affil[14]{University of Tokyo}
-\affil[15]{Colorado State University}
-\affil[16]{Naval Research Laboratory}
-\affil[17]{University of Miami}
+\affil[6]{University of Colorado, Boulder}
+\affil[7]{Naval Research Laboratory}
+\affil[8]{Colorado State University}
+\affil[9]{Geophysical Fluid Dynamics Laboratory}
+\affil[10]{Max Planck Institute for Meteorology}
+\affil[11]{Deutscher Wetterdienst (DWD)}
+\affil[12]{European Center for Medium-Range Weather Forecasting}
+\affil[13]{University of Miami}
+\affil[14]{Environment Canada}
+\affil[15]{University of Tokyo}
+\affil[16]{RIKEN}
+\affil[17]{Institut Pierre-Simon Laplace (IPSL)}
+\affil[18]{Lawrence Berkeley National Laboratory}
 
@@
 
-{\color{red}INSTRUCTIONS FOR AUTHORS}
+{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}
 
-{\color{red} Fill in text in section 3, 4, 5, 6, 7 and 8 below.}
+The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.
+Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) throughout the atmosphere, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.
+These factors include the choice of model grid, vertical coordinates, representation of topography, numerical method, physics/dynamics coupler, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.
+
+To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP has hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
+It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
+Special attention is paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling.
+The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.
+
+The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.
+The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.
+Modeling groups have already begun to leverage this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.
+
+The proposed workshop has advanced our knowledge of (a) the relative behaviors exhibited by atmospheric dynamical cores, (b) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (c) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
+Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.
+A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.
 
@@
 
+\section{Surface Fluxes on an Aqua-Planet with Prescribed Sea Surface Temperatures} \label{sec:OceanSurfaceFluxes}
+
+The forcing by surface fluxes from an idealized ocean is described in \cite{reed2012idealized} and is partly reproduced here.
+We use a model configuration which corresponds to an aqua-planet setup with prescribed sea surface temperatures (SSTs).
+This forcing by the surface fluxes is applied to the state variables in the lowermost model level using a partially implicit formulation to avoid numerical instabilities.
+Throughout this section we use the subscript $a$ to denote variables defined on the lowermost model level.
+
+The surface fluxes depend on the \textit{drag coefficient} $C_d$, defined as
+\begin{equation} \label{eq:DragCoefficient}
+\begin{array}{ll}
+C_d = C_{d0} + C_{d1} \vert \vec{v}_a \vert & \mbox{for $\vert \vec{v}_a \vert < 20$ m s$^{-1}$} \\ 
+C_d = 0.002 & \mbox{for $\vert \vec{v}_a \vert \geq 20$ m s$^{-1}$}, 
+\end{array}
+\end{equation}
+where $C_{d0}$ and $C_{d1}$ are $7.0 \times 10^{-4}$ (unitless) and $6.5 \times 10^{-5}$ s m$^{-1}$, respectively, and $\vert \vec{v}_a \vert$ is the magnitude of the horizontal wind at the lowermost model level.
+In terms of the zonal wind $u_a$ and meridional wind $v_a$, it is defined as
+\begin{equation}
+\vert \vec{v}_a \vert = \sqrt{u_a^2 + v_a^2}.
+\end{equation}  For both evaporation and sensible heat the bulk coefficient is set to
+\begin{equation} \label{eq:BulkTransferCoefficient}
+C_E = C_H = 0.0011.
+\end{equation}
+
+The formulation of the surface fluxes makes use of the height of the lowermost full model level $z_a$ (in m).
+For pressure-based models, $z_a$ can be expressed with the help of the hydrostatic equation in terms of pressure
+\begin{eqnarray} \label{eqn:za}
+z_a = \frac{R_d T_{\nu,a}}{g} \frac{(\ln p_s- \ln p_-)}{2},
+\end{eqnarray}
+where $T_{\nu,a} = T_a (1+0.608 q_a)$ is the virtual temperature at the lowermost full model level and $p_-$ is the edge pressure at the model level interface between the lowest and second lowest full model levels.
+This notation and all following equations assume that the temperature, horizontal wind components and the specific humidity in the physical parameterization package are co-located in both the vertical and horizontal directions, as is the case for the Lorenz grid.
+The height of the lowest full model level should ideally lie between 60-70m above the ground to make the results comparable to those in the literature.
+
+As described in \cite{reed2012idealized}, the surface fluxes can be written as
+\begin{eqnarray}
+\frac{\partial \vec{v}_a}{\partial t} &=& - \frac{C_d \vert \vec{v}_a \vert \vec{v}_a}{z_a} \label{sfu} \\
+\frac{\partial T_a}{\partial t} &=& \frac{C_H \vert \vec{v}_a \vert (T_s-T_a)}{z_a} \label{sft} \\
+\frac{\partial q_a}{\partial t} &=& \frac{C_E \vert \vec{v}_a \vert(q_{sat,s}-q_a)}{z_a}. \label{sfq}
+\end{eqnarray}
+We note that the wind at the surface is taken to be zero and therefore does not appear explicitly in (\ref{sfu}).
+In these equations $T_s$  denotes the prescribed sea surface temperature (SST) and $q_{sat,s}$ is the saturation specific humidity defined by the Clausius-Clapeyron equation
+\begin{eqnarray} \label{qsatcalc}
+q_{sat}(p) \approx \varepsilon\frac{e_s(T_s)}{p} \approx \frac{\varepsilon}{p}e^{*}_{0}e^{-(L/R_\nu)[(1/T_s)-(1/T_0)]},
+\end{eqnarray}
+where $e^{*}_{0}$ ($=610.78$ Pa) is the saturation vapor pressure at $T_0=273.16$ K.
+
+The final form of the surface fluxes will vary for models with other choices of prognostic variables. For example, if potential temperature $\Theta_a$ is used (\ref{sft}) takes the form
+\begin{eqnarray}
+\frac{\partial \Theta_a}{\partial t} &=& \frac{C_H \vert \vec{v}_a \vert (T_s-T_a)}{z_a} \left (\frac{p_0}{p_a} \right )^{R_d/c_p} \label{sftheta}
+\end{eqnarray}
+where $p_0 = 1000$ hPa is a reference pressure.
+This conversion uses the assumption that the pressure is time-invariant when individual physics parameterizations are applied.
+For other choices of prognostic variables like $(\rho u)_a$, $(\rho v)_a$, $(\rho \Theta)_a$ and $(\rho q)_a$ the right-hand-side of (\ref{sfu}), (\ref{sftheta}) and (\ref{sfq}) would need to be multiplied by the density of the air $\rho$.
+
+In order to ensure numerical stability, each of the aforementioned surface fluxes are applied via a semi-implicit operator.
+We demonstrate this procedure on the temperature evolution equation (\ref{sft}).
+First, the time derivative is expanded using a backward Euler operator,
+\begin{eqnarray}
+\frac{T_a^{n+1}-T_a^n}{\Delta t} &=& \frac{C_H \vert \vec{v}_a^n \vert (T_s-T_a^{n+1})}{z_a}.
+\end{eqnarray}
+The superscripts $n$ and $n+1$ represent the current time step (after the update from the large-scale condensation scheme) and the future time step, respectively.
+Note, that on the right-hand-side of the equation the only variable taken implicitly is $T_a$.
+$\vert \vec{v}_a^n \vert$ is evaluated at the current time step and $C_H$ is constant.
+The equation can now be solved for $T_a^{n+1}$
+\begin{eqnarray} \label{Tsurfflux}
+T_a^{n+1} &=& \frac{T_a^n + C_H \vert \vec{v}_a^n \vert T_s \frac{\Delta t}{z_a}}{1 + C_H \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}}.
+\end{eqnarray}
+Similar equations for $\vec{v}_a$ and $q_a$ can be calculated
+\begin{eqnarray}
+\vec{v}_a^{n+1} &=& \frac{\vec{v}_a^n}{1 + C_d^n \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}} \label{qusurfflux}\\
+q_a^{n+1} &=& \frac{q_a^n + C_E \vert \vec{v}_a^n \vert q_{sat,s}^n \frac{\Delta t}{z_a}}{1 + C_E \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}}, \label{qsurfflux}
+\end{eqnarray} 
+with the time-level dependent coefficient $C_d^n$.
+Notice that the second term in the numerator of (\ref{qusurfflux}) is absent in the case of the zonal and merdional wind.
+This is because the wind is set to zero at the surface.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Simplified Mixing in the Planetary Boundary Layer} \label{sec:PlanetaryBoundaryLayer}
@@
 \begin{acknowledgements}
-{\color{blue}[Include a complete list of DCMIP2016 student participants here along with sponsors]}
+DCMIP2016 is sponsored by the National Center for Atmospheric Research Computational Information Systems Laboratory, the Department of Energy Office of Science (award no.
+DE-SC0016015), the National Science Foundation (award no.
+1629819), the National Aeronautics and Space Administration (award no.
+{\color{red}??}), the National Oceanic and Atmospheric Administration (award no.
+{\color{red}??}), the Office of Naval Research and CU Boulder Research Computing.
+This work was made possible with support from our student and postdoctoral participants: Sabina Abba Omar, Scott Bachman, Amanda Back, Tobias Bauer, Vinicius Capistrano, Spencer Clark, Ross Dixon, Christopher Eldred, Robert Fajber, Jared Ferguson, Emily Foshee, Ariane Frassoni, Alexander Goldstein, Jorge Guerra, Chasity Henson, Adam Herrington, Tsung-Lin Hsieh, Dave Lee, Theodore Letcher, Weiwei Li, Laura Mazzaro, Maximo Menchaca, Jonathan Meyer, Farshid Nazari, John O'Brien, Bjarke Tobias Olsen, Hossein Parishani, Charles Pelletier, Thomas Rackow, Kabir Rasouli, Cameron Rencurrel, Koichi Sakaguchi, G\"okhan Sever, James Shaw, Konrad Simon, Abhishekh Srivastava, Nicholas Szapiro, Kazushi Takemura, Pushp Raj Tiwari, Chii-Yun Tsai, Richard Urata, Karin van der Wiel, Lei Wang, Eric Wolf, Zheng Wu, Haiyang Yu, Sungduk Yu and Jiawei Zhuang.
+We would also like to thank Rich Loft, Cecilia Banner, Kathryn Peczkowicz and Rory Kelly (NCAR), Carmen Ho, Perla Dinger, and Gina Skyberg (UC Davis) and Kristi Hansen (University of Michigan) for administrative support during the workshop and summer school.
 \end{acknowledgements}
@@
 \bibliographystyle{copernicus}
-\bibliography{DCMIP2016.bib}
+\bibliography{DCMIP2016-Part1.bib}
 %%

### 2016-08-18 contributions/FV3/FV3-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

--- 2016-05-03 DCMIP2016-Part1-Models.tex
@@
 
+%%%GFDL start
+
+FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}.
+The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm.
+The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
+The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.
+%%%GFDL end
+
 \subsection{Icosahedral grid}
@@
 {\color{red}[ALL] Include the continuous equation set that you use for your model here.}
+
+\subsection{FV$^3$} \label{sec:FV3Equations}
+
+%%%GFDL start
+FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
+%%%GFDL end
+
+%They can be written as:%
+%
+%\begin{align}
+%\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
+%\pdiff{\vb{u}}{t} &= \left( f\vg{\hat k + \zeta} \right ) \times \vb{u} - \frac{1}{\rho}\nabla p |_z, \\
+%\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v, \\
+%\pdiff{w}{t} + \vb{u} \cdot \nabla w &= -\frac{1}{\rho}\pdiff{p}{z} - g,
+%\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$, and local Coriolis parameter $f$.
+Here, gradient operators (except for that in the pressure-gradient term) are along Lagrangian surfaces.
 
@@
 {\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
+%%%GFDL start
+
+Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations.
+The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.
+FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion.
+The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
+If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
+This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping.
+Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces.
+An option to convert kinetic energy lost to the hyperdiffusion to heat is available.
+FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.
+
+A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
+An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
+
+%%%GFDL end
 
@@
 \subsection{Energy fixers}
+%%%GFDL start
+
+FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
+This is only done before the physics is called and is not used in idealized simulations.
+%%%GFDL end
 
@@
 {\color{red}[ALL] Describe the time-stepping scheme / temporal discretization employed by your dynamical core here.}
+
+\subsection{Forward-backward vertically-Lagrangian dynamics}
+
+%%%GFDL start
+
+FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
+With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
+The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the nonhydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
+This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
+Physics tendencies are applied ``impulsively'' at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
+%%%GFDL end
 
@@
 
-\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV3)}
+\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}
+%%%GFDL start
 
-{\color{red}[HARRIS]}
+FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
+The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
+Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport.
+The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
+Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+
+Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
+Explicit viscosity is applied every acoustic timestep.
+
+The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
+The nonhydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
+
+%%%GFDL end
 
@@
 \bibliographystyle{copernicus}
-\bibliography{DCMIP2016.bib}
+\bibliography{DCMIP2016.bib,FV3.bib}
 %%

### 2016-08-18 contributions/FVM/FVM-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

--- 2016-05-03 DCMIP2016-Part1-Models.tex
@@
 %\usepackage{float}
+\usepackage{subcaption}
 %\usepackage{subfig}
@@
 \Author[8]{Hans}{Johansen}
-\Author[11]{Christian}{Kuehnlein}
+\Author[11]{Christian}{K\"uhnlein}
 \Author[12]{Vivian}{Lee}
@@
 \affil[10]{Geophysical Fluid Dynamics Laboratory}
-\affil[11]{European Center for Medium-Range Weather Forecasting}
+\affil[11]{European Centre for Medium-Range Weather Forecasting}
 \affil[12]{Environment Canada}
@@
 \correspondence{Paul A. Ullrich (paullrich@ucdavis.edu)}
+
+% User-defined mathematical symbols
+\renewcommand{\emph}[1]{{\color{red}\textbf{#1}}}
+\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
+\newcommand{\p}[1]{\partial #1}
+%Then, writing a partial is as simple as $\pd{J}{x}$.
+\newcommand{\ol}[1]{{\overline #1}}
+\newcommand{\mbf}[1]{{\mathbf #1}}
+\newcommand{\mbfol}[1]{\overline{{\mathbf #1}}}
+\newcommand{\wdh}[1]{{\widehat #1}}
+\newcommand{\wdt}[1]{{\widetilde #1}}
+\newcommand{\sss}[1]{{\scriptscriptstyle #1}}
+\newcommand{\scs}[1]{{\scriptstyle #1}}
 
@@
 $w$ & Vertical velocity \\
-$\omega$ & Vertical pressure velocity  \\
+$\omega$ & Pressure velocity  \\
 $\delta$ & Divergence\\
@@
 $T_v$ & Virtual temperature \\
-$\Theta$ & Potential temperature \\
-$\Theta_v$ & Virtual potential temperature \\
+$\theta$ & Potential temperature \\
+$\theta_v$ & Virtual potential temperature \\
 $q$ & Specific humidity \\
 $P_{ls}$ & Large-scale precipitation rate \\
+$q_v$ & Water vapor water mixing ratio \\
 $q_c$ & Cloud water mixing ratio \\
@@
 
+\subsection{Octahedral reduced Gaussian grid}
+\label{subsection:octahedralreducedgaussiangrid}
+
+As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
+reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
+roots of the Legendre polynomials. 
+The two grids differ in the arrangement of the points along the latitudes, which follows a simple
+rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
+are added with every latitude towards the equator, whereby the spacing between points along the 
+latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
+is suitable for transformations involving spherical harmonics, and has been introduced for global 
+medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
+Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
+the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
+
+\begin{figure}
+\centering
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=5cm]{octahedral_gauss_nodes}
+ \end{subfigure}%
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=4.7cm]{octahedral_RGG_white}
+ \end{subfigure}
+\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
+connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
+24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
+reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
+\label{fig:octahedralgrid}
+\end{figure}
+
+%octahedral_RGG_white.pdf
+
+%Fig.~\ref{fig:octahedralgridFVM}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 {\color{red}[ALL] Include the continuous equation set that you use for your model here.}
+
+\subsection{FVM} \label{sec:FVMEquations}
+
+The fully compressible Euler equations solved in FVM are given as
+\begin{subequations} \label{fvm:compressible}
+\begin{align}
+ & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
+\label{fvm:mass}  \\
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\epsilon q'_v - q_c - q_r) \right)
+- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
++ \mathbf{M} \right)~,
+\label{fvm:momentum} \\
+%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
+& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
+\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
+- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
++  \mathcal{H} \right)~,
+\label{fvm:thermodynamic} \\
+& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
+\label{fvm:gaslaw}
+\end{align}
+\end{subequations}
+which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
+and dry entropy \eqref{fvm:thermodynamic}.
+Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
+physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
+pressure perturbation $\pi'$, 
+with the thermodynamic variables related by the gas law 
+\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
+Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
+respectively. All primed variables correspond to deviations from an ambient state (denoted by
+subscript "a")  that satisfies 
+a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
+see \cite{prusa2008} and \cite{smolarkiewiczJCP2014}. 
+The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
+Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
+coefficient 
+\begin{equation}
+\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
+\end{equation}
+in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
+the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
+form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
+$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
+$\epsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
+The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
+coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
+simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
+Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
+$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
+and the transformation of the physical to the contravariant velocity 
+$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003} and \cite{kuehnleinJCP2012} for discussion.
+%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
+The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
+of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
+in \eqref{fvm:thermodynamic} represents the diffusion of heat.
+
+%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
+%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
 
@@
 
-\subsection{Semi-Implicit time integration}
+\subsection{Semi-implicit time integration of the fully compressible Euler equations in FVM}
+
+In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
+compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
+A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
+for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
+moist dynamics. 
+The generic two-time-level second-order template algorithm employed in the integration is given as
+\begin{equation}
+\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
++0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
+\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
+\label{fvm:solscalar}
+\end{equation}
+In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
+and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
+finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
+\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}~\footnote{Furthermore, 
+in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
+on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.}.
+
+The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
+mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
+and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
+momentum \eqref{fvm:momentum}, and moisture equations enter
+\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
+which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
+is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
+velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
+Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
+The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
+an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
+equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
+The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
+Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
+Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
+executed in an outer iteration.
+
+%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
+%slow rotational modes.
+
+%The integration scheme of FVM approximates the generic conservation law
+%\begin{equation}
+%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
+%\label{scalar}
+%\end{equation}
+%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
+%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
+%second-order template algorithm
+%\begin{equation}
+%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
+%+0.5\,\delta t\,R^{\Psi}|_{\mbf{i}}^{\scs n+1}~,
+%\hspace{1em} \widetilde{\Psi}^{n}\,\equiv\,\Psi^n+0.5\,\delta t\,R^{\Psi}|^{\scs n}~,
+%%\hspace{1em} \widetilde{\psi}^{n}$\,$\equiv$\,$\psi^n+0.5\,\delta\ol{t}\,R^{\psi}|^{\scs n}~,
+%\label{solscalar}
+%\end{equation}
+%where $\mathcal{A}$ is a shorthand for the advective transport operator MPDATA. The index $\mathbf{i}$ 
+%denotes the spatial position on the computational grid, $\delta t$ is the time step size between levels $n$
+%and $n+1$,
 
@@
 
-\subsection{ECMWF PentaRei Finite Volume Model}
+\subsection{Finite-volume module of the Integrated Forecasting System}
+\label{sec:FVM}
 
-{\color{red}[KUEHNLEIN]}
+%{\color{red}[KUEHNLEIN]}
+
+The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is developed at 
+ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
+equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
+are available by means of simple switches. The formulation incorporates 
+a generalised, optionally time-dependent, terrain-following vertical coordinate based on height. 
+A centred two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
+of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
+Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
+Conjugate Residual approach. 
+The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
+The horizontal spatial discretisation is fully unstructured finite-volume using the 
+median-dual approach. This is combined with a structured-grid finite-difference approach in the 
+vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
+In both the horizontal and the vertical discretisation, all prognostic variables are 
+co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
+of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
+reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
+IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
+not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
+
+No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
+diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
+boundary layer parametrisation and the constant-coefficient second-order dissipation/diffusion in 
+the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
+the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
+at every time step according to a given maximum advective CFL number (typically somewhat smaller 
+than 1). The physics time step is identical to the dynamics time step. 
+%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
+%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
+
+%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
+%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
+%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
+%efficiency and accuracy.
+
+%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
+%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
+%framework developed at ECMWF.
 
@@
 
-%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
+%% Since the Copernicus LaTeX package includes the TeX style file copernicus.bst,
 %% authors experienced with BibTeX only have to include the following two lines:
@@
 \bibliographystyle{copernicus}
-\bibliography{DCMIP2016.bib}
+\bibliography{DCMIP2016,../../bib/fvm}
+%\bibliography{fvm}
 %%

### 2016-08-18 contributions/GEM/GEM-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

--- 2016-05-03 DCMIP2016-Part1-Models.tex
@@
 
+\subsection{Yin-Yang grid} \label{sec:GEM_model_grid}
+In GEM model, we use spherical coordinates on the overset Yin-Yang grid.
+The Yin-Yang grid  \citep{kageama2004yinyang}  has two grid components which are geometrically
+identical (see Fig. \ref{fig:gem_yinyang}). They are combined to cover a spherical surface with partial
+overlap on their borders. Each component is in fact a part of the
+latitude-longitude grid and is defined in spherical
+polar coordinates by
+\begin{equation}\label{eq:gem_yinyang}
+(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
+\frac{\pi}{4}+\delta_{\theta})  \cap
+(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
+\frac{3\pi}{4}+\delta_{\lambda}),
+\end{equation}
+where $\delta_{\lambda}, \delta_{\theta}$ are small buffers, which are
+proportional to the respective grid-spacings, required for a minimum
+overlap in the overset methodology. In our DCMIP experiments, we use 
+$ \delta_{\theta}=2 $ degrees and
+$ \delta_{\lambda}=3 \delta_{\theta} $.
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[height=6cm]{gem_yinyang.png}
+\caption{ Yin-Yang grid }
+\label{fig:gem_yinyang}
+\end{figure}
+%----------------------------------------------------
+
 \subsection{Latitude-longitude grid}
@@
 
+\subsection{GEM} \label{sec:GEM_equations}
+The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
+\begin{eqnarray}
+\frac{ {d \bf V}_{h} }{dt}+ f {\bf k} \times {\bf{V}_{h}} + R_d T_v\nabla
+_\zeta \,  \ln\, p+ \left( 1 + \mu \right) \nabla
+_\zeta {\phi} = 0,
+\label{eq:gem_eq_1} \\
+\frac{d w}{dt}- g \mu = 0,
+\label{eq:gem_eq_2} \\
+\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right)
++ \nabla
+_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial
+\dot{ \zeta }}{\partial
+\zeta }  = 0,
+\label{eq:gem_eq_3} \\
+\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt}=0,
+\label{eq:gem_eq_4} \\
+R_dT_v+  \frac{p}{\pi} \frac{\partial \phi}{\partial \ln \pi}=0,
+\label{eq:gem_eq_5} \\
+\frac{d \phi}{dt}-  g w  = 0,
+\label{eq:gem_eq_6} \\
+1+ \mu - \frac{p}{\pi}   \frac{\partial \ln p}{\partial \ln  \pi}=0,
+\label{eq:gem_eq_7} \\
+\ln \pi= \zeta+Bs ,
+\label{eq:gem_eq_8}
+\end{eqnarray}
+
+\noindent with horizontal velocity ${\bf V}_h$, 
+generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+vertical velocity $w$,
+geopotential $\phi$, 
+virtual temperature $T_v$, 
+pressure $p$ and
+hydrostatic pressure $\pi$. The variable 
+$\mu$ is a measure of departure from hydrostatic balance, 
+$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.
+
 \subsection{Tempest} \label{sec:TempestEquations}
@@
 \subsection{Tracer transport}
+In GEM model, the transport equation for tracers uses the Lagrangian form:
+\begin{align}
+\diff{q}{t} &= 0.
+\end{align}
+
+\vfill
+\vfill
 
@@
 
-\subsection{Mass-based coordinates}
+\subsection{Mass-based coordinates} \label{sec:GEM_zeta}
+The vertical coordinate of GEM model named $ \zeta$ is of a log-hydrostatic-pressure type and is defined by:
+\begin{align}\label{eq:gem_zeta}
+ \ln \pi= \zeta+B(\zeta) s \, \, ; \, \, B(\zeta) =\left( \frac {\zeta-\zeta_T} {\zeta_s-\zeta_T}\right)^r \,; \, s= \ln ( \frac {\pi_s} {p_{o}}) \,; \, \, p_{o}=1000 \,\, \mbox{hPa},
+\end{align}
+with $\zeta_T= \ln \pi_T$, $\zeta_s= \ln {p_{o}}$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+Figure \ref{fig:gem_charney-phillips} shows the Charney-Phillips grid in GEM model, giving the position occupied by each variable in the vertical domain.
+Horizontal velocity ${\bf V}_h$, geopotential $\phi$ and $q=\ln(p/\pi)$ are on momentum levels.
+Temperature $T$, vertical velocity $w$,
+generalized vertical velocity $\dot{ \zeta }$ and tracers are on thermodynamic levels.
+
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[height=10cm]{gem_charney-phillips.png}
+\caption{ Charney-Phillips grid }
+\label{fig:gem_charney-phillips}
+\end{figure}
+%----------------------------------------------------
 
@@
 {\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
+
+Via an application of (\ref{eq:scalar_viscosity_direct}),
+scalar viscosity is employed for wind components and tracers. 
+A vertical sponge layer, also via an application of (\ref{eq:scalar_viscosity_direct}),
+is employed on wind components and $T_v$ with a vertical modulation on the 6 top levels 
+and a maximum damping coefficient of $.380000 \times 10^{6}\ \mbox{m\textsuperscript{2}/s}$ at the top.
+For stabilization purpose, 
+the temporal discretization of GEM presented  
+in section \ref{sec:GEM_temporal} uses an off-centering parameter $b^{A} = 0.6$.
 
@@
 \subsection{Mass borrowing (positive definite preservation)}
+In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}
+
+Consider a frictionless adiabatic prognostic equation of the form:
+\begin{equation} 
+\frac {dF}{dt}+G=0,
+\label{eq:gem_temporal_O}
+\end{equation}
+\noindent where $F$ represents one of the prognostic variables and $G$ represents the remaining terms, some of which are non-linear.
+Based on the semi-Lagrangian scheme,
+The equation (\ref {eq:gem_temporal_O}) is discretized as:
+\begin{equation} 
+\frac{F^A-F^{D}} {\Delta t}+ b^A G^A + (1-b^A) G^D = 0,
+\label{eq:gem_temporal_D}
+\end{equation}
+with $A \, \mbox{(Arrival)} = (\vg{r},t)$, $D \, \mbox{(Departure)} = (\vg{r}-\Delta \vg{r},t-\Delta t)$ 
+and $\vg{r}$ is the 3D grid position. 
+The displacements $\Delta \vg{r}$ are obtained by the iterative process:
+\begin{equation} 
+\Delta  \vg{r}^{i} =  \left [ b^A \vg{V}(\vg{r},t) + (1-b^A) \vg{V}(\vg{r}-\Delta \vg{r}^{i-1},t-\Delta t) \right ] \Delta t,
+\label{eq:gem_temporal_traj} 
+\end{equation} 
+where  $\vg{V}$ is the 3D wind vector.
+From (\ref {eq:gem_temporal_D}), we regroup unknown terms on the left-hand side 
+and known terms on the right-hand side, that is:
+\begin{align} 
+\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D}= R,
+\label{eq:gem_temporal_AD}
+\end{align}
+\noindent where $\tau= \Delta t b^A$ and $\beta = (1-b^A)/b^A$. 
+Cubic Lagrange interpolation is used for upstream evaluations $(F^{D}, G^{D})$.
+Each left-hand side term $\frac{F^A} {\tau}+ G^A$ is then split into a linear part $L$ and a nonlinear residual part $N$:   
+\begin{align} 
+\frac{F^A} {\tau} + G^A = L + N = R. 
+\label{eq:gem_temporal_LN}
+\end{align}
+The linearization is done around a constant in time, horizontally homogeneous, reference state. 
+This results in the non-linear problem:
+\begin{align}
+L^{i}=R-N^{i-1},
+\label{eq:gem_temporal_solve}
+\end{align}
+which is solved with two iterations.   
+In each iteration, the
+linear system (\ref{eq:gem_temporal_solve})  
+is reduced to an Helmholtz problem 
+for one composite variable.
+It 
+is solved 
+with a direct solver
+using the Schwarz-type domain decomposition method on Yin-Yang grid \citep{Qaddouri2008schwarz}.
+The composite variable solution is used to update the prognostic variables at time $t$.  
+Our procedure involves another iterative process linked to the estimation of $\vg{V}(\vg{r},t)$ in the displacement calculations (\ref{eq:gem_temporal_traj}). 
+This results in redoing the sequence of operations (\ref{eq:gem_temporal_traj})-(\ref{eq:gem_temporal_solve}). 
+At each time step, the cubic Lagrange interpolation is used to update the static halo region of both panels of the Yin-Yang grid 
+\citep{Qaddouri2011operational}.
 
@@
 
-\subsection{Global Environmental Multiscale (GEM) Model}
+\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
 
-{\color{red}[LEE]}
+In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
+The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
+on the Charney-Phillips grid.
+A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
+A scalar viscosity is employed for wind components and tracers 
+via an application of (\ref{eq:scalar_viscosity_direct}). 
+Viscosity operations are applied after the completion of the dynamic time step.
 

### 2016-08-18 contributions/HOMME/HOMME-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

--- 2016-05-03 DCMIP2016-Part1-Models.tex
@@
 
+HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
+It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
+Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
+It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
+A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
+After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
+Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
+The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
+The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
+Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
+
+In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
+Fields are current co-located in the vertical and explicit time-stepping is used.
+Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
+Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+
 \subsection{Model for Prediction Across Scales (MPAS)}

### 2016-08-18 contributions/ICON/ICON-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

--- 2016-08-18 contributions/HOMME/HOMME-DCMIP2016-Part1-Models.tex
@@
 \Author[16]{Alex}{Reinecke}
+\Author[18]{Daniel}{Reinert}
 \Author[4]{William}{Skamarock}
@@
 \affil[17]{University of Miami}
+\affil[18]{Deutscher Wetterdienst}
 
@@
 
+The ICON model uses a triangular grid, which for the common global applications is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
+Each vertex shares 5 triangles.
+These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
+For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle.
+By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
+The ICON grid generator allows an "arbitrary" subdivision factor $n$ for the first refinement step only, the so-called root refinement.
+Typical choices are $n$ = 2, 3 or 5.
+All additional $m$ refinement steps use $n$ = 2, i.e.
+are bisection steps.
+A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
+The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.
+
+For practical purposes the icosahedron is oriented such that one of the major axes coincides with the Earth's rotation axis, with one arc at the 0 meridian.
+The mean resolution on a global grid is defined as $dx_{mean} = SQRT(4 \cdot \pi \cdot a_ref^2 / n_c)$.
+For a discussion of the effective resolution see Dipankar et al.
+(2016).
+As a specialty the ICON grid generator can produce additive regional grids.
+These are regional grids produced by additional refinement steps limited to a region of choice on a given parent grid.
+If needed several regions can be added to a base grid, and each region can be refined by one or more steps.
+During a simulation these regions are coupled by one-way or two-way nesting to the global simulation.
+The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.
+
 \subsection{Centroidal Voronoi tessellation grid}
@@
 {\color{red}[ALL] Include the continuous equation set that you use for your model here.}
+
+\subsection{ICON} \label{sec:ICONEquations}
+
+The equation system of the ICON model is based upon the prognostic variables suggested by \cite{gassmann2008} 
+but uses the two-dimensional rather than the three-dimensional Lamb transformation to convert the nonlinear momentum
+advection into a vector-invariant form. So far, the shallow-atmosphere approximation is applied, but there
+are plans to add an option for the unapproximated deep-atmosphere equations. The basic equation system reads as follows
+\begin{align}\label{1eqmo}
+\pdiff{v_n}{t} + \pdiff{K_h}{n} + (\zeta + f)v_t + w\pdiff{v_n}{z} = -c_{pd} \theta_v \pdiff{ \pi}{n} + F(v_n)
+\end{align}
+\begin{align}\label{3eqmo}
+\pdiff{w}{t} + {\bf v_h} \cdot \nabla w + w \pdiff{w}{z} = -c_{pd} \theta_v \pdiff{\pi}{z} - g  
+\end{align}
+\begin{align}\label{conteq}
+\pdiff{\rho}{t} + \nabla \cdot ({\bf v} \rho) = 0
+\end{align}
+\begin{align}\label{traceq}
+\pdiff{\rho q}{t} + \nabla \cdot ({\bf v} \rho q) = S(\rho q)
+\end{align}
+\begin{align}\label{thdeq1}
+\pdiff{\rho \theta_v}{t} + \nabla \cdot ({\bf v} \rho \theta_v) = \tilde{Q}
+\end{align}
+\rrv{As in \cite{gassmann2013},} Eq. (\ref{thdeq1}) is reformulated using the time derivative of
+\begin{align}\label{def-exner}
+\pi = \left( \frac{R_d}{p_{00}} \rho \theta_v \right)^\frac{R_d}{c_{vd}}
+\end{align}
+in order to obtain
+\begin{align}\label{thdeq}
+\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot ({\bf v} \rho \theta_v) = \hat{Q} \,,
+\end{align}
+\rrv{which simplifies the implicit numerical treatment of the terms representing vertical sound wave propagation.}
+The prognostic variables of the (untransformed) equation system are the horizontal velocity component normal to the triangle edges $v_n$, the
+vertical wind component $w$, density $\rho$, and virtual potential temperature $\theta_v$. When coupled with moisture physics,
+$\rho$ represents the full air density including liquid and solid hydrometeors. The reconstructed
+tangential velocity component is denoted as $v_t$, and in accordance with the model code, it is assumed here that 
+$(v_t,v_n,w)$ form a right-handed system. $\bf v_h$ and $\bf v$ denote the horizontal and three-dimensional wind vector,
+respectively. The remaining symbols are the vertical vorticity component $\zeta$, 
+the Coriolis parameter $f$, the horizontal part of the kinetic energy $K_h = \frac{1}{2} (v_n^2 + v_t^2)$,
+the Exner function $\pi$, the specific heat capacities of dry air at constant pressure $c_{pd}$ and volume $c_{vd}$,
+the gas constant of dry air $R_d = c_{pd}-c_{vd}$, the gravitational acceleration $g$, \rv{and the reference pressure
+$p_{00}$ = 1000 hPa conventionally used for defining the potential temperature.} $\pdiff{}{n}$ denotes a horizontal derivative
+in edge-normal direction, i.e. between the two adjacent mass points. Moreover, $F(v_n)$ denotes source
+terms for horizontal momentum (to be specified below), and $\tilde{Q}$ and $\hat{Q}$ denote appropriately 
+formulated diabatic heat source terms. In the tracer equation $q$ denotes the mixing ration with respect to the full air density $\rho$, and $S(\rho q)$ represents the source term for a tracer $q$.
 
@@
 
+\subsubsection{ICON}
+
+The vertical discretization is formulated in a generalized height-based system, employing the Lorenz staggering.
+The default implementation uses the generalized smooth-level vertical coordinate described by  \cite{leuenberger2010}, but also the \cite{galchen1975} formulation is available.
+
 \subsection{Mass-based coordinates}
@@
 {\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
+
+\subsection{ICON}
+
+The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure. Details are described in sections 2.4 and 2.5 of \cite{zangl2014}. Here a brief summary:
+
+\emph{Damping: }In the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps only without incurring numerical stability problems under extreme conditions.
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
+\end{align}
+$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
+
+Another artificial damping term is Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
+The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
+
+\emph{Diffusion: }The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
+\begin{align} \label{smagn2}
+F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
+\end{align}
+\begin{align}
+F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
+\end{align}
+\begin{align}
+F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
+\end{align}
+An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
+
+A fourth-order computational diffusion is also available for vertical wind speed $w$.
+This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
+Consequently diffusion of $w$ is employed for the DCMIP SCELL test.
+\begin{align}
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
+\end{align}
 
@@
 
+\subsubsection{ICON}
+
+The time stepping scheme of ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
+No time splitting is used with respect to sound waves, because the ration of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
+Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
+Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
+
 \subsection{Runge-Kutta}
@@
 
+\subsection{ICON}
+
+The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
+The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
+The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
+The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
+Further divergence damping is applied to the normal wind at every sub-step.
+Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
+The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
+
 \subsection{Tempest}
@@
 {\color{red}[HALL]}
-
-HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
-It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
-Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
-It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
-A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
-A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
-After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
-Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
-The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
-Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
-The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
-Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
-
-In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
-Fields are current co-located in the vertical and explicit time-stepping is used.
-Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
-Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
 
@@
 \bibliographystyle{copernicus}
-\bibliography{DCMIP2016.bib}
+\bibliography{./DCMIP2016.bib}
 %%

### 2016-11-08 DCMIP2016-Part1-Models.tex
- Updates to descriptions of model grids

@@
 
+\usepackage{subcaption}
+
 % Custom commands
@@
 \newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
+
+% User-defined mathematical symbols
+\renewcommand{\emph}[1]{{\color{red}\textbf{#1}}}
+\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
+\newcommand{\p}[1]{\partial #1}
+%Then, writing a partial is as simple as $\pd{J}{x}$.
+\newcommand{\ol}[1]{{\overline #1}}
+\newcommand{\mbf}[1]{{\mathbf #1}}
+\newcommand{\mbfol}[1]{\overline{{\mathbf #1}}}
+\newcommand{\wdh}[1]{{\widehat #1}}
+\newcommand{\wdt}[1]{{\widetilde #1}}
+\newcommand{\sss}[1]{{\scriptscriptstyle #1}}
+\newcommand{\scs}[1]{{\scriptstyle #1}}
 
@@
 
-{\color{red}[ALL] Add a short description of your model grid here.}
-
 \subsection{Latitude-longitude grid}
 
+The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude {\color{red}[Figure]}.
+Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
+This grid is employed by the UK Met Office \citep{davies2005new}.
+
 \subsection{Cubed-sphere grid}
 
-The equiangular cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.
+The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.
 More information on this choice of grid can be found in \cite{ullrich2014global}.
-On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $i_p \in \{1,2,3,4,5,6\}$.
-By convention, we choose panels 1--4 to be along the equator and panels $5$ and $6$ to be centered on the northern and southern pole, respectively.
-
-\subsection{Icosahedral grid}
-
-\subsection{Centroidal Voronoi tessellation grid}
+On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.
+The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
+The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.
+Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
+{\color{red}[Cubed-sphere grid figure]}
+
+%FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}.
+The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm.
+The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
+The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.
+
+\subsection{Icosahedral (triangular) grid}
+
+For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
+These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
+For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle.
+By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
+
+Several methods are available for subdividing the triangular regions.
+One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement.
+Typical choices are $n$ = 2, 3 or 5.
+All additional $m$ refinement steps use $n$ = 2, i.e.
+are bisection steps.
+A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
+The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.
+For a discussion of the effective resolution see \cite{dipankar2015large}.
+The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.
+The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model.
+The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.
+
+\subsection{Icosahedral (hexagonal) grid}
+
+\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids}
+
+Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.
+For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.
+However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.
+This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.
+Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e.
+for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
+\end{center}
+\caption{A constrained centroidal Voronoi tessellation with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
+\end{figure}
+
+\subsection{Octahedral reduced Gaussian grid}
+\label{subsection:octahedralreducedgaussiangrid}
+
+As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
+reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
+roots of the Legendre polynomials. 
+The two grids differ in the arrangement of the points along the latitudes, which follows a simple
+rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
+are added with every latitude towards the equator, whereby the spacing between points along the 
+latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
+is suitable for transformations involving spherical harmonics, and has been introduced for global 
+medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
+Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
+the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
+
+\begin{figure}
+\centering
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
+ \end{subfigure}%
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
+ \end{subfigure}
+\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
+connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
+24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
+reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
+\label{fig:octahedralgrid}
+\end{figure}
+
+\subsection{Yin-Yang grid} \label{sec:GEM_model_grid}
+
+The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).
+These components are combined to cover a spherical surface with partial overlap along their borders.
+The Yin component covers the latitude-longitude region
+\begin{equation}\label{eq:gem_yinyang}
+(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
+\frac{\pi}{4}+\delta_{\theta})  \cap
+(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
+\frac{3\pi}{4}+\delta_{\lambda}),
+\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.
+For instance, a common configuration employed by the GEM model for DCMIP fixes
+$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.
+The Yang component covers an analogous area, but is rotated so as to cover the region of the sphere outside of the Yin grid.
+This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
+
+\begin{figure}[]
+\includegraphics[height=6cm]{gem_yinyang.png}
+\caption{ Yin-Yang grid }
+\label{fig:gem_yinyang}
+\end{figure}
 
@@
 {\color{red}[ALL] Include the continuous equation set that you use for your model here.}
+
+\subsection{FV$^3$} \label{sec:FV3Equations}
+
+%%%GFDL start
+FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
+%%%GFDL end
+
+\subsection{FVM} \label{sec:FVMEquations}
+
+The fully compressible Euler equations solved in FVM are given as
+\begin{subequations} \label{fvm:compressible}
+\begin{align}
+ & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
+\label{fvm:mass}  \\
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\epsilon q'_v - q_c - q_r) \right)
+- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
++ \mathbf{M} \right)~,
+\label{fvm:momentum} \\
+%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
+& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
+\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
+- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
++  \mathcal{H} \right)~,
+\label{fvm:thermodynamic} \\
+& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
+\label{fvm:gaslaw}
+\end{align}
+\end{subequations}
+which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
+and dry entropy \eqref{fvm:thermodynamic}.
+Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
+physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
+pressure perturbation $\pi'$, 
+with the thermodynamic variables related by the gas law 
+\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
+Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
+respectively. All primed variables correspond to deviations from an ambient state (denoted by
+subscript "a")  that satisfies 
+a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
+see \cite{prusa2008} and \cite{smolarkiewiczJCP2014}. 
+The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
+Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
+coefficient 
+\begin{equation}
+\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
+\end{equation}
+in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
+the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
+form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
+$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
+$\epsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
+The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
+coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
+simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
+Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
+$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
+and the transformation of the physical to the contravariant velocity 
+$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003} and \cite{kuehnleinJCP2012} for discussion.
+%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
+The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
+of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
+in \eqref{fvm:thermodynamic} represents the diffusion of heat.
+
+%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
+%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
+
+\subsection{GEM} \label{sec:GEM_equations}
+The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
+\begin{eqnarray}
+\frac{ {d \bf V}_{h} }{dt}+ f {\bf k} \times {\bf{V}_{h}} + R_d T_v\nabla
+_\zeta \,  \ln\, p+ \left( 1 + \mu \right) \nabla
+_\zeta {\phi} = 0,
+\label{eq:gem_eq_1} \\
+\frac{d w}{dt}- g \mu = 0,
+\label{eq:gem_eq_2} \\
+\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right)
++ \nabla
+_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial
+\dot{ \zeta }}{\partial
+\zeta }  = 0,
+\label{eq:gem_eq_3} \\
+\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt}=0,
+\label{eq:gem_eq_4} \\
+R_dT_v+  \frac{p}{\pi} \frac{\partial \phi}{\partial \ln \pi}=0,
+\label{eq:gem_eq_5} \\
+\frac{d \phi}{dt}-  g w  = 0,
+\label{eq:gem_eq_6} \\
+1+ \mu - \frac{p}{\pi}   \frac{\partial \ln p}{\partial \ln  \pi}=0,
+\label{eq:gem_eq_7} \\
+\ln \pi= \zeta+Bs ,
+\label{eq:gem_eq_8}
+\end{eqnarray}
+
+\noindent with horizontal velocity ${\bf V}_h$, 
+generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+vertical velocity $w$,
+geopotential $\phi$, 
+virtual temperature $T_v$, 
+pressure $p$ and
+hydrostatic pressure $\pi$. The variable 
+$\mu$ is a measure of departure from hydrostatic balance, 
+$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.
+
+\subsection{ICON} \label{sec:ICONEquations}
+
+The equation system of the ICON model is based upon the prognostic variables suggested by \cite{gassmann2008} 
+but uses the two-dimensional rather than the three-dimensional Lamb transformation to convert the nonlinear momentum
+advection into a vector-invariant form. So far, the shallow-atmosphere approximation is applied, but there
+are plans to add an option for the unapproximated deep-atmosphere equations. The basic equation system reads as follows
+\begin{align}\label{1eqmo}
+\pdiff{v_n}{t} + \pdiff{K_h}{n} + (\zeta + f)v_t + w\pdiff{v_n}{z} = -c_{pd} \theta_v \pdiff{ \pi}{n} + F(v_n)
+\end{align}
+\begin{align}\label{3eqmo}
+\pdiff{w}{t} + {\bf v_h} \cdot \nabla w + w \pdiff{w}{z} = -c_{pd} \theta_v \pdiff{\pi}{z} - g  
+\end{align}
+\begin{align}\label{conteq}
+\pdiff{\rho}{t} + \nabla \cdot ({\bf v} \rho) = 0
+\end{align}
+\begin{align}\label{traceq}
+\pdiff{\rho q}{t} + \nabla \cdot ({\bf v} \rho q) = S(\rho q)
+\end{align}
+\begin{align}\label{thdeq1}
+\pdiff{\rho \theta_v}{t} + \nabla \cdot ({\bf v} \rho \theta_v) = \tilde{Q}
+\end{align}
+{As in \cite{gassmann2013},} Eq. (\ref{thdeq1}) is reformulated using the time derivative of
+\begin{align}\label{def-exner}
+\pi = \left( \frac{R_d}{p_{00}} \rho \theta_v \right)^\frac{R_d}{c_{vd}}
+\end{align}
+in order to obtain
+\begin{align}\label{thdeq}
+\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot ({\bf v} \rho \theta_v) = \hat{Q} \,,
+\end{align}
+{which simplifies the implicit numerical treatment of the terms representing vertical sound wave propagation.}
+The prognostic variables of the (untransformed) equation system are the horizontal velocity component normal to the triangle edges $v_n$, the
+vertical wind component $w$, density $\rho$, and virtual potential temperature $\theta_v$. When coupled with moisture physics,
+$\rho$ represents the full air density including liquid and solid hydrometeors. The reconstructed
+tangential velocity component is denoted as $v_t$, and in accordance with the model code, it is assumed here that 
+$(v_t,v_n,w)$ form a right-handed system. $\bf v_h$ and $\bf v$ denote the horizontal and three-dimensional wind vector,
+respectively. The remaining symbols are the vertical vorticity component $\zeta$, 
+the Coriolis parameter $f$, the horizontal part of the kinetic energy $K_h = \frac{1}{2} (v_n^2 + v_t^2)$,
+the Exner function $\pi$, the specific heat capacities of dry air at constant pressure $c_{pd}$ and volume $c_{vd}$,
+the gas constant of dry air $R_d = c_{pd}-c_{vd}$, the gravitational acceleration $g$, {and the reference pressure
+$p_{00}$ = 1000 hPa conventionally used for defining the potential temperature.} $\pdiff{}{n}$ denotes a horizontal derivative
+in edge-normal direction, i.e. between the two adjacent mass points. Moreover, $F(v_n)$ denotes source
+terms for horizontal momentum (to be specified below), and $\tilde{Q}$ and $\hat{Q}$ denote appropriately 
+formulated diabatic heat source terms. In the tracer equation $q$ denotes the mixing ration with respect to the full air density $\rho$, and $S(\rho q)$ represents the source term for a tracer $q$.
 
@@
 {\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
+
+\subsection{FV3}
+
+Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations.
+The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.
+FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion.
+The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
+If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
+This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping.
+Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces.
+An option to convert kinetic energy lost to the hyperdiffusion to heat is available.
+FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.
+
+A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
+An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
+
+\subsection{GEM}
+
+Via an application of (\ref{eq:scalar_viscosity_direct}),
+scalar viscosity is employed for wind components and tracers. 
+A vertical sponge layer, also via an application of (\ref{eq:scalar_viscosity_direct}),
+is employed on wind components and $T_v$ with a vertical modulation on the 6 top levels 
+and a maximum damping coefficient of $.380000 \times 10^{6}\ \mbox{m\textsuperscript{2}/s}$ at the top.
+For stabilization purpose, 
+the temporal discretization of GEM presented  
+in section \ref{sec:GEM_temporal} uses an off-centering parameter $b^{A} = 0.6$.
+
+\subsection{ICON}
+
+The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure. Details are described in sections 2.4 and 2.5 of \cite{zangl2014}. Here a brief summary:
+
+\emph{Damping: }In the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps only without incurring numerical stability problems under extreme conditions.
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
+\end{align}
+$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
+
+Another artificial damping term is Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
+The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
+
+\emph{Diffusion: }The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
+\begin{align} \label{smagn2}
+F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
+\end{align}
+\begin{align}
+F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
+\end{align}
+\begin{align}
+F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
+\end{align}
+An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
+
+A fourth-order computational diffusion is also available for vertical wind speed $w$.
+This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
+Consequently diffusion of $w$ is employed for the DCMIP SCELL test.
+\begin{align}
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
+\end{align}
 
@@
 
+In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
+
 \subsection{Mass fixers}
@@
 \subsection{Energy fixers}
+
+FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
+This is only done before the physics is called and is not used in idealized simulations.
 
@@
 
+\subsection{Semi-implicit time integration of the fully compressible Euler equations in FVM}
+
+In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
+compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
+A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
+for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
+moist dynamics. 
+The generic two-time-level second-order template algorithm employed in the integration is given as
+\begin{equation}
+\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
++0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
+\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
+\label{fvm:solscalar}
+\end{equation}
+In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
+and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
+finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
+\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}~\footnote{Furthermore, 
+in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
+on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.}.
+
+The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
+mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
+and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
+momentum \eqref{fvm:momentum}, and moisture equations enter
+\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
+which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
+is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
+velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
+Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
+The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
+an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
+equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
+The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
+Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
+Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
+executed in an outer iteration.
+
+%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
+%slow rotational modes.
+
+%The integration scheme of FVM approximates the generic conservation law
+%\begin{equation}
+%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
+%\label{scalar}
+%\end{equation}
+%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
+%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
+%second-order template algorithm
+%\begin{equation}
+%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
+%+0.5\,\delta t\,R^{\Psi}|
+
+\subsection{Forward-backward vertically-Lagrangian dynamics}
+
+%%%GFDL start
+
+FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
+With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
+The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the nonhydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
+This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
+Physics tendencies are applied ``impulsively'' at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
+%%%GFDL end
+
+\subsubsection{ICON}
+
+The time stepping scheme of ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
+No time splitting is used with respect to sound waves, because the ration of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
+Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
+Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
+
 \subsection{Semi-Implicit time integration}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}
+
+Consider a frictionless adiabatic prognostic equation of the form:
+\begin{equation} 
+\frac {dF}{dt}+G=0,
+\label{eq:gem_temporal_O}
+\end{equation}
+\noindent where $F$ represents one of the prognostic variables and $G$ represents the remaining terms, some of which are non-linear.
+Based on the semi-Lagrangian scheme,
+The equation (\ref {eq:gem_temporal_O}) is discretized as:
+\begin{equation} 
+\frac{F^A-F^{D}} {\Delta t}+ b^A G^A + (1-b^A) G^D = 0,
+\label{eq:gem_temporal_D}
+\end{equation}
+with $A \, \mbox{(Arrival)} = (\vg{r},t)$, $D \, \mbox{(Departure)} = (\vg{r}-\Delta \vg{r},t-\Delta t)$ 
+and $\vg{r}$ is the 3D grid position. 
+The displacements $\Delta \vg{r}$ are obtained by the iterative process:
+\begin{equation} 
+\Delta  \vg{r}^{i} =  \left [ b^A \vg{V}(\vg{r},t) + (1-b^A) \vg{V}(\vg{r}-\Delta \vg{r}^{i-1},t-\Delta t) \right ] \Delta t,
+\label{eq:gem_temporal_traj} 
+\end{equation} 
+where  $\vg{V}$ is the 3D wind vector.
+From (\ref {eq:gem_temporal_D}), we regroup unknown terms on the left-hand side 
+and known terms on the right-hand side, that is:
+\begin{align} 
+\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D}= R,
+\label{eq:gem_temporal_AD}
+\end{align}
+\noindent where $\tau= \Delta t b^A$ and $\beta = (1-b^A)/b^A$. 
+Cubic Lagrange interpolation is used for upstream evaluations $(F^{D}, G^{D})$.
+Each left-hand side term $\frac{F^A} {\tau}+ G^A$ is then split into a linear part $L$ and a nonlinear residual part $N$:   
+\begin{align} 
+\frac{F^A} {\tau} + G^A = L + N = R. 
+\label{eq:gem_temporal_LN}
+\end{align}
+The linearization is done around a constant in time, horizontally homogeneous, reference state. 
+This results in the non-linear problem:
+\begin{align}
+L^{i}=R-N^{i-1},
+\label{eq:gem_temporal_solve}
+\end{align}
+which is solved with two iterations.   
+In each iteration, the
+linear system (\ref{eq:gem_temporal_solve})  
+is reduced to an Helmholtz problem 
+for one composite variable.
+It 
+is solved 
+with a direct solver
+using the Schwarz-type domain decomposition method on Yin-Yang grid \citep{Qaddouri2008schwarz}.
+The composite variable solution is used to update the prognostic variables at time $t$.  
+Our procedure involves another iterative process linked to the estimation of $\vg{V}(\vg{r},t)$ in the displacement calculations (\ref{eq:gem_temporal_traj}). 
+This results in redoing the sequence of operations (\ref{eq:gem_temporal_traj})-(\ref{eq:gem_temporal_solve}). 
+At each time step, the cubic Lagrange interpolation is used to update the static halo region of both panels of the Yin-Yang grid 
+\citep{Qaddouri2011operational}.
 
@@
 
+\subsection{Colorado State University Model (CSU)}
+
+The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
+The model is based on the non-hydrostatic Unified System of equations proposed by Arakawa and Konor (2009), which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
+The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
+Time-differencing is based on the third-order Adams-Bashforth  scheme.
+Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.
+\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}
+
+FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
+The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
+Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport.
+The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
+Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+
+Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
+Explicit viscosity is applied every acoustic timestep.
+
+The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
+The nonhydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
+
+\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
+
+In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
+The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
+on the Charney-Phillips grid.
+A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
+A scalar viscosity is employed for wind components and tracers 
+via an application of (\ref{eq:scalar_viscosity_direct}). 
+Viscosity operations are applied after the completion of the dynamic time step.
+
+\subsection{High-Order Method Modeling Environment (HOMME)}
+
+{\color{red}[HALL]}
+
+HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
+It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
+Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
+It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
+A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
+After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
+Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
+The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
+The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
+Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
+
+In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
+Fields are current co-located in the vertical and explicit time-stepping is used.
+Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
+Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+
+\subsection{ICON}
+
+The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
+The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
+The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
+The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
+Further divergence damping is applied to the normal wind at every sub-step.
+Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
+The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
+
+\subsection{Model for Prediction Across Scales (MPAS)}
+
+{\color{red}[SKAMAROCK]}
+
+\subsection{Naval Research Laboratory NEPTUNE Model}
+
+{\color{red}[VINER, REINECKE]}
+
+\subsection{Ocean-Land-Atmosphere Model (OLAM)}
+
+{\color{red}[WALKO]}
+
+\subsection{DYNAMICO}
+
+{\color{red}[DUBOS]}
+
 \subsection{Tempest}
-
-{\color{red}[ULLRICH]}
 
@@
 
-\subsection{High-Order Method Modeling Environment (HOMME)}
-
-{\color{red}[HALL]}
-
-\subsection{Model for Prediction Across Scales (MPAS)}
-
-{\color{red}[SKAMAROCK]}
-
-\subsection{Colorado State University Model (CSU)}
-
-{\color{red}[RANDALL]}
-
-\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV3)}
-
-{\color{red}[HARRIS]}
-
-\subsection{Chombo}
-
-{\color{red}[JOHANSEN]}
-
-\subsection{Naval Research Laboratory NEPTUNE Model}
-
-{\color{red}[VINER, REINECKE]}
-
-\subsection{Global Environmental Multiscale (GEM) Model}
-
-{\color{red}[LEE]}
-
-\subsection{Ocean-Land-Atmosphere Model (OLAM)}
-
-{\color{red}[WALKO]}
-
-\subsection{DYNAMICO}
-
-{\color{red}[DUBOS]}
-
-\subsection{ECMWF PentaRei Finite Volume Model}
-
-{\color{red}[KUEHNLEIN]}
-
-\subsection{Icosahedral Non-hydrostatic (ICON) Model}
-
-{\color{red}[GIORGETTA]}
+\subsection{Finite-volume module of the Integrated Forecasting System}
+\label{sec:FVM}
+
+%{\color{red}[KUEHNLEIN]}
+
+The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is developed at 
+ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
+equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
+are available by means of simple switches. The formulation incorporates 
+a generalised, optionally time-dependent, terrain-following vertical coordinate based on height. 
+A centred two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
+of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
+Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
+Conjugate Residual approach. 
+The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
+The horizontal spatial discretisation is fully unstructured finite-volume using the 
+median-dual approach. This is combined with a structured-grid finite-difference approach in the 
+vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
+In both the horizontal and the vertical discretisation, all prognostic variables are 
+co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
+of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
+reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
+IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
+not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
+
+No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
+diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
+boundary layer parametrisation and the constant-coefficient second-order dissipation/diffusion in 
+the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
+the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
+at every time step according to a given maximum advective CFL number (typically somewhat smaller 
+than 1). The physics time step is identical to the dynamics time step. 
+%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
+%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
+
+%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
+%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
+%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
+%efficiency and accuracy.
+
+%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
+%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
+%framework developed at ECMWF.
 

### 2016-11-11 DCMIP2016-Part1-Models.tex
- Added about half of the equations section

@@
 $p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
-$c_p$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_v$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
 $R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$R_\nu$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
-$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
-$\varepsilon$ & Ratio of $R_d$ to $R_\nu$ & $0.622$ \\
+$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
+%$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
+$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
 $M_v$ & Constant for virtual temperature conversion & $0.608$ \\
@@
 
-\section{Equation Sets} \label{sec:EquationSets}
-
-{\color{red}[ALL] Include the continuous equation set that you use for your model here.}
+\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}
+
+In this section we describe the fluid equations utilized by nonhydrostatic models.
+The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
+\begin{align}
+\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
+\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
+\begin{align} \label{eq:TracerTransport}
+\diff{q_i}{t} = 0.
+\end{align}
+
+\subsection{Diagnostic relationships}
+
+The atmospheric fluid is assumed to be an ideal gas.
+For moist air, the ideal gas constant $R^\ast$, specific heat capacity at constant pressure $c_p^\ast$ and specific heat capacity at constant volume $c_v^\ast$ are given by
+\begin{align}
+R^\ast =& R_d + (R_w - R_d) q, & c_p^\ast =& c_{pd} + (c_{pw} - c_{pd}) q, & c_v^\ast =& c_{vd} + (c_{vw} - c_{vd}) q.
+\end{align}  Note that in many models, $R^\ast$, $c_p^\ast$ and $c_v^\ast$ are approximated by $R_d$, $c_{pd}$ and $c_{vd}$, respectively.
+Dry air, water vapor and moist air quantities all satisfy the linear relationship $R = c_p - c_v$.
+For a two-fluid system (dry air plus water vapor), two independent variables plus the specific humidity $q$ are needed to describe the thermodynamic state of the system.
+Key thermodynamic variables include dry air density $\rho_d$, moist density $\rho$, pressure $p$, vapor pressure $e$, temperature $T$, virtual temperature $T_v$, Exner pressure $\pi$, potential temperature $\theta$, and virtual potential temperature $\theta_v$.
+Common ratios $\kappa = R^\ast / c_p^\ast$, $\epsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
+
+Relationships between key thermodynamic variables arise from the ideal gas law, along with definitions of Exner pressure, potential temperature and virtual potential temperature
+\begin{align} \label{eq:DiagnosticRelationships1}
+p =& \rho R_d T_v, & \pi =& \left( \frac{p}{p_0} \right)^{\kappa}, & \theta =& T \left( \frac{p_0}{p} \right)^{\kappa}, & \theta_v =& T_v \left( \frac{p_0}{p} \right)^{\kappa},
+\end{align}  which further give rise to
+\begin{align}
+p =& \left( \frac{\rho R_d \theta_v}{p_0^\kappa} \right)^{\kappa - 1}, & \pi =& \left( \frac{\rho R_d \theta_v}{p_0} \right)^{- R^\ast / c_v^\ast}, & \theta =& \frac{T}{\pi}, & \theta_v =& \frac{T_v}{\pi}.
+\end{align}
+
+Note that virtual temperature is typically approximated by
+\begin{align}
+T_v \approx& T \left(1 + \frac{(1-\epsilon)}{\epsilon} q \right),
+\end{align} which arises from the exact relationship
+\begin{align}
+T_v = \frac{T}{1 - \frac{e}{p} (1 - \epsilon)},
+\end{align} upon applying $e/p = q / \epsilon$ and using a Taylor expansion around $q = 0$.
+
+\subsection{Prognostic equations for thermodynamic variables}
+
+Note that, as a consequence of (\ref{eq:TracerTransport}), the following simplifications can be applied:
+\begin{align}
+\frac{1}{T_v} \diff{T_v}{t} =& \frac{1}{T} \diff{T}{t}, & \diff{R^\ast}{t} =& 0, & \diff{c_p^\ast}{t} =& 0, & \diff{c_v^\ast}{t} =& 0.
+\end{align}
+
+Mass conservation is typically represented through the continuity equation, which can be written in the Lagrangian frame as
+\begin{align} \label{eq:ContinuityEquation}
+\diff{\rho}{t} =& - \rho \nabla \cdot \vb{u},
+\end{align} or equivalently in the Eulerian frame,
+\begin{align}
+\pdiff{\rho}{t} =& - \nabla \cdot (\rho \vb{u}).
+\end{align}
+
+Further prognostic relationships can be derived from the thermodynamic equation, including the diabatic heating rate $J$,
+\begin{align}
+\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
+\end{align} which can be alternatively written as
+\begin{align}
+\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
+\end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
+\begin{align}
+\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
+\end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
+\begin{align}
+\frac{1}{p} \diff{p}{t} = \frac{1}{\rho} \diff{\rho}{t} + \frac{1}{T_v} \diff{T_v}{t},
+\end{align} the thermodynamic equation can be written in the form
+\begin{align}
+\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
+\end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
+\begin{align}
+\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
+\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
+\begin{align} \label{eq:PressurePrognosticEq}
+\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
+\end{align}  And similarly for Exner pressure,
+\begin{align}
+\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
+\end{align}
+
+\subsection{Momentum Equations}
+
+In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
+\begin{align} \label{eq:PrognosticVelocity1}
+\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
+\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.
+The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.
+In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since for an arbitrary coordinate frame will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.
+Note that it is common to rewrite the pressure gradient force using the relationship
+\begin{align}
+- \frac{1}{\rho} \nabla p = - c_p^\ast \theta_v \nabla \pi,
+\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
+\begin{align}
+\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
+\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
+This gives rise to the vector-invariant form,
+\begin{align} \label{eq:PrognosticVelocity2}
+\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} - \nabla (K + \Phi). 
+\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).
+In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
+\begin{align}
+\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
+\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.
+
+The equations above still provide some flexibility with regards to the choice of $\Phi$ and $\vg{\Omega}$.  For \textit{deep atmosphere} models, one typically chooses
+\begin{align}
+\Phi = \frac{g r^3}{3 a^2}, \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \phi + \vb{j} \cos \phi),
+\end{align} where $g$ is gravitational acceleration at the surface, $r$ is the radial coordinate, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\phi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
+For \textit{shallow atmosphere} models, these expressions take the simpler form
+\begin{align}
+\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \phi \vb{k},
+\end{align} where $z$ is the altitude above the surface.  In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \phi$ is the Coriolis parameter.
+
+\subsection{Conservation of Energy}
+
+\subsection{Discretization}
+
+Horizontal momentum equations:
+
+Lagrangian form:
+\begin{align}
+\diff{\vb{u}_h}{t} + f \vb{k} \times \vb{u}_h + \frac{1}{\rho} \nabla_s p + \nabla_s \phi = 0.
+\end{align}
+
+Clark form:
+\begin{align}
+\pdiff{\vb{u}_h}{t} + \nabla_h (K + \phi) + (\zeta + f) \vb{k} \times \vb{u}_h + \ldots + \dot{s} \nabla_h \left( \pdiff{z}{s} w \right)
+\end{align}
+
+Vertical velocity:
+\begin{align}
+\diff{w}{t} = 
+\end{align}
+
+\subsection{Discretization}
+
+The equations in this section make no assumptions about the horizontal coordinate frame, and are written in terms of an arbitrary vertical coordinate $s(t,x,y,z)$.
+In this context, the material derivative takes the form
+\begin{align}
+\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \dot{s} \pdiff{}{s},
+\end{align} where $\vb{u}_h$ is the 2D horizontal velocity vector, $\nabla_s$ is the horizontal gradient operator along surfaces of constant $s$, and $\dot{s}$ denotes the flow velocity perpendicular to surfaces of constant $s$, which relates the physical velocity components $\vb{u}_h$ and $w$ via
+\begin{align}
+\dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
+\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.
 
@@
 $s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
-The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.
 
@@
 
-\subsection{Tracer transport}
-
-\noindent Lagrangian form:
-\begin{align}
-\diff{q}{t} &= 0.
-\end{align}
-
-\noindent Non-conservative Eulerian form:
-\begin{align}
-\pdiff{q}{t} &= - \vb{u} \cdot \nabla q.
-\end{align}
-
-\noindent Flux form:
-\begin{align}
-\pdiff{}{t} (\rho q) &= - \nabla \cdot (\rho q \vb{u}).
-\end{align}
-
 \subsection{Height-based coordinates}
@@
 %----------------------------------------------------
+
+\subsection{Tracer Transport}
+
+\noindent Lagrangian form:
+\begin{align}
+\diff{q}{t} &= 0.
+\end{align}
+
+\noindent Non-conservative Eulerian form:
+\begin{align}
+\pdiff{q}{t} &= - \vb{u} \cdot \nabla q.
+\end{align}
+
+\noindent Flux form:
+\begin{align}
+\pdiff{}{t} (\rho q) &= - \nabla \cdot (\rho q \vb{u}).
+\end{align}
 

### 2016-11-16 DCMIP2016-Part1-Models.tex
- Added some more equation text

@@
 \begin{align} \label{eq:PrognosticVelocity2}
-\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} - \nabla (K + \Phi). 
+\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
 \end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).
@@
 
+%Finally, a prognostic equation for geopotential follows directly from the material derivative,
+%\begin{align} \label{eq:Geopotential}
+%\diff{\Phi}{t} = \vb{u} \cdot \nabla \Phi.
+%\end{align}
+
 \subsection{Conservation of Energy}
 
-\subsection{Discretization}
-
-Horizontal momentum equations:
-
-Lagrangian form:
-\begin{align}
-\diff{\vb{u}_h}{t} + f \vb{k} \times \vb{u}_h + \frac{1}{\rho} \nabla_s p + \nabla_s \phi = 0.
-\end{align}
-
-Clark form:
-\begin{align}
-\pdiff{\vb{u}_h}{t} + \nabla_h (K + \phi) + (\zeta + f) \vb{k} \times \vb{u}_h + \ldots + \dot{s} \nabla_h \left( \pdiff{z}{s} w \right)
-\end{align}
-
-Vertical velocity:
-\begin{align}
-\diff{w}{t} = 
-\end{align}
-
-\subsection{Discretization}
+Total energy, equal to the sum of Kinetic, potential and internal energy, is given by
+\begin{align}
+E = \rho (K + \Phi + c_v^\ast T_v).
+\end{align}  Differentiating total energy and using (\ref{eq:ContinuityEquation}) and (\ref{eq:PrognosticVelocity2}) then leads to
+\begin{align}
+\pdiff{E}{t} + \nabla \cdot (\rho \vb{u} (K + \Phi + c_p^\ast T_v)) = 0.
+\end{align} {\color{red}[This section needs work]}
+
+\subsection{Contravariant Formulation}
+
+The vector velocity $\vb{u}$ is expanded as
+\begin{align}
+\vb{u} = \vb{u}_h + w \vb{k},
+\end{align} where $\vb{k} = \nabla z$ is the unit vector in the vertical direction and $\vb{u}_h \cdot \vb{k} = 0$  ($\vb{u}_h$ is aligned with surfaces of constant $z$).
+The vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
+\begin{align}
+\pdiff{w}{t} =& - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{}{z} (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k},
+\end{align} which, from $\vb{u}_h = \vb{u} - w \vb{k}$, then gives rise to
+\begin{align} \label{eq:HorizontalVelocityOrthogonal}
+\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z,
+\end{align} where the subscript $z$ denotes projection perpendicular to $\vb{k}$.
+
+It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$, which admits the relationships
+\begin{align}
+\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
+(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})
+\end{align} where $\zeta_h = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the vorticity scalar.
+Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.
+In this case
+\begin{align}
+\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) + w \vb{k} \cdot \nabla \vb{u}_h + w \vb{k} \cdot \nabla (w \vb{k}) \\
+=& \vb{u}_h \cdot \nabla \vb{u}_h + w \vb{u}_h \cdot \nabla \vb{k} + w \vb{k} \cdot \nabla \vb{u}_h + (\vb{u} \cdot \nabla w)\vb{k} \\
+=& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w)\vb{k}
+\end{align} where we have used $\nabla \times \vb{k} = 0$.  Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$ takes the form
+\begin{align}
+\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} 
+\end{align}  which, from $\vb{u}_h = \vb{u} - w \vb{k}$, then gives rise to
+\begin{align}
+\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - (\nabla \times \vb{u}_h) \times (w \vb{k}).
+\end{align}
+
+The dynamics equations are then formulated in terms of the vertical coordinate $s(t,x,y,z)$ with $\partial s / \partial z \neq 0$ everywhere.
+It can be shown that horizontal derivatives on $z$ surfaces are related to horizontal derivatives on $s$ surfaces via the relationship
+\begin{align}
+\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
+\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) in terms of derivatives over $s$.
+
+In the new coordinate only $s$ is a function of $z$, and so by chain rule we observe
+\begin{align}
+\pdiff{}{z} = \pdiff{s}{z} \pdiff{}{s}.
+\end{align}  The flow velocity $\dot{s}$ perpendicular to surfaces of constants $s$, which is needed to compute the vertical flux, is then given by
+\begin{align}
+\dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
+\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.
+
+{\color{blue}Divergence:  By construction, between surfaces of constant $z$,
+\begin{align}
+\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \pdiff{w}{z}
+\end{align}
+%And between surfaces of constant $s$, [check in the time dependent case]
+%\begin{align}
+%\nabla \cdot \vb{u} = \nabla_s \cdot \vb{v}_h + \pdiff{\dot{s}}{s}
+%\end{align}
+
+Curl: ?}
+
+\subsection{Covariant Formulation}
+
+The covariant velocity vector $\vb{v}_h$ is defined so as to be the projection of $\vb{u}$ along $\vb{k}$ onto the surface of constant $s$,
+\begin{align} \label{eq:CovariantProjection}
+\vb{v}_h = \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} = \vb{u} - \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \vb{k}.
+\end{align}  Note that $\vb{u}_h$ and $\vb{v}_h$ only vary in their vertical components.  Using (\ref{eq:CovariantProjection}), the material derivative along $s$ surfaces takes the form
+\begin{align}
+\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z} = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
+\end{align}  The evolution equation for the covariant velocity vector can then be derived:
+\begin{align}
+\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] \\
+=& \pdiff{\vb{u}}{t} - \left( \pdiff{\vb{u}}{t} \cdot \nabla s + \vb{u} \cdot \nabla \pdiff{s}{t} \right) \left( \vb{k} \cdot \nabla s \right)^{-1} \vb{k}  + (\vb{u} \cdot \nabla s) (\vb{k} \cdot \nabla s)^{-2} \left(\vb{k} \cdot \nabla \pdiff{s}{t} \right) \vb{k} \\
+=& \left[ \pdiff{\vb{u}}{t} \right]_s - (\vb{k} \cdot \nabla s)^{-1} \left( \vb{u} - \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \vb{k} \right) \cdot \nabla \pdiff{s}{t} \vb{k} \\
+=& \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
+\end{align}  Consequently, the evolution equation is simply the projection of the evolution equation for $\vb{u}$ onto surfaces of constant $s$ plus a vertical term associated with shifting $s$ surfaces.
+In practice, only the horizontal components of $\vb{v}_h$ are treated prognostically, so this term does not appear in the evolution equations.
+
+{\color{blue}
+Taking (\ref{eq:PrognosticVelocity2}) $\cdot \nabla s$ (no time dependence) expands as
+\begin{align}
+\pdiff{\dot{s}}{t} (\vb{s} \cdot \nabla s) =& - \frac{1}{\rho} (\nabla p \cdot \nabla s) - \nabla (K + \Phi) \cdot \nabla s - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \nabla s
+\end{align}
+\begin{align}
+\pdiff{\vb{v}_h}{t} = \pdiff{}{t} (\vb{u} - \dot{s} \vb{s}) =& - \frac{1}{\rho} (\nabla p - \frac{\nabla p \cdot \nabla s}{\vb{s} \cdot \nabla s} \vb{s}) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] + \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \frac{(\nabla s) \vb{s}}{\vb{s} \cdot \nabla s}
+\end{align} Which leads to the covariant evolution equation:
+\begin{align}
+\pdiff{\vb{v}_h}{t} = \pdiff{}{t} (\vb{u} - \dot{s} \vb{s}) =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s
+\end{align}
+
+Analogously, the contravariant evolution equation looks like:
+\begin{align}
+\pdiff{\vb{u}_h}{t} = \pdiff{}{t} (\vb{u} - w \vb{k}) =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[(\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z \\
+=& - \frac{1}{\rho} \left[ \nabla_s p - \pdiff{s}{z} (\nabla_s z) \pdiff{p}{s} \right]
+\end{align}}
+
+%\begin{align}
+%\nabla \cdot \vb{u} = \nabla_s \cdot \vb{u}_h - \pdiff{s}{z} (\nabla_s z) \pdiff{\vb{u}_h}{s} + \pdiff{s}{z} \pdiff{}{s} \left( \vb{v}_h \cdot \nabla_s z + \dot{s} \pdiff{z}{s} \right)
+%\end{align}
+{\color{blue}
+\begin{align}
+\diff{\vb{u}}{t} = \diff{\vb{u}_h}{t} + \diff{w}{t} \vb{k} + w \diff{\vb{k}}{t}.
+\end{align}
 
 The equations in this section make no assumptions about the horizontal coordinate frame, and are written in terms of an arbitrary vertical coordinate $s(t,x,y,z)$.
-In this context, the material derivative takes the form
+In this context, the scalar material derivative takes the form
 \begin{align}
@@
 \dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
-\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.
+\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.  Note that in any Eulerian formulation, $\partial s / \partial t = 0$.}
 

### 2016-11-18 DCMIP2016-Part1-Models.tex
- Added cubed-sphere figure - Added more equations text

@@
 Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
-{\color{red}[Cubed-sphere grid figure]}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{CubedSphereMesh.pdf}
+\end{center}
+\caption{An example cubed-sphere grid.} \label{fig:CubedSphereMesh}
+\end{figure}
 
@@
 \end{align} or equivalently in the Eulerian frame,
-\begin{align}
+\begin{align} \label{eq:EulerianContinuityEquation}
 \pdiff{\rho}{t} =& - \nabla \cdot (\rho \vb{u}).
@@
 The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.
-In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since for an arbitrary coordinate frame will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.
+In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.
 Note that it is common to rewrite the pressure gradient force using the relationship
@@
 
-The equations above still provide some flexibility with regards to the choice of $\Phi$ and $\vg{\Omega}$.  For \textit{deep atmosphere} models, one typically chooses
-\begin{align}
-\Phi = \frac{g r^3}{3 a^2}, \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \phi + \vb{j} \cos \phi),
-\end{align} where $g$ is gravitational acceleration at the surface, $r$ is the radial coordinate, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\phi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
-For \textit{shallow atmosphere} models, these expressions take the simpler form
-\begin{align}
-\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \phi \vb{k},
-\end{align} where $z$ is the altitude above the surface.  In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \phi$ is the Coriolis parameter.
-
 %Finally, a prognostic equation for geopotential follows directly from the material derivative,
@@
 
-\subsection{Conservation of Energy}
-
-Total energy, equal to the sum of Kinetic, potential and internal energy, is given by
-\begin{align}
-E = \rho (K + \Phi + c_v^\ast T_v).
-\end{align}  Differentiating total energy and using (\ref{eq:ContinuityEquation}) and (\ref{eq:PrognosticVelocity2}) then leads to
-\begin{align}
-\pdiff{E}{t} + \nabla \cdot (\rho \vb{u} (K + \Phi + c_p^\ast T_v)) = 0.
-\end{align} {\color{red}[This section needs work]}
-
-\subsection{Contravariant Formulation}
+The equations above still provide some flexibility with regards to the choice of $\Phi$ and $\vg{\Omega}$.  For \textit{deep atmosphere} models, one typically chooses
+\begin{align} \label{eq:DeepAtmosphereGeopotential}
+\Phi = g a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \phi + \vb{j} \cos \phi),
+\end{align} where $g$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\phi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
+For models that don't utilize a height-based vertical coordinate, the geopotential is generally treated as a prognostic variable, with an evolution equation that emerges from the definition $w = dz/dt$,
+\begin{align}
+\diff{\Phi}{t} = \frac{a^2 g w}{(a + z)^2}.
+\end{align}  For \textit{shallow atmosphere} models, the geopotential takes the simpler form
+\begin{align} \label{eq:ShallowAtmosphereGeopotential}
+\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \phi \vb{k},
+\end{align} where $z$ is the altitude above the surface.
+In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \phi$ is the Coriolis parameter.
+The evolution equation for the shallow atmosphere geopotential is then
+\begin{align} \label{eq:PrognosticShallowGeopotential}
+\diff{\Phi}{t} = g w.
+\end{align}
+
+\subsection{Orthogonal Formulation}
 
@@
 \end{align} where $\vb{k} = \nabla z$ is the unit vector in the vertical direction and $\vb{u}_h \cdot \vb{k} = 0$  ($\vb{u}_h$ is aligned with surfaces of constant $z$).
-The vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
-\begin{align}
+Note that this decomposition of $\vb{u}$ is not exactly the contravariant form, which instead uses $\nabla s$ in place of $\vb{k}$.
+In the orthogonal formulation, the vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
+\begin{align} \label{eq:VerticalVelocityOrthogonal}
 \pdiff{w}{t} =& - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{}{z} (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k},
@@
 
-It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$, which admits the relationships
+Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$, which admits the relationships
 \begin{align}
@@
 =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w)\vb{k}
-\end{align} where we have used $\nabla \times \vb{k} = 0$.  Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$ takes the form
-\begin{align}
-\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} 
+\end{align} where we have used $\nabla \times \vb{k} = 0$.  Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
+\begin{align} \label{eq:PrognosticWEquationK2}
+\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}
 \end{align}  which, from $\vb{u}_h = \vb{u} - w \vb{k}$, then gives rise to
-\begin{align}
-\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - (\nabla \times \vb{u}_h) \times (w \vb{k}).
-\end{align}
+\begin{align} \label{eq:PrognosticUhEquationK2}
+\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
+\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquation}) is set equal to zero in accordance with \cite{phillips1966equations}.
 
@@
 \end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.
-
-{\color{blue}Divergence:  By construction, between surfaces of constant $z$,
-\begin{align}
-\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \pdiff{w}{z}
-\end{align}
-%And between surfaces of constant $s$, [check in the time dependent case]
-%\begin{align}
-%\nabla \cdot \vb{u} = \nabla_s \cdot \vb{v}_h + \pdiff{\dot{s}}{s}
-%\end{align}
-
-Curl: ?}
 
@@
 \end{align}  Note that $\vb{u}_h$ and $\vb{v}_h$ only vary in their vertical components.  Using (\ref{eq:CovariantProjection}), the material derivative along $s$ surfaces takes the form
-\begin{align}
+\begin{align} \label{eq:MaterialDerivative}
 \diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z} = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
-\end{align}  The evolution equation for the covariant velocity vector can then be derived:
-\begin{align}
-\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] \\
-=& \pdiff{\vb{u}}{t} - \left( \pdiff{\vb{u}}{t} \cdot \nabla s + \vb{u} \cdot \nabla \pdiff{s}{t} \right) \left( \vb{k} \cdot \nabla s \right)^{-1} \vb{k}  + (\vb{u} \cdot \nabla s) (\vb{k} \cdot \nabla s)^{-2} \left(\vb{k} \cdot \nabla \pdiff{s}{t} \right) \vb{k} \\
-=& \left[ \pdiff{\vb{u}}{t} \right]_s - (\vb{k} \cdot \nabla s)^{-1} \left( \vb{u} - \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \vb{k} \right) \cdot \nabla \pdiff{s}{t} \vb{k} \\
-=& \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
+\end{align}  This equation is typically used to obtain a diagnostic relationship for $\dot{s}$, utilizing $w = dz/dt$,
+\begin{align} \label{eq:DiagnosticSDot}
+\dot{s} = \pdiff{s}{t} + \left( \pdiff{s}{z} \right) (w - \vb{v}_h \cdot \nabla_s z).
+\end{align}
+The evolution equation for the covariant velocity vector can then be derived:
+\begin{align} \label{eq:CovariantVelocityUpdate}
+\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] = \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
 \end{align}  Consequently, the evolution equation is simply the projection of the evolution equation for $\vb{u}$ onto surfaces of constant $s$ plus a vertical term associated with shifting $s$ surfaces.
 In practice, only the horizontal components of $\vb{v}_h$ are treated prognostically, so this term does not appear in the evolution equations.
-
-{\color{blue}
-Taking (\ref{eq:PrognosticVelocity2}) $\cdot \nabla s$ (no time dependence) expands as
-\begin{align}
-\pdiff{\dot{s}}{t} (\vb{s} \cdot \nabla s) =& - \frac{1}{\rho} (\nabla p \cdot \nabla s) - \nabla (K + \Phi) \cdot \nabla s - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \nabla s
-\end{align}
-\begin{align}
-\pdiff{\vb{v}_h}{t} = \pdiff{}{t} (\vb{u} - \dot{s} \vb{s}) =& - \frac{1}{\rho} (\nabla p - \frac{\nabla p \cdot \nabla s}{\vb{s} \cdot \nabla s} \vb{s}) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] + \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \frac{(\nabla s) \vb{s}}{\vb{s} \cdot \nabla s}
-\end{align} Which leads to the covariant evolution equation:
-\begin{align}
-\pdiff{\vb{v}_h}{t} = \pdiff{}{t} (\vb{u} - \dot{s} \vb{s}) =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s
-\end{align}
-
-Analogously, the contravariant evolution equation looks like:
-\begin{align}
-\pdiff{\vb{u}_h}{t} = \pdiff{}{t} (\vb{u} - w \vb{k}) =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[(\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z \\
-=& - \frac{1}{\rho} \left[ \nabla_s p - \pdiff{s}{z} (\nabla_s z) \pdiff{p}{s} \right]
-\end{align}}
-
 %\begin{align}
-%\nabla \cdot \vb{u} = \nabla_s \cdot \vb{u}_h - \pdiff{s}{z} (\nabla_s z) \pdiff{\vb{u}_h}{s} + \pdiff{s}{z} \pdiff{}{s} \left( \vb{v}_h \cdot \nabla_s z + \dot{s} \pdiff{z}{s} \right)
+%=& \pdiff{\vb{u}}{t} - \left( \pdiff{\vb{u}}{t} \cdot \nabla s + \vb{u} \cdot \nabla \pdiff{s}{t} \right) \left( \vb{k} \cdot \nabla s \right)^{-1} \vb{k}  + (\vb{u} \cdot \nabla s) (\vb{k} \cdot \nabla s)^{-2} \left(\vb{k} \cdot \nabla \pdiff{s}{t} \right) \vb{k} \\
+%=& \left[ \pdiff{\vb{u}}{t} \right]_s - (\vb{k} \cdot \nabla s)^{-1} \left( \vb{u} - \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \vb{k} \right) \cdot \nabla \pdiff{s}{t} \vb{k} \\
 %\end{align}
-{\color{blue}
-\begin{align}
-\diff{\vb{u}}{t} = \diff{\vb{u}_h}{t} + \diff{w}{t} \vb{k} + w \diff{\vb{k}}{t}.
-\end{align}
-
-The equations in this section make no assumptions about the horizontal coordinate frame, and are written in terms of an arbitrary vertical coordinate $s(t,x,y,z)$.
-In this context, the scalar material derivative takes the form
-\begin{align}
-\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \dot{s} \pdiff{}{s},
-\end{align} where $\vb{u}_h$ is the 2D horizontal velocity vector, $\nabla_s$ is the horizontal gradient operator along surfaces of constant $s$, and $\dot{s}$ denotes the flow velocity perpendicular to surfaces of constant $s$, which relates the physical velocity components $\vb{u}_h$ and $w$ via
-\begin{align}
-\dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
-\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.  Note that in any Eulerian formulation, $\partial s / \partial t = 0$.}
-
-\subsection{FV$^3$} \label{sec:FV3Equations}
-
-%%%GFDL start
-FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
-%%%GFDL end
+
+\subsection{Conservation Laws}
+
+For arbitrary vertical coordinates, it is common to adopt \textit{pseudo-density} in place of physical density,
+\begin{align}
+\rho_s = \rho (\vb{k} \cdot \nabla s)^{-1} = \rho \left( \pdiff{s}{z} \right)^{-1}.
+\end{align}  Whereas the density $\rho$ is associated with the total mass in an infinitesmal volume $\delta V = \delta x \times \delta y \times \delta z$, the pseudo-density instead describes the mass in the volume bounded above and below by $s$ surfaces, namely $\delta V_s = \delta x \times \delta y \times \delta s$.
+The advantage of this formulation is that it automatically accounts for shifting $s$ surfaces and allows the continuity equation (\ref{eq:ContinuityEquation}) to be formulated as
+\begin{align} \label{eq:PrognosticPseudoDensity}
+\pdiff{\rho_s}{t} + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
+\end{align}  Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
+\begin{align}
+\pdiff{\rho_s q}{t} + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
+\end{align}  In particular, the prognostic equation for virtual potential temperature (or equivalently for potential temperature) reads
+\begin{align} \label{eq:PrognosticPseudoDensityTheta}
+\pdiff{\rho_s \theta_v}{t} + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
+\end{align}
+
+\subsection{Examples of vertical coordinates}
+
+\subsubsection{Terrain-following $z$ coordinates}
+
+In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.  This simplifies the treatment of the bottom boundary condition.
+
+\subsubsection{Mass-based coordinates}
+
+Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
+\begin{align}
+s = \int^{\infty}_{z} \rho g dz.
+\end{align}  Under this definition,
+\begin{align}
+\pdiff{s}{z} = - \rho g.
+\end{align}  Geopotential is adopted as a prognostic variable and $\dot{s}$ is diagnosed using (\ref{eq:DiagnosticSDot}),
+\begin{align}
+\dot{s} = - \rho g \left( w - \vb{v}_h \cdot \nabla_s z \right).
+\end{align}
+
+\subsubsection{Floating Lagrangian coordinates}
+
+Under Lagrangian coordinates the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
+This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).
+
+A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
+Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.
+
+\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
+
+{\color{red}[Fix notation]}
+
+The vertical coordinate of GEM model (denoted $\zeta$ in the model documentation) is of a log-hydrostatic-pressure type and is defined by
+\begin{align}\label{eq:gem_zeta}
+ \log \pi_h = s + B(s) \sigma_p, \qquad B(s) =\left( \frac {s-s_T} {(\log p_0) - s_T}\right)^r, \qquad \sigma_p = \log \left( \frac {\pi_h} {p_{0}} \right),
+\end{align}
+with $s_T= \log \pi_T$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+Figure \ref{fig:gem_charney-phillips} shows the Charney-Phillips grid in GEM model, giving the position occupied by each variable in the vertical domain.
+Horizontal velocity $\vb{v}_h$, geopotential $\Phi$ and $q = \log(p/\pi_h)$ are on momentum levels.
+Temperature $T$, vertical velocity $w$,
+generalized vertical velocity $\dot{s}$ and tracers are on thermodynamic levels.
+
+\subsection{Model Implementations}
+
+\subsubsection{Tempest} \label{sec:TempestEquations}
+
+Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
+
+In the horizontal, Tempest utilizes a spectral-element formulation.
+
+\subsubsection{ICON} \label{sec:ICONEquations}
+
+ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
+The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
+\begin{align}
+\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
+\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
+
+The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
+
+\subsubsection{FV$^3$} \label{sec:FV3Equations}
+
+The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{L2004MWR}.
+Prognostic equations include horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
+The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
+
+\subsubsection{GEM} \label{sec:GEM_equations}
+
+GEM is a semi-Lagrangian model with Eulerian log-hydrostatic-pressure coordinate $\pi_h$ described in section \ref{sec:GEM_zeta} \citep{Girard2014}.
+The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance, using
+\begin{align}
+\mu = \pdiff{p}{\pi_h} - 1,
+\end{align} in which case 
+\begin{align}
+\frac{d \vb{v}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
+\frac{d w}{dt}- g \mu =& 0, \\
+\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
+\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt} =& 0, \\
+R_dT_v+  \frac{p}{\pi} \frac{\partial \Phi}{\partial \log \pi} =& 0, \\
+\frac{d \phi}{dt}-  g w  =& 0, \\
+1+ \mu - \frac{p}{\pi}   \frac{\partial \log p}{\partial \log  \pi} =& 0, \\
+\ln \pi =& \zeta + Bs.
+\end{align}
+
+\subsection{GEM} \label{sec:GEM_equations}
+The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
+
+\noindent with horizontal velocity ${\bf V}_h$, 
+generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+vertical velocity $w$,
+geopotential $\phi$, 
+virtual temperature $T_v$, 
+pressure $p$ and
+hydrostatic pressure $\pi$. The variable 
+$\mu$ is a measure of departure from hydrostatic balance, 
+$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
 
@@
 
-\subsection{GEM} \label{sec:GEM_equations}
-The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
-\begin{eqnarray}
-\frac{ {d \bf V}_{h} }{dt}+ f {\bf k} \times {\bf{V}_{h}} + R_d T_v\nabla
-_\zeta \,  \ln\, p+ \left( 1 + \mu \right) \nabla
-_\zeta {\phi} = 0,
-\label{eq:gem_eq_1} \\
-\frac{d w}{dt}- g \mu = 0,
-\label{eq:gem_eq_2} \\
-\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right)
-+ \nabla
-_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial
-\dot{ \zeta }}{\partial
-\zeta }  = 0,
-\label{eq:gem_eq_3} \\
-\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt}=0,
-\label{eq:gem_eq_4} \\
-R_dT_v+  \frac{p}{\pi} \frac{\partial \phi}{\partial \ln \pi}=0,
-\label{eq:gem_eq_5} \\
-\frac{d \phi}{dt}-  g w  = 0,
-\label{eq:gem_eq_6} \\
-1+ \mu - \frac{p}{\pi}   \frac{\partial \ln p}{\partial \ln  \pi}=0,
-\label{eq:gem_eq_7} \\
-\ln \pi= \zeta+Bs ,
-\label{eq:gem_eq_8}
-\end{eqnarray}
-
-\noindent with horizontal velocity ${\bf V}_h$, 
-generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
-vertical velocity $w$,
-geopotential $\phi$, 
-virtual temperature $T_v$, 
-pressure $p$ and
-hydrostatic pressure $\pi$. The variable 
-$\mu$ is a measure of departure from hydrostatic balance, 
-$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
-
-\subsection{ICON} \label{sec:ICONEquations}
-
-The equation system of the ICON model is based upon the prognostic variables suggested by \cite{gassmann2008} 
-but uses the two-dimensional rather than the three-dimensional Lamb transformation to convert the nonlinear momentum
-advection into a vector-invariant form. So far, the shallow-atmosphere approximation is applied, but there
-are plans to add an option for the unapproximated deep-atmosphere equations. The basic equation system reads as follows
-\begin{align}\label{1eqmo}
-\pdiff{v_n}{t} + \pdiff{K_h}{n} + (\zeta + f)v_t + w\pdiff{v_n}{z} = -c_{pd} \theta_v \pdiff{ \pi}{n} + F(v_n)
-\end{align}
-\begin{align}\label{3eqmo}
-\pdiff{w}{t} + {\bf v_h} \cdot \nabla w + w \pdiff{w}{z} = -c_{pd} \theta_v \pdiff{\pi}{z} - g  
-\end{align}
-\begin{align}\label{conteq}
-\pdiff{\rho}{t} + \nabla \cdot ({\bf v} \rho) = 0
-\end{align}
-\begin{align}\label{traceq}
-\pdiff{\rho q}{t} + \nabla \cdot ({\bf v} \rho q) = S(\rho q)
-\end{align}
-\begin{align}\label{thdeq1}
-\pdiff{\rho \theta_v}{t} + \nabla \cdot ({\bf v} \rho \theta_v) = \tilde{Q}
-\end{align}
-{As in \cite{gassmann2013},} Eq. (\ref{thdeq1}) is reformulated using the time derivative of
-\begin{align}\label{def-exner}
-\pi = \left( \frac{R_d}{p_{00}} \rho \theta_v \right)^\frac{R_d}{c_{vd}}
-\end{align}
-in order to obtain
-\begin{align}\label{thdeq}
-\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot ({\bf v} \rho \theta_v) = \hat{Q} \,,
-\end{align}
-{which simplifies the implicit numerical treatment of the terms representing vertical sound wave propagation.}
-The prognostic variables of the (untransformed) equation system are the horizontal velocity component normal to the triangle edges $v_n$, the
-vertical wind component $w$, density $\rho$, and virtual potential temperature $\theta_v$. When coupled with moisture physics,
-$\rho$ represents the full air density including liquid and solid hydrometeors. The reconstructed
-tangential velocity component is denoted as $v_t$, and in accordance with the model code, it is assumed here that 
-$(v_t,v_n,w)$ form a right-handed system. $\bf v_h$ and $\bf v$ denote the horizontal and three-dimensional wind vector,
-respectively. The remaining symbols are the vertical vorticity component $\zeta$, 
-the Coriolis parameter $f$, the horizontal part of the kinetic energy $K_h = \frac{1}{2} (v_n^2 + v_t^2)$,
-the Exner function $\pi$, the specific heat capacities of dry air at constant pressure $c_{pd}$ and volume $c_{vd}$,
-the gas constant of dry air $R_d = c_{pd}-c_{vd}$, the gravitational acceleration $g$, {and the reference pressure
-$p_{00}$ = 1000 hPa conventionally used for defining the potential temperature.} $\pdiff{}{n}$ denotes a horizontal derivative
-in edge-normal direction, i.e. between the two adjacent mass points. Moreover, $F(v_n)$ denotes source
-terms for horizontal momentum (to be specified below), and $\tilde{Q}$ and $\hat{Q}$ denote appropriately 
-formulated diabatic heat source terms. In the tracer equation $q$ denotes the mixing ration with respect to the full air density $\rho$, and $S(\rho q)$ represents the source term for a tracer $q$.
-
-\subsection{Tempest} \label{sec:TempestEquations}
-
-The continuity, momentum and thermodynamic equations can be written as:
-\begin{align}
-\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
-\pdiff{\vb{u}}{t} &= - \nabla (K + \Phi) - \theta \nabla \Pi + \vg{\eta} \times \vb{u}, \\
-\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v,
-\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, geopotential $\Phi = g_c z$ and absolute vorticity $\vg{\eta} = \vg{\zeta} + \vg{\Omega}$, which consists of relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$ and planetary vorticity $\vg{\Omega}$.
-The Exner pressure is related to the prognosed density and potential temperature via
-\begin{equation} \label{eq:BaseNonhydrostaticEOS}
-\Pi = c_p \left( \frac{p_0}{p} \right)^{R_d/c_p} = c_p \left( \frac{R_d \rho \theta_v}{p_0} \right)^{R_d/c_v}.
-\end{equation}
+\subsection{Conservation of Energy}
+
+Total energy, equal to the sum of Kinetic, potential and internal energy, is given by
+\begin{align}
+E = \rho (K + \Phi + c_v^\ast T_v).
+\end{align}  Differentiating total energy and using (\ref{eq:ContinuityEquation}) and (\ref{eq:PrognosticVelocity2}) then leads to
+\begin{align}
+\pdiff{E}{t} + \nabla \cdot (\rho \vb{u} (K + \Phi + c_p^\ast T_v)) = 0.
+\end{align} {\color{red}[This section needs work]}
 
@@
 {\color{blue}Define Unstaggered, Lorenz and Charney-Phillips staggering here}
-
-\subsection{Mass-based coordinates} \label{sec:GEM_zeta}
-The vertical coordinate of GEM model named $ \zeta$ is of a log-hydrostatic-pressure type and is defined by:
-\begin{align}\label{eq:gem_zeta}
- \ln \pi= \zeta+B(\zeta) s \, \, ; \, \, B(\zeta) =\left( \frac {\zeta-\zeta_T} {\zeta_s-\zeta_T}\right)^r \,; \, s= \ln ( \frac {\pi_s} {p_{o}}) \,; \, \, p_{o}=1000 \,\, \mbox{hPa},
-\end{align}
-with $\zeta_T= \ln \pi_T$, $\zeta_s= \ln {p_{o}}$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-Figure \ref{fig:gem_charney-phillips} shows the Charney-Phillips grid in GEM model, giving the position occupied by each variable in the vertical domain.
-Horizontal velocity ${\bf V}_h$, geopotential $\phi$ and $q=\ln(p/\pi)$ are on momentum levels.
-Temperature $T$, vertical velocity $w$,
-generalized vertical velocity $\dot{ \zeta }$ and tracers are on thermodynamic levels.
 

### 2016-11-29 DCMIP2016-Part1-Models.tex
- More revisions to equation section

@@
 \begin{align}
-p =& \left( \frac{\rho R_d \theta_v}{p_0^\kappa} \right)^{\kappa - 1}, & \pi =& \left( \frac{\rho R_d \theta_v}{p_0} \right)^{- R^\ast / c_v^\ast}, & \theta =& \frac{T}{\pi}, & \theta_v =& \frac{T_v}{\pi}.
+p =& \left( \frac{\rho R_d \theta_v}{p_0^\kappa} \right)^{\gamma}, & \pi =& \left( \frac{\rho R_d \theta_v}{p_0} \right)^{R^\ast / c_v^\ast}, & \theta =& \frac{T}{\pi}, & \theta_v =& \frac{T_v}{\pi}.
 \end{align}
@@
 \frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
-\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
+\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.
+An analogous equations for pressure can be obtained through a similar procedure, {\color{red}[Double check these two]}
 \begin{align} \label{eq:PressurePrognosticEq}
@@
 
-The vector velocity $\vb{u}$ is expanded as
+Under the orthogonal formulation, projection of a vector field $\vb{b}$ onto its horizontal components is defined via
+\begin{align}
+[\vb{b}]_z = \vb{b} - (\vb{b} \cdot \vb{k}) \vb{k}.
+\end{align}  When applied to the velocity vector this gives rise to the decomposition
 \begin{align}
 \vb{u} = \vb{u}_h + w \vb{k},
-\end{align} where $\vb{k} = \nabla z$ is the unit vector in the vertical direction and $\vb{u}_h \cdot \vb{k} = 0$  ($\vb{u}_h$ is aligned with surfaces of constant $z$).
-Note that this decomposition of $\vb{u}$ is not exactly the contravariant form, which instead uses $\nabla s$ in place of $\vb{k}$.
-In the orthogonal formulation, the vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
+\end{align} where $\vb{k} = \nabla z$ is the unit vector in the vertical direction and $\vb{u}_h = [\vb{u}]_z$ ($\vb{u}_h$ is aligned with surfaces of constant $z$).
+In the orthogonal formulation, the material derivative expands as
+\begin{align} \label{eq:MaterialDerivative}
+\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla + w \vb{k} \cdot \nabla.
+\end{align}  For the special case of the material derivative applied to scalars, this equation can also be written as
+\begin{align} \label{eq:MaterialDerivativeScalar}
+\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla_z + w \pdiff{}{z}.
+\end{align} where $\nabla_z b = [\nabla b]_z$ denotes the gradient along constant $z$ surfaces.
+From here, the vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
 \begin{align} \label{eq:VerticalVelocityOrthogonal}
@@
 \begin{align} \label{eq:HorizontalVelocityOrthogonal}
-\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z,
-\end{align} where the subscript $z$ denotes projection perpendicular to $\vb{k}$.
-
-Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$, which admits the relationships
+\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z.\end{align}
+
+%{\color{red}[Using $K = K_2 + w^2/2$ does this give rise to a simpler version of the $K_2$ evolution equations?]}
+
+Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
+Decomposing the momentum transport term into horizontal and vertical components gives
+\begin{align}
+\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
+\end{align}  The first term in this expression admits the relationships
 \begin{align}
 \left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
-(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})
-\end{align} where $\zeta_h = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the vorticity scalar.
+(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
+\end{align} where $\zeta_h = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.
 Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.
-In this case
+
+{\color{blue}
+\noindent \textbf{Derivation.}
 \begin{align}
 \vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) + w \vb{k} \cdot \nabla \vb{u}_h + w \vb{k} \cdot \nabla (w \vb{k}) \\
-=& \vb{u}_h \cdot \nabla \vb{u}_h + w \vb{u}_h \cdot \nabla \vb{k} + w \vb{k} \cdot \nabla \vb{u}_h + (\vb{u} \cdot \nabla w)\vb{k} \\
-=& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w)\vb{k}
-\end{align} where we have used $\nabla \times \vb{k} = 0$.  Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
+=& \vb{u}_h \cdot \nabla \vb{u}_h + \nabla( \vb{u}_h \cdot (w \vb{k}) ) - (w \vb{k}) \times (\nabla \times \vb{u}_h) - \vb{u}_h \times (\nabla \times w \vb{k}) + \nabla ( \tfrac{1}{2} w^2) + ((\nabla \times (w \vb{k})) \times (w \vb{k})) \\
+=& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}
+\end{align} where we have used
+\begin{align}
+(\nabla \times (w \vb{k})) \times (w \vb{k}) =& (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) \times (w \vb{k}) = - w \vb{k} \times (\nabla w \times \vb{k}) = (- w \vb{k} \cdot \vb{k}) \nabla w - (- w \vb{k} \cdot \nabla w) \vb{k} \\
+- \vb{u}_h \times (\nabla \times w \vb{k}) =& - \vb{u}_h \times (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) = - \vb{u}_h \times (\nabla w \times \vb{k}) = (\vb{u}_h \cdot \nabla w) \vb{k}
+\end{align}}
+
+Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
 \begin{align} \label{eq:PrognosticWEquationK2}
 \pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}
-\end{align}  which, from $\vb{u}_h = \vb{u} - w \vb{k}$, then gives rise to
+\end{align}  and
 \begin{align} \label{eq:PrognosticUhEquationK2}
 \pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
-\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquation}) is set equal to zero in accordance with \cite{phillips1966equations}.
+\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.
 
 The dynamics equations are then formulated in terms of the vertical coordinate $s(t,x,y,z)$ with $\partial s / \partial z \neq 0$ everywhere.
-It can be shown that horizontal derivatives on $z$ surfaces are related to horizontal derivatives on $s$ surfaces via the relationship
-\begin{align}
+In the new coordinate only $s$ is a function of $z$, and so by chain rule we have
+\begin{align} \label{eq:VerticalChainRule}
+\pdiff{}{z} = \pdiff{s}{z} \pdiff{}{s}.
+\end{align}  Chain rule also motivates the definition of the horizontal gradient operator on $s$ surfaces, denoted $\nabla_s$, 
+\begin{align}
+\nabla_s \equiv \nabla_z - \left( \pdiff{s}{z} \right)^{-1} ( \nabla_z s ) \pdiff{}{z}.
+\end{align}  Applying this operator to $z$ gives
+\begin{align}
+\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z,
+\end{align} which in conjunction with (\ref{eq:VerticalChainRule}) admits the more commonly observed relationship
+\begin{align} \label{eq:HorizontalChainRule}
 \nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
-\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) in terms of derivatives over $s$.
-
-In the new coordinate only $s$ is a function of $z$, and so by chain rule we observe
-\begin{align}
-\pdiff{}{z} = \pdiff{s}{z} \pdiff{}{s}.
-\end{align}  The flow velocity $\dot{s}$ perpendicular to surfaces of constants $s$, which is needed to compute the vertical flux, is then given by
-\begin{align}
-\dot{s} = \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z},
-\end{align} where $\nabla_z$ is the horizontal gradient operator along surfaces of constant $z$.
+\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.
+Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z = 0$ and $\vb{k} \cdot \nabla_s = 0$.
+
+The generalized velocity $\dot{s}$ following a fluid parcel is defined by
+\begin{align} \label{eq:DefinitionDotS}
+\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z}.
+\end{align}  A diagnostic relationship for $w$ in terms of $\dot{s}$ then follows,
+\begin{align}
+w = \left( \pdiff{s}{z} \right)^{-1} \left( \dot{s} - \pdiff{s}{t} - \vb{u}_h \cdot \nabla_z s \right).
+\end{align} Substituting this expression back into (\ref{eq:MaterialDerivativeScalar}) gives the material derivative of scalars on $s$ surfaces,
+\begin{align}
+\diff{}{t} =& \pdiff{}{t} + \vb{u}_h \cdot \left( \nabla_z - \left( \pdiff{s}{z} \right)^{-1} (\nabla_z s) \pdiff{}{z} \right) + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z}, \nonumber \\
+=& \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
+\end{align}
 
@@
 
-The covariant velocity vector $\vb{v}_h$ is defined so as to be the projection of $\vb{u}$ along $\vb{k}$ onto the surface of constant $s$,
-\begin{align} \label{eq:CovariantProjection}
-\vb{v}_h = \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} = \vb{u} - \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \vb{k}.
-\end{align}  Note that $\vb{u}_h$ and $\vb{v}_h$ only vary in their vertical components.  Using (\ref{eq:CovariantProjection}), the material derivative along $s$ surfaces takes the form
-\begin{align} \label{eq:MaterialDerivative}
-\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z} = \pdiff{}{t} + \vb{v}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
-\end{align}  This equation is typically used to obtain a diagnostic relationship for $\dot{s}$, utilizing $w = dz/dt$,
-\begin{align} \label{eq:DiagnosticSDot}
-\dot{s} = \pdiff{s}{t} + \left( \pdiff{s}{z} \right) (w - \vb{v}_h \cdot \nabla_s z).
-\end{align}
-The evolution equation for the covariant velocity vector can then be derived:
-\begin{align} \label{eq:CovariantVelocityUpdate}
-\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] = \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
-\end{align}  Consequently, the evolution equation is simply the projection of the evolution equation for $\vb{u}$ onto surfaces of constant $s$ plus a vertical term associated with shifting $s$ surfaces.
+In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.
+This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.
+The horizontal covariance operator is defined by
+\begin{align}
+[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
+\end{align} Hence the covariant horizontal velocity is given by
+\begin{align}
+\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
+\end{align}  For a time-dependent $s$ coordinate,
+\begin{align}
+\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
+=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
+=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
+\end{align}  And so the evolution equation for the covariant velocity is
+\begin{align}
+\pdiff{\vb{v}_h}{t} =& \pdiff{\vb{u}_h}{t} + \pdiff{w}{t} (\nabla_s z) + w \pdiff{}{t} (\nabla_s z) \\
+=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right).
+\end{align}  Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.
+
+\subsection{Floating Lagrangian Formulation}
+
+Under the floating Lagrangian formulation, the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
+This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).
+
+A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
+Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.
+
+In the case of time-varying $s$ coordinates, degrees of freedom are usually stored on $s$ surfaces.  To track floating surfaces of constant $s$ we define the floating Lagrangian derivative,
+\begin{align}
+\frac{d_s}{dt} = \pdiff{}{t} + w \vb{k} \cdot \nabla,
+\end{align} which is used to define the tendency in variables that are allowed to ``float'' vertically along surfaces that adjust with the advection velocity.
+From (\ref{eq:VerticalVelocityOrthogonal}), the vertical velocity equation then takes the form
+\begin{align}
+\frac{d_s w}{d t} =& \pdiff{w}{t} + w \pdiff{w}{z} \\
+=& - \frac{1}{\rho} \left( \pdiff{s}{z} \right) \pdiff{p}{s} - \left( \pdiff{s}{z} \right) \pdiff{}{s} (K_2 + \Phi) - [ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} ] \cdot \vb{k}.
+\end{align}  From (\ref{eq:PrognosticUhEquationK2}), the horizontal velocity equation takes the form
+\begin{align}
+\frac{d_s \vb{u}_h}{d t} =& \pdiff{\vb{u}_h}{t} + w \vb{k} \cdot \nabla \vb{u}_h \\
+=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}),
+\end{align} where we have also used
+\begin{align}
+w \vb{k} \cdot \nabla \vb{u}_h = - w \vb{u}_h \cdot \nabla \vb{k} - (w \vb{k}) \times (\nabla \times \vb{u}_h).
+\end{align}  Note that under the shallow atmosphere approximation, the $(\vb{u}_h \cdot \nabla \vb{k})$ term is typically dropped.
+
+%\begin{align}
+%=& \pdiff{\vb{u}}{t} + \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) \\
+%=& \pdiff{\vb{u}}{t} + \vb{u}_h \cdot \nabla \vb{u}_h + w (\vb{u}_h \cdot \nabla \vb{k}) + (\vb{u}_h \cdot \nabla w) \vb{k}.
+%\end{align}
+
+%Using (\ref{eq:CovariantProjection}), the material derivative along $s$ surfaces takes the form
+%\begin{align} \label{eq:MaterialDerivative}
+%\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla = \pdiff{}{t} + \vb{v}_h \cdot \nabla + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z} = \pdiff{}{t} + \vb{v}_h \cdot \nabla + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
+%\end{align}  This equation is typically used to obtain a diagnostic relationship for $\dot{s}$, utilizing $w = dz/dt$,
+%\begin{align} \label{eq:DiagnosticSDot}
+%\dot{s} = \pdiff{s}{t} + \left( \pdiff{s}{z} \right) (w - \vb{v}_h \cdot \vb{k}).
+%\end{align}
+
+%The evolution equation for the covariant velocity vector can then be derived:
+%\begin{align} \label{eq:CovariantVelocityUpdate}
+%\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] = \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
+%\end{align}  Consequently, the evolution equation is simply the projection of the evolution equation for $\vb{u}$ onto surfaces of constant $s$ plus a vertical term associated with shifting $s$ surfaces.
 In practice, only the horizontal components of $\vb{v}_h$ are treated prognostically, so this term does not appear in the evolution equations.
 %\begin{align}
+
 %=& \pdiff{\vb{u}}{t} - \left( \pdiff{\vb{u}}{t} \cdot \nabla s + \vb{u} \cdot \nabla \pdiff{s}{t} \right) \left( \vb{k} \cdot \nabla s \right)^{-1} \vb{k}  + (\vb{u} \cdot \nabla s) (\vb{k} \cdot \nabla s)^{-2} \left(\vb{k} \cdot \nabla \pdiff{s}{t} \right) \vb{k} \\
@@
 \end{align}
-
-\subsubsection{Floating Lagrangian coordinates}
-
-Under Lagrangian coordinates the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
-This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).
-
-A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
-Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.
 

### 2016-12-22 DCMIP2016-Part1-Models.tex
- Substantial updates to write-up

@@
 
-{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}
+%{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}
 
@@
 The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.
+An example of such a grid is depicted in Figure \ref{fig:ICONMesh}.
 For a discussion of the effective resolution see \cite{dipankar2015large}.
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf}
+\end{center}
+\caption{An icosahedral (triangular) grid with additional refinement over Europe (shown in red).} \label{fig:ICONMesh}
+\end{figure}
+
 \subsection{Icosahedral (hexagonal) grid}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
+\end{center}
+\caption{An icosahedral (hexagonal) grid.} \label{fig:ICONIcosahedlalHexGrid}
+\end{figure}
 
@@
 \frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
-\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.
-An analogous equations for pressure can be obtained through a similar procedure, {\color{red}[Double check these two]}
+\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
 \begin{align} \label{eq:PressurePrognosticEq}
@@
 \begin{align}
-- \frac{1}{\rho} \nabla p = - c_p^\ast \theta_v \nabla \pi,
+- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
 \end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
@@
 (\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
-\end{align} where $\zeta_h = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.
+\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.
 Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.
@@
 
-The dynamics equations are then formulated in terms of the vertical coordinate $s(t,x,y,z)$ with $\partial s / \partial z \neq 0$ everywhere.
-In the new coordinate only $s$ is a function of $z$, and so by chain rule we have
-\begin{align} \label{eq:VerticalChainRule}
-\pdiff{}{z} = \pdiff{s}{z} \pdiff{}{s}.
-\end{align}  Chain rule also motivates the definition of the horizontal gradient operator on $s$ surfaces, denoted $\nabla_s$, 
-\begin{align}
-\nabla_s \equiv \nabla_z - \left( \pdiff{s}{z} \right)^{-1} ( \nabla_z s ) \pdiff{}{z}.
-\end{align}  Applying this operator to $z$ gives
-\begin{align}
-\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z,
-\end{align} which in conjunction with (\ref{eq:VerticalChainRule}) admits the more commonly observed relationship
+\subsection{Arbitrary vertical coordinates}
+
+The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e.
+following \cite{Kasahara1974} (hereafter K74).
+Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
+\begin{align} \label{eq:ChainRule}
+\pdiff{}{z} =& \pdiff{s}{z} \pdiff{}{s}, & \nabla_s =& \nabla_z + (\nabla_s z) (\vb{k} \cdot \nabla), & \left( \pdiff{}{t} \right)_s =& \pdiff{}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla),
+\end{align} which correspond to derivatives in the vertical, in the horizontal and in time.
+This final expression is used to describe the rate of change of a quantity on $s$ surfaces.
+When applied to $s$, these operators yields the useful identities
+\begin{align} \label{eq:NablaZSIdentity}
+\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z, \qquad \mbox{and} \qquad \pdiff{s}{t} = - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
+\end{align}  From here (\ref{eq:ChainRule}) also gives rise to
 \begin{align} \label{eq:HorizontalChainRule}
@@
 \begin{align} \label{eq:DefinitionDotS}
-\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \pdiff{s}{t} + \vb{u}_h \cdot \nabla_z s + w \pdiff{s}{z}.
-\end{align}  A diagnostic relationship for $w$ in terms of $\dot{s}$ then follows,
-\begin{align}
-w = \left( \pdiff{s}{z} \right)^{-1} \left( \dot{s} - \pdiff{s}{t} - \vb{u}_h \cdot \nabla_z s \right).
-\end{align} Substituting this expression back into (\ref{eq:MaterialDerivativeScalar}) gives the material derivative of scalars on $s$ surfaces,
-\begin{align}
-\diff{}{t} =& \pdiff{}{t} + \vb{u}_h \cdot \left( \nabla_z - \left( \pdiff{s}{z} \right)^{-1} (\nabla_z s) \pdiff{}{z} \right) + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z}, \nonumber \\
-=& \pdiff{}{t} + \vb{u}_h \cdot \nabla_s + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
-\end{align}
+\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \vb{u}_h \cdot \nabla_z s + \left[ w - \left( \pdiff{z}{t} \right)_s \right] \pdiff{s}{z}.
+\end{align}  Then using (\ref{eq:ChainRule}) and (\ref{eq:DefinitionDotS}) to rewrite (\ref{eq:MaterialDerivativeScalar}) gives an expression for the material derivative for scalars on $s$ surfaces,
+\begin{align}
+\diff{A}{t} =& \left( \pdiff{A}{t} \right)_s - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s \pdiff{A}{s} + \vb{u}_h \cdot \left[ \nabla_s A + (\nabla_z s) \pdiff{A}{s} \right] + w \pdiff{s}{z} \pdiff{A}{s} \\
+=& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
+\end{align}
+
+The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
+\begin{align} \label{eq:VerticalVelocityOrthogonalS}
+\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}.
+\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
+\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
+\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
+\end{align}  Note that the last term of this expression can then be expanded using
+\begin{align}
+(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
+\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.
+
+The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived from (\ref{eq:PrognosticWEquationK2}),
+\begin{align} \label{eq:PrognosticWEquationK2S}
+\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k},
+\end{align} and from (\ref{eq:PrognosticUhEquationK2}),
+\begin{align} \label{eq:PrognosticUhEquationK2S}
+\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}).
+\end{align}  Observe that both of these equations are simplified if $(\partial z / \partial t)_s = w$.
+
+{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}
 
@@
 \begin{align}
-\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
-=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
-=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
-\end{align}  And so the evolution equation for the covariant velocity is
-\begin{align}
-\pdiff{\vb{v}_h}{t} =& \pdiff{\vb{u}_h}{t} + \pdiff{w}{t} (\nabla_s z) + w \pdiff{}{t} (\nabla_s z) \\
-=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right).
-\end{align}  Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.
-
-\subsection{Floating Lagrangian Formulation}
-
-Under the floating Lagrangian formulation, the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
+\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
+\end{align} and so, using (\ref{eq:VerticalVelocityOrthogonalS}) and (\ref{eq:PrognosticUhEquationK2S}) the evolution equation for the covariant horizontal velocity reads
+\begin{align}
+\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
+%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
+%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
+%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
+=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [(2 \vg{\Omega} + \vg{\zeta}) \times \vb{u}]_s \nonumber \\
+& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) + (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right].
+\label{eq:CovariantVelocityUpdateS}
+\end{align}  Again, observe that this equation simplifies significantly if $(\partial z / \partial t)_s = w$.
+
+%For a time-dependent $s$ coordinate,
+%\begin{align}
+%\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
+%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
+%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
+%\end{align}
+%For a time-dependent $s$ coordinate,
+%\begin{align}
+%\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) (\nabla_s z) \\
+%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
+%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-1} \nabla_z \left( \pdiff{s}{z} \right) + \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-2} (\nabla_z s) \pdiff{}{z} \pdiff{s}{z} \\
+%=& \left( \pdiff{s}{z} \right)^{-1} \left[ - \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \nabla_s \left( \pdiff{s}{z} \right)  \right] \\
+%=& \nabla_s \left( \pdiff{z}{t} \right)_s
+%\end{align}
+
+%Then using (\ref{eq:MaterialDerivativeS}) gives
+%\begin{align}
+%\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{v}_h)
+%\end{align}
+
+%Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.
+
+%\subsection{Floating Lagrangian Formulation}
+
+%Under the floating Lagrangian formulation, the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
 This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).
 
-A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
+%A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
 Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.
 
-In the case of time-varying $s$ coordinates, degrees of freedom are usually stored on $s$ surfaces.  To track floating surfaces of constant $s$ we define the floating Lagrangian derivative,
-\begin{align}
-\frac{d_s}{dt} = \pdiff{}{t} + w \vb{k} \cdot \nabla,
-\end{align} which is used to define the tendency in variables that are allowed to ``float'' vertically along surfaces that adjust with the advection velocity.
+%In the case of time-varying $s$ coordinates, degrees of freedom are usually stored on $s$ surfaces.  To track floating surfaces of constant $s$ we define the floating Lagrangian derivative,
+%\begin{align}
+%\frac{d_s}{dt} = \pdiff{}{t} + w \vb{k} \cdot \nabla,
+%\end{align} which is used to define the tendency in variables that are allowed to ``float'' vertically along surfaces that adjust with the advection velocity.
 From (\ref{eq:VerticalVelocityOrthogonal}), the vertical velocity equation then takes the form
-\begin{align}
-\frac{d_s w}{d t} =& \pdiff{w}{t} + w \pdiff{w}{z} \\
-=& - \frac{1}{\rho} \left( \pdiff{s}{z} \right) \pdiff{p}{s} - \left( \pdiff{s}{z} \right) \pdiff{}{s} (K_2 + \Phi) - [ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} ] \cdot \vb{k}.
-\end{align}  From (\ref{eq:PrognosticUhEquationK2}), the horizontal velocity equation takes the form
-\begin{align}
-\frac{d_s \vb{u}_h}{d t} =& \pdiff{\vb{u}_h}{t} + w \vb{k} \cdot \nabla \vb{u}_h \\
-=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}),
-\end{align} where we have also used
-\begin{align}
-w \vb{k} \cdot \nabla \vb{u}_h = - w \vb{u}_h \cdot \nabla \vb{k} - (w \vb{k}) \times (\nabla \times \vb{u}_h).
-\end{align}  Note that under the shallow atmosphere approximation, the $(\vb{u}_h \cdot \nabla \vb{k})$ term is typically dropped.
+%\begin{align}
+%\frac{d_s w}{d t} =& \pdiff{w}{t} + w \pdiff{w}{z} \\
+%=& - \frac{1}{\rho} \left( \pdiff{s}{z} \right) \pdiff{p}{s} - \left( \pdiff{s}{z} \right) \pdiff{}{s} (K_2 + \Phi) - [ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} ] \cdot \vb{k}.
+%\end{align}  From (\ref{eq:PrognosticUhEquationK2}), the horizontal velocity equation takes the form
+%\begin{align}
+%\frac{d_s \vb{u}_h}{d t} =& \pdiff{\vb{u}_h}{t} + w \vb{k} \cdot \nabla \vb{u}_h \\
+%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}),
+%\end{align} where we have also used
+%\begin{align}
+%w \vb{k} \cdot \nabla \vb{u}_h = - w \vb{u}_h \cdot \nabla \vb{k} - (w \vb{k}) \times (\nabla \times \vb{u}_h).
+%\end{align}  Note that under the shallow atmosphere approximation, the $(\vb{u}_h \cdot \nabla \vb{k})$ term is typically dropped.
 
@@
 
-\subsection{Conservation Laws}
-
-For arbitrary vertical coordinates, it is common to adopt \textit{pseudo-density} in place of physical density,
-\begin{align}
-\rho_s = \rho (\vb{k} \cdot \nabla s)^{-1} = \rho \left( \pdiff{s}{z} \right)^{-1}.
-\end{align}  Whereas the density $\rho$ is associated with the total mass in an infinitesmal volume $\delta V = \delta x \times \delta y \times \delta z$, the pseudo-density instead describes the mass in the volume bounded above and below by $s$ surfaces, namely $\delta V_s = \delta x \times \delta y \times \delta s$.
-The advantage of this formulation is that it automatically accounts for shifting $s$ surfaces and allows the continuity equation (\ref{eq:ContinuityEquation}) to be formulated as
+\subsection{Conservation Laws in Arbitrary Vertical Coordinates}
+
+In this section we rewrite the continuity equation in arbitrary vertical coordinates.
+The divergence of the contravariant velocity field on surfaces of constant $s$ can be obtained from (\ref{eq:ChainRule}), which is valid for both shallow- and deep-atmosphere formulations and equal to K74 eq.
+(3.17),
+\begin{align} \label{eq:DivergenceS}
+\nabla_s \cdot \vb{u}_h = \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \cdot \left( \pdiff{\vb{u}_h}{s} \right).
+\end{align}
+
+On the sphere the 3D divergence takes the form
+\begin{align} \label{eq:ThreeDDivergence}
+\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \frac{1}{\alpha} \pdiff{}{z} (\alpha w),
+\end{align} where $\alpha = 1$ for shallow-atmosphere models and $\alpha = r^2 = (a + z)^2$ for deep-atmosphere models.  Using $w = dz/dt$, this last term also takes the form
+\begin{align} \label{eq:GeneralVerticalTermX}
+\frac{1}{\alpha} \pdiff{}{z} (\alpha w) = \pdiff{w}{z} + \frac{w}{\alpha} \pdiff{\alpha}{z} = \pdiff{w}{z} + \frac{1}{\alpha} \diff{\alpha}{t}.
+\end{align}
+
+Using (\ref{eq:NablaZSIdentity}) to rewrite (\ref{eq:DefinitionDotS}) gives rise to
+\begin{align}
+w = \left( \pdiff{z}{t} \right)_s + \vb{u}_h \cdot \nabla_s z + \dot{s} \left( \pdiff{s}{z} \right)^{-1},
+\end{align} which is then differentiated to yield K74 eq. (3.16),
+\begin{align} \label{eq:WDerivativeS}
+\pdiff{w}{z} = \left( \pdiff{s}{z} \right) \left[ \diff{}{t} \left( \pdiff{s}{z} \right)^{-1} + \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \pdiff{\dot{s}}{s} = 0.
+\end{align}  Substituting this expression into the continuity equation (\ref{eq:ContinuityEquation}), and using (\ref{eq:ThreeDDivergence}), (\ref{eq:GeneralVerticalTermX}), and (\ref{eq:WDerivativeS}) then leads to
+\begin{align}
+\diff{}{t} \left[ \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \left[ \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \pdiff{\dot{s}}{s} = 0.
+\end{align}  Defining the \textit{pseudodensity} as 
+\begin{align}
+\rho_s = \alpha \left( \pdiff{s}{z} \right)^{-1} \rho,
+\end{align} and using (\ref{eq:MaterialDerivativeS}) in the form
 \begin{align} \label{eq:PrognosticPseudoDensity}
-\pdiff{\rho_s}{t} + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
-\end{align}  Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
+\diff{\rho_s}{t} = \left( \diff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
+\end{align} along with (\ref{eq:DivergenceS}) leads to
+\begin{align}
+\left( \diff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
+\end{align}
+
+Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
 \begin{align}
@@
 
-In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.  This simplifies the treatment of the bottom boundary condition.
+In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
+This simplifies the treatment of the bottom boundary condition.
+Common terrain-following coordinates include \cite{galchen1975}, which takes the form
+\begin{align}
+s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
+\end{align}  Another coordinate with smoother variation away from the surface was also introduced by \cite{schar2002new}.
 
@@
 \pdiff{s}{z} = - \rho g.
-\end{align}  Geopotential is adopted as a prognostic variable and $\dot{s}$ is diagnosed using (\ref{eq:DiagnosticSDot}),
-\begin{align}
-\dot{s} = - \rho g \left( w - \vb{v}_h \cdot \nabla_s z \right).
 \end{align}
@@
 
+\subsection{Vertical staggering}
+
+{\color{blue}Define Unstaggered, Lorenz and Charney-Phillips staggering here with figure}
+
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[height=6cm]{gem_charney-phillips.png}
+\caption{A modification of the Charney-Phillips grid found in GEM}
+\label{fig:gem_charney-phillips}
+\end{figure}
+%----------------------------------------------------
+
 \subsection{Model Implementations}
@@
 
-Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum with $(\partial z / \partial t)_s = 0$ (\ref{{eq:CovariantVelocityUpdateS}}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
 The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
 
-In the horizontal, Tempest utilizes a spectral-element formulation.
+%In the horizontal, Tempest utilizes a spectral-element formulation.
 
@@
 
-ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
 As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
@@
 
-\subsection{Conservation of Energy}
-
-Total energy, equal to the sum of Kinetic, potential and internal energy, is given by
-\begin{align}
-E = \rho (K + \Phi + c_v^\ast T_v).
-\end{align}  Differentiating total energy and using (\ref{eq:ContinuityEquation}) and (\ref{eq:PrognosticVelocity2}) then leads to
-\begin{align}
-\pdiff{E}{t} + \nabla \cdot (\rho \vb{u} (K + \Phi + c_p^\ast T_v)) = 0.
-\end{align} {\color{red}[This section needs work]}
-
-\subsection{Height-based coordinates}
-
-{\color{blue}Define Unstaggered, Lorenz and Charney-Phillips staggering here}
-
-%----------------------------------------------------
-\begin{figure}[]
-\includegraphics[height=10cm]{gem_charney-phillips.png}
-\caption{ Charney-Phillips grid }
-\label{fig:gem_charney-phillips}
-\end{figure}
-%----------------------------------------------------
-
-\subsection{Tracer Transport}
-
-\noindent Lagrangian form:
-\begin{align}
-\diff{q}{t} &= 0.
-\end{align}
-
-\noindent Non-conservative Eulerian form:
-\begin{align}
-\pdiff{q}{t} &= - \vb{u} \cdot \nabla q.
-\end{align}
-
-\noindent Flux form:
-\begin{align}
-\pdiff{}{t} (\rho q) &= - \nabla \cdot (\rho q \vb{u}).
-\end{align}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Diffusion and Stabilization} \label{sec:DiffusionStabilization}
-
-{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
+\section{Diffusion, Stabilization, Filters and Fixers} \label{sec:DiffusionStabilization}
+
+Most dynamical cores implement specialized techniques for diffusion or stabilization.
+Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
+Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
+Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, or wave-mode filters.
+Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.
+A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.
+In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.
+
+\subsection{ICON}
+
+The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014} and summarized here:
+
+For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
+\end{align}
+$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
+
+ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
+The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
+
+The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
+\begin{align} \label{smagn2}
+F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
+\end{align}
+\begin{align}
+F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
+\end{align}
+\begin{align}
+F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
+\end{align}
+An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
+
+A fourth-order computational diffusion is also available for vertical wind speed $w$.
+This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
+This term appears as
+\begin{align}
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
+\end{align}
 
@@
 
+FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
+This is only done before the physics is called and is not used in idealized simulations.
+
 \subsection{GEM}
@@
 
-\subsection{ICON}
-
-The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure. Details are described in sections 2.4 and 2.5 of \cite{zangl2014}. Here a brief summary:
-
-\emph{Damping: }In the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps only without incurring numerical stability problems under extreme conditions.
-\begin{align}
-F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
-\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
-\end{align}
-$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
-
-Another artificial damping term is Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
-The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
-
-\emph{Diffusion: }The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
-\begin{align} \label{smagn2}
-F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
-\end{align}
-\begin{align}
-F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
-\end{align}
-\begin{align}
-F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
-\end{align}
-An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
-
-A fourth-order computational diffusion is also available for vertical wind speed $w$.
-This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
-Consequently diffusion of $w$ is employed for the DCMIP SCELL test.
-\begin{align}
-F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
-\end{align}
-
-\subsection{Scalar viscosity}
-
-\noindent Scalar viscosity (direct):
-\begin{align} \label{eq:scalar_viscosity_direct}
-\diff{s}{t} &= \ldots + \nu \nabla \cdot \nabla s.
-\end{align}
-
-\noindent Scalar viscosity (conservative):
-\begin{align} \label{eq:scalar_viscosity_conservative}
-\diff{}{t} (\rho q) &= \ldots + \nu \nabla \cdot ( \rho \nabla q ).
-\end{align}
-
-\subsection{Smagorinsky eddy viscosity}
-
-\subsection{Vector viscosity, divergence and vorticity damping}
-
-\noindent Vector viscosity:
-\begin{align} \label{eq:vector_viscosity}
-\diff{\vb{u}}{t} &= \ldots + \nu \nabla^2 \vb{u}
-\end{align}
-
-\noindent Divergence damping:
-\begin{align} \label{eq:divergence_damping}
-\diff{\vb{u}}{t} &= \ldots + \nu_{div} \nabla (\nabla \cdot \vb{u})
-\end{align}
-
-\noindent Vorticity damping:
-\begin{align} \label{eq:vorticity_damping}
-\diff{\vb{u}}{t} &= \ldots + \nu_{vort} \nabla \times (\nabla \times \vb{u})
-\end{align}
-
-\subsection{Hyperviscosity} \label{sec:Diffusion_Hyperviscosity}
+In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
+
+\subsection{Tempest}
 
 {\color{blue}Repeated application of the scalar and vector viscosity operators}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Filters and Fixers} \label{sec:FiltersFixers}
-
-{\color{red}[ALL] Include explicit filters and fixers that you have utilized in the dynamical core here.}
-
-\subsection{Mass borrowing (positive definite preservation)}
-
-In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
-
-\subsection{Mass fixers}
-
-\subsection{Energy fixers}
-
-FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
-This is only done before the physics is called and is not used in idealized simulations.
 

### 2016-12-26 DCMIP2016-Part1-Models.tex
- Added DYNAMICO text - More paper cleanup

@@
 
+Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.
+In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
+This section describes many of the model grids that have been employed globally.
+
 \subsection{Latitude-longitude grid}
@@
 Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
+This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).
 
@@
 \begin{center}
-\includegraphics[width=2in]{CubedSphereMesh.pdf}
+\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
 \end{center}
-\caption{An example cubed-sphere grid.} \label{fig:CubedSphereMesh}
+\caption{(Left) A cubed-sphere grid with 16 elements along each cubed-sphere edge.
+(Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.
+(Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
 \end{figure}
@@
 The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.
-An example of such a grid is depicted in Figure \ref{fig:ICONMesh}.
+An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
 For a discussion of the effective resolution see \cite{dipankar2015large}.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf}
-\end{center}
-\caption{An icosahedral (triangular) grid with additional refinement over Europe (shown in red).} \label{fig:ICONMesh}
-\end{figure}
-
-\subsection{Icosahedral (hexagonal) grid}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
-\end{center}
-\caption{An icosahedral (hexagonal) grid.} \label{fig:ICONIcosahedlalHexGrid}
-\end{figure}
+\subsection{Icosahedral (hexagonal) grid / geodesic grid} \label{sec:IcosahedralHexagonalGrid}
+
+The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.
+Their cells are hexagonal, except for twelve pentagonal cells.
+Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.
+A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.
+The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).
+Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this the geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
+
+It is useful to optimize icosahedral-hexagonal grids as well.
+DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh.
+This improves the homogeneity of the grid (e.g.
+ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
 
@@
 for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
+CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.
+However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.
 
@@
 \end{center}
-\caption{A constrained centroidal Voronoi tessellation with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
+\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
 \end{figure}
@@
 
-Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
-Decomposing the momentum transport term into horizontal and vertical components gives
+It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
 \begin{align}
@@
 \begin{align} \label{eq:PrognosticWEquationK2}
-\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}
+\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}) \cdot \vb{k}
 \end{align}  and
@@
 \begin{align} \label{eq:VerticalVelocityOrthogonalS}
-\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}.
+\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}_h \right] \cdot \vb{k}.
 \end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
@@
 This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.
-The horizontal covariance operator is defined by
+The horizontal partial covariance operator is defined by
 \begin{align}
@@
 \left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
-%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
-%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
-%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
-=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [(2 \vg{\Omega} + \vg{\zeta}) \times \vb{u}]_s \nonumber \\
-& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) + (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right].
+=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
+& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
+& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
+=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [( \vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_s \nonumber \\
+& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) - (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right].
 \label{eq:CovariantVelocityUpdateS}
@@
 
-%Then using (\ref{eq:MaterialDerivativeS}) gives
+%In the case of an Eulerian formulation with $(\partial z / \partial t)_s = 0$, we can use
+%\begin{align}
+%0 = \nabla (\vb{k} \cdot \vb{u}_h) = (\vb{k} \cdot \nabla \vb{u}_h) + (\vb{u}_h \cdot \nabla \vb{k}) + \vb{k} \times (\nabla \times \vb{u}_h)
+%\end{align} to obtain
 %\begin{align}
@@
 %In the horizontal, Tempest utilizes a spectral-element formulation.
+
+\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
+
+DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014).
+Prognostic variables are pseudo-density ?s, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.
 
@@
 
-\subsubsection{ICON}
+\subsection{DYNAMICO}
+
+DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
+A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
+As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
+The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
+The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
+Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
+
+\subsection{ICON}
 
@@
 
-{\color{red}In this section provide a short description (approximately 0.5 pages) of the dynamical core, focusing on unique features or design specifications.
-Do not include information on the physical parameterizations used by the modeling system.
+This section provide a short description of unique features or design specifications from several modern dynamical cores.
+
+%Do not include information on the physical parameterizations used by the modeling system.
 Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
@@
 Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.
-\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}
-
-FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
+\subsection{DYNAMICO}
+
+DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
+The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical.
+DYNAMICO's design uniquely combines:
+ \begin{itemize}
+\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}
+\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature \citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization
+\item A novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
+\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
+
+\subsection{FV Cubed (FV$^3$)}
+
+The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
 The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
@@
 All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
-
-\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
-
-In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
-The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
-on the Charney-Phillips grid.
-A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
-A scalar viscosity is employed for wind components and tracers 
-via an application of (\ref{eq:scalar_viscosity_direct}). 
-Viscosity operations are applied after the completion of the dynamic time step.
-
-\subsection{High-Order Method Modeling Environment (HOMME)}
-
-{\color{red}[HALL]}
-
-HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
-It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
-Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
-It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
-A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
-A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
-After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
-Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
-The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
-Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
-The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
-Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
-
-In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
-Fields are current co-located in the vertical and explicit time-stepping is used.
-Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
-Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
-
-\subsection{ICON}
-
-The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
-The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
-The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
-For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
-Further divergence damping is applied to the normal wind at every sub-step.
-Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
-
-\subsection{Model for Prediction Across Scales (MPAS)}
-
-{\color{red}[SKAMAROCK]}
-
-\subsection{Naval Research Laboratory NEPTUNE Model}
-
-{\color{red}[VINER, REINECKE]}
-
-\subsection{Ocean-Land-Atmosphere Model (OLAM)}
-
-{\color{red}[WALKO]}
-
-\subsection{DYNAMICO}
-
-{\color{red}[DUBOS]}
-
-\subsection{Tempest}
-
-The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.
-The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
-These continuous equations are given in section \ref{sec:TempestEquations}.
-The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
-Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
-Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
 
@@
 %framework developed at ECMWF.
+
+\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
+
+In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
+The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
+on the Charney-Phillips grid.
+A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
+A scalar viscosity is employed for wind components and tracers 
+via an application of (\ref{eq:scalar_viscosity_direct}). 
+Viscosity operations are applied after the completion of the dynamic time step.
+
+\subsection{High-Order Method Modeling Environment (HOMME)}
+
+HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
+It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
+Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
+It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
+A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
+After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
+Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
+The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
+The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
+Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
+
+In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
+Fields are current co-located in the vertical and explicit time-stepping is used.
+Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
+Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+
+\subsection{ICON}
+
+The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
+The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
+The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
+The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
+Further divergence damping is applied to the normal wind at every sub-step.
+Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
+The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
+
+\subsection{Model for Prediction Across Scales (MPAS)}
+
+{\color{red}[SKAMAROCK]}
+
+\subsection{Naval Research Laboratory NEPTUNE Model}
+
+{\color{red}[VINER, REINECKE]}
+
+\subsection{Ocean-Land-Atmosphere Model (OLAM)}
+
+{\color{red}[WALKO]}
+
+\subsection{Tempest}
+
+The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.
+The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
+These continuous equations are given in section \ref{sec:TempestEquations}.
+The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
+Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).
+Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
+Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
 

### 2016-12-28 DCMIP2016-Part1-Models.tex
- Further cleanup and additions - Moved dynamical core overviews to top

@@
 Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) throughout the atmosphere, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.
-These factors include the choice of model grid, vertical coordinates, representation of topography, numerical method, physics/dynamics coupler, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.
-
-To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP has hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
+These factors include the choice of model grid, choice of prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.
+
+To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
 It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
-Special attention is paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling.
+Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling.
 The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.
+Modeling groups were then invited to contribute model descriptions and results to
 
@@
 The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.
-Modeling groups have already begun to leverage this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.
-
-The proposed workshop has advanced our knowledge of (a) the relative behaviors exhibited by atmospheric dynamical cores, (b) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (c) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
+Modeling groups have leveraged this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.
+
+The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
 Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.
@@
 
-\section{Notation}
-
-\subsection{List of Symbols}
-
+This paper is the first in a series of papers documenting the results of this workshop.
+This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.
+Section \ref{sec:Notation} describes the unified notation that our project has adopted that will be relevant throughout all papers in this series.
+The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; section \ref{sec:TemporalDiscretizations} describes temporal discretizations; and section \ref{sec:DynamicalCores} provides an overview of how these techniques are tied together in each modeling system.
+
+\section{Notation} \label{sec:Notation}
+
+This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
 Table \ref{tab:symbols} lists the symbols used in this paper.
+A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
+Great circle distance is used throughout the project and is computed via
+\begin{equation}
+R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
+\end{equation}
 
 \begin{table}[h]
-\caption{List of symbols used in this manuscript} \label{tab:symbols}
+\caption{List of symbols used by DCMIP} \label{tab:symbols}
 \begin{center}
@@
 $T_v$ & Virtual temperature \\
-$\Theta$ & Potential temperature \\
-$\Theta_v$ & Virtual potential temperature \\
+$\theta$ & Potential temperature \\
+$\theta_v$ & Virtual potential temperature \\
 $q$ & Specific humidity \\
@@
 \end{table}
-
-\subsection{List of Physical Constants}
-A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
-Constants which are specific to each test case are similarly tabulated at the beginning of each section.
 
@@
 \end{tabular*}
-
 \end{table}
 
-\subsection{Great Circle Distance}
-
-The great circle distance is used throughout the document and is given by
-\begin{equation}
-R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
-\end{equation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Dynamical Cores} \label{sec:DynamicalCores}
+
+This section provide a short description of unique features or design specifications from several modern dynamical cores.
+A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.
+
+\begin{table}[h]
+\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
+\begin{center}
+\begin{tabular}{cll}
+\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
+CSU & Colorado State University Model & Colorado State University \\
+DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
+FV$^3$ & Finite Volume Cubed Model & Geophysical Fluid Dynamics Laboratory \\
+FVM & Finite Volume Module of the Integrated Forecast System & European Center for Medium-Range Weather Forecasting \\
+GEM & Global Environmental Multiscale model & Environment Canada \\
+HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
+ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
+MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
+NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
+Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
+\hline 
+\end{tabular}
+\end{center}
+\end{table}
+
+%Do not include information on the physical parameterizations used by the modeling system.
+Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
+
+\subsection{Colorado State University Model (CSU)}
+
+The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
+The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
+The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
+Time-differencing is based on the third-order Adams-Bashforth scheme.
+Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.
+\subsection{DYNAMICO}
+
+DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
+The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical.
+DYNAMICO's design uniquely combines:
+ \begin{itemize}
+\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
+\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature,\citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
+\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
+\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
+
+\subsection{FV Cubed (FV$^3$)}
+
+The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
+The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
+Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
+The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
+Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+
+Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
+Explicit viscosity is applied every acoustic timestep.
+
+The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
+The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
+
+\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
+\label{sec:FVM}
+
+The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
+ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
+equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
+are available by means of simple switches. The formulation incorporates 
+a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
+A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
+of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
+Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
+Conjugate Residual approach. 
+The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
+The horizontal spatial discretization is fully unstructured finite-volume using the 
+median-dual approach. This is combined with a structured-grid finite-difference approach in the 
+vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
+In both the horizontal and the vertical discretization, all prognostic variables are 
+co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
+of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
+reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
+IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
+not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
+
+No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
+diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
+boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
+the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
+the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
+at every time step according to a given maximum advective CFL number (typically somewhat smaller 
+than 1). The physics time step is identical to the dynamics time step. 
+%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
+%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
+
+%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
+%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
+%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
+%efficiency and accuracy.
+
+%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
+%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
+%framework developed at ECMWF.
+
+\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
+
+The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.
+The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.
+A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
+A scalar viscosity is employed for wind components and tracers
+via an application of (\ref{eq:scalar_viscosity_direct}).  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}
+
+\subsection{High-Order Method Modeling Environment (HOMME)}
+
+The High Order Method Modeling Environment (HOMME) is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
+As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present.
+The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.
+The model employs a hybrid terrain-following / pressure coordinate.
+The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+\subsection{ICON}
+
+The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
+The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
+The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
+The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
+Further divergence damping is applied to the normal wind at every sub-step.
+Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
+The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
+
+\subsection{Model for Prediction Across Scales (MPAS)}
+
+{\color{red}[SKAMAROCK]}
+
+\subsection{Tempest}
+
+The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.
+The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
+These continuous equations are given in section \ref{sec:TempestEquations}.
+The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
+Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).
+Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
+Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
 
@@
 In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
-This section describes many of the model grids that have been employed globally.
+The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.
+The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.
+Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.
+To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.
 
@@
 
-The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude {\color{red}[Figure]}.
+The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
 Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
@@
 A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
-The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.
+The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001shallow}.
 An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
@@
 for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
+Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.
 CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.
@@
 
-As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
-reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
+As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
+reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
 roots of the Legendre polynomials. 
@@
 
-\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}
-
-In this section we describe the fluid equations utilized by nonhydrostatic models.
+\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}
+
+In this section we describe the fluid equations utilized by non-hydrostatic models.
 The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
@@
 The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e.
-following \cite{Kasahara1974} (hereafter K74).
+following \cite{kasahara1974various} (hereafter K74).
 Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
@@
 
-{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}
+%{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}
 
@@
 \left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
-=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
-& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
-& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
+%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
+%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
+%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
 =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [( \vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_s \nonumber \\
@@
 
+\subsection{Vorticity-Divergence Form}
+
+The vorticity-divergence form of the dynamical equations in the height vertical coordinate predicts the (vertical component of) vorticity and (velocity) divergence given by
+\begin{align}
+\omega_z \equiv \vb{k} \cdot \nabla_s \times \vb{u}_h,
+\end{align} and
+\begin{align}
+D \equiv \nabla_s \cdot \vb{u}_h,
+\end{align} respectively, instead of the horizontal velocity.
+
+The vorticity prediction equation is obtained by taking the curl ($\vb{k} \cdot \nabla_s \times$) of the horizontal momentum equation as
+\begin{align}
+\pdiff{\omega_z}{t} + \nabla_s \cdot (\omega_z^\ast \vb{u}_h) + \pdiff{}{z} (\omega_z^\ast w) - \omega_z \pdiff{w}{z} - \pdiff{}{z} (\vb{k} \times \vb{u}_h) \cdot \nabla_s w + J(c_p \theta_v, \pi) = 0.
+\end{align} The Jacobian   is a compact form for the curl of the pressure gradient force, where is the specific heat of air at constant pressure.
+
 \subsection{Conservation Laws in Arbitrary Vertical Coordinates}
@@
 
-\subsubsection{Tempest} \label{sec:TempestEquations}
-
-Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum with $(\partial z / \partial t)_s = 0$ (\ref{{eq:CovariantVelocityUpdateS}}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
-
-%In the horizontal, Tempest utilizes a spectral-element formulation.
-
 \subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
@@
 
-\subsubsection{ICON} \label{sec:ICONEquations}
-
-ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
-The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
-\begin{align}
-\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
-\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
-
-The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
-
 \subsubsection{FV$^3$} \label{sec:FV3Equations}
@@
 The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
-
-\subsubsection{GEM} \label{sec:GEM_equations}
-
-GEM is a semi-Lagrangian model with Eulerian log-hydrostatic-pressure coordinate $\pi_h$ described in section \ref{sec:GEM_zeta} \citep{Girard2014}.
-The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance, using
-\begin{align}
-\mu = \pdiff{p}{\pi_h} - 1,
-\end{align} in which case 
-\begin{align}
-\frac{d \vb{v}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
-\frac{d w}{dt}- g \mu =& 0, \\
-\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
-\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt} =& 0, \\
-R_dT_v+  \frac{p}{\pi} \frac{\partial \Phi}{\partial \log \pi} =& 0, \\
-\frac{d \phi}{dt}-  g w  =& 0, \\
-1+ \mu - \frac{p}{\pi}   \frac{\partial \log p}{\partial \log  \pi} =& 0, \\
-\ln \pi =& \zeta + Bs.
-\end{align}
-
-\subsection{GEM} \label{sec:GEM_equations}
-The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
-
-\noindent with horizontal velocity ${\bf V}_h$, 
-generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
-vertical velocity $w$,
-geopotential $\phi$, 
-virtual temperature $T_v$, 
-pressure $p$ and
-hydrostatic pressure $\pi$. The variable 
-$\mu$ is a measure of departure from hydrostatic balance, 
-$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
 
@@
 a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
-see \cite{prusa2008} and \cite{smolarkiewiczJCP2014}. 
+see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014}. 
 The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
@@
 and the transformation of the physical to the contravariant velocity 
-$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003} and \cite{kuehnleinJCP2012} for discussion.
+$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003all} and \cite{kuehnleinJCP2012} for discussion.
 %(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
@@
 %momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
+
+\subsubsection{GEM} \label{sec:GEM_equations}
+
+GEM is a semi-Lagrangian model with Eulerian log-hydrostatic-pressure coordinate $\pi_h$ described in section \ref{sec:GEM_zeta} \citep{Girard2014}.
+The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance, using
+\begin{align}
+\mu = \pdiff{p}{\pi_h} - 1,
+\end{align} in which case 
+\begin{align}
+\frac{d \vb{v}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
+\frac{d w}{dt}- g \mu =& 0, \\
+\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
+\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt} =& 0, \\
+R_dT_v+  \frac{p}{\pi} \frac{\partial \Phi}{\partial \log \pi} =& 0, \\
+\frac{d \phi}{dt}-  g w  =& 0, \\
+1+ \mu - \frac{p}{\pi}   \frac{\partial \log p}{\partial \log  \pi} =& 0, \\
+\ln \pi =& \zeta + Bs.
+\end{align}
+
+The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
+
+\noindent with horizontal velocity ${\bf V}_h$, 
+generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+vertical velocity $w$,
+geopotential $\phi$, 
+virtual temperature $T_v$, 
+pressure $p$ and
+hydrostatic pressure $\pi$. The variable 
+$\mu$ is a measure of departure from hydrostatic balance, 
+$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+
+\subsubsection{HOMME}
+
+HOMME uses a floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics.
+A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
+Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
+The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
+The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
+Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
+
+In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
+Fields are current co-located in the vertical and explicit time-stepping is used.
+Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+
+%This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
+Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+
+\subsubsection{ICON} \label{sec:ICONEquations}
+
+ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
+The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
+\begin{align}
+\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
+\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
+
+The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
+
+\subsubsection{Tempest} \label{sec:TempestEquations}
+
+Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum with $(\partial z / \partial t)_s = 0$ (\ref{{eq:CovariantVelocityUpdateS}}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
+
+%In the horizontal, Tempest utilizes a spectral-element formulation.
 
@@
 
-\subsection{Runge-Kutta}
-
-\subsubsection{Ullrich-Kinnmark-Gray 5 step 3rd order scheme} \label{sec:UKG53Scheme}
+\subsection{Ullrich-Kinnmark-Gray 5 step 3rd order scheme} \label{sec:UKG53Scheme}
 
@@
 With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
-The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the nonhydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
+The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
 This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
@@
 Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
-
-\subsection{Semi-Implicit time integration}
 
@@
 
-\section{Dynamical Cores}
-
-This section provide a short description of unique features or design specifications from several modern dynamical cores.
-
-%Do not include information on the physical parameterizations used by the modeling system.
-Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
-
-\subsection{Colorado State University Model (CSU)}
-
-The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
-The model is based on the non-hydrostatic Unified System of equations proposed by Arakawa and Konor (2009), which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
-The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
-Time-differencing is based on the third-order Adams-Bashforth  scheme.
-Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.
-\subsection{DYNAMICO}
-
-DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
-The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical.
-DYNAMICO's design uniquely combines:
- \begin{itemize}
-\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}
-\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature \citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization
-\item A novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
-\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
-
-\subsection{FV Cubed (FV$^3$)}
-
-The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
-The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
-Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
-Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
-
-Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
-Explicit viscosity is applied every acoustic timestep.
-
-The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
-The nonhydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
-All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
-
-\subsection{Finite-volume module of the Integrated Forecasting System}
-\label{sec:FVM}
-
-%{\color{red}[KUEHNLEIN]}
-
-The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is developed at 
-ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-are available by means of simple switches. The formulation incorporates 
-a generalised, optionally time-dependent, terrain-following vertical coordinate based on height. 
-A centred two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
-Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-Conjugate Residual approach. 
-The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
-The horizontal spatial discretisation is fully unstructured finite-volume using the 
-median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-In both the horizontal and the vertical discretisation, all prognostic variables are 
-co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
-reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
-
-No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-boundary layer parametrisation and the constant-coefficient second-order dissipation/diffusion in 
-the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
-at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-than 1). The physics time step is identical to the dynamics time step. 
-%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
-%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
-
-%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
-%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
-%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
-%efficiency and accuracy.
-
-%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
-%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
-%framework developed at ECMWF.
-
-\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
-
-In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
-The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
-on the Charney-Phillips grid.
-A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
-A scalar viscosity is employed for wind components and tracers 
-via an application of (\ref{eq:scalar_viscosity_direct}). 
-Viscosity operations are applied after the completion of the dynamic time step.
-
-\subsection{High-Order Method Modeling Environment (HOMME)}
-
-HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model.
-It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed.
-Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth.
-It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom.
-A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
-A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid.
-After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
-Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
-The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
-Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
-The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
-Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
-
-In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
-Fields are current co-located in the vertical and explicit time-stepping is used.
-Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
-Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
-
-\subsection{ICON}
-
-The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
-The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
-The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
-For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
-Further divergence damping is applied to the normal wind at every sub-step.
-Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
-
-\subsection{Model for Prediction Across Scales (MPAS)}
-
-{\color{red}[SKAMAROCK]}
-
-\subsection{Naval Research Laboratory NEPTUNE Model}
-
-{\color{red}[VINER, REINECKE]}
-
-\subsection{Ocean-Land-Atmosphere Model (OLAM)}
-
-{\color{red}[WALKO]}
-
-\subsection{Tempest}
-
-The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.
-The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
-These continuous equations are given in section \ref{sec:TempestEquations}.
-The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
-Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
-Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
-
-\subsection{Nonhydrostatic ICosahedral Atmospheric Model (NICAM) Model}
-
-{\color{red}[MIURA]}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Idealized Physical Parameterizations}
+\section{Idealized Physical Parameterizations} \label{sec:IdealizedPhysics}
+
+{\color{red}[Perhaps move this to paper 2?]}
 

### 2017-02-18 DCMIP2016-Part1-Models.tex
Cleanup and updates

@@
 
-\Author[18]{Elijah}{Goodfriend}
-\Author[18]{Hans}{Johansen}
+%\Author[18]{Elijah}{Goodfriend}
+%\Author[18]{Hans}{Johansen}
 
@@
 \item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
-\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature,\citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
+\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
 \item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
@@
 A scalar viscosity is employed for wind components and tracers
-via an application of (\ref{eq:scalar_viscosity_direct}).  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}
+via an application of a scalar Laplacian.  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}
 
@@
 
-{\color{red}[SKAMAROCK]}
+The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation, in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
+Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
+In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
+The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
+Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
+Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
 
@@
 
-The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.
+The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
 The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
@@
 The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
-Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
-Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
 
@@
 
-\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}
-
-In this section we describe the fluid equations utilized by non-hydrostatic models.
+\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}
+
+In this section we describe the fluid equations utilized by nonhydrostatic models.
 The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
@@
 The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.
-In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.
+In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis.
 Note that it is common to rewrite the pressure gradient force using the relationship
@@
 \end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
-This gives rise to the vector-invariant form,
+This gives rise to the 3D vector-invariant form,
 \begin{align} \label{eq:PrognosticVelocity2}
@@
 
-It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
+Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
+Decomposing the momentum transport term into horizontal and vertical components gives
 \begin{align}
@@
 \begin{align}
-\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
-(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
+\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \label{eq:UhDotNablaUhRelationship1} \\
+(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right] \label{eq:UhDotNablaUhRelationship2}
 \end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.
@@
 
-{\color{blue}
-\noindent \textbf{Derivation.}
-\begin{align}
-\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) + w \vb{k} \cdot \nabla \vb{u}_h + w \vb{k} \cdot \nabla (w \vb{k}) \\
-=& \vb{u}_h \cdot \nabla \vb{u}_h + \nabla( \vb{u}_h \cdot (w \vb{k}) ) - (w \vb{k}) \times (\nabla \times \vb{u}_h) - \vb{u}_h \times (\nabla \times w \vb{k}) + \nabla ( \tfrac{1}{2} w^2) + ((\nabla \times (w \vb{k})) \times (w \vb{k})) \\
-=& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}
-\end{align} where we have used
-\begin{align}
-(\nabla \times (w \vb{k})) \times (w \vb{k}) =& (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) \times (w \vb{k}) = - w \vb{k} \times (\nabla w \times \vb{k}) = (- w \vb{k} \cdot \vb{k}) \nabla w - (- w \vb{k} \cdot \nabla w) \vb{k} \\
-- \vb{u}_h \times (\nabla \times w \vb{k}) =& - \vb{u}_h \times (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) = - \vb{u}_h \times (\nabla w \times \vb{k}) = (\vb{u}_h \cdot \nabla w) \vb{k}
-\end{align}}
-
-Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
+Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1})$\cdot \vb{k}$, is
 \begin{align} \label{eq:PrognosticWEquationK2}
-\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}) \cdot \vb{k}
-\end{align}  and
+\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{w}{z} - \vb{u}_h \cdot \nabla_z w - \pdiff{\Phi}{z} - \frac{1}{\rho} \pdiff{p}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
+\end{align}  Then subtracting (\ref{eq:PrognosticWEquationK2})$\vb{k}$ from (\ref{eq:PrognosticVelocity1}) gives
 \begin{align} \label{eq:PrognosticUhEquationK2}
-\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
+\pdiff{\vb{u}_h}{t} =& - w (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{\vb{u}_h}{z} - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
 \end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.
@@
 This final expression is used to describe the rate of change of a quantity on $s$ surfaces.
-When applied to $s$, these operators yields the useful identities
+These operators then yield the useful identities
 \begin{align} \label{eq:NablaZSIdentity}
-\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z, \qquad \mbox{and} \qquad \pdiff{s}{t} = - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
+\pdiff{s}{z} =& \left( \pdiff{z}{s} \right)^{-1}, & \nabla_z s =& - \left( \pdiff{s}{z} \right) \nabla_s z, & \pdiff{s}{t} =& - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
 \end{align}  From here (\ref{eq:ChainRule}) also gives rise to
 \begin{align} \label{eq:HorizontalChainRule}
-\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
-\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.
-Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z = 0$ and $\vb{k} \cdot \nabla_s = 0$.
+\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s},
+\end{align} which can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.
+Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z A = 0$ and $\vb{k} \cdot \nabla_s A = 0$ for any scalar field $A$.
+
+From (\ref{eq:ChainRule}), it can be shown that the 2D divergence on $s$ surfaces (given by K74 eq. (3.17)) is
+\begin{align} \label{eq:DivergenceS}
+\nabla_s \cdot \vb{u}_h = \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \cdot \left( \pdiff{\vb{u}_h}{s} \right),
+\end{align} and that the 2D curl is given by
+\begin{align} \label{eq:CurlS}
+\nabla_s \times \vb{u}_h = \nabla_z \times \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \times \left( \pdiff{\vb{u}_h}{s} \right),
+\end{align} where $\nabla_z \times \vb{u}_h = \vb{k} (\vb{k} \cdot (\nabla \times \vb{u}_h))$.  Notably, these expressions are valid for both shallow- and deep-atmosphere formulations.
 
@@
 =& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
-\end{align}
-
-The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
-\begin{align} \label{eq:VerticalVelocityOrthogonalS}
-\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}_h \right] \cdot \vb{k}.
-\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
-\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
-\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
-\end{align}  Note that the last term of this expression can then be expanded using
-\begin{align}
-(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
-\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.
-
-The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived from (\ref{eq:PrognosticWEquationK2}),
-\begin{align} \label{eq:PrognosticWEquationK2S}
-\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k},
-\end{align} and from (\ref{eq:PrognosticUhEquationK2}),
-\begin{align} \label{eq:PrognosticUhEquationK2S}
-\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}).
-\end{align}  Observe that both of these equations are simplified if $(\partial z / \partial t)_s = w$.
-
-%{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}
-
-\subsection{Covariant Formulation}
-
-In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.
+\end{align}  A similar expression arises for vectors, although in this case $\vb{u}_h \cdot \nabla \vb{a} \neq \vb{u}_h \cdot \nabla_z \vb{a}$ implies we cannot use the operator $\nabla_s$ in the form (\ref{eq:ChainRule}), and instead obtain
+\begin{align} \label{eq:MaterialDerivativeVectorS}
+\diff{\vb{a}}{t} =& \left( \pdiff{\vb{a}}{t} \right)_s + \left[ \vb{u}_h \cdot \nabla \vb{a} + (\vb{u}_h \cdot \nabla_s z) (\vb{k} \cdot \nabla \vb{a}) \right] + \dot{s} \pdiff{\vb{a}}{s}.
+\end{align}
+
+%{\color{red}[Is this needed?]} Note that
+%\begin{align}
+%\nabla A = \nabla_z A + \vb{k} \pdiff{A}{z} = \nabla_s A - \pdiff{s}{z} (\nabla_s z) \pdiff{A}{s} + \vb{k} \pdiff{s}{z} \pdiff{A}{s} = \nabla_s A + ( \vb{k} - \nabla_s z) \pdiff{s}{z} \pdiff{A}{s}
+%\end{align}
+
+\subsection{2D Vector Invariant Form}
+
+The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived by applying (\ref{eq:ChainRule}) to (\ref{eq:PrognosticWEquationK2}), yielding
+\begin{align} \label{eq:PrognosticWEquationK2Sa}
+\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
+\end{align}  Similarly, from (\ref{eq:PrognosticUhEquationK2}),
+\begin{align} \label{eq:PrognosticUhEquationK2Sa}
+\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) + \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \zeta_h \vb{k} \times \vb{u}_h - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z.
+\end{align}  Observe that both of these equations simplify when $w = (\partial z / \partial t)_s$, \textit{i.e.} model levels are advected with the vertical wind.
+
+An alternative form of these equation can similarly be obtained in terms of $\dot{s}$.  Substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticWEquationK2Sa}) then gives
+\begin{align} \label{eq:PrognosticWEquationK2Sb}
+\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla_s w - \dot{s} \pdiff{w}{s} + \left( \pdiff{s}{z} \right) \left[ - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right] - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
+\end{align}  Similarly, substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticUhEquationK2Sa}) and using the identity
+\begin{align}
+(\vb{u}_h \cdot \nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} = (\nabla_s \times \vb{u}_h) \times \vb{u}_h - \zeta_h \vb{k} \times \vb{u}_h + (\nabla_s z) \pdiff{K_2}{z}
+\end{align}
+ then gives
+\begin{align} \label{eq:PrognosticUhEquationK2Sb}
+\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) - \nabla_s K_2 - \zeta_s \vb{k} \times \vb{u}_h - \dot{s} \pdiff{\vb{u}_h}{s} - \nabla_z \Phi - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z,
+\end{align} where
+\begin{align}
+\nabla_s \times \vb{u}_h = \vb{k} \zeta_s, \quad \mbox{and} \quad \zeta_s = \vb{k} \cdot (\nabla_s \times \vb{u}_h).
+\end{align}  In this case the vertical advection terms are removed when $\dot{s} = 0$, \textit{i.e.} the vertical coordinate is advected with the 3D wind $\vb{u}$.
+
+Note that under the shallow-atmosphere approximation, the first metric terms (those that include $(\vb{u}_h \cdot \nabla \vb{k})$) in (\ref{eq:PrognosticWEquationK2Sa})-(\ref{eq:PrognosticUhEquationK2Sb}) are typically dropped.
+
+\subsection{3D Vector Invariant Form}
+
+From (\ref{eq:PrognosticVelocity2}) and (\ref{eq:ChainRule}) the evolution equation for the 3D velocity vector takes the form
+%\begin{align}
+%\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}
+%\end{align} we have
+\begin{align}
+\left( \pdiff{\vb{u}}{t} \right)_s = \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}) - \nabla (K + \Phi) - \frac{1}{\rho} \nabla p - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}
+\end{align}  Then taking the dot product of this expression with $\vb{k}$ gives
+\begin{align} \label{eq:PrognosticWEquation}
+\left(\pdiff{w}{t} \right)_s =& \left( \pdiff{s}{z} \right) \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \pdiff{}{s} (K + \Phi) - \frac{1}{\rho} \pdiff{p}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}
+\end{align} where we have used $\vb{k} \cdot (\vb{k} \cdot \nabla \vb{u}) = \vb{k} \cdot \nabla w$.
+ Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
+\begin{align} \label{eq:PrognosticUhEquationKS}
+\left( \pdiff{\vb{u}_h}{t} \right)_s =& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \nabla_z (K + \Phi) - \frac{1}{\rho} \nabla_z p - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z.
+\end{align}  Note that the vorticity term in this expression can be simplified further using
+\begin{align}
+[(\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_z = - (\zeta_h + \vb{k} \cdot 2 \vg{\Omega}) (\vb{u}_h \times \vb{k}) - w \vb{k} \times (\vg{\zeta} + 2 \vg{\Omega}),
+\end{align} and
+\begin{align}
+- \vb{k} \times \vg{\zeta} =& \vb{k} \cdot \nabla \vb{u} - \nabla (\vb{k} \cdot \vb{u}) + \vb{u} \cdot \nabla \vb{k} = \pdiff{\vb{u}_h}{z} - \nabla_z w + \vb{u}_h \cdot \nabla \vb{k}.
+\end{align}
+%so that (\ref{eq:PrognosticUhEquationKS}) takes the form
+%\begin{align}
+%\left( \pdiff{\vb{u}_h}{t} \right)_s =& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \nabla_z (K + \Phi) - \frac{1}{\rho} \nabla_z p - \left[ 2 \vg{\Omega} \times \vb{u} \right]_z - \zeta_h (\vb{u}_h \times \vb{k}) + w \left[ \pdiff{\vb{u}_h}{z} - \nabla_z w + \vb{u}_h \cdot \nabla \vb{k} \right]
+%\end{align}
+
+\subsection{Covariant Component Formulation}
+
+In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (in 2D vector invariant form as  (\ref{eq:PrognosticUhEquationK2Sa}) or (\ref{eq:PrognosticUhEquationK2Sb}), or in 3D vector invariant form as (\ref{eq:PrognosticUhEquationKS})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.
 This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.
-The horizontal partial covariance operator is defined by
+We define a horizontal covariance operator by
 \begin{align}
 [\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
-\end{align} Hence the covariant horizontal velocity is given by
+\end{align} Applying this operator to the horizontal velocity gives
 \begin{align}
 \vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
-\end{align}  For a time-dependent $s$ coordinate,
+\end{align}  For a time-dependent $s$ coordinate, we obtain the identity
 \begin{align}
 \left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
-\end{align} and so, using (\ref{eq:VerticalVelocityOrthogonalS}) and (\ref{eq:PrognosticUhEquationK2S}) the evolution equation for the covariant horizontal velocity reads
-\begin{align}
-\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
-%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
-%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
-%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
-=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [( \vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_s \nonumber \\
-& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) - (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right].
-\label{eq:CovariantVelocityUpdateS}
-\end{align}  Again, observe that this equation simplifies significantly if $(\partial z / \partial t)_s = w$.
-
-%For a time-dependent $s$ coordinate,
-%\begin{align}
-%\pdiff{}{t} (\nabla_s z) =& \pdiff{}{t} \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
-%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} \\
-%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right)
-%\end{align}
-%For a time-dependent $s$ coordinate,
-%\begin{align}
-%\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) (\nabla_s z) \\
-%=& - \left( \nabla_z \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} + (\nabla_z s) \left( \pdiff{s}{z} \right)^{-2} \pdiff{}{z} \pdiff{s}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla) \left[ - (\nabla_z s) \left( \pdiff{s}{z} \right)^{-1} \right] \\
-%=& - \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-1} \nabla_z \left( \pdiff{s}{z} \right) + \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right)^{-2} (\nabla_z s) \pdiff{}{z} \pdiff{s}{z} \\
-%=& \left( \pdiff{s}{z} \right)^{-1} \left[ - \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s \nabla_s \left( \pdiff{s}{z} \right)  \right] \\
-%=& \nabla_s \left( \pdiff{z}{t} \right)_s
-%\end{align}
-
-%In the case of an Eulerian formulation with $(\partial z / \partial t)_s = 0$, we can use
-%\begin{align}
-%0 = \nabla (\vb{k} \cdot \vb{u}_h) = (\vb{k} \cdot \nabla \vb{u}_h) + (\vb{u}_h \cdot \nabla \vb{k}) + \vb{k} \times (\nabla \times \vb{u}_h)
-%\end{align} to obtain
-%\begin{align}
-%\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{v}_h)
-%\end{align}
-
-%Although a time-dependent $s$ is admitted by this equation, its fully Eulerian nature makes it difficult to use in this case since degrees of freedom are not guaranteed to align with constant $s$ surfaces.
-
-%\subsection{Floating Lagrangian Formulation}
-
-%Under the floating Lagrangian formulation, the coordinate $s$ is chosen so as to satisfy $\dot{s} = 0$.
-This can be accomplished by choosing $s$ to equal potential temperature, although this has issues with monotonicity, or an imaginary conservative tracer \citep{L2004MWR} (for instance, the altitude of the fluid parcel at its initial time).
-
-%A consequence of the use of floating Lagrangian coordinates is the need for vertical remapping to avoid severe deformation of coordinate surfaces.
-Several options are available, with the piecewise-parabolic method being perhaps the most widely adopted \citep{colella1984piecewise, L2004MWR}.
-
-%In the case of time-varying $s$ coordinates, degrees of freedom are usually stored on $s$ surfaces.  To track floating surfaces of constant $s$ we define the floating Lagrangian derivative,
-%\begin{align}
-%\frac{d_s}{dt} = \pdiff{}{t} + w \vb{k} \cdot \nabla,
-%\end{align} which is used to define the tendency in variables that are allowed to ``float'' vertically along surfaces that adjust with the advection velocity.
-From (\ref{eq:VerticalVelocityOrthogonal}), the vertical velocity equation then takes the form
-%\begin{align}
-%\frac{d_s w}{d t} =& \pdiff{w}{t} + w \pdiff{w}{z} \\
-%=& - \frac{1}{\rho} \left( \pdiff{s}{z} \right) \pdiff{p}{s} - \left( \pdiff{s}{z} \right) \pdiff{}{s} (K_2 + \Phi) - [ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} ] \cdot \vb{k}.
-%\end{align}  From (\ref{eq:PrognosticUhEquationK2}), the horizontal velocity equation takes the form
-%\begin{align}
-%\frac{d_s \vb{u}_h}{d t} =& \pdiff{\vb{u}_h}{t} + w \vb{k} \cdot \nabla \vb{u}_h \\
-%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}),
-%\end{align} where we have also used
-%\begin{align}
-%w \vb{k} \cdot \nabla \vb{u}_h = - w \vb{u}_h \cdot \nabla \vb{k} - (w \vb{k}) \times (\nabla \times \vb{u}_h).
-%\end{align}  Note that under the shallow atmosphere approximation, the $(\vb{u}_h \cdot \nabla \vb{k})$ term is typically dropped.
-
-%\begin{align}
-%=& \pdiff{\vb{u}}{t} + \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) \\
-%=& \pdiff{\vb{u}}{t} + \vb{u}_h \cdot \nabla \vb{u}_h + w (\vb{u}_h \cdot \nabla \vb{k}) + (\vb{u}_h \cdot \nabla w) \vb{k}.
-%\end{align}
-
-%Using (\ref{eq:CovariantProjection}), the material derivative along $s$ surfaces takes the form
-%\begin{align} \label{eq:MaterialDerivative}
-%\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla = \pdiff{}{t} + \vb{v}_h \cdot \nabla + \left( \dot{s} - \pdiff{s}{t} \right) \left( \pdiff{s}{z} \right)^{-1} \pdiff{}{z} = \pdiff{}{t} + \vb{v}_h \cdot \nabla + \left( \dot{s} - \pdiff{s}{t} \right) \pdiff{}{s}.
-%\end{align}  This equation is typically used to obtain a diagnostic relationship for $\dot{s}$, utilizing $w = dz/dt$,
-%\begin{align} \label{eq:DiagnosticSDot}
-%\dot{s} = \pdiff{s}{t} + \left( \pdiff{s}{z} \right) (w - \vb{v}_h \cdot \vb{k}).
-%\end{align}
-
-%The evolution equation for the covariant velocity vector can then be derived:
-%\begin{align} \label{eq:CovariantVelocityUpdate}
-%\pdiff{\vb{v}_h}{t} =& \pdiff{}{t} \left[ \vb{u} - \left( \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \right) \vb{k} \right] = \left[ \pdiff{\vb{u}}{t} \right]_s - \left( \pdiff{s}{z} \right)^{-1} \left( \vb{v}_h \cdot \nabla \pdiff{s}{t} \right) \vb{k}.
-%\end{align}  Consequently, the evolution equation is simply the projection of the evolution equation for $\vb{u}$ onto surfaces of constant $s$ plus a vertical term associated with shifting $s$ surfaces.
-In practice, only the horizontal components of $\vb{v}_h$ are treated prognostically, so this term does not appear in the evolution equations.
-%\begin{align}
-
-%=& \pdiff{\vb{u}}{t} - \left( \pdiff{\vb{u}}{t} \cdot \nabla s + \vb{u} \cdot \nabla \pdiff{s}{t} \right) \left( \vb{k} \cdot \nabla s \right)^{-1} \vb{k}  + (\vb{u} \cdot \nabla s) (\vb{k} \cdot \nabla s)^{-2} \left(\vb{k} \cdot \nabla \pdiff{s}{t} \right) \vb{k} \\
-%=& \left[ \pdiff{\vb{u}}{t} \right]_s - (\vb{k} \cdot \nabla s)^{-1} \left( \vb{u} - \frac{\vb{u} \cdot \nabla s}{\vb{k} \cdot \nabla s} \vb{k} \right) \cdot \nabla \pdiff{s}{t} \vb{k} \\
-%\end{align}
+\end{align} and so can write
+\begin{align} \label{eq:CovariantExpansion}
+\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + (\nabla_s z) \left( \pdiff{w}{t} \right)_s + w \nabla_s \left( \pdiff{z}{t} \right)_s.
+\end{align}  Then using (\ref{eq:CovariantExpansion}), (\ref{eq:PrognosticWEquation}) and (\ref{eq:PrognosticUhEquationKS}) and identity
+\begin{align}
+& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \left[ \pdiff{\vb{u}_h}{s} + (\nabla_s z) \pdiff{w}{s} \right] + w \nabla_s \left( \pdiff{z}{t} \right)_s \nonumber \\
+& \qquad \qquad  = \left( \pdiff{s}{z} \right) \left( \pdiff{z}{t} \right)_s \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\} + \nabla_s \left[ \left( \pdiff{z}{t} \right)_s w \right]
+\end{align} gives
+\begin{align}
+\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \nabla_s \left[ K - w \left( \pdiff{z}{t} \right)_s + \Phi \right] - \frac{1}{\rho} \nabla_s p - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s \\
+& \qquad + \left( \pdiff{s}{z} \right) \left( \pdiff{z}{t} \right)_s \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\}.
+\end{align}  Finally, we can expand the vorticity term and hence obtain
+\begin{align} \label{eq:CovariantUpdateEquation}
+\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \nabla_s \left[ K - w \left( \pdiff{z}{t} \right)_s + \Phi \right] - \frac{1}{\rho} \nabla_s p - \left[ 2 \vg{\Omega} \times \vb{u} \right]_s \nonumber \\
+& \qquad + \left[ \vb{k} \cdot \nabla_s \times \vb{v}_h \right] (\vb{u}_h \times \vb{k}) - \dot{s} \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\} - \left( \pdiff{z}{t} \right)_s \vb{u}_h \cdot \nabla \vb{k}.
+\end{align}
 
@@
 
-The vorticity-divergence form of the dynamical equations in the height vertical coordinate predicts the (vertical component of) vorticity and (velocity) divergence given by
-\begin{align}
-\omega_z \equiv \vb{k} \cdot \nabla_s \times \vb{u}_h,
+The vorticity-divergence form of the dynamical equations in an arbitrary vertical coordinate predicts the absolute vorticity ($\zeta_h^\ast$) and velocity divergence ($D$) given by
+\begin{align} \label{eq:AbsoluteVorticityS}
+\zeta_h^\ast = (\nabla_s \times \vb{u}_h + 2 \vg{\Omega}) \cdot \vb{k},
 \end{align} and
-\begin{align}
+\begin{align} \label{eq:DivergenceS}
 D \equiv \nabla_s \cdot \vb{u}_h,
-\end{align} respectively, instead of the horizontal velocity.
-
-The vorticity prediction equation is obtained by taking the curl ($\vb{k} \cdot \nabla_s \times$) of the horizontal momentum equation as
-\begin{align}
-\pdiff{\omega_z}{t} + \nabla_s \cdot (\omega_z^\ast \vb{u}_h) + \pdiff{}{z} (\omega_z^\ast w) - \omega_z \pdiff{w}{z} - \pdiff{}{z} (\vb{k} \times \vb{u}_h) \cdot \nabla_s w + J(c_p \theta_v, \pi) = 0.
-\end{align} The Jacobian   is a compact form for the curl of the pressure gradient force, where is the specific heat of air at constant pressure.
+\end{align} respectively, instead of the horizontal velocity.  The horizontal velocity can be obtained from the streamfunction $\psi$ and the velocity potential $\chi$ following
+\begin{align} \label{eq:VelocityStreamfunctionPotentialRelationship}
+\vb{u}_h = \vb{k} \times \nabla_s \psi + \nabla_s \chi.
+\end{align}  By using (\ref{eq:VelocityStreamfunctionPotentialRelationship}) in (\ref{eq:AbsoluteVorticityS}) and (\ref{eq:DivergenceS}), we obtain the elliptic equations that diagnose the streamfunction and velocity potential from the predicted velocity and divergence as
+\begin{align} \label{eq:LaplacianStreamfunctionPotential}
+\nabla_s^2 = \zeta_h^\ast - 2 \vg{\Omega} \cdot \vb{k}, \quad \mbox{and} \quad \nabla_s^2 \chi = D,
+\end{align} respectively.
+
+By taking the material derivative (\ref{eq:MaterialDerivativeS}) of (\ref{eq:AbsoluteVorticityS}) and using the horizontal momentum equation (\ref{eq:HorizontalVelocityOrthogonalS}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), the absolute vorticity prediction equation emerges,
+\begin{align}
+\left( \pdiff{\zeta_h^\ast}{t} \right)_s - J_s(\zeta_h^\ast, \psi) + \nabla_s \cdot (\zeta_h^\ast \nabla_s \chi) + \nabla_s \cdot \left( \dot{s} \pdiff{}{s} \nabla_s \psi \right) + \vb{k} \cdot \nabla_s \times \left( \dot{s} \pdiff{}{s} \nabla_s \chi \right) + J_s(\rho^{-1}, p) = 0,
+\end{align} where $J_s(a,b) = \vb{k} \cdot \nabla_s \times (a \nabla_s b)$ is the Jacobian operator.  It can also be shown that $\dot{s}$ relates to the vertical velocity $w$ through
+\begin{align}
+\dot{s} = \left( \pdiff{s}{z} \right) (w - w_c),
+\end{align} where
+\begin{align}
+w_c \equiv \left( \pdiff{z}{t} \right)_s + \left( \vb{k} \times \nabla_s \psi + \nabla_s \chi \right) \cdot (\nabla_s z).
+\end{align}
+
+By taking the material derivative of (\ref{eq:DivergenceS}) and using (\ref{eq:HorizontalVelocityOrthogonalS}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), we can obtain the divergence prediction equation
+\begin{align}
+\left( \pdiff{D}{t} \right)_s - J_s(\zeta_h^\ast, \chi) - \nabla_s \cdot (\zeta_h^\ast \nabla_s \psi) + \nabla_s \cdot \left( \dot{s} \pdiff{}{s} \nabla_s \chi \right) + \left( \vb{k} \times \pdiff{}{s} \nabla_s \psi \right) \cdot \nabla_s \dot{s} & \\
++ \nabla_s \cdot (\nabla_s K_2 + g \nabla_s z) + \nabla_s \cdot \left( \frac{1}{\rho} \nabla_s p \right) &= 0,
+\end{align}  where $K_2$ can be reformulated in terms of streamfunction and velocity potential as
+\begin{align}
+K_2 = \frac{1}{2} \left[ \nabla_s \cdot (\psi \nabla_s \psi) - \psi \nabla_s^2 \psi + \nabla_s \cdot (\chi \nabla_s \chi) - \chi \nabla_s^2 \chi \right] + J_s(\psi, \chi).
+\end{align}
 
@@
 \end{align} and using (\ref{eq:MaterialDerivativeS}) in the form
+\begin{align}
+\diff{\rho_s}{t} = \left( \pdiff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
+\end{align} along with (\ref{eq:DivergenceS}) leads to
 \begin{align} \label{eq:PrognosticPseudoDensity}
-\diff{\rho_s}{t} = \left( \diff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
-\end{align} along with (\ref{eq:DivergenceS}) leads to
-\begin{align}
-\left( \diff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
+\left( \pdiff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
 \end{align}
@@
 \begin{align}
-\pdiff{\rho_s q}{t} + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
+\left( \pdiff{\rho_s q}{t} \right)_s + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
 \end{align}  In particular, the prognostic equation for virtual potential temperature (or equivalently for potential temperature) reads
 \begin{align} \label{eq:PrognosticPseudoDensityTheta}
-\pdiff{\rho_s \theta_v}{t} + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
+\left( \pdiff{\rho_s \theta_v}{t} \right)_s + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
 \end{align}
@@
 generalized vertical velocity $\dot{s}$ and tracers are on thermodynamic levels.
-
-\subsection{Vertical staggering}
-
-{\color{blue}Define Unstaggered, Lorenz and Charney-Phillips staggering here with figure}
 
@@
 
+\subsection{Vertical staggering}
+
+Along with the choice of prognostic variables, the vertical discretization of the equations of motion also allows for the staggered placement of prognostic variables.
+As with hydrostatic models, certain discretizations give rise to spurious computational modes that can contaminate the solution \citep{tokioka1978some,arakawa1988baroclinic}.
+These issues are also present in non-hydrostatic models \citep{thuburn2005vertical}, suggesting care should be taken when selecting the discretization.
+
 \subsection{Model Implementations}
@@
 DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014).
-Prognostic variables are pseudo-density ?s, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
 Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.
@@
 The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{L2004MWR}.
-Prognostic equations include horizontal momentum (\ref{eq:CovariantVelocityUpdate}) in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
+Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
 The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
 
-\subsection{FVM} \label{sec:FVMEquations}
+\subsubsection{FVM} \label{sec:FVMEquations}
 
@@
 
-ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
 As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
@@
 
-Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum with $(\partial z / \partial t)_s = 0$ (\ref{{eq:CovariantVelocityUpdateS}}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for covariant horizontal velocity (\ref{eq:CovariantUpdateEquation}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
 The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
@@
 Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
-Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, or wave-mode filters.
+Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.
 Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.
@@
 
+\subsection{HOMME / Tempest}
+
+In both HOMME and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
+Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
+Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
+
 \subsection{ICON}
@@
 
-\subsection{FV3}
+\subsection{FV$^3$}
 
@@
 
-A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
+Wave absorption at the model top is also provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
 An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
@@
 
-Via an application of (\ref{eq:scalar_viscosity_direct}),
-scalar viscosity is employed for wind components and tracers. 
-A vertical sponge layer, also via an application of (\ref{eq:scalar_viscosity_direct}),
-is employed on wind components and $T_v$ with a vertical modulation on the 6 top levels 
-and a maximum damping coefficient of $.380000 \times 10^{6}\ \mbox{m\textsuperscript{2}/s}$ at the top.
-For stabilization purpose, 
-the temporal discretization of GEM presented  
-in section \ref{sec:GEM_temporal} uses an off-centering parameter $b^{A} = 0.6$.
-
-In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
-
-\subsection{Tempest}
-
-{\color{blue}Repeated application of the scalar and vector viscosity operators}
+Diffusion in GEM is handled via application of a Laplacian, for both wind components and tracers. 
+A vertical sponge layer, which also uses the formulation (\ref{eq:scalar_viscosity_direct}), is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
+For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
+Shape-preserving solutions to the tracer advection equation are also computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
 
@@
 
-{\color{red}[ALL] Describe the time-stepping scheme / temporal discretization employed by your dynamical core here.}
-
-\subsection{Ullrich-Kinnmark-Gray 5 step 3rd order scheme} \label{sec:UKG53Scheme}
-
-Explicit terms are evolved using a Runge-Kutta method which supports a large stability bound for spatial discretizations with purely imaginary eigenvalues.
-This particular scheme is based on \cite{kinnmark1984onestepA, kinnmark1984onestepB} and takes the form
-\begin{align}
-\psi^{(1)} &= \psi^{(0)} + \tfrac{\Delta t}{5} f(\psi^{(0)}), \nonumber \\
-\psi^{(2)} &= \psi^{(0)} + \tfrac{\Delta t}{5} f(\psi^{(1)}), \nonumber \\
-\psi^{(3)} &= \psi^{(0)} + \tfrac{\Delta t}{3} f(\psi^{(2)}), \\
-\psi^{(4)} &= \psi^{(0)} + \tfrac{2 \Delta t}{3} f(\psi^{(3)}), \nonumber \\
-\psi^{(5)} &= -\tfrac{1}{4} \psi^{(0)} + \tfrac{5}{4} \psi^{(1)} + \tfrac{3 \Delta t}{4} f(\psi^{(4)}). \nonumber
-\end{align}
-
-\subsection{Semi-implicit time integration of the fully compressible Euler equations in FVM}
+There is a wide diversity of temporal discretizations available among global atmospheric models
+
+\subsection{Split-Explicit Methods}
+
+\subsection{Mixed Implicit-Explicit, Forward-Backward, Semi-Implicit and Additive Runge-Kutta Schemes}
+
+Implicit-explicit schemes are a broad category of time integration schemes that divide the terms of the prognostic equations into a set of explicitly integrated terms and implicitly integrated terms.
+At the very least, terms associated with vertically propagating sound waves are included among the implicit terms.
+For the remaining terms, there is some freedom in choosing how to integrate terms associated with vertical advection and horizontally propagating sound waves.
+Semi-implicit schemes are one such class of schemes that typically incorporate horizontally propagating sound waves into the implicit solve, and so rely on a global Helmholtz-type solve.
+Additive Runge-Kutta schemes are another mechanism to ensure high-order temporal accuracy, and many such schemes have been described throughout the literature (see, for example, \cite{weller2013runge, ullrich2012operator}).
+Several examples of these schemes can be found among the DCMIP models:
+
+FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
+With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
+The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
+This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
+Physics tendencies are applied impulsively at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
+
+DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
+A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
+As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
+The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
+The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
+Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
+
+ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
+No time splitting is used with respect to sound waves, because the ratio of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
+Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
+Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
+
+Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
+
+\subsection{Semi-Implicit methods}
 
@@
 finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
-\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}~\footnote{Furthermore, 
+\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}.  Furthermore, 
 in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
-on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.}.
+on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.
 
@@
 
-\subsection{Forward-backward vertically-Lagrangian dynamics}
-
-%%%GFDL start
-
-FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
-With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
-The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
-This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
-Physics tendencies are applied ``impulsively'' at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
-%%%GFDL end
-
-\subsection{DYNAMICO}
-
-DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
-A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
-As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
-The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
-The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
-Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
-
-\subsection{ICON}
-
-The time stepping scheme of ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
-No time splitting is used with respect to sound waves, because the ration of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
-Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
-Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
-Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}
+\subsubsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}
 
@@
 
-{\color{red}[Perhaps move this to paper 2?]}
+{\color{red}[This will be moved to paper 2?]}
 

### 2017-03-06 DCMIP2016-Part1-Models.tex
- First official draft of paper

@@
 \begin{abstract}
-This paper provides a comprehensive review of the design of modern non-hydrostatic atmospheric dynamical cores, including relevant equation sets, numerical stabilization techniques and idealized physics routines.
+Atmospheric dynamical cores are a core component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
+These systems have existed in one form or another for over half of a century, with the earliest strategies now evolved into a complex ecosystem of algorithms and computational strategies.
+In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.
+To better understand modern dynamical cores, this paper aims to provide a comprehensive review of their design, including model grids, relevant equation sets, numerical stabilization techniques and idealized physics routines.
+Examples are taken from operational dynamical cores that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
 \end{abstract}
@@
 
-%{\color{red}INSTRUCTIONS FOR AUTHORS: Fill in text in section 3, 4, 5, 6, 7 and 8 below.}
-
 The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.
-Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) throughout the atmosphere, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.
+Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.
 These factors include the choice of model grid, choice of prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.
@@
 The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.
-Modeling groups were then invited to contribute model descriptions and results to
+Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.
 
@@
 Section \ref{sec:Notation} describes the unified notation that our project has adopted that will be relevant throughout all papers in this series.
-The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; section \ref{sec:TemporalDiscretizations} describes temporal discretizations; and section \ref{sec:DynamicalCores} provides an overview of how these techniques are tied together in each modeling system.
+Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models.
+The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
+Conclusions and discussion then follow in section \ref{sec:Conclusions}.
 
@@
 These continuous equations are given in section \ref{sec:TempestEquations}.
-The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
+The implementation includes both fully explicit time integration, using a third-order scheme based on \cite{kinnmark1984onestepB} and described in \cite{guerra2016high}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
 
@@
 \end{align} which can be alternatively written as
-\begin{align}
+\begin{align} \label{eq:PotentialTemperaturePrognostic}
 \diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
@@
 \end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
-\begin{align}
+\begin{align} \label{eq:VirtualTemperaturePrognosticEq}
 \frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
@@
 
+\subsection{Conservation Laws in Arbitrary Vertical Coordinates}
+
+Using (\ref{eq:DivergenceS}), we observe that the 3D divergence on the sphere takes the form
+\begin{align} \label{eq:ThreeDDivergence}
+\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \frac{1}{\alpha} \pdiff{}{z} (\alpha w),
+\end{align} where $\alpha = 1$ for shallow-atmosphere models and $\alpha = r^2 = (a + z)^2$ for deep-atmosphere models.  Using $w = dz/dt$, this last term also takes the form
+\begin{align} \label{eq:GeneralVerticalTermX}
+\frac{1}{\alpha} \pdiff{}{z} (\alpha w) = \pdiff{w}{z} + \frac{w}{\alpha} \pdiff{\alpha}{z} = \pdiff{w}{z} + \frac{1}{\alpha} \diff{\alpha}{t}.
+\end{align}
+
+Using (\ref{eq:NablaZSIdentity}) to rewrite (\ref{eq:DefinitionDotS}) gives rise to
+\begin{align}
+w = \left( \pdiff{z}{t} \right)_s + \vb{u}_h \cdot \nabla_s z + \dot{s} \left( \pdiff{s}{z} \right)^{-1},
+\end{align} which is then differentiated to yield K74 eq. (3.16),
+\begin{align} \label{eq:WDerivativeS}
+\pdiff{w}{z} = \left( \pdiff{s}{z} \right) \left[ \diff{}{t} \left( \pdiff{s}{z} \right)^{-1} + \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \pdiff{\dot{s}}{s} = 0.
+\end{align}  Substituting this expression into the continuity equation (\ref{eq:ContinuityEquation}), and using (\ref{eq:ThreeDDivergence}), (\ref{eq:GeneralVerticalTermX}), and (\ref{eq:WDerivativeS}) then leads to
+\begin{align}
+\diff{}{t} \left[ \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \left[ \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \pdiff{\dot{s}}{s} = 0.
+\end{align}  Defining the \textit{pseudodensity} as 
+\begin{align}
+\rho_s = \alpha \left( \pdiff{s}{z} \right)^{-1} \rho,
+\end{align} and using (\ref{eq:MaterialDerivativeS}) in the form
+\begin{align}
+\diff{\rho_s}{t} = \left( \pdiff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
+\end{align} along with (\ref{eq:DivergenceS}) leads to
+\begin{align} \label{eq:PrognosticPseudoDensity}
+\left( \pdiff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
+\end{align}
+
+Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
+\begin{align}
+\left( \pdiff{\rho_s q}{t} \right)_s + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
+\end{align}  In particular, the prognostic equation for virtual potential temperature (or equivalently for potential temperature) reads
+\begin{align} \label{eq:PrognosticPseudoDensityTheta}
+\left( \pdiff{\rho_s \theta_v}{t} \right)_s + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
+\end{align}
+
 \subsection{2D Vector Invariant Form}
 
-The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived by applying (\ref{eq:ChainRule}) to (\ref{eq:PrognosticWEquationK2}), yielding
+The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are derived by applying (\ref{eq:ChainRule}) to (\ref{eq:PrognosticWEquationK2}), yielding
 \begin{align} \label{eq:PrognosticWEquationK2Sa}
@@
 \end{align} and
-\begin{align} \label{eq:DivergenceS}
+\begin{align} \label{eq:DivergenceSx}
 D \equiv \nabla_s \cdot \vb{u}_h,
@@
 \vb{u}_h = \vb{k} \times \nabla_s \psi + \nabla_s \chi.
-\end{align}  By using (\ref{eq:VelocityStreamfunctionPotentialRelationship}) in (\ref{eq:AbsoluteVorticityS}) and (\ref{eq:DivergenceS}), we obtain the elliptic equations that diagnose the streamfunction and velocity potential from the predicted velocity and divergence as
+\end{align}  By using (\ref{eq:VelocityStreamfunctionPotentialRelationship}) in (\ref{eq:AbsoluteVorticityS}) and (\ref{eq:DivergenceSx}), we obtain the elliptic equations that diagnose the streamfunction and velocity potential from the predicted velocity and divergence as
 \begin{align} \label{eq:LaplacianStreamfunctionPotential}
@@
 
-By taking the material derivative (\ref{eq:MaterialDerivativeS}) of (\ref{eq:AbsoluteVorticityS}) and using the horizontal momentum equation (\ref{eq:HorizontalVelocityOrthogonalS}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), the absolute vorticity prediction equation emerges,
+By taking the material derivative (\ref{eq:MaterialDerivativeS}) of (\ref{eq:AbsoluteVorticityS}) and using the horizontal momentum equation (\ref{eq:PrognosticUhEquationK2}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), the absolute vorticity prediction equation emerges,
 \begin{align}
@@
 
-By taking the material derivative of (\ref{eq:DivergenceS}) and using (\ref{eq:HorizontalVelocityOrthogonalS}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), we can obtain the divergence prediction equation
+By taking the material derivative of (\ref{eq:DivergenceSx}) and using (\ref{eq:PrognosticUhEquationK2}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), we can obtain the divergence prediction equation
 \begin{align}
@@
 
-\subsection{Conservation Laws in Arbitrary Vertical Coordinates}
-
-In this section we rewrite the continuity equation in arbitrary vertical coordinates.
-The divergence of the contravariant velocity field on surfaces of constant $s$ can be obtained from (\ref{eq:ChainRule}), which is valid for both shallow- and deep-atmosphere formulations and equal to K74 eq.
-(3.17),
-\begin{align} \label{eq:DivergenceS}
-\nabla_s \cdot \vb{u}_h = \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \cdot \left( \pdiff{\vb{u}_h}{s} \right).
-\end{align}
-
-On the sphere the 3D divergence takes the form
-\begin{align} \label{eq:ThreeDDivergence}
-\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \frac{1}{\alpha} \pdiff{}{z} (\alpha w),
-\end{align} where $\alpha = 1$ for shallow-atmosphere models and $\alpha = r^2 = (a + z)^2$ for deep-atmosphere models.  Using $w = dz/dt$, this last term also takes the form
-\begin{align} \label{eq:GeneralVerticalTermX}
-\frac{1}{\alpha} \pdiff{}{z} (\alpha w) = \pdiff{w}{z} + \frac{w}{\alpha} \pdiff{\alpha}{z} = \pdiff{w}{z} + \frac{1}{\alpha} \diff{\alpha}{t}.
-\end{align}
-
-Using (\ref{eq:NablaZSIdentity}) to rewrite (\ref{eq:DefinitionDotS}) gives rise to
-\begin{align}
-w = \left( \pdiff{z}{t} \right)_s + \vb{u}_h \cdot \nabla_s z + \dot{s} \left( \pdiff{s}{z} \right)^{-1},
-\end{align} which is then differentiated to yield K74 eq. (3.16),
-\begin{align} \label{eq:WDerivativeS}
-\pdiff{w}{z} = \left( \pdiff{s}{z} \right) \left[ \diff{}{t} \left( \pdiff{s}{z} \right)^{-1} + \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \pdiff{\dot{s}}{s} = 0.
-\end{align}  Substituting this expression into the continuity equation (\ref{eq:ContinuityEquation}), and using (\ref{eq:ThreeDDivergence}), (\ref{eq:GeneralVerticalTermX}), and (\ref{eq:WDerivativeS}) then leads to
-\begin{align}
-\diff{}{t} \left[ \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \left[ \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \pdiff{\dot{s}}{s} = 0.
-\end{align}  Defining the \textit{pseudodensity} as 
-\begin{align}
-\rho_s = \alpha \left( \pdiff{s}{z} \right)^{-1} \rho,
-\end{align} and using (\ref{eq:MaterialDerivativeS}) in the form
-\begin{align}
-\diff{\rho_s}{t} = \left( \pdiff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
-\end{align} along with (\ref{eq:DivergenceS}) leads to
-\begin{align} \label{eq:PrognosticPseudoDensity}
-\left( \pdiff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
-\end{align}
-
-Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
-\begin{align}
-\left( \pdiff{\rho_s q}{t} \right)_s + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
-\end{align}  In particular, the prognostic equation for virtual potential temperature (or equivalently for potential temperature) reads
-\begin{align} \label{eq:PrognosticPseudoDensityTheta}
-\left( \pdiff{\rho_s \theta_v}{t} \right)_s + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
+\subsection{Momentum Form}
+
+The momentum form of the prognostic equations emerges by combining the prognostic velocity equations with a continuity equation.
+Essentially any of the continuity equations can be chosen, as long as the mass field represented by the equation is everywhere non-zero.
+However, the most common options are moist psuedo-density \citep{ullrich2012mcore} or dry pseudo-density \citep{skamarock2012multiscale}.
+Here we denote our density variable by $\tilde{\rho}_s$, and assume no external sources or sinks of $\tilde{\rho}$.
+Multiplying (\ref{eq:PrognosticWEquationK2Sb}) through by $\tilde{\rho}_s$ and using (\ref{eq:PrognosticPseudoDensity}) gives
+\begin{align} \label{eq:PrognosticRhoWEquationK2Sb}
+\left( \pdiff{\tilde{\rho}_s w}{t} \right)_s =& \tilde{\rho}_s \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \nabla_s \cdot (\tilde{\rho}_s \vb{u}_h w) - \pdiff{}{s} (\tilde{\rho}_s \dot{s} w) + \tilde{\rho}_s \left( \pdiff{s}{z} \right) \left[ - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right] - (2 \vg{\Omega} \times \tilde{\rho}_s \vb{u}_h) \cdot \vb{k}.
+\end{align}  Similarly, from (\ref{eq:PrognosticUhEquationK2Sb}) we have
+\begin{align} \label{eq:PrognosticRhoUhEquationK2Sb}
+\left( \pdiff{\tilde{\rho}_s \vb{u}_h}{t} \right)_s =& - \tilde{\rho}_s w (\vb{u}_h \cdot \nabla \vb{k}) - \tilde{\rho}_s \nabla_s K_2 - \zeta_s \vb{k} \times \tilde{\rho}_s \vb{u}_h - \vb{u}_h \nabla_s \cdot (\tilde{\rho}_s \vb{u}_h) - \pdiff{}{s} (\tilde{\rho}_s \dot{s} \vb{u}_h) - \tilde{\rho}_s \left( \nabla_z \Phi + \frac{1}{\rho} \nabla_z p \right) - [2 \vg{\Omega} \times \tilde{\rho}_s \vb{u}]_z,
 \end{align}
@@
 \subsection{Examples of vertical coordinates}
+
+We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
 
@@
 This simplifies the treatment of the bottom boundary condition.
-Common terrain-following coordinates include \cite{galchen1975}, which takes the form
+Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
 \begin{align}
 s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
-\end{align}  Another coordinate with smoother variation away from the surface was also introduced by \cite{schar2002new}.
+\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
+Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
 
@@
 
+%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
+%\begin{align}
+%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
+%\end{align} where the integrals are taken over the column.
+
 \subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
 
-{\color{red}[Fix notation]}
+{\color{red}[Please assist with fixing notation]}
 
@@
 with $s_T= \log \pi_T$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-Figure \ref{fig:gem_charney-phillips} shows the Charney-Phillips grid in GEM model, giving the position occupied by each variable in the vertical domain.
-Horizontal velocity $\vb{v}_h$, geopotential $\Phi$ and $q = \log(p/\pi_h)$ are on momentum levels.
-Temperature $T$, vertical velocity $w$,
-generalized vertical velocity $\dot{s}$ and tracers are on thermodynamic levels.
-
-%----------------------------------------------------
-\begin{figure}[]
-\includegraphics[height=6cm]{gem_charney-phillips.png}
-\caption{A modification of the Charney-Phillips grid found in GEM}
-\label{fig:gem_charney-phillips}
-\end{figure}
-%----------------------------------------------------
+
+\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
+
+In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
+The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
+\begin{align}
+s = \int^{\infty}_{z} \rho_0 g dz,
+\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
+\begin{align}
+\dot{s} = \frac{ds}{dt} = 0,
+\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
+Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
 
@@
 These issues are also present in non-hydrostatic models \citep{thuburn2005vertical}, suggesting care should be taken when selecting the discretization.
+For this reason, co-located discretizations, where all degrees of freedom are geometrically co-located, are not frequently adopted in global atmospheric models.
+In particular, it is common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
+These approaches can be further augmented as needed, for instance by shifting the vertical velocity and thermodynamic variables from the bottom boundary to an intermediate level, as in the GEM model.
+Note that, in general, tracer variables are co-located with the buoyancy variable.
+
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[width=5in]{VerticalStaggering.pdf}
+\caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
+\label{fig:verticalstaggering}
+\end{figure}
+%----------------------------------------------------
 
 \subsection{Model Implementations}
+
+In this section we describe the prognostic equations that have been chosen for the dynamical cores that participated in DCMIP.
 
@@
 \subsubsection{FVM} \label{sec:FVMEquations}
+
+{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
 
@@
 
+{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
+
 GEM is a semi-Lagrangian model with Eulerian log-hydrostatic-pressure coordinate $\pi_h$ described in section \ref{sec:GEM_zeta} \citep{Girard2014}.
@@
 \mu = \pdiff{p}{\pi_h} - 1,
-\end{align} in which case 
-\begin{align}
-\frac{d \vb{v}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
+\end{align} in which case the equations of GEM model \citep{Girard2014} are
+\begin{align}
+\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
 \frac{d w}{dt}- g \mu =& 0, \\
-\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
+\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
 \frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt} =& 0, \\
@@
 1+ \mu - \frac{p}{\pi}   \frac{\partial \log p}{\partial \log  \pi} =& 0, \\
-\ln \pi =& \zeta + Bs.
-\end{align}
-
-The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
-
-\noindent with horizontal velocity ${\bf V}_h$, 
-generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
-vertical velocity $w$,
-geopotential $\phi$, 
-virtual temperature $T_v$, 
-pressure $p$ and
-hydrostatic pressure $\pi$. The variable 
-$\mu$ is a measure of departure from hydrostatic balance, 
-$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+\ln \pi =& \zeta + B \tilde{s}.
+\end{align}
+
+%\noindent with horizontal velocity ${\bf V}_h$, 
+%generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+%vertical velocity $w$,
+%geopotential $\phi$, 
+%virtual temperature $T_v$, 
+%pressure $p$ and
+%hydrostatic pressure $\pi$. The variable 
+%$\mu$ is a measure of departure from hydrostatic balance, 
+%$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
 
@@
 
-HOMME uses a floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics.
-A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing.
-Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form.
-The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics.
+The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure {\color{red}[Equation?]}.
+
+In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ {\color{red}[Equation?]}, potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
+Fields are co-located in the vertical and explicit time-stepping is used.
+
+%Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+
+% The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
 Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
@@
 
-In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$.
-Fields are current co-located in the vertical and explicit time-stepping is used.
-Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-
 %This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
@@
 
-ICON is a shallow-atmosphere Eulerian model based on \cite{gassmann2008} with terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-As with tempest, specific heat capacities and ideal gas constant are approximated by their dry values.
+ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
 The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
@@
 \end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
-
 The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
@@
 Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for covariant horizontal velocity (\ref{eq:CovariantUpdateEquation}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
 
@@
 In both HOMME and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.
-Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
+Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
+\begin{align}
+\nabla^2 \vb{u}_h = \nabla \nabla \cdot \vb{u}_h + \nabla \times \nabla \times \vb{u}_h.
+\end{align}  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
+Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
 
@@
 
+{\color{red}[Check notation and clean up]}
+
 The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014} and summarized here:
-
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and is summarized here.
 For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
@@
 
-The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
+The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
 \begin{align} \label{smagn2}
@@
 Diffusion in GEM is handled via application of a Laplacian, for both wind components and tracers. 
-A vertical sponge layer, which also uses the formulation (\ref{eq:scalar_viscosity_direct}), is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
+A vertical sponge layer, which also uses the Laplacian, is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
 For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
@@
 
-There is a wide diversity of temporal discretizations available among global atmospheric models
-
-\subsection{Split-Explicit Methods}
+A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.
+In this section we briefly discuss several modern approaches that are presently employed.
 
@@
 
+MPAS uses a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
+At the beginning of each Runge-Kutta sub-cycle, tendencies are computed for each of the prognostic variables and stored for the duration of the sub-cycle.
+Several iterations of an acoustic loop are then performed with a time-step much smaller than require for the Runge-Kutta sub-cycle.
+Within the acoustic loop, an implicit solve for vertically-integrated sound waves is performed to avoid timestep constraints that may arise from vertically-propagating sound waves.
+
 Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
 
-\subsection{Semi-Implicit methods}
+\subsection{The FVM Semi-Implicit method}
+
+{\color{red}[Could we emphasize here how this approach differs from the approaches described in section 7.1?  Or should it be included in section 7.1?]}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}
-
-Consider a frictionless adiabatic prognostic equation of the form:
+\subsubsection {A semi-Lagrangian implicit time discretization in the GEM model} \label{sec:GEM_temporal}
+
+{\color{red}[Introductory sentence that states the difference between this scheme and the approaches above.]}   Consider a frictionless adiabatic prognostic equation of the form:
 \begin{equation} 
@@
 
+{\color{blue}[Blue text to be moved to paper 2]
 \section{Idealized Physical Parameterizations} \label{sec:IdealizedPhysics}
-
-{\color{red}[This will be moved to paper 2?]}
 
@@
 The details of this discretization are somewhat complicated, and so we refer to implementation details in Appendix D of \cite{reed2012idealized}.
-In addition, we supply the DCMIP modeling groups with the complete ``simple-physics'' package as used in the model CAM which can serve as a template routine.
-
-\conclusions  %% \conclusions[modified heading if necessary]
-TEXT
+In addition, we supply the DCMIP modeling groups with the complete ``simple-physics'' package as used in the model CAM which can serve as a template routine.}
+
+\conclusions \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
+
+{\color{red}[To be written]}
 

### 2017-03-14 DCMIP2016-Part1-Models.tex
corrections inst+model names, references

--- 2017-02-18 DCMIP2016-Part1-Models.tex
@@
 
-\Author[12]{Christian}{Kuehnlein}
+\Author[12]{Christian}{K\"uhnlein}
 
@@
 \affil[11]{Deutscher Wetterdienst (DWD)}
-\affil[12]{European Center for Medium-Range Weather Forecasting}
+\affil[12]{European Centre for Medium-Range Weather Forecasts}
 \affil[13]{University of Miami}
@@
 FV$^3$ & Finite Volume Cubed Model & Geophysical Fluid Dynamics Laboratory \\
-FVM & Finite Volume Module of the Integrated Forecast System & European Center for Medium-Range Weather Forecasting \\
+FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
 GEM & Global Environmental Multiscale model & Environment Canada \\
@@
 The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}. 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
 The horizontal spatial discretization is fully unstructured finite-volume using the 
@@
 finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
-\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}.  Furthermore, 
+\citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  Furthermore, 
 in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 

### 2017-03-21 DCMIP2016-Part1-Models.tex
- Added edits from James Kent and Vivian Lee / GEM

@@
 \affil[13]{University of Miami}
-\affil[14]{Environment Canada}
+\affil[14]{Environment and Climate Change Canada}
 \affil[15]{University of Tokyo}
@@
 To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
-It was built on previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
+It was built on earlier DCMIP workshops (held in 2008 and 2012), and previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
 Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling.
@@
 A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.
+The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.
 
@@
 $\omega$ & Vertical pressure velocity  \\
-$\delta$ & Divergence\\
-$\zeta$ & Relative vorticity\\
+$D$ & Divergence\\
+$\zeta$ & Vertical component of relative vorticity\\
 $p$ & Pressure (pressure of moist air if $q>0$) \\
@@
 $P_{ls}$ & Large-scale precipitation rate \\
+$q$ & Water vapor mixing ratio \\
 $q_c$ & Cloud water mixing ratio \\
 $q_r$ & Rain water mixing ratio \\
+$q_i$ & General tracer mixing ratio \\
 \hline 
@@
 DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
-FV$^3$ & Finite Volume Cubed Model & Geophysical Fluid Dynamics Laboratory \\
+FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory \\
 FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
-GEM & Global Environmental Multiscale model & Environment Canada \\
+GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
 HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
@@
 \item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
-\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
+\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, {\color{red}to a certain extent [what does this mean?]}, an energy-conserving spatial discretization, and
 \item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
@@
 
-The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate.
-The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms.
+The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations.
+The discretization is done in the horizontal on the equiangular gnomonic cubed-sphere grid \citep{PL2007JCP}, and a flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
+The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
 Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
+The discretization is on the C-D grid as described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
+that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
 Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
@@
 
-The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
+The prognostic variables are horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
 The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
@@
 
-The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.
-The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.
+The GEM model \citep{Girard2014}, which is used for operational forecasting at Environment and Climate Change Canada, uses the Yin-Yang grid \citep{kageama2004yinyang} and horizontal discretization on an Arakawa C grid.
+The vertical coordinate is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type
+and the vertical discretization is based on the Charney-Phillips grid.
 A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
-A scalar viscosity is employed for wind components and tracers
-via an application of a scalar Laplacian.  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}
+It gives rise to an {\it{iterative}} process where each step requires the solution of a
+linear system of equations that is reduced to a Helmholtz problem for one composite variable.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid \citep{Qaddouri2008schwarz}.
+The dynamics and physics are time split.
+An explicit hyperviscosity is employed for wind components and tracers
+via applications of the Laplacian operator.
+Viscosity operations are applied after the completion of the physics time step.
 
@@
 
-The High Order Method Modeling Environment (HOMME) is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
+The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
 As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present.
@@
 
-The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
+The ICON model {\color{red}[Reference?]} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
 The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
@@
 
-Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.
+Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.
 In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
@@
 $ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.
-The Yang component covers an analogous area, but is rotated so as to cover the region of the sphere outside of the Yin grid.
+The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
 This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
@@
 \begin{align}
-\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
-\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
+\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
+\end{align} where $\vb{u}$ denotes the 3D vector velocity.  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
 \begin{align} \label{eq:TracerTransport}
@@
 \pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
-\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).
+\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis  that would arise from the momentum transport term $\vb{u} \cdot \nabla \vb{u}$ in (\ref{eq:PrognosticVelocity1}) (note that these derivatives sometimes appear as Christoffel symbols).
 In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
@@
 
-{\color{red}[Please assist with fixing notation]}
-
-The vertical coordinate of GEM model (denoted $\zeta$ in the model documentation) is of a log-hydrostatic-pressure type and is defined by
+The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
+Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
 \begin{align}\label{eq:gem_zeta}
- \log \pi_h = s + B(s) \sigma_p, \qquad B(s) =\left( \frac {s-s_T} {(\log p_0) - s_T}\right)^r, \qquad \sigma_p = \log \left( \frac {\pi_h} {p_{0}} \right),
-\end{align}
-with $s_T= \log \pi_T$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
+\end{align}
+\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
+and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
+and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
 
@@
 As with hydrostatic models, certain discretizations give rise to spurious computational modes that can contaminate the solution \citep{tokioka1978some,arakawa1988baroclinic}.
-These issues are also present in non-hydrostatic models \citep{thuburn2005vertical}, suggesting care should be taken when selecting the discretization.
+The choice of vertical staggering may also impact many physically-relevant properties of the model near the grid scale, such as the phase speed of Rossby waves \citep{thuburn2005vertical}.
+Finally, the choice of vertical staggering can have impacts on the physics-dynamics coupling \cite{holdaway2013comparison1,holdaway2013comparison2}.
+Taken altogether, these issues suggest care should be taken when selecting the discretization.
 For this reason, co-located discretizations, where all degrees of freedom are geometrically co-located, are not frequently adopted in global atmospheric models.
@@
 
-The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{L2004MWR}.
+The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{lin2004vertically}.
 Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
@@
 
-{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
-
-GEM is a semi-Lagrangian model with Eulerian log-hydrostatic-pressure coordinate $\pi_h$ described in section \ref{sec:GEM_zeta} \citep{Girard2014}.
-The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance, using
-\begin{align}
-\mu = \pdiff{p}{\pi_h} - 1,
-\end{align} in which case the equations of GEM model \citep{Girard2014} are
-\begin{align}
-\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{v}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} =& 0, \\
-\frac{d w}{dt}- g \mu =& 0, \\
-\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta }  =& 0, \\
-\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt} =& 0, \\
-R_dT_v+  \frac{p}{\pi} \frac{\partial \Phi}{\partial \log \pi} =& 0, \\
-\frac{d \phi}{dt}-  g w  =& 0, \\
-1+ \mu - \frac{p}{\pi}   \frac{\partial \log p}{\partial \log  \pi} =& 0, \\
-\ln \pi =& \zeta + B \tilde{s}.
-\end{align}
-
-%\noindent with horizontal velocity ${\bf V}_h$, 
-%generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
-%vertical velocity $w$,
-%geopotential $\phi$, 
-%virtual temperature $T_v$, 
-%pressure $p$ and
-%hydrostatic pressure $\pi$. The variable 
-%$\mu$ is a measure of departure from hydrostatic balance, 
-%$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$,
+described in section \ref{sec:GEM_zeta}.
+The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
+\begin{align}
+\mu = \pdiff{p}{s} - 1,
+\end{align}
+\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
+and in which case
+the equations of GEM model \citep{Girard2014} are:
+\begin{align}
+\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
+\frac{d w}{dt}- g \mu &= 0, \\
+\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
+\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
+\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
+\frac{d \Phi}{dt}-  g w & = 0.
+\end{align}
 
@@
 The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
-If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
-This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping.
-Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces.
-An option to convert kinetic energy lost to the hyperdiffusion to heat is available.
-FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.
+If additional damping is desired, or if the non-monotonic advection is used, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
+The hyperdiffusion is of the same order as but much smaller than the divergence damping.
+Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep.
 
@@
 
-Diffusion in GEM is handled via application of a Laplacian, for both wind components and tracers. 
-A vertical sponge layer, which also uses the Laplacian, is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
+An explicit hyperviscosity in GEM is handled, via applications of the Laplacian operator,
+for both wind components and tracers.
+A vertical sponge layer, which uses a Laplacian operator,
+is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
 For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
-Shape-preserving solutions to the tracer advection equation are also computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.
+The quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}
+is used operationally to ensure tracer monotonicity
+for specific humidity and different hydrometeors.
+Other options are now available
+in GEM including a mass conserving monotonic scheme \citep{Sorenson2013} and
+a global mass fixer \citep{Bermejo2002MASS}. Those approaches have been evaluated using
+chemical constituents such as ozone \citep{deGrandpre2016OZONE}.
 
@@
 In this section we briefly discuss several modern approaches that are presently employed.
+{\color{red}[reword?]}
 
@@
 
-{\color{red}[Introductory sentence that states the difference between this scheme and the approaches above.]}   Consider a frictionless adiabatic prognostic equation of the form:
-\begin{equation} 
+GEM differs from the approaches above by using a semi-Lagrangian advection.
+Any model equations, prognostic or diagnostic, is written in the form
+\begin{align}
 \frac {dF}{dt}+G=0,
 \label{eq:gem_temporal_O}
-\end{equation}
-\noindent where $F$ represents one of the prognostic variables and $G$ represents the remaining terms, some of which are non-linear.
-Based on the semi-Lagrangian scheme,
-The equation (\ref {eq:gem_temporal_O}) is discretized as:
-\begin{equation} 
-\frac{F^A-F^{D}} {\Delta t}+ b^A G^A + (1-b^A) G^D = 0,
+\end{align}
+\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
+The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
+\begin{align}
+\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
 \label{eq:gem_temporal_D}
-\end{equation}
-with $A \, \mbox{(Arrival)} = (\vg{r},t)$, $D \, \mbox{(Departure)} = (\vg{r}-\Delta \vg{r},t-\Delta t)$ 
-and $\vg{r}$ is the 3D grid position. 
-The displacements $\Delta \vg{r}$ are obtained by the iterative process:
-\begin{equation} 
-\Delta  \vg{r}^{i} =  \left [ b^A \vg{V}(\vg{r},t) + (1-b^A) \vg{V}(\vg{r}-\Delta \vg{r}^{i-1},t-\Delta t) \right ] \Delta t,
-\label{eq:gem_temporal_traj} 
-\end{equation} 
-where  $\vg{V}$ is the 3D wind vector.
-From (\ref {eq:gem_temporal_D}), we regroup unknown terms on the left-hand side 
-and known terms on the right-hand side, that is:
-\begin{align} 
-\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D}= R,
+\end{align}
+where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
+and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
+due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
+having occurred during the timestep $\Delta t$.
+$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
+The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
+\begin{align}
+\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
+\label{eq:gem_temporal_d_O}
+\end{align}
+discretized in the same way (trapezoidal method) though without off-centering:
+\begin{align}
+\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
+\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
+\label{eq:gem_temporal_d_D}
+\end{align}
+
+\noindent The process is of course a {\it{multi-step iterative}}
+one
+since both positions and velocities at departure positions (past time $t-\Delta t$)
+are unknown as well as, of course, the velocities at arrival positions (time $t$).
+Once a first estimate of the departure positions is obtained,
+the model equations are solved to obtain a first estimate of the velocities at time $t$.
+The model equations must be solved simultaneously and this is only possible for the linear part $L$
+which becomes a {\it {matrix inversion problem}}.
+Hence a suitable linearization is considered. The unknown (arrival) linear $L$ and non-linear $N$
+parts are then separated from the known (first departure estimate) remaining $R$ part.
+Thus, first separating space-times, (\ref{eq:gem_temporal_D}) is rewritten as follows
+\begin{align}
+\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D} \equiv R^{D},
 \label{eq:gem_temporal_AD}
 \end{align}
-\noindent where $\tau= \Delta t b^A$ and $\beta = (1-b^A)/b^A$. 
-Cubic Lagrange interpolation is used for upstream evaluations $(F^{D}, G^{D})$.
-Each left-hand side term $\frac{F^A} {\tau}+ G^A$ is then split into a linear part $L$ and a nonlinear residual part $N$:   
-\begin{align} 
-\frac{F^A} {\tau} + G^A = L + N = R. 
+where $\tau = \left ( 1/2 + \epsilon\right)\Delta t$ and $\beta=\left ( 1/2 - \epsilon\right)/\left ( 1/2 + \epsilon\right)$.
+Second separating linear from non-linear parts, we get:
+\begin{align}
+L^A + N^A = R^D,
 \label{eq:gem_temporal_LN}
 \end{align}
-The linearization is done around a constant in time, horizontally homogeneous, reference state. 
-This results in the non-linear problem:
-\begin{align}
-L^{i}=R-N^{i-1},
-\label{eq:gem_temporal_solve}
-\end{align}
-which is solved with two iterations.   
-In each iteration, the
-linear system (\ref{eq:gem_temporal_solve})  
-is reduced to an Helmholtz problem 
-for one composite variable.
-It 
-is solved 
-with a direct solver
-using the Schwarz-type domain decomposition method on Yin-Yang grid \citep{Qaddouri2008schwarz}.
-The composite variable solution is used to update the prognostic variables at time $t$.  
-Our procedure involves another iterative process linked to the estimation of $\vg{V}(\vg{r},t)$ in the displacement calculations (\ref{eq:gem_temporal_traj}). 
-This results in redoing the sequence of operations (\ref{eq:gem_temporal_traj})-(\ref{eq:gem_temporal_solve}). 
-At each time step, the cubic Lagrange interpolation is used to update the static halo region of both panels of the Yin-Yang grid 
-\citep{Qaddouri2011operational}.
+\noindent
+with $L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}$ and, of course, $N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}$.
+Note that both $F$ and $G$ may require linearization. $L^A$ may then be obtained if $N^A$ is first guessed:
+Once $L^A$ is found, an estimate of $N^A$ is obtained and $L^A$ is recalculated.
+This is called the {\it{non-linear iteration process}}
+(one iteration is usually sufficient).
+The overall process is then repeated once starting from a new estimate of the departure positions.
+
+There are two intensive calculation sections in this process:
+the so-called semi-Lagrangian calculations
+(twice estimating departure positions, twice interpolating right-hand sides $R$ on departure positions),
+and solution of the linear system (four times).
+Each time, the linear system is reduced to a Helmholtz problem for one composite variable.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid
+\citep{Qaddouri2008schwarz}.
+The composite variable solution is then used to update the prognostic variables (back substitution).
+At the end of the time-step, the static halo region of both panels of the Yin-Yang grid is updated \citep{Qaddouri2011operational}.
+All required interpolations throughout the semi-Lagrangian process and between Yin and Yang grids are cubic interpolations.
 

### 2017-04-13 DCMIP2016-Part1-Models.tex
- Cleaned up section 2 by tabulating important model properties, moved equation derivations to an appendix, and began framing a section on horizontal discretization

@@
 This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.
-Section \ref{sec:Notation} describes the unified notation that our project has adopted that will be relevant throughout all papers in this series.
-Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models.
+For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
+Section \ref{sec:DynamicalCores} provides a brief overview of each of the participating models.
 The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
 Conclusions and discussion then follow in section \ref{sec:Conclusions}.
-
-\section{Notation} \label{sec:Notation}
-
-This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
-Table \ref{tab:symbols} lists the symbols used in this paper.
-A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
-Great circle distance is used throughout the project and is computed via
-\begin{equation}
-R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
-\end{equation}
 
@@
 $w$ & Vertical velocity \\
+$\vb{u}$ & 3D wind vector \\
+$\vb{u}_h$ & Horizontal wind vector \\
+$\vb{v}_h$ & Horizontal wind vector with covariant components \\
 $\omega$ & Vertical pressure velocity  \\
-$D$ & Divergence\\
+$D$ & Divergence of the horizontal wind vector\\
 $\zeta$ & Vertical component of relative vorticity\\
 $p$ & Pressure (pressure of moist air if $q>0$) \\
+$e$ & Internal energy \\
 $\rho$ & Total air density \\
 $\rho_d$ & Dry air density \\
+$\rho_s$ & Pseudo-density \\
 $T$ &Temperature \\
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Dynamical Cores} \label{sec:DynamicalCores}
+
+This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.
+Further details on these choices can be found in subsequent sections.
+Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).
+Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
+{\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The particular choice of equation set,  that have been made by these models.
+
+A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.
+
+\begin{table}[h]
+\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
+\begin{center}
+\begin{tabular}{cll}
+\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
+CSU & Colorado State University Model & Colorado State University, USA \\
+DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
+FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
+FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
+GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
+HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
+ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
+MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
+NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
+Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
+\hline 
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.
+Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).
+Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).
+More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
+\begin{center}
+\begin{tabular}{cccccc}
+\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
+& & & & method & staggering \\ \hline
+CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
+DYNAMICO & H/NH {\color{red}(D)} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
+FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
+GEM & NH & $\vb{u}_h$, $w$, $\log \partial s / \partial \zeta$, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
+HOMME & H & $u$, $v$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
+MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
+NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
+Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+\hline 
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
+\begin{center}
+\begin{tabular}{cccccc}
+\hline Acronym & Vertical Coordinate & Vertical Staggering \\ \hline 
+CSU & fixed height & Lorenz \\
+DYNAMICO & {\color{red}floating mass?} & Lorenz \\
+FV$^3$ & floating mass & Lorenz \\
+FVM & (floating) height & Co-located \\
+GEM & log pressure & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering})  \\
+HOMME & floating mass & Co-located \\
+ICON & fixed height & Lorenz \\
+MPAS & fixed height & Lorenz \\
+NICAM & fixed height & Lorenz \\
+Tempest & fixed height & Lorenz \\
+\hline 
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{Standard options for diffusion, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
+\begin{center}
+\begin{tabular}{cccccc}
+\hline Acronym & Standard options for diffusion, & Temporal discretization \\
+& stabilization, filters, or fixers & \\ \hline 
+CSU & 4th-order horizontal hyperdiffusion & 3rd-order Adams-Bashforth (AB3) \\
+DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
+FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
+FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
+GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
+HOMME & Hyperdiffusion & KGU53 \citep{guerra2016high} \\
+ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
+MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
+NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
+Tempest & Hyperdiffusion & ARS232 \citep{ascher1997implicit} \\
+\hline 
+\end{tabular}
+\end{center}
+\end{table}
+
+%Do not include information on the physical parameterizations used by the modeling system.
+Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
+
+\subsection{Colorado State University Model (CSU)}
+
+The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
+The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
+The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
+Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.
+The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
+\subsection{DYNAMICO}
+
+DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
+DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.
+It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature,\citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
+It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
+In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
+
+\subsection{FV Cubed (FV$^3$)}
+
+The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
+The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
+Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
+The discretization is on the C-D grid as described by \citet{LR1997QJR} (see section \ref{sec:CDGrid}).
+Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
+Explicit viscosity is applied every acoustic timestep.
+
+The prognostic horizontal winds are stored in the native Gnomonic local coordinate.
+The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.
+
+\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
+\label{sec:FVM}
+
+The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
+ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
+equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
+are available by means of simple switches. The formulation incorporates 
+a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
+A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
+of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
+Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
+Conjugate Residual approach. The dynamics time step is adapted 
+at every time step according to a given maximum advective CFL number (typically somewhat smaller 
+than 1). The physics time step is identical to the dynamics time step.
+
+The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
+The horizontal spatial discretization is fully unstructured finite-volume using the 
+median-dual approach. This is combined with a structured-grid finite-difference approach in the 
+vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
+In both the horizontal and the vertical discretization, all prognostic variables are 
+co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
+of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
+reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
+IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
+not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
+
+No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
+diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
+boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
+the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
+the form of a Rayleigh-type forcing to the prognostic variables.
+
+%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
+%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
+
+%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
+%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
+%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
+%efficiency and accuracy.
+
+%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
+%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
+%framework developed at ECMWF.
+
+\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
+
+The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.
+The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang}, with Arakawa C-grid staggering.
+The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
+and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
+A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
+It gives rise to an {\it{iterative}} process where each step requires the solution of a
+linear system of equations that is reduced to a Helmholtz problem for one composite variable.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
+The dynamics and physics are time split.
+To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
+via applications of the Laplacian operator.
+Viscosity operations are applied after the completion of the physics time step.
+
+\subsection{High-Order Method Modeling Environment (HOMME)}
+
+The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
+As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present.
+The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.
+The model employs a hybrid terrain-following pressure coordinate in the vertical.
+The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.
+
+%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
+The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
+
+\subsection{ICON}
+
+The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
+Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
+Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels.
+The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
+The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
+Further divergence damping is applied to the normal wind at every sub-step.
+Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
+The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
+
+\subsection{Model for Prediction Across Scales (MPAS)}
+
+The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
+Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
+In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
+The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
+Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
+Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
+
+\subsection{NICAM}
+
+NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}.
+The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}.
+Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}.
+All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids.
+The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}.
+The Arakawa C-grid staggering is used vertically.
+A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended.
+The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}.
+The implicit time discretization is adopted for the vertically propagating one.
+A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.
+
+\subsection{Tempest}
+
+The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
+These continuous equations are given in section \ref{sec:TempestEquations}.
+The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Model Grids} \label{sec:ModelGrids}
+
+Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.
+In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
+The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.
+The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.
+Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.
+To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.
+
+\subsection{Latitude-longitude grid} \label{sec:grid_rll}
+
+The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
+Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
+This grid is employed by the UK Met Office \citep{davies2005new}.
+
+\subsection{Cubed-sphere grid} \label{sec:grid_cs}
+
+The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed, PL2007JCP} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.
+More information on this choice of grid can be found in \cite{ullrich2014global}.
+On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.
+The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
+The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.
+Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
+This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
+\end{center}
+\caption{(Left) A cubed-sphere grid with 16 elements along each cubed-sphere edge.
+(Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.
+(Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
+\end{figure}
+
+%FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}.
+The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm.
+The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
+The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.
+
+\subsection{Icosahedral (triangular) grid} \label{sec:grid_ico_tri}
+
+For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
+These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
+For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle.
+By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
+
+Several methods are available for subdividing the triangular regions.
+One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement.
+Typical choices are $n$ = 2, 3 or 5.
+All additional $m$ refinement steps use $n$ = 2, i.e.
+are bisection steps.
+A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
+The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001shallow}.
+An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
+For a discussion of the effective resolution see \cite{dipankar2015large}.
+The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.
+The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model.
+The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.
+
+\subsection{Icosahedral (hexagonal) grid / geodesic grid} \label{sec:grid_geo}
+
+The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.
+Their cells are hexagonal, except for twelve pentagonal cells.
+Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.
+A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.
+The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).
+Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this the geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
+
+It is useful to optimize icosahedral-hexagonal grids as well.
+DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh.
+This improves the homogeneity of the grid (e.g.
+ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
+
+\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids} \label{sec:grid_ccvt}
+
+Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.
+For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.
+However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.
+This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.
+Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e.
+for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
+Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.
+CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:grid_geo}, since both typically contain a large number of hexagonal elements.
+However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
+\end{center}
+\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
+\end{figure}
+
+\subsection{Octahedral reduced Gaussian grid} \label{sec:grid_oct}
+\label{subsection:octahedralreducedgaussiangrid}
+
+As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
+reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
+roots of the Legendre polynomials. 
+The two grids differ in the arrangement of the points along the latitudes, which follows a simple
+rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
+are added with every latitude towards the equator, whereby the spacing between points along the 
+latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
+is suitable for transformations involving spherical harmonics, and has been introduced for global 
+medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
+Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
+the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
+
+\begin{figure}
+\centering
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
+ \end{subfigure}%
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
+ \end{subfigure}
+\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
+connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
+24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
+reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
+\label{fig:octahedralgrid}
+\end{figure}
+
+\subsection{Yin-Yang grid} \label{sec:grid_yinyang}
+
+The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).
+These components are combined to cover a spherical surface with partial overlap along their borders.
+The Yin component covers the latitude-longitude region
+\begin{equation}\label{eq:gem_yinyang}
+(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
+\frac{\pi}{4}+\delta_{\theta})  \cap
+(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
+\frac{3\pi}{4}+\delta_{\lambda}),
+\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.
+For instance, a common configuration employed by the GEM model for DCMIP fixes
+$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.
+The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
+This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
+
+\begin{figure}[]
+\includegraphics[height=6cm]{gem_yinyang.png}
+\caption{ Yin-Yang grid }
+\label{fig:gem_yinyang}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Variable staggering and vertical coordinates}
+
+\subsection{Horizontal staggering} \label{sec:HorizontalStaggering}
+
+More information on horizontal staggerings can be found in \cite{arakawa1977computational}.
+An analysis of the dispersion properties of each staggering in the context of geophysical flows can be found in \cite{randall1994geostrophic}.
+
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
+\caption{Various options for horizontal staggering of variables among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
+\label{fig:horizontalstaggering}
+\end{figure}
+%----------------------------------------------------
+
+\subsubsection{C-D-Grid in the FV$^3$ model} \label{sec:CDGrid}
+
+which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
+that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep
+
+\subsection{Vertical coordinates}
+
+We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
+
+\subsubsection{Terrain-following $z$ coordinates}
+
+In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
+This simplifies the treatment of the bottom boundary condition.
+Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
+\begin{align}
+s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
+\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
+Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
+
+\subsubsection{Mass-based coordinates}
+
+Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
+\begin{align}
+s = \int^{\infty}_{z} \rho g dz.
+\end{align}  Under this definition,
+\begin{align}
+\pdiff{s}{z} = - \rho g.
+\end{align}
+
+%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
+%\begin{align}
+%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
+%\end{align} where the integrals are taken over the column.
+
+\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
+
+The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
+Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
+\begin{align}\label{eq:gem_zeta}
+\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
+\end{align}
+\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
+and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
+and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+
+\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
+
+In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
+The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
+\begin{align}
+s = \int^{\infty}_{z} \rho_0 g dz,
+\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
+\begin{align}
+\dot{s} = \frac{ds}{dt} = 0,
+\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
+Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
+
+\subsection{Vertical staggering} \label{sec:VerticalStaggering}
+
+Along with the choice of prognostic variables, the vertical discretization of the equations of motion also allows for the staggered placement of prognostic variables.
+As with hydrostatic models, certain discretizations give rise to spurious computational modes that can contaminate the solution \citep{tokioka1978some,arakawa1988baroclinic}.
+The choice of vertical staggering may also impact many physically-relevant properties of the model near the grid scale, such as the phase speed of Rossby waves \citep{thuburn2005vertical}.
+Finally, the choice of vertical staggering can have impacts on the physics-dynamics coupling \cite{holdaway2013comparison1,holdaway2013comparison2}.
+Taken altogether, these issues suggest care should be taken when selecting the discretization.
+For this reason, co-located discretizations, where all degrees of freedom are geometrically co-located, are not frequently adopted in global atmospheric models.
+In particular, it is common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
+These approaches can be further augmented as needed, for instance by shifting the vertical velocity and thermodynamic variables from the bottom boundary to an intermediate level, as in the GEM model.
+Note that, in general, tracer variables are co-located with the buoyancy variable.
+
+%----------------------------------------------------
+\begin{figure}[]
+\includegraphics[width=5in]{VerticalStaggering.pdf}
+\caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
+\label{fig:verticalstaggering}
+\end{figure}
+%----------------------------------------------------
+
+\subsection{Model Implementations}
+
+In this section we describe the prognostic equations that have been chosen for the dynamical cores that participated in DCMIP.
+
+\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
+
+DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014).
+Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.
+
+\subsubsection{FV$^3$} \label{sec:FV3Equations}
+
+The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{lin2004vertically}.
+Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
+The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
+
+\subsubsection{FVM} \label{sec:FVMEquations}
+
+{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
+
+The fully compressible Euler equations solved in FVM are given as
+\begin{subequations} \label{fvm:compressible}
+\begin{align}
+ & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
+\label{fvm:mass}  \\
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\varepsilon q'_v - q_c - q_r) \right)
+- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
++ \mathbf{M} \right)~,
+\label{fvm:momentum} \\
+%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
+& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
+\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
+- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
++  \mathcal{H} \right)~,
+\label{fvm:thermodynamic} \\
+& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
+\label{fvm:gaslaw}
+\end{align}
+\end{subequations}
+which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
+and dry entropy \eqref{fvm:thermodynamic}.
+Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
+physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
+pressure perturbation $\pi'$, 
+with the thermodynamic variables related by the gas law 
+\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
+Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
+respectively. All primed variables correspond to deviations from an ambient state (denoted by
+subscript "a")  that satisfies 
+a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
+see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014}. 
+The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
+Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
+coefficient 
+\begin{equation}
+\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
+\end{equation}
+in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
+the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
+form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
+$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
+$\varepsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
+The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
+coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
+simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
+Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
+$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
+and the transformation of the physical to the contravariant velocity 
+$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003all} and \cite{kuehnleinJCP2012} for discussion.
+%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
+The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
+of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
+in \eqref{fvm:thermodynamic} represents the diffusion of heat.
+
+%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
+%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
+
+\subsubsection{GEM} \label{sec:GEM_equations}
+
+GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$,
+described in section \ref{sec:GEM_zeta}.
+The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
+\begin{align}
+\mu = \pdiff{p}{s} - 1,
+\end{align}
+\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
+and in which case
+the equations of GEM model \citep{Girard2014} are:
+\begin{align}
+\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
+\frac{d w}{dt}- g \mu &= 0, \\
+\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
+\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
+\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
+\frac{d \Phi}{dt}-  g w & = 0.
+\end{align}
+
+\subsubsection{HOMME}
+
+HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics.
+The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure {\color{red}[Equation?]}.
+
+In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
+This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}.
+In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
+Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ {\color{red}[Equation?]}, potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
+Fields are co-located in the vertical and explicit time-stepping is used.
+
+%Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
+
+% The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
+Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
+The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
+Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
+
+%This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
+Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+
+\subsubsection{ICON} \label{sec:ICONEquations}
+
+ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
+The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
+\begin{align}
+\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
+\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
+The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
+
+\subsubsection{Tempest} \label{sec:TempestEquations}
+
+Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for covariant horizontal velocity (\ref{eq:CovariantUpdateEquation}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+
+%In the horizontal, Tempest utilizes a spectral-element formulation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Diffusion, Stabilization, Filters and Fixers} \label{sec:DiffusionStabilization}
+
+Most dynamical cores implement specialized techniques for diffusion or stabilization.
+Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
+Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
+Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.
+Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.
+A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.
+In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.
+
+\subsection{HOMME / Tempest}
+
+In both HOMME and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
+Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
+\begin{align}
+\nabla^2 \vb{u}_h = \nabla \nabla \cdot \vb{u}_h + \nabla \times \nabla \times \vb{u}_h.
+\end{align}  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
+Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+
+\subsection{ICON}
+
+{\color{red}[Check notation and clean up]}
+
+The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and is summarized here.
+For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
+\end{align}
+$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
+
+ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
+The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
+
+The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
+\begin{align} \label{smagn2}
+F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
+\end{align}
+\begin{align}
+F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
+\end{align}
+\begin{align}
+F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
+\end{align}
+An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
+
+A fourth-order computational diffusion is also available for vertical wind speed $w$.
+This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
+This term appears as
+\begin{align}
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
+\end{align}
+
+\subsection{FV$^3$}
+
+Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations.
+The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.
+FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion.
+The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
+If additional damping is desired, or if the non-monotonic advection is used, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
+The hyperdiffusion is of the same order as but much smaller than the divergence damping.
+Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep.
+
+Wave absorption at the model top is also provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
+An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
+
+FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
+This is only done before the physics is called and is not used in idealized simulations.
+
+\subsection{GEM}
+
+An explicit hyperviscosity in GEM is handled, via applications of the Laplacian operator,
+for both wind components and tracers.
+A vertical sponge layer, which uses a Laplacian operator,
+is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
+For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
+The quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}
+is used operationally to ensure tracer monotonicity
+for specific humidity and different hydrometeors.
+Other options are now available
+in GEM including a mass conserving monotonic scheme \citep{Sorenson2013} and
+a global mass fixer \citep{Bermejo2002MASS}. Those approaches have been evaluated using
+chemical constituents such as ozone \citep{deGrandpre2016OZONE}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Temporal Discretizations} \label{sec:TemporalDiscretizations}
+
+A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.
+In this section we briefly discuss several modern approaches that are presently employed.
+{\color{red}[reword?]}
+
+\subsection{Mixed Implicit-Explicit, Forward-Backward, Semi-Implicit and Additive Runge-Kutta Schemes}
+
+Implicit-explicit schemes are a broad category of time integration schemes that divide the terms of the prognostic equations into a set of explicitly integrated terms and implicitly integrated terms.
+At the very least, terms associated with vertically propagating sound waves are included among the implicit terms.
+For the remaining terms, there is some freedom in choosing how to integrate terms associated with vertical advection and horizontally propagating sound waves.
+Semi-implicit schemes are one such class of schemes that typically incorporate horizontally propagating sound waves into the implicit solve, and so rely on a global Helmholtz-type solve.
+Additive Runge-Kutta schemes are another mechanism to ensure high-order temporal accuracy, and many such schemes have been described throughout the literature (see, for example, \cite{weller2013runge, ullrich2012operator}).
+Several examples of these schemes can be found among the DCMIP models:
+
+FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
+With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
+The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
+This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
+Physics tendencies are applied impulsively at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
+
+DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
+A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
+As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
+The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
+The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
+Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
+
+ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
+No time splitting is used with respect to sound waves, because the ratio of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
+Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
+Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
+
+MPAS uses a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
+At the beginning of each Runge-Kutta sub-cycle, tendencies are computed for each of the prognostic variables and stored for the duration of the sub-cycle.
+Several iterations of an acoustic loop are then performed with a time-step much smaller than require for the Runge-Kutta sub-cycle.
+Within the acoustic loop, an implicit solve for vertically-integrated sound waves is performed to avoid timestep constraints that may arise from vertically-propagating sound waves.
+
+Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
+
+\subsection{The FVM Semi-Implicit method} \label{sec:FVMSemiImplicit}
+
+{\color{red}[Could we emphasize here how this approach differs from the approaches described in section 7.1?  Or should it be included in section 7.1?]}
+
+In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
+compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
+A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
+for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
+moist dynamics. 
+The generic two-time-level second-order template algorithm employed in the integration is given as
+\begin{equation}
+\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
++0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
+\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
+\label{fvm:solscalar}
+\end{equation}
+In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
+and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
+finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
+\citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  Furthermore, 
+in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
+on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.
+
+The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
+mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
+and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
+momentum \eqref{fvm:momentum}, and moisture equations enter
+\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
+which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
+is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
+velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
+Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
+The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
+an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
+equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
+The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
+Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
+Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
+executed in an outer iteration.
+
+%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
+%slow rotational modes.
+
+%The integration scheme of FVM approximates the generic conservation law
+%\begin{equation}
+%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
+%\label{scalar}
+%\end{equation}
+%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
+%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
+%second-order template algorithm
+%\begin{equation}
+%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
+%+0.5\,\delta t\,R^{\Psi}|
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection {A semi-Lagrangian implicit time discretization in the GEM model} \label{sec:GEM_temporal}
+
+GEM differs from the approaches above by using a semi-Lagrangian advection.
+Any model equations, prognostic or diagnostic, is written in the form
+\begin{align}
+\frac {dF}{dt}+G=0,
+\label{eq:gem_temporal_O}
+\end{align}
+\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
+The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
+\begin{align}
+\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
+\label{eq:gem_temporal_D}
+\end{align}
+where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
+and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
+due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
+having occurred during the timestep $\Delta t$.
+$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
+The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
+\begin{align}
+\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
+\label{eq:gem_temporal_d_O}
+\end{align}
+discretized in the same way (trapezoidal method) though without off-centering:
+\begin{align}
+\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
+\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
+\label{eq:gem_temporal_d_D}
+\end{align}
+
+\noindent The process is of course a {\it{multi-step iterative}}
+one
+since both positions and velocities at departure positions (past time $t-\Delta t$)
+are unknown as well as, of course, the velocities at arrival positions (time $t$).
+Once a first estimate of the departure positions is obtained,
+the model equations are solved to obtain a first estimate of the velocities at time $t$.
+The model equations must be solved simultaneously and this is only possible for the linear part $L$
+which becomes a {\it {matrix inversion problem}}.
+Hence a suitable linearization is considered. The unknown (arrival) linear $L$ and non-linear $N$
+parts are then separated from the known (first departure estimate) remaining $R$ part.
+Thus, first separating space-times, (\ref{eq:gem_temporal_D}) is rewritten as follows
+\begin{align}
+\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D} \equiv R^{D},
+\label{eq:gem_temporal_AD}
+\end{align}
+where $\tau = \left ( 1/2 + \epsilon\right)\Delta t$ and $\beta=\left ( 1/2 - \epsilon\right)/\left ( 1/2 + \epsilon\right)$.
+Second separating linear from non-linear parts, we get:
+\begin{align}
+L^A + N^A = R^D,
+\label{eq:gem_temporal_LN}
+\end{align}
+\noindent
+with $L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}$ and, of course, $N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}$.
+Note that both $F$ and $G$ may require linearization. $L^A$ may then be obtained if $N^A$ is first guessed:
+Once $L^A$ is found, an estimate of $N^A$ is obtained and $L^A$ is recalculated.
+This is called the {\it{non-linear iteration process}}
+(one iteration is usually sufficient).
+The overall process is then repeated once starting from a new estimate of the departure positions.
+
+There are two intensive calculation sections in this process:
+the so-called semi-Lagrangian calculations
+(twice estimating departure positions, twice interpolating right-hand sides $R$ on departure positions),
+and solution of the linear system (four times).
+Each time, the linear system is reduced to a Helmholtz problem for one composite variable.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid
+\citep{Qaddouri2008schwarz}.
+The composite variable solution is then used to update the prognostic variables (back substitution).
+At the end of the time-step, the static halo region of both panels of the Yin-Yang grid is updated \citep{Qaddouri2011operational}.
+All required interpolations throughout the semi-Lagrangian process and between Yin and Yang grids are cubic interpolations.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+
+\section{Notation} \label{sec:Notation}
+
+This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
+Table \ref{tab:symbols} lists the symbols used in this paper.
+A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
+Great circle distance is used throughout the project and is computed via
+\begin{equation}
+R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
+\end{equation}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Dynamical Cores} \label{sec:DynamicalCores}
-
-This section provide a short description of unique features or design specifications from several modern dynamical cores.
-A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.
-
-\begin{table}[h]
-\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
-\begin{center}
-\begin{tabular}{cll}
-\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
-CSU & Colorado State University Model & Colorado State University \\
-DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
-FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory \\
-FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
-GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
-HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
-ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
-MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
-NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
-Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-%Do not include information on the physical parameterizations used by the modeling system.
-Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
-
-\subsection{Colorado State University Model (CSU)}
-
-The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
-The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
-The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
-Time-differencing is based on the third-order Adams-Bashforth scheme.
-Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.
-\subsection{DYNAMICO}
-
-DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
-The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical.
-DYNAMICO's design uniquely combines:
- \begin{itemize}
-\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
-\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, {\color{red}to a certain extent [what does this mean?]}, an energy-conserving spatial discretization, and
-\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
-\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
-
-\subsection{FV Cubed (FV$^3$)}
-
-The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations.
-The discretization is done in the horizontal on the equiangular gnomonic cubed-sphere grid \citep{PL2007JCP}, and a flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
-The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
-Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid as described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
-that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
-Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
-
-Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
-Explicit viscosity is applied every acoustic timestep.
-
-The prognostic variables are horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass.
-The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
-All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.
-
-\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
-\label{sec:FVM}
-
-The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
-ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-are available by means of simple switches. The formulation incorporates 
-a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
-A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
-Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-Conjugate Residual approach. 
-The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-The horizontal spatial discretization is fully unstructured finite-volume using the 
-median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-In both the horizontal and the vertical discretization, all prognostic variables are 
-co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
-reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
-
-No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
-the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
-at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-than 1). The physics time step is identical to the dynamics time step. 
-%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
-%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
-
-%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
-%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
-%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
-%efficiency and accuracy.
-
-%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
-%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
-%framework developed at ECMWF.
-
-\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
-
-The GEM model \citep{Girard2014}, which is used for operational forecasting at Environment and Climate Change Canada, uses the Yin-Yang grid \citep{kageama2004yinyang} and horizontal discretization on an Arakawa C grid.
-The vertical coordinate is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type
-and the vertical discretization is based on the Charney-Phillips grid.
-A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
-It gives rise to an {\it{iterative}} process where each step requires the solution of a
-linear system of equations that is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid \citep{Qaddouri2008schwarz}.
-The dynamics and physics are time split.
-An explicit hyperviscosity is employed for wind components and tracers
-via applications of the Laplacian operator.
-Viscosity operations are applied after the completion of the physics time step.
-
-\subsection{High-Order Method Modeling Environment (HOMME)}
-
-The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
-As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present.
-The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.
-The model employs a hybrid terrain-following / pressure coordinate.
-The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
-%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
-\subsection{ICON}
-
-The ICON model {\color{red}[Reference?]} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid.
-The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density.
-The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
-For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
-Further divergence damping is applied to the normal wind at every sub-step.
-Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.
-
-\subsection{Model for Prediction Across Scales (MPAS)}
-
-The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation, in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
-Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
-In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
-The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
-Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
-Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
-
-\subsection{Tempest}
-
-The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
-The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.
-These continuous equations are given in section \ref{sec:TempestEquations}.
-The implementation includes both fully explicit time integration, using a third-order scheme based on \cite{kinnmark1984onestepB} and described in \cite{guerra2016high}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Model Grids} \label{sec:ModelGrids}
-
-Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.
-In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
-The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.
-The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.
-Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.
-To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.
-
-\subsection{Latitude-longitude grid}
-
-The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
-Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
-This grid is employed by the UK Met Office \citep{davies2005new}.
-
-\subsection{Cubed-sphere grid}
-
-The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.
-More information on this choice of grid can be found in \cite{ullrich2014global}.
-On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.
-The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
-The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.
-Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
-This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
-\end{center}
-\caption{(Left) A cubed-sphere grid with 16 elements along each cubed-sphere edge.
-(Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.
-(Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
-\end{figure}
-
-%FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}.
-The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm.
-The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
-The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.
-
-\subsection{Icosahedral (triangular) grid}
-
-For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
-These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
-For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle.
-By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
-
-Several methods are available for subdividing the triangular regions.
-One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement.
-Typical choices are $n$ = 2, 3 or 5.
-All additional $m$ refinement steps use $n$ = 2, i.e.
-are bisection steps.
-A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
-The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001shallow}.
-An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
-For a discussion of the effective resolution see \cite{dipankar2015large}.
-The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.
-The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model.
-The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.
-
-\subsection{Icosahedral (hexagonal) grid / geodesic grid} \label{sec:IcosahedralHexagonalGrid}
-
-The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.
-Their cells are hexagonal, except for twelve pentagonal cells.
-Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.
-A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.
-The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).
-Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this the geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
-
-It is useful to optimize icosahedral-hexagonal grids as well.
-DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh.
-This improves the homogeneity of the grid (e.g.
-ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
-
-\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids}
-
-Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.
-For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.
-However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.
-This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.
-Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e.
-for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
-Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.
-CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.
-However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
-\end{center}
-\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
-\end{figure}
-
-\subsection{Octahedral reduced Gaussian grid}
-\label{subsection:octahedralreducedgaussiangrid}
-
-As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
-reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
-roots of the Legendre polynomials. 
-The two grids differ in the arrangement of the points along the latitudes, which follows a simple
-rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
-are added with every latitude towards the equator, whereby the spacing between points along the 
-latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
-is suitable for transformations involving spherical harmonics, and has been introduced for global 
-medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
-Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
-the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
- \end{subfigure}%
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
- \end{subfigure}
-\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
-connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
-24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
-reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
-\label{fig:octahedralgrid}
-\end{figure}
-
-\subsection{Yin-Yang grid} \label{sec:GEM_model_grid}
-
-The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).
-These components are combined to cover a spherical surface with partial overlap along their borders.
-The Yin component covers the latitude-longitude region
-\begin{equation}\label{eq:gem_yinyang}
-(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
-\frac{\pi}{4}+\delta_{\theta})  \cap
-(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
-\frac{3\pi}{4}+\delta_{\lambda}),
-\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.
-For instance, a common configuration employed by the GEM model for DCMIP fixes
-$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.
-The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
-This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
-
-\begin{figure}[]
-\includegraphics[height=6cm]{gem_yinyang.png}
-\caption{ Yin-Yang grid }
-\label{fig:gem_yinyang}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}
 
-In this section we describe the fluid equations utilized by nonhydrostatic models.
+In this appendix we provide a detailed derivation of the fluid equations utilized by nonhydrostatic models.
 The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
@@
 Key thermodynamic variables include dry air density $\rho_d$, moist density $\rho$, pressure $p$, vapor pressure $e$, temperature $T$, virtual temperature $T_v$, Exner pressure $\pi$, potential temperature $\theta$, and virtual potential temperature $\theta_v$.
-Common ratios $\kappa = R^\ast / c_p^\ast$, $\epsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
+Common ratios $\kappa = R^\ast / c_p^\ast$, $\varepsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
 
@@
 \begin{align}
-T_v \approx& T \left(1 + \frac{(1-\epsilon)}{\epsilon} q \right),
+T_v \approx& T \left(1 + \frac{(1-\varepsilon)}{\varepsilon} q \right),
 \end{align} which arises from the exact relationship
 \begin{align}
-T_v = \frac{T}{1 - \frac{e}{p} (1 - \epsilon)},
-\end{align} upon applying $e/p = q / \epsilon$ and using a Taylor expansion around $q = 0$.
+T_v = \frac{T}{1 - \frac{e}{p} (1 - \varepsilon)},
+\end{align} upon applying $e/p = q / \varepsilon$ and using a Taylor expansion around $q = 0$.
 
@@
 \begin{align} \label{eq:DeepAtmosphereGeopotential}
-\Phi = g a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \phi + \vb{j} \cos \phi),
-\end{align} where $g$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\phi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
+\Phi = g a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \varphi + \vb{j} \cos \varphi),
+\end{align} where $g$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\varphi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
 For models that don't utilize a height-based vertical coordinate, the geopotential is generally treated as a prognostic variable, with an evolution equation that emerges from the definition $w = dz/dt$,
@@
 \begin{align} \label{eq:ShallowAtmosphereGeopotential}
-\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \phi \vb{k},
+\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \varphi \vb{k},
 \end{align} where $z$ is the altitude above the surface.
-In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \phi$ is the Coriolis parameter.
+In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \varphi$ is the Coriolis parameter.
 The evolution equation for the shallow atmosphere geopotential is then
@@
 \pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{w}{z} - \vb{u}_h \cdot \nabla_z w - \pdiff{\Phi}{z} - \frac{1}{\rho} \pdiff{p}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
-\end{align}  Then subtracting (\ref{eq:PrognosticWEquationK2})$\vb{k}$ from (\ref{eq:PrognosticVelocity1}) gives
+\end{align}  Then subtracting (\ref{eq:PrognosticWEquationK2})$\cdot \vb{k}$ from (\ref{eq:PrognosticVelocity1}) gives
 \begin{align} \label{eq:PrognosticUhEquationK2}
@@
 \end{align}
-
-\subsection{Examples of vertical coordinates}
-
-We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
-
-\subsubsection{Terrain-following $z$ coordinates}
-
-In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
-This simplifies the treatment of the bottom boundary condition.
-Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
-\begin{align}
-s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
-\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
-Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
-
-\subsubsection{Mass-based coordinates}
-
-Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
-\begin{align}
-s = \int^{\infty}_{z} \rho g dz.
-\end{align}  Under this definition,
-\begin{align}
-\pdiff{s}{z} = - \rho g.
-\end{align}
-
-%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
-%\begin{align}
-%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
-%\end{align} where the integrals are taken over the column.
-
-\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
-
-The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
-Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
-\begin{align}\label{eq:gem_zeta}
-\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
-\end{align}
-\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
-and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
-and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-
-\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
-
-In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
-The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
-\begin{align}
-s = \int^{\infty}_{z} \rho_0 g dz,
-\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
-\begin{align}
-\dot{s} = \frac{ds}{dt} = 0,
-\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
-Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
-
-\subsection{Vertical staggering}
-
-Along with the choice of prognostic variables, the vertical discretization of the equations of motion also allows for the staggered placement of prognostic variables.
-As with hydrostatic models, certain discretizations give rise to spurious computational modes that can contaminate the solution \citep{tokioka1978some,arakawa1988baroclinic}.
-The choice of vertical staggering may also impact many physically-relevant properties of the model near the grid scale, such as the phase speed of Rossby waves \citep{thuburn2005vertical}.
-Finally, the choice of vertical staggering can have impacts on the physics-dynamics coupling \cite{holdaway2013comparison1,holdaway2013comparison2}.
-Taken altogether, these issues suggest care should be taken when selecting the discretization.
-For this reason, co-located discretizations, where all degrees of freedom are geometrically co-located, are not frequently adopted in global atmospheric models.
-In particular, it is common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
-These approaches can be further augmented as needed, for instance by shifting the vertical velocity and thermodynamic variables from the bottom boundary to an intermediate level, as in the GEM model.
-Note that, in general, tracer variables are co-located with the buoyancy variable.
-
-%----------------------------------------------------
-\begin{figure}[]
-\includegraphics[width=5in]{VerticalStaggering.pdf}
-\caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
-\label{fig:verticalstaggering}
-\end{figure}
-%----------------------------------------------------
-
-\subsection{Model Implementations}
-
-In this section we describe the prognostic equations that have been chosen for the dynamical cores that participated in DCMIP.
-
-\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
-
-DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014).
-Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
-Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.
-
-\subsubsection{FV$^3$} \label{sec:FV3Equations}
-
-The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{lin2004vertically}.
-Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
-The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
-
-\subsubsection{FVM} \label{sec:FVMEquations}
-
-{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
-
-The fully compressible Euler equations solved in FVM are given as
-\begin{subequations} \label{fvm:compressible}
-\begin{align}
- & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
-\label{fvm:mass}  \\
-& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
-\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\epsilon q'_v - q_c - q_r) \right)
-- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
-+ \mathbf{M} \right)~,
-\label{fvm:momentum} \\
-%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
-& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
-\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
-- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
-+  \mathcal{H} \right)~,
-\label{fvm:thermodynamic} \\
-& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
-\label{fvm:gaslaw}
-\end{align}
-\end{subequations}
-which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
-and dry entropy \eqref{fvm:thermodynamic}.
-Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
-physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
-pressure perturbation $\pi'$, 
-with the thermodynamic variables related by the gas law 
-\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
-Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
-respectively. All primed variables correspond to deviations from an ambient state (denoted by
-subscript "a")  that satisfies 
-a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
-see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014}. 
-The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
-Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
-coefficient 
-\begin{equation}
-\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
-\end{equation}
-in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
-the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
-form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
-$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
-$\epsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
-The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
-coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
-simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
-Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
-$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
-and the transformation of the physical to the contravariant velocity 
-$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003all} and \cite{kuehnleinJCP2012} for discussion.
-%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
-The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
-of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
-in \eqref{fvm:thermodynamic} represents the diffusion of heat.
-
-%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
-%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
-
-\subsubsection{GEM} \label{sec:GEM_equations}
-
-GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$,
-described in section \ref{sec:GEM_zeta}.
-The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
-\begin{align}
-\mu = \pdiff{p}{s} - 1,
-\end{align}
-\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
-and in which case
-the equations of GEM model \citep{Girard2014} are:
-\begin{align}
-\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
-\frac{d w}{dt}- g \mu &= 0, \\
-\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
-\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
-\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
-\frac{d \Phi}{dt}-  g w & = 0.
-\end{align}
-
-\subsubsection{HOMME}
-
-HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics.
-The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure {\color{red}[Equation?]}.
-
-In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ {\color{red}[Equation?]}, potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
-Fields are co-located in the vertical and explicit time-stepping is used.
-
-%Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-
-% The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
-Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
-The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
-Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
-
-%This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
-Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
-
-\subsubsection{ICON} \label{sec:ICONEquations}
-
-ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
-The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
-\begin{align}
-\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
-\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
-The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
-
-\subsubsection{Tempest} \label{sec:TempestEquations}
-
-Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for covariant horizontal velocity (\ref{eq:CovariantUpdateEquation}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-
-%In the horizontal, Tempest utilizes a spectral-element formulation.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Diffusion, Stabilization, Filters and Fixers} \label{sec:DiffusionStabilization}
-
-Most dynamical cores implement specialized techniques for diffusion or stabilization.
-Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
-Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
-Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.
-Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.
-A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.
-In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.
-
-\subsection{HOMME / Tempest}
-
-In both HOMME and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
-\begin{align}
-\nabla^2 \vb{u}_h = \nabla \nabla \cdot \vb{u}_h + \nabla \times \nabla \times \vb{u}_h.
-\end{align}  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
-Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-
-\subsection{ICON}
-
-{\color{red}[Check notation and clean up]}
-
-The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and is summarized here.
-For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
-\begin{align}
-F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
-\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
-\end{align}
-$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
-
-ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
-The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
-
-The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
-\begin{align} \label{smagn2}
-F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
-\end{align}
-\begin{align}
-F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
-\end{align}
-\begin{align}
-F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
-\end{align}
-An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
-
-A fourth-order computational diffusion is also available for vertical wind speed $w$.
-This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
-This term appears as
-\begin{align}
-F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
-\end{align}
-
-\subsection{FV$^3$}
-
-Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations.
-The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.
-FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion.
-The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
-If additional damping is desired, or if the non-monotonic advection is used, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
-The hyperdiffusion is of the same order as but much smaller than the divergence damping.
-Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep.
-
-Wave absorption at the model top is also provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
-An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
-
-FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
-This is only done before the physics is called and is not used in idealized simulations.
-
-\subsection{GEM}
-
-An explicit hyperviscosity in GEM is handled, via applications of the Laplacian operator,
-for both wind components and tracers.
-A vertical sponge layer, which uses a Laplacian operator,
-is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
-For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
-The quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}
-is used operationally to ensure tracer monotonicity
-for specific humidity and different hydrometeors.
-Other options are now available
-in GEM including a mass conserving monotonic scheme \citep{Sorenson2013} and
-a global mass fixer \citep{Bermejo2002MASS}. Those approaches have been evaluated using
-chemical constituents such as ozone \citep{deGrandpre2016OZONE}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Temporal Discretizations} \label{sec:TemporalDiscretizations}
-
-A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.
-In this section we briefly discuss several modern approaches that are presently employed.
-{\color{red}[reword?]}
-
-\subsection{Mixed Implicit-Explicit, Forward-Backward, Semi-Implicit and Additive Runge-Kutta Schemes}
-
-Implicit-explicit schemes are a broad category of time integration schemes that divide the terms of the prognostic equations into a set of explicitly integrated terms and implicitly integrated terms.
-At the very least, terms associated with vertically propagating sound waves are included among the implicit terms.
-For the remaining terms, there is some freedom in choosing how to integrate terms associated with vertical advection and horizontally propagating sound waves.
-Semi-implicit schemes are one such class of schemes that typically incorporate horizontally propagating sound waves into the implicit solve, and so rely on a global Helmholtz-type solve.
-Additive Runge-Kutta schemes are another mechanism to ensure high-order temporal accuracy, and many such schemes have been described throughout the literature (see, for example, \cite{weller2013runge, ullrich2012operator}).
-Several examples of these schemes can be found among the DCMIP models:
-
-FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
-With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
-The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
-This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
-Physics tendencies are applied impulsively at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
-
-DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
-A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
-As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
-The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
-The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
-Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
-
-ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
-No time splitting is used with respect to sound waves, because the ratio of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
-Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
-Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
-Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
-
-MPAS uses a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
-At the beginning of each Runge-Kutta sub-cycle, tendencies are computed for each of the prognostic variables and stored for the duration of the sub-cycle.
-Several iterations of an acoustic loop are then performed with a time-step much smaller than require for the Runge-Kutta sub-cycle.
-Within the acoustic loop, an implicit solve for vertically-integrated sound waves is performed to avoid timestep constraints that may arise from vertically-propagating sound waves.
-
-Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
-
-\subsection{The FVM Semi-Implicit method}
-
-{\color{red}[Could we emphasize here how this approach differs from the approaches described in section 7.1?  Or should it be included in section 7.1?]}
-
-In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
-compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
-A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
-for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
-moist dynamics. 
-The generic two-time-level second-order template algorithm employed in the integration is given as
-\begin{equation}
-\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
-+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
-\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
-\label{fvm:solscalar}
-\end{equation}
-In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
-and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
-finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
-\citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  Furthermore, 
-in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
-on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.
-
-The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
-mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
-and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
-momentum \eqref{fvm:momentum}, and moisture equations enter
-\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
-which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
-is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
-velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
-Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
-The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
-an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
-equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
-The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
-Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
-Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
-executed in an outer iteration.
-
-%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
-%slow rotational modes.
-
-%The integration scheme of FVM approximates the generic conservation law
-%\begin{equation}
-%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
-%\label{scalar}
-%\end{equation}
-%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
-%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
-%second-order template algorithm
-%\begin{equation}
-%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
-%+0.5\,\delta t\,R^{\Psi}|
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection {A semi-Lagrangian implicit time discretization in the GEM model} \label{sec:GEM_temporal}
-
-GEM differs from the approaches above by using a semi-Lagrangian advection.
-Any model equations, prognostic or diagnostic, is written in the form
-\begin{align}
-\frac {dF}{dt}+G=0,
-\label{eq:gem_temporal_O}
-\end{align}
-\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
-The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
-\begin{align}
-\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
-\label{eq:gem_temporal_D}
-\end{align}
-where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
-and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
-due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
-having occurred during the timestep $\Delta t$.
-$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
-The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
-\begin{align}
-\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
-\label{eq:gem_temporal_d_O}
-\end{align}
-discretized in the same way (trapezoidal method) though without off-centering:
-\begin{align}
-\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
-\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
-\label{eq:gem_temporal_d_D}
-\end{align}
-
-\noindent The process is of course a {\it{multi-step iterative}}
-one
-since both positions and velocities at departure positions (past time $t-\Delta t$)
-are unknown as well as, of course, the velocities at arrival positions (time $t$).
-Once a first estimate of the departure positions is obtained,
-the model equations are solved to obtain a first estimate of the velocities at time $t$.
-The model equations must be solved simultaneously and this is only possible for the linear part $L$
-which becomes a {\it {matrix inversion problem}}.
-Hence a suitable linearization is considered. The unknown (arrival) linear $L$ and non-linear $N$
-parts are then separated from the known (first departure estimate) remaining $R$ part.
-Thus, first separating space-times, (\ref{eq:gem_temporal_D}) is rewritten as follows
-\begin{align}
-\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D} \equiv R^{D},
-\label{eq:gem_temporal_AD}
-\end{align}
-where $\tau = \left ( 1/2 + \epsilon\right)\Delta t$ and $\beta=\left ( 1/2 - \epsilon\right)/\left ( 1/2 + \epsilon\right)$.
-Second separating linear from non-linear parts, we get:
-\begin{align}
-L^A + N^A = R^D,
-\label{eq:gem_temporal_LN}
-\end{align}
-\noindent
-with $L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}$ and, of course, $N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}$.
-Note that both $F$ and $G$ may require linearization. $L^A$ may then be obtained if $N^A$ is first guessed:
-Once $L^A$ is found, an estimate of $N^A$ is obtained and $L^A$ is recalculated.
-This is called the {\it{non-linear iteration process}}
-(one iteration is usually sufficient).
-The overall process is then repeated once starting from a new estimate of the departure positions.
-
-There are two intensive calculation sections in this process:
-the so-called semi-Lagrangian calculations
-(twice estimating departure positions, twice interpolating right-hand sides $R$ on departure positions),
-and solution of the linear system (four times).
-Each time, the linear system is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid
-\citep{Qaddouri2008schwarz}.
-The composite variable solution is then used to update the prognostic variables (back substitution).
-At the end of the time-step, the static halo region of both panels of the Yin-Yang grid is updated \citep{Qaddouri2011operational}.
-All required interpolations throughout the semi-Lagrangian process and between Yin and Yang grids are cubic interpolations.
 

### 2017-04-14 DCMIP2016-Part1-Models.tex
- Added text for section on horizontal staggerings

@@
 Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
-{\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The particular choice of equation set,  that have been made by these models.
-
-A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.
+{\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
+The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.
+Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
 
@@
 CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
-DYNAMICO & H/NH {\color{red}(D)} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
 FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
 FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
-GEM & NH & $\vb{u}_h$, $w$, $\log \partial s / \partial \zeta$, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
-HOMME & H & $u$, $v$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
+HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
 ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
@@
 \begin{table}[h]
-\caption{Standard options for diffusion, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
+\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
 \begin{center}
@@
 
-The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
-These continuous equations are given in section \ref{sec:TempestEquations}.
+The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
 The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.
@@
 
+In this section we discuss the placement of prognostic variables on the grid and the choice of vertical coordinate.
+
 \subsection{Horizontal staggering} \label{sec:HorizontalStaggering}
 
-More information on horizontal staggerings can be found in \cite{arakawa1977computational}.
-An analysis of the dispersion properties of each staggering in the context of geophysical flows can be found in \cite{randall1994geostrophic}.
+The horizontal placement of variables impacts a number of properties of the numerical method, including the ability of the model to conserve invariant such as energy and enstrophy, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.
+The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.
+Later other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.
+New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.
+Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).
+
+There are arguments to be made for and against each of these grids:  the A-grid is relatively simple to implement and tends to support large time step sizes, but also supports a $2 \Delta x$ computational mode (even in the context of finite element methods); the C-grid has excellent dispersion properties, and does not support extraneous computational modes when the number of edges is equal to twice the number of volumes, but also has imposes a reduced maximum timestep size relative to the A-grid; the D-grid provides a better representation of vorticity, but suffers to similar computational modes as the A-grid; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
 
@@
 
-which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
-that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep
-
-\subsection{Vertical coordinates}
-
-We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
-
-\subsubsection{Terrain-following $z$ coordinates}
-
-In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
-This simplifies the treatment of the bottom boundary condition.
-Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
-\begin{align}
-s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
-\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
-Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
-
-\subsubsection{Mass-based coordinates}
-
-Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
-\begin{align}
-s = \int^{\infty}_{z} \rho g dz.
-\end{align}  Under this definition,
-\begin{align}
-\pdiff{s}{z} = - \rho g.
-\end{align}
-
-%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
-%\begin{align}
-%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
-%\end{align} where the integrals are taken over the column.
-
-\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
-
-The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
-Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
-\begin{align}\label{eq:gem_zeta}
-\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
-\end{align}
-\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
-and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
-and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-
-\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
-
-In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
-The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
-\begin{align}
-s = \int^{\infty}_{z} \rho_0 g dz,
-\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
-\begin{align}
-\dot{s} = \frac{ds}{dt} = 0,
-\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
-Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
+Although velocities in the FV$^3$ model are stored in accordance with the D-grid arrangement, at the intermediate stages of the forward-backwards timestepping scheme, velocities are actually prognosed on the C-grid.
+The intermediate velocities then act as a simplified Riemann solver:  the intermediate stage velocities are time-centered and can be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
+More details on this approach can be found in \cite{LR1997QJR}.
 
@@
 %----------------------------------------------------
+
+\subsection{Vertical coordinates}
+
+We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
+
+\subsubsection{Terrain-following $z$ coordinates}
+
+In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
+This simplifies the treatment of the bottom boundary condition.
+Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
+\begin{align}
+s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
+\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
+Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
+
+\subsubsection{Mass-based coordinates}
+
+Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
+\begin{align}
+s = \int^{\infty}_{z} \rho g dz.
+\end{align}  Under this definition,
+\begin{align}
+\pdiff{s}{z} = - \rho g.
+\end{align}
+
+%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
+%\begin{align}
+%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
+%\end{align} where the integrals are taken over the column.
+
+\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
+
+The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
+Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
+\begin{align}\label{eq:gem_zeta}
+\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
+\end{align}
+\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
+and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
+and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+
+\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
+
+In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
+The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
+\begin{align}
+s = \int^{\infty}_{z} \rho_0 g dz,
+\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
+\begin{align}
+\dot{s} = \frac{ds}{dt} = 0,
+\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
+Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
 

### 2017-04-14 DCMIP2016-Part1-Models.tex
- Cleanup of section 4 - Incorporated changes from FVM and GEM in section 5

@@
 Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
-{\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
+The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.
+The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
 The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.
@@
 FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
-GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
+GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
 HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
@@
 \begin{table}[h]
-\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
+\caption{Vertical coordinates and staggering for participating dynamical cores.
+More details on vertical coordinates can be found in section \ref{sec:VerticalCoordinates}, and details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
 \begin{center}
 \begin{tabular}{cccccc}
-\hline Acronym & Vertical Coordinate & Vertical Staggering \\ \hline 
-CSU & fixed height & Lorenz \\
-DYNAMICO & {\color{red}floating mass?} & Lorenz \\
-FV$^3$ & floating mass & Lorenz \\
-FVM & (floating) height & Co-located \\
-GEM & log pressure & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering})  \\
-HOMME & floating mass & Co-located \\
-ICON & fixed height & Lorenz \\
-MPAS & fixed height & Lorenz \\
-NICAM & fixed height & Lorenz \\
-Tempest & fixed height & Lorenz \\
+\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
+CSU & Lorenz & fixed height \\
+DYNAMICO & Lorenz & {\color{red}floating mass?} \\
+FV$^3$ & Lorenz & floating mass \\
+FVM & Co-located & (floating) height \\
+GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
+HOMME & Co-located & floating mass \\
+ICON & Lorenz & fixed height \\
+MPAS & Lorenz & fixed height \\
+NICAM & Lorenz & fixed height \\
+Tempest & Lorenz & fixed height \\
 \hline 
@@
 
-\subsection{ICON}
+\subsection{Icosahedral Non-hydrostatic model (ICON)}
 
@@
 
-\subsection{NICAM}
+\subsection{Non-hydrostatic ICosahedral Atmospheric Model (NICAM)}
 
@@
 
-In this section we discuss the placement of prognostic variables on the grid and the choice of vertical coordinate.
+Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.
+Hence, in this section we turn our attention to the placement of prognostic variables on the grid and the choice of vertical coordinate.
 
@@
 \includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
-\caption{Various options for horizontal staggering of variables among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
+\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
 \label{fig:horizontalstaggering}
@@
 
-\subsection{Vertical coordinates}
-
-We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.
-
-\subsubsection{Terrain-following $z$ coordinates}
-
-In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.
-This simplifies the treatment of the bottom boundary condition.
-Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
-\begin{align}
-s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
-\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
+\subsection{Vertical Coordinates} \label{sec:VerticalCoordinates}
+
+Most atmospheric models use dimension splitting in order to separate the evolution equations for momentum into horizontal and vertical directions.
+Whereas ``horizontal'' in this context either refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g.
+terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.
+In this context, the use of dimension splitting allows us to choose an arbitrary vertical coordinate $s$ with the only requirement being that it is a monotonic function of the altitude $z$.
+Typically $s$ is chosen so that the Earth's surface (the bottom boundary of the atmosphere) is a coordinate surface, allowing for easy specification of boundary conditions for the prognostic equations -- this leads to the so-called ``terrain-following'' family of vertical coordinates.
+Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which is in terms of the altitude $z$ and takes the form
+\begin{align}
+s(\vb{x},z) = z_{\footnotesize \mbox{top}} \left[ \frac{z - z_s(\vb{x})}{z_{\footnotesize \mbox{top}} - z_s(\vb{x})} \right],
+\end{align} where $\vb{x}$ denotes the horizontal position, $z_s(\vb{x})$ is the height of the topography at that position, and $z_{\footnotesize \mbox{top}}$ denotes the height of the model top (typically independent of position).
+Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
 Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
+All models in this paper use some variant of terrain-following coordinates, although work on cut-cell, embedded boundary and immersed boundary representations is ongoing (e.g.
+\cite{lock2012demonstration}).
+Note that time-dependent vertical coordinates are allowed and are typically referred to as ``floating'' coordinates.
+Several examples of vertical coordinates are now given.
 
@@
 Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
-\begin{align}
+\begin{align} \label{eq:s_mass-based}
 s = \int^{\infty}_{z} \rho g dz.
@@
 
-%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
-%\begin{align}
-%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
-%\end{align} where the integrals are taken over the column.
+The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
+\begin{align} \label{eq:SurfacePressureEquation}
+\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
+\end{align} where the integrals are taken over the column.
 
@@
 \noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
-and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
+and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
 and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
 
-\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}
+\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and HOMME)} \label{sec:FloatingLagrangian}
 
@@
 
-\subsection{Model Implementations}
-
-In this section we describe the prognostic equations that have been chosen for the dynamical cores that participated in DCMIP.
+\subsection{Prognostic equations} \label{sec:PrognosticEquations}
+
+The Navi\'er-Stokes equations that govern atmospheric motion can take on many forms, depending on the choice of prognostic variables and coordinate system.
+A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.
+The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.
+The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models.
 
@@
 
-DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014).
+DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation \citep{Dubos2014Equations}.
 Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
@@
 
-{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}
-
-The fully compressible Euler equations solved in FVM are given as
+FVM solves conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, and dry 
+entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
+which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
+\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Underlying the governing equations 
+in FVM is a perturbational form with respect to a balanced ambient state, and a generalized curvilinear 
+coordinate formulation in a geospherical framework. These governing equations can 
+concisely be written as
+\citep{smolarkiewiczetalJCP2017}
 \begin{subequations} \label{fvm:compressible}
 \begin{align}
- & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
+ & \pdiff{\mathcal{G}\rho_d}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
 \label{fvm:mass}  \\
-& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
-\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\varepsilon q'_v - q_c - q_r) \right)
-- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
-+ \mathbf{M} \right)~,
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+\mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
+- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a (\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r) \right)
+- 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
++ \mathcal{M}' \right)~,
+%+ \mathbf{M} \right)~,
 \label{fvm:momentum} \\
 %- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
-& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
-\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
-- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
-+  \mathcal{H} \right)~,
+& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \overline{\nabla} \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
+- \mathcal{G}\rho_d\,\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \overline{\nabla} \theta_{a} ~,
+%- \frac{L}{c_{pd}\,\pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
+%+  \mathcal{H} \right)~,
 \label{fvm:thermodynamic} \\
-& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
+& \phi' = c_{pd} \left[\left( \frac{R_{d}}{p_{0}} \rho_d\,\theta\,(1+q_v/\varepsilon) \right)^{R_{d}/c_{vd}} - \pi_{a} \right]~.
 \label{fvm:gaslaw}
@@
 \end{subequations}
-which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
-and dry entropy \eqref{fvm:thermodynamic}.
-Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
-physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
-pressure perturbation $\pi'$, 
+Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, 3D
+physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and a modified
+Exner pressure perturbation $\phi'$, 
 with the thermodynamic variables related by the gas law 
-\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
-Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
-respectively. All primed variables correspond to deviations from an ambient state (denoted by
-subscript "a")  that satisfies 
-a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
-see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014}. 
-The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
-Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
-coefficient 
-\begin{equation}
-\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
-\end{equation}
-in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
-the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
-form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
-$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
-$\varepsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
-The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
-coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
-simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
-Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
-$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
-and the transformation of the physical to the contravariant velocity 
-$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003all} and \cite{kuehnleinJCP2012} for discussion.
-%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
-The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
-of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
-in \eqref{fvm:thermodynamic} represents the diffusion of heat.
-
-%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
-%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.
+\eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
+subscript "a"; see also \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. 
+In the shallow- versus deep-atmosphere form of \eqref{fvm:compressible}, gravity is constant 
+$g \equiv g_c$ or varies with height as $g = g_c\,(a/(a+z))^2$, respectively.
+As far as geometric aspects are concerned, the nabla operator $\overline{\nabla}$ represents
+the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
+a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
+the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see also \cite{prusa2003all} 
+and \cite{kuehnleinJCP2012}.
+The symbol $\mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
+subsumes spherical Christoffel symbols \citep[][]{smolarkiewiczetalJCP2017}.
 
@@
 
-GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$,
-described in section \ref{sec:GEM_zeta}.
-The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
+%GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$, described in section \ref{sec:GEM_zeta}.
+In GEM the non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
 \begin{align}
@@
 
+%\noindent with horizontal velocity ${\bf V}_h$, 
+%generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
+%vertical velocity $w$,
+%geopotential $\phi$, 
+%virtual temperature $T_v$, 
+%pressure $p$ and
+%hydrostatic pressure $\pi$. The variable 
+%$\mu$ is a measure of departure from hydrostatic balance, 
+%$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
+
 \subsubsection{HOMME}
@@
 HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics.
-The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure {\color{red}[Equation?]}.
+The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure (\ref{eq:SurfacePressureEquation}).
 
@@
 In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ {\color{red}[Equation?]}, potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
-Fields are co-located in the vertical and explicit time-stepping is used.
+Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
 
@@
 
-%In the horizontal, Tempest utilizes a spectral-element formulation.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Most dynamical cores implement specialized techniques for diffusion or stabilization.
+Most dynamical cores implement specialized techniques for diffusion or stabilization (see Table \ref{tab:ModelsTemporalDiffusion}).
 Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
@@
 
-{\color{red}[Check notation and clean up]}
-
 The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and is summarized here.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and are summarized here.
 For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
@@
 \begin{align} \label{smagn2}
-F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
-\end{align}
-\begin{align}
-F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
-\end{align}
-\begin{align}
-F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
+F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
+& F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
 \end{align}
@@
 
-An explicit hyperviscosity in GEM is handled, via applications of the Laplacian operator,
+An explicit hyperviscosity in GEM is handled via applications of the Laplacian operator
 for both wind components and tracers.
@@
 
-{\color{red}[Could we emphasize here how this approach differs from the approaches described in section 7.1?  Or should it be included in section 7.1?]}
-
-In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
-compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
-A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
-for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
-moist dynamics. 
-The generic two-time-level second-order template algorithm employed in the integration is given as
+A characteristic feature of the FVM (Section~\ref{sec:FVM}) time-stepping scheme is the 3D implicit treatment 
+of the fast buoyant and acoustic modes, and the slow rotational modes. Therefore, the model time step is 
+identical for all processes and selected according to the stability of the advective transport scheme---i.e.~%
+the time step is continuously adapted according to a given maximum advective Courant number 
+permitted by the MPDATA scheme.
+A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016,kuehnleinJCP2017} for dry dynamics, 
+and in \cite{kurowskiJAS2014,smolarkiewiczetalJCP2017} for moist-precipitating dynamics. Here,
+we provide an outline of the basic integration scheme for the compressible Euler equations
+\eqref{fvm:compressible}. The scheme employs the two-time-level second-order-accurate template 
+algorithm given as
 \begin{equation}
@@
 \end{equation}
-In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
-and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
-finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
-\citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  Furthermore, 
-in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
-on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.
+where $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
+$\mathcal{A}$ symbolises the advective transport operator---the non-oscillatory 
+finite-volume MPDATA bespoke for the compressible dynamics on the hybrid horizontally-unstructured
+vertically-structured mesh \citep{kuehnleinJCP2017}---and the vector index $\mathbf{i} \equiv (k,i)$ denotes the 
+spatial position on the computational mesh.
 
@@
 \eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
-which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
+which is generally depending on all prognostic variables. The high degree of implicitness in the representation of the rhs forcings
 is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
-velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
+velocity updates---the procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
 Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term.
-The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
-an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
-equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
-The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
-Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
-Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
-executed in an outer iteration.
-
-%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
-%slow rotational modes.
-
-%The integration scheme of FVM approximates the generic conservation law
-%\begin{equation}
-%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
-%\label{scalar}
-%\end{equation}
-%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
-%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
-%second-order template algorithm
-%\begin{equation}
-%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
-%+0.5\,\delta t\,R^{\Psi}|
+The subsequent third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
+an advective form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
+equation with a Euler backward scheme, in the spirit of \eqref{fvm:solscalar}, leads a Helmholtz equation
+\citep{smolarkiewiczJCP2014}.
+The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
+Residual approach; see \cite{smolarkiewiczAG2011} and references therein. Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
+solution-dependent coefficients of the Helmholtz equation are lagged behind and executed in an outer iteration.
 
@@
 \noindent
-with $L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}$ and, of course, $N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}$.
+with
+\begin{align}
+L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}, \quad and \quad N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}.
+\end{align}
 Note that both $F$ and $G$ may require linearization. $L^A$ may then be obtained if $N^A$ is first guessed:
@@
 \end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
-\begin{align}
+\begin{align} \label{eq:PotentialTemperatureFluxForm}
 \pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
@@
 In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
-\begin{align}
+\begin{align} \label{eq:PrognosticVelocityFluxForm}
 \pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,

### 2017-04-15 DCMIP2016-Part1-Models.tex
- More revisions - Added OLAM text

@@
 \begin{abstract}
-Atmospheric dynamical cores are a core component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
-These systems have existed in one form or another for over half of a century, with the earliest strategies now evolved into a complex ecosystem of algorithms and computational strategies.
+Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
+These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.
 In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.
-To better understand modern dynamical cores, this paper aims to provide a comprehensive review of their design, including model grids, relevant equation sets, numerical stabilization techniques and idealized physics routines.
-Examples are taken from operational dynamical cores that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
+To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
+This review includes choice of model grid, variable placement, vertical coordinate choices, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
 \end{abstract}
@@
 The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.
-Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.
-These factors include the choice of model grid, choice of prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.
+Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.
+These factors include the choice of model grid, variable placement, prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.
 
 To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
-It was built on earlier DCMIP workshops (held in 2008 and 2012), and previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
-Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling.
+The 2016 workshop and summer school followed from earlier DCMIP and dynamical core workshops (held in 2012 and 2008, respectively), and other model intercomparison efforts.
+Its goals were to provide a forum for discussing outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
+Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling, and variable-resolution global modeling.
 The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.
@@
 The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.
-Modeling groups have leveraged this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.
-
-The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
-Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.
-A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.
+Past and present intercomparison efforts have been leveraged by modeling groups to improve their own models, in turn leading to a positive impact on the quality of weather and climate simulations.
+The workshop component of DCMIP has also advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
+Notably, the use of idealized test cases to isolate specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.
+Another important outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.
 The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.
@@
 This paper is the first in a series of papers documenting the results of this workshop.
-This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.
+It is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling.
 For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
-Section \ref{sec:DynamicalCores} provides a brief overview of each of the participating models.
-The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
-Conclusions and discussion then follow in section \ref{sec:Conclusions}.
-
-\begin{table}[h]
-\caption{List of symbols used by DCMIP} \label{tab:symbols}
+Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.
+The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
+The summary and conclusions then follow in section \ref{sec:Conclusions}.
+Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.
+
+\begin{table}[p]
+\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
 \begin{center}
@@
 $z$ & Height with respect to mean sea level (set to zero) \\
+$s$ & Vertical model coordinate \\
 $p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
@@
 $\theta_v$ & Virtual potential temperature \\
-$q$ & Specific humidity \\
-$P_{ls}$ & Large-scale precipitation rate \\
+$\theta_{il}$ & Ice-liquid potential temperature \\
+%$P_{ls}$ & Large-scale precipitation rate \\
 $q$ & Water vapor mixing ratio \\
@@
 The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
-The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.
+The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.
 Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
-
-\begin{table}[h]
+The participant models are briefly described through the remainder of this section.
+
+\begin{table}[p]
 \caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
@@
 NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
+OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
 Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
@@
 
-\begin{table}[h]
+\begin{table}[p]
 \caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.
-Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).
+Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).
 Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).
@@
 NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
+OLAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, {\color{red} $q$} & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
 Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
@@
 
-\begin{table}[h]
-\caption{Vertical coordinates and staggering for participating dynamical cores.
-More details on vertical coordinates can be found in section \ref{sec:VerticalCoordinates}, and details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
+\begin{table}[p]
+\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
 \begin{center}
@@
 CSU & Lorenz & fixed height \\
-DYNAMICO & Lorenz & {\color{red}floating mass?} \\
-FV$^3$ & Lorenz & floating mass \\
+DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
+FV$^3$ & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 FVM & Co-located & (floating) height \\
 GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
-HOMME & Co-located & floating mass \\
+HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 ICON & Lorenz & fixed height \\
@@
 NICAM & Lorenz & fixed height \\
+OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
 Tempest & Lorenz & fixed height \\
@@
 
-\begin{table}[h]
-\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
+\begin{table}[p]
+\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
 \begin{center}
@@
 & stabilization, filters, or fixers & \\ \hline 
-CSU & 4th-order horizontal hyperdiffusion & 3rd-order Adams-Bashforth (AB3) \\
+CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
 DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
@@
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
-HOMME & Hyperdiffusion & KGU53 \citep{guerra2016high} \\
+HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
 ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
@@
 NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
-Tempest & Hyperdiffusion & ARS232 \citep{ascher1997implicit} \\
+OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
+Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
 \hline 
@@
 \subsection{Colorado State University Model (CSU)}
+
+{\color{red}[More references?]}
 
@@
 The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.
-The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang}, with Arakawa C-grid staggering.
+The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
 The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
@@
 
-The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
+{\color{red}[More references?]}
+
+The ICON model \citep{zangl2014icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
 Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
@@
 
+\subsection{Ocean-Land-Atmosphere Model (OLAM)}
+
+OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.
+Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting.
+Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
+Height is the vertical coordinate, and a Lorenz vertical grid stagger is used.
+A unique feature of OLAM is that grid levels are horizontal and intersect topography.
+This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms.
+Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground.
+The OLAM cut-cell formulation conserves mass and momentum.
+
 \subsection{Tempest}
@@
 \end{center}
-\caption{(Left) A cubed-sphere grid with 16 elements along each cubed-sphere edge.
+\caption{(Left) A cubed-sphere grid.
 (Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.
@@
 The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).
-Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this the geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
+Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
 
@@
 ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
+
+OLAM optimizes by applying the spring dynamics method of \cite{tomita2001shallow} to the dual triangular mesh prior to its mapping to the Voronoi grid.
+When local mesh refinement is applied, which OLAM achieves in a series of one or more resolution-doubling steps, each spanning a transition zone that is three grid rows wide (Figure \ref{fig:OLAMRefinement}), the equilibrium spring length is scaled to the target grid cell size in each refinement level and is varied incrementally across the transition zone.
+Spring dynamics is further modified by forcing angles on the dual triangular mesh in the transition zone in order to move the triangle edges closer to the centers of the hexagon edges they intersect.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.5in]{OLAM_Grid_Refinement.png}
+\end{center}
+\caption{Detail of one step of local mesh refinement used by OLAM Voronoi mesh.
+The transition zone is constructed by explicit topological reconnection of the grid lines, which produces pairings of heptagons (red dots) and pentagons (blue dots) along the refinement perimeter.} \label{fig:OLAMRefinement}
+\end{figure}
 
@@
 
-\section{Variable staggering and vertical coordinates}
+\section{Variable placement and prognosis} \label{sec:VariablePlacement}
 
 Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.
-Hence, in this section we turn our attention to the placement of prognostic variables on the grid and the choice of vertical coordinate.
+Hence, in this section we turn our attention to the placement of prognostic variables on the grid, the choice of vertical coordinate, and the choice of corresponding prognostic equations.
 
@@
 
-There are arguments to be made for and against each of these grids:  the A-grid is relatively simple to implement and tends to support large time step sizes, but also supports a $2 \Delta x$ computational mode (even in the context of finite element methods); the C-grid has excellent dispersion properties, and does not support extraneous computational modes when the number of edges is equal to twice the number of volumes, but also has imposes a reduced maximum timestep size relative to the A-grid; the D-grid provides a better representation of vorticity, but suffers to similar computational modes as the A-grid; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
+There are arguments to be made for and against each of these grids:  the A-grid is relatively simple to implement and tends to support large time step sizes, but also supports a $2 \Delta x$ computational mode (even in the context of finite element methods); the C-grid has excellent dispersion properties, and does not support extraneous computational modes (as long as the number of edges is equal to twice the number of volumes), but also requires a timestep that is half that of the A-grid; the D-grid provides a better representation of vorticity, but suffers to similar computational modes as the A-grid that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
 
@@
 Taken altogether, these issues suggest care should be taken when selecting the discretization.
-For this reason, co-located discretizations, where all degrees of freedom are geometrically co-located, are not frequently adopted in global atmospheric models.
-In particular, it is common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
+Since co-located discretizations of the non-hydrostatic equations generally require some additional effort to control spurious computational modes, it is more common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
 These approaches can be further augmented as needed, for instance by shifting the vertical velocity and thermodynamic variables from the bottom boundary to an intermediate level, as in the GEM model.
@@
 \begin{figure}[]
-\includegraphics[width=5in]{VerticalStaggering.pdf}
+\includegraphics[width=5in]{VerticalStaggering_v2.pdf}
 \caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
@@
 terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.
-In this context, the use of dimension splitting allows us to choose an arbitrary vertical coordinate $s$ with the only requirement being that it is a monotonic function of the altitude $z$.
+In this context, the use of dimension splitting allows us to choose an arbitrary function $s(t,\vb{x},z)$ that is monotonic in $z$, so that model interfaces are equally spaced with respect to $s$.
 Typically $s$ is chosen so that the Earth's surface (the bottom boundary of the atmosphere) is a coordinate surface, allowing for easy specification of boundary conditions for the prognostic equations -- this leads to the so-called ``terrain-following'' family of vertical coordinates.
@@
 Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
-All models in this paper use some variant of terrain-following coordinates, although work on cut-cell, embedded boundary and immersed boundary representations is ongoing (e.g.
+All models in this paper except for OLAM use some variant of terrain-following coordinates, although work on developing modern cut-cell, embedded boundary and immersed boundary representations is ongoing (e.g.
 \cite{lock2012demonstration}).
@@
 
-The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
+The vertical coordinate in the GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
 Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
@@
 \end{align}
-\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
-and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
+\noindent with
+\begin{align}
+A(\zeta) = \zeta, \quad \mbox{and} \quad B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r.
+\end{align}
+Here  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
 and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
@@
 Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
+
+\subsubsection{Cut-cells in OLAM} \label{sec:CutCells}
+
+A pure z coordinate with horizontal grid levels is used in OLAM \citep{walko2008ocean2} in order to completely avoid topographic imprinting on the model grid levels (Figure \ref{fig:OLAM_CutCell}).
+This implies that grid levels intersect the topographic surface, leading to some grid cells being partially above and partly below the surface.
+The face areas of these so-called cut cells are reduced accordingly, which in turn regulates cell-to-cell flux transport in accordance with the kinematic constraint imposed by the topography.
+Cut cell volumes are also reduced, and volumes and surface areas of all cells appear explicitly in the finite-volume formulation of the mass and momentum conservation equations.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2in]{OLAM_CutCellA.png} \quad \includegraphics[width=2in]{OLAM_CutCellB.png}
+\end{center}
+\caption{(Left) A terrain-following coordinate passing over rough topography.  (Right) A cut-cell coordinate used for representing the same topography.} \label{fig:OLAM_CutCell}
+\end{figure}
 
@@
 
-DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation \citep{Dubos2014Equations}.
-Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
-Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.
+The prognostic equations employed by DYNAMICO are based on a Hamiltonian formulation \citep{Dubos2014Equations}.
+The specific prognostic variables employed pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+Prognostic equations are in flux-form for mass (\ref{eq:PrognosticPseudoDensity}) and $W$ (\ref{eq:PrognosticVelocityFluxForm}), in advective form for $\Phi$ (\ref{eq:PrognosticShallowGeopotential}), and in vector-invariant form for covariant horizontal momentum (\ref{eq:CovariantUpdateEquation}).
 
@@
 \label{fvm:mass}  \\
-& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left[\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
 \mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
-- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a (\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r) \right)
+- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
 - 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
-+ \mathcal{M}' \right)~,
++ \mathcal{M}' \right]~,
 %+ \mathbf{M} \right)~,
@@
 \eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
-subscript "a"; see also \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. 
-In the shallow- versus deep-atmosphere form of \eqref{fvm:compressible}, gravity is constant 
-$g \equiv g_c$ or varies with height as $g = g_c\,(a/(a+z))^2$, respectively.
-As far as geometric aspects are concerned, the nabla operator $\overline{\nabla}$ represents
+subscript `a' (see also \cite{prusa2008eulag,smolarkiewiczJCP2014}).   As far as geometric aspects are concerned, the nabla operator $\overline{\nabla}$ represents
 the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
@@
 The symbol $\mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
-subsumes spherical Christoffel symbols \citep[][]{smolarkiewiczetalJCP2017}.
+subsumes the spherical Christoffel symbols \citep[][]{smolarkiewiczetalJCP2017}.
 
@@
 \end{align}
-\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
-and in which case
-the equations of GEM model \citep{Girard2014} are:
+\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based}).  In this case the equations of GEM model \citep{Girard2014} are concisely given by
 \begin{align}
@@
 In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
+Prognostic variables include the 3d wind velocity components $u,v,w$ ((\ref{eq:PrognosticWEquationK2Sb}) and (\ref{eq:PrognosticUhEquationK2Sb})), the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
 
@@
 \end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
-The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangental velocity components $v_t$.
+The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangential velocity components $v_t$.
+
+\subsubsection{OLAM} \label{sec:OLAMEquations}
+
+OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:ContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and
+(\ref{eq:PotentialTemperatureFluxForm}).
+Prognostic variables are the 3 components of momentum, ice-liquid potential temperature $\theta_{il}$ \citep{walko2000efficient}, total density $\rho$, specific density of total water, and specific bulk density and/or bulk number concentration of various scalar quantities including liquid and ice hydrometeors, aerosols, and trace gases.
+For DCMIP, the latter are limited to cloud and rain specific bulk densities.
+Water vapor density $\rho q$ is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$.
+The latter is in turn diagnosed from $\theta_{il}$ and from the latent heat required to convert any hydrometeors present to the vapor phase.
+Velocity components are diagnosed by dividing momentum components by total density.
+
+Momentum is C-staggered in the horizontal and vertical (Lorenz vertical staggering is used), meaning that prognosed components live on the grid cell faces and are each normal to the respective face and the pressure gradient force is evaluated and applied at those locations.
+However, evaluation of advective and turbulent momentum transport (as well as the Coriolis force) involves a diagnostic reconstruction of the total momentum vector at the centers of scalar grid cells \citep{perot2000conservation}, and cell-to-cell flux of momentum is computed from that reconstruction using the same A- grid control volumes as for scalars.
+This arrangement is particularly convenient for the cut-cell formulation at the topographic surface where reduced cell face areas and volumes regulate momentum and scalar fluxes in an identical manner.
 
@@
 The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014}, and are summarized here.
-For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions.
-\begin{align}
-F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
-\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014icon}, and are summarized here.
+For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the (relatively) computationally  expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions,
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left\{\nabla \left[\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right]\right\} \,.
 \end{align}
@@
 \begin{align}
-F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)).
 \end{align}
@@
 
+\subsection{OLAM}
+
+OLAM requires two types of artificial damping.
+In the upper layers of the model, vertical velocity and small-scale horizontal divergence are damped in order to attenuate gravity waves and thereby mitigate their reflection off the rigid top boundary of the domain.
+The damping layer is commonly applied in the uppermost 10 km of the domain, where the model top is 35 or 40 km above sea level.
+The damping rate is zero at the bottom of the damping layer and increases upward, usually linearly.
+
+Throughout the model domain, vertical vorticity is filtered horizontally at the smallest resolvable scale in order to control a spurious computational mode.
+This vertical vorticity mode is inherent in C-staggered momentum formulations on hexagonal meshes because horizontal velocities are more numerous than twice the number of scalar (mass) values and are thus under-constrained.
+The vorticity filter is constructed so as to have zero impact on divergence at any scale.
+
+Upwinding in the Lax-Wendroff formulation of the advection operator provides sufficient damping that no other type of filtering is required.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-A wide variety of temporal discretizations available among global atmospheric models, particularly to address issues related to the large horizontal-vertical aspect ratio typical in these systems.
-In this section we briefly discuss several modern approaches that are presently employed.
-{\color{red}[reword?]}
+Temporal discretizations are important for capturing the discrete dynamical evolution of the global atmosphere.
+In the past two decades, a variety of new temporal discretizations have been developed, leaving behind the days when the leapfrog scheme was ubiquitous across models.
+This diversity is in part because of the demands of non-hydrostatic models: unlike their hydrostatic counterparts, non-hydrostatic atmospheric models that do not use the anelastic \citep{ogura1962scale}, quasi-hydrostatic \citep{orlanski1981quasi}, pseudo-incompressible \citep{durran1989improving}, or unified approximation \citep{arakawa2009unification} must include a mechanism in the temporal discretization for dealing with vertically propagating sound waves.
+These waves are meteorologically insignificant, but with a vertical grid spacing of 100 meters, a purely explicit temporal discretization would require a time step size on the order of one second or less.
+In this section we discuss the timestepping schemes that have been employed across the DCMIP suite of models.
 
@@
 Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
-Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014icon}.
 
@@
 Within the acoustic loop, an implicit solve for vertically-integrated sound waves is performed to avoid timestep constraints that may arise from vertically-propagating sound waves.
+
+OLAM uses a unique temporal discretization that combines elements of the Adams-Bashforth (AB2) scheme and a Lax- Wendroff formulation for advected quantities.
+However, instead of extrapolating all prognostic tendencies forward to the half-future time level as in AB2, the horizontal momentum components alone (not their tendencies) are extrapolated in time at the cell boundaries where they reside.
+The extrapolated momentum provides the time-centered cell-to-cell total mass flux across the grid cell faces that is responsible for advective transport.
+Advection of all quantities, including all 3 velocity components that are diagnostically reconstructed at scalar cell centers, and advancement in time from the current to the future time level is based on the time- and space-centered Lax-Wendroff formulation.
+This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
+A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection.
+Scalar advection is performed afterward and uses the same total mass fluxes as momentum advection.
+A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.
 
@@
 
+\conclusions[Summary and Conclusions] \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
+
+As discussed earlier, this paper represents the first in a series of papers documenting the results from the 2016 Dynamical Core Model Intercomparison Project workshop and summer school.
+In this paper we have provided a description of the differences and similarities between participating models, including the choice of computational grid, horizontal staggering, vertical staggering, vertical coordinates, prognostic equations, choice of diffusion, stabilization, filters and fixers, and temporal discretization.
+The literature on dynamical core development is vast, with origins that go back over half a century.
+Consequently, the models discussed in this paper only represent a sample of the many dynamical cores that have been developed for general circulation modeling.
+Some of the models that have not been discussed include \cite{fox1997finite, davies2005new, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global} and \cite{doyle2014next}.
+
+The vast diversity of the modern dynamical core ecosystem suggests that there is no singular approach that is intrinsically superior to other methods.
+Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.
+The repercussions that emerge from these choices can then be explored in the context of idealized test cases, such as the ones that have been proposed as part of DCMIP.
+The remaining papers in this series investigate how the models described in this paper are able to simulate three idealized test cases, which each incorporate simplified model physics:  a moist baroclinic wave, an idealized tropical cyclone, and a splitting supercell storm on a small planet.
+Where appropriate, metrics have been included that may be indicative of model performance.
+These tests can also be used for future dynamical core development to identify where a new dynamical core diverges from a suite of modern models.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \appendix
 
-\section{Notation} \label{sec:Notation}
-
-This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
+\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}
+
+%This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
 Table \ref{tab:symbols} lists the symbols used in this paper.
-A list of physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
-Great circle distance is used throughout the project and is computed via
-\begin{equation}
-R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
-\end{equation}
+%Great circle distance is used throughout the project and is computed via
+%\begin{equation}
+%R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
+%\end{equation}
+
+In this appendix we provide a detailed derivation of the fluid equations utilized by non-hydrostatic models.
+The physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
+The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
+\begin{align}
+\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
+\end{align} where $\vb{u}$ denotes the 3D vector velocity.  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
+\begin{align} \label{eq:TracerTransport}
+\diff{q_i}{t} = 0.
+\end{align}
 
@@
 
-\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}
-
-In this appendix we provide a detailed derivation of the fluid equations utilized by nonhydrostatic models.
-The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
-\begin{align}
-\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
-\end{align} where $\vb{u}$ denotes the 3D vector velocity.  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
-\begin{align} \label{eq:TracerTransport}
-\diff{q_i}{t} = 0.
-\end{align}
-
 \subsection{Diagnostic relationships}
@@
 Essentially any of the continuity equations can be chosen, as long as the mass field represented by the equation is everywhere non-zero.
-However, the most common options are moist psuedo-density \citep{ullrich2012mcore} or dry pseudo-density \citep{skamarock2012multiscale}.
+However, the most common options are moist pseudo-density \citep{ullrich2012mcore} or dry pseudo-density \citep{skamarock2012multiscale}.
 Here we denote our density variable by $\tilde{\rho}_s$, and assume no external sources or sinks of $\tilde{\rho}$.
@@
 
-\conclusions \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
-
-{\color{red}[To be written]}
-
 %\appendix

### 2017-04-22 DCMIP2016-Part1-Models.tex
For FVM: tab, fig and ref updates and corrections

@@
 FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
-FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
+FVM & NH, D & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
 GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
@@
 FV$^3$ & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
-FVM & Co-located & (floating) height \\
+FVM & Co-located & fixed height \\
 GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
@@
 FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
-FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
+FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\

### 2017-04-24 DCMIP2016-Part1-Models.tex
- Merged updates from FVM

@@
 This paper is the first in a series of papers documenting the results of this workshop.
-It is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling.
+Its purposes is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.
 For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
@@
 DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
-FVM & NH, D & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
+FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
+FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
 GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
@@
 DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
-FV$^3$ & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
+FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 FVM & Co-located & fixed height \\
@@
 DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
-FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
-FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
+FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
+FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
@@
 
-The horizontal placement of variables impacts a number of properties of the numerical method, including the ability of the model to conserve invariant such as energy and enstrophy, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.
+The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.
 The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.
-Later other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.
+Later, other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.
 New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.
@@
 \includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
-\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
+\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.  Here $\eta$ denotes the buoyancy variable.}
 \label{fig:horizontalstaggering}
@@
 
-In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
+In the floating Lagrangian formulation \citep{starr1945quasi, lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
 The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
@@
 
-The vast diversity of the modern dynamical core ecosystem suggests that there is no singular approach that is intrinsically superior to other methods.
+The vast diversity of the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.
 Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.

### 2017-04-26 DCMIP2016-Part1-Models.tex
- Some small edits to text

@@
 
-\title{DCMIP2016, Part 1: Models and Equation Sets}
+\title{DCMIP2016:  A Review of Nonhydrostatic Dynamical Core Design and Intercomparison of Participating Models}
 
@@
 
-\Author[6]{David A.}{Hall}
-
-\Author[7]{Alex}{Reinecke}
-\Author[7]{Kevin}{Viner}
-
-\Author[8]{Don}{Dazlich}
-\Author[8]{Ross}{Heikes}
-\Author[8]{Celal}{Konor}
-\Author[8]{David}{Randall}
-
-\Author[9]{Xi}{Chen}
-\Author[9]{Lucas}{Harris}
-
-\Author[10]{Marco}{Giorgetta}
-\Author[11]{Daniel}{Reinert}
-
-\Author[12]{Christian}{K\"uhnlein}
-
-\Author[13]{Robert}{Walko}
-
-\Author[14]{Vivian}{Lee}
-\Author[14]{Abdessamad}{Qaddouri}
-\Author[14]{Claude}{Girard}
-
-\Author[15]{Hiroaki}{Miura}
-\Author[15]{Tomoki}{Ohno}
-\Author[16]{Ryuji}{Yoshida}
+\Author[6]{Don}{Dazlich}
+\Author[6]{Ross}{Heikes}
+\Author[6]{Celal}{Konor}
+\Author[6]{David}{Randall}
+
+\Author[7]{Thomas}{Dubos}
+\Author[7]{Yann}{Meurdesoif}
+
+\Author[8]{Xi}{Chen}
+\Author[8]{Lucas}{Harris}
+
+\Author[9]{Christian}{K\"uhnlein}
+
+\Author[10]{Vivian}{Lee}
+\Author[10]{Abdessamad}{Qaddouri}
+\Author[10]{Claude}{Girard}
+
+\Author[11]{David A.}{Hall}
+
+\Author[12]{Marco}{Giorgetta}
+\Author[13]{Daniel}{Reinert}
 
@@
 
-\Author[17]{Thomas}{Dubos}
-\Author[17]{Yann}{Meurdesoif}
+\Author[14]{Hiroaki}{Miura}
+\Author[14]{Tomoki}{Ohno}
+\Author[15]{Ryuji}{Yoshida}
+
+\Author[16]{Robert}{Walko}
+
+\Author[17]{Alex}{Reinecke}
+\Author[17]{Kevin}{Viner}
 
@@
 \affil[5]{Stony Brook University}
-\affil[6]{University of Colorado, Boulder}
-\affil[7]{Naval Research Laboratory}
-\affil[8]{Colorado State University}
-\affil[9]{Geophysical Fluid Dynamics Laboratory}
-\affil[10]{Max Planck Institute for Meteorology}
-\affil[11]{Deutscher Wetterdienst (DWD)}
-\affil[12]{European Centre for Medium-Range Weather Forecasts}
-\affil[13]{University of Miami}
-\affil[14]{Environment and Climate Change Canada}
-\affil[15]{University of Tokyo}
-\affil[16]{RIKEN}
-\affil[17]{Institut Pierre-Simon Laplace (IPSL)}
-\affil[18]{Lawrence Berkeley National Laboratory}
+\affil[6]{Colorado State University}
+\affil[7]{Institut Pierre-Simon Laplace (IPSL)}
+\affil[8]{Geophysical Fluid Dynamics Laboratory (GFDL)}
+\affil[9]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
+\affil[10]{Environment and Climate Change Canada}
+\affil[11]{University of Colorado, Boulder}
+\affil[12]{Max Planck Institute for Meteorology}
+\affil[13]{Deutscher Wetterdienst (DWD)}
+\affil[14]{University of Tokyo}
+\affil[15]{RIKEN}
+\affil[16]{University of Miami}
+\affil[17]{Naval Research Laboratory}
 
@@
 
-\runningtitle{DCMIP2016: Models and Equation Sets}
+\runningtitle{DCMIP2016: A Review of Nonhydrostatic Dynamical Core Design and Intercomparison}
 
@@
 Further details on these choices can be found in subsequent sections.
-Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).
+Simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).
 Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
@@
 Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).
-Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).
+Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).
 More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
@@
 CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
-DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
 FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
 FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
-GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
+GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
 HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
@@
 NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
-OLAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, {\color{red} $q$} & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
 Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
@@
 CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
-DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
+DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
 FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
-FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
+FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
@@
 
-{\color{red}[More references?]}
-
-The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate.
+The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate.
 The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
@@
 The model employs a hybrid terrain-following pressure coordinate in the vertical.
-The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
 Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.
@@
 %A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions.
+The PDEs are discretized using a split approximation, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
 Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
@@
 \subsection{Icosahedral Non-hydrostatic model (ICON)}
-
-{\color{red}[More references?]}
 
@@
 Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th order Laplacian diffusion operator.
+The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator.
 For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-{\color{blue}[Blue text to be moved to paper 2]
-\section{Idealized Physical Parameterizations} \label{sec:IdealizedPhysics}
-
-\subsection{Kessler Physics} \label{sec:KesslerPhysics}
-
-The cloud microphysics update according to the following equation set:
-\begin{alignat}{5}
-\frac{\Delta \theta}{\Delta t} = & - \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big) & \\
-\frac{\Delta q_v}{\Delta t} = & & \frac{\Delta q_{vs}}{\Delta t} & + E_r \\
-\frac{\Delta q_c}{\Delta t} = & & - \frac{\Delta q_{vs}}{\Delta t} & & - A_r & - C_r \\
-\frac{\Delta q_r}{\Delta t} = & & & - E_r & + A_r & + C_r & - V_r \pdiff{q_r}{z},
-\end{alignat} where $L$ is the latent heat of condensation, $A_r$ is the autoconversion rate of cloud water to rain water, $C_r$ is the collection rate of rain water, $E_r$ is the rain water evaporation rate, and $V_r$ is the rain water terminal velocity.
-
-The pressure follows from the equation of state
-\begin{equation}
-p=\rho R_dT(1+0.61q_v)
-\end{equation} with $p$ the pressure, $\rho$ the density of moist air, $R_d$ the gas constant for dry air, $T$ the temperature and $q_v$ the mixing ratio of water vapor.
-The equation is rewritten as a nondimensional pressure $\Pi$ equation.
-\begin{equation}
-\pi = \left(\frac{p}{p_0}\right)^{\frac{R_dT}{cp}}
-\end{equation}
-
-To determine the saturation vapor mixing ratio the Teten's formula is used,
-\begin{equation}
-q_{vs}(p,T) = \left( \frac{380.0}{p} \right) \exp\left(17.27 \times \frac{T-273.0}{T-36.0}\right)
-\end {equation}
-
-The autoconvection rate ($A_r$) and collection rate ($C_r$) follow Kessler parametrization and are defined by:
-\begin{align}
-A_r &= k_1(q_c-a) \\
-C_r &= k_2q_cq_r^{0.875}
-\end{align} With $k_1=0.001 \text{s}^{-1}$, $a=0.001 \text{g}.\text{g}^{-1}$ and $k_2=2.2 \text{s}^{-1}$
-
-Deriving from \cite{klemp1978simulation} description of cloud water,rain water and water vapor mixing ratios. they are define as followed:
-\begin{equation}
-q_c^{n+1}=\mbox{max}(q_c^r-\Delta q_r,0)
-\end{equation}
-\begin{equation}
-q_r^{n+1}=\mbox{max}(q_r^r-\Delta q_r+S,0)
-\end{equation} where $S$ is the sedimentation term and $\Delta q_r$ is defined as
-\begin{equation}
-\Delta q_r=q_c^n-\frac{q_c^n-\Delta \text{t}\ \mbox{max}(A_r,0)}{1+\Delta \text{t} C_r}
-\end{equation}
-
-The Rain evaporation equation is defined similarly to \cite{ogura1971numerical} description:
-\begin{equation}
-E_r=\frac{1}{\rho}\frac{\left(1-\frac{q_v}{q_{vs}}\right)C(\rho q_r)^{0.525}}{5.4\times10^5+\frac{2.55\times10^6}{pq_{vs}}}
-\end{equation}  With ventilation factor C define as 
-\begin{equation}
-C_r=1.6+124.9(\rho q_r)^{0.2046}
-\label{venti}
-\end{equation}
-
-The liquid water terminal velocity is similar to \cite{soong1973comparison} description with a mean density adjustment as suggested by \cite{kessler1969distribution}:
-\begin{equation}
-V_r = 36349(\rho q_r)^{0.1346}\left(\frac{\rho}{\rho_0}\right)^{-\frac{1}{2}}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Surface Fluxes on an Aqua-Planet with Prescribed Sea Surface Temperatures} \label{sec:OceanSurfaceFluxes}
-
-The forcing by surface fluxes from an idealized ocean is described in \cite{reed2012idealized} and is partly reproduced here.
-We use a model configuration which corresponds to an aqua-planet setup with prescribed sea surface temperatures (SSTs).
-This forcing by the surface fluxes is applied to the state variables in the lowermost model level using a partially implicit formulation to avoid numerical instabilities.
-Throughout this section we use the subscript $a$ to denote variables defined on the lowermost model level.
-
-The surface fluxes depend on the \textit{drag coefficient} $C_d$, defined as
-\begin{equation} \label{eq:DragCoefficient}
-\begin{array}{ll}
-C_d = C_{d0} + C_{d1} \vert \vec{v}_a \vert & \mbox{for $\vert \vec{v}_a \vert < 20$ m s$^{-1}$} \\ 
-C_d = 0.002 & \mbox{for $\vert \vec{v}_a \vert \geq 20$ m s$^{-1}$}, 
-\end{array}
-\end{equation}
-where $C_{d0}$ and $C_{d1}$ are $7.0 \times 10^{-4}$ (unitless) and $6.5 \times 10^{-5}$ s m$^{-1}$, respectively, and $\vert \vec{v}_a \vert$ is the magnitude of the horizontal wind at the lowermost model level.
-In terms of the zonal wind $u_a$ and meridional wind $v_a$, it is defined as
-\begin{equation}
-\vert \vec{v}_a \vert = \sqrt{u_a^2 + v_a^2}.
-\end{equation}  For both evaporation and sensible heat the bulk coefficient is set to
-\begin{equation} \label{eq:BulkTransferCoefficient}
-C_E = C_H = 0.0011.
-\end{equation}
-
-The formulation of the surface fluxes makes use of the height of the lowermost full model level $z_a$ (in m).
-For pressure-based models, $z_a$ can be expressed with the help of the hydrostatic equation in terms of pressure
-\begin{eqnarray} \label{eqn:za}
-z_a = \frac{R_d T_{\nu,a}}{g} \frac{(\ln p_s- \ln p_-)}{2},
-\end{eqnarray}
-where $T_{\nu,a} = T_a (1+0.608 q_a)$ is the virtual temperature at the lowermost full model level and $p_-$ is the edge pressure at the model level interface between the lowest and second lowest full model levels.
-This notation and all following equations assume that the temperature, horizontal wind components and the specific humidity in the physical parameterization package are co-located in both the vertical and horizontal directions, as is the case for the Lorenz grid.
-The height of the lowest full model level should ideally lie between 60-70m above the ground to make the results comparable to those in the literature.
-
-As described in \cite{reed2012idealized}, the surface fluxes can be written as
-\begin{eqnarray}
-\frac{\partial \vec{v}_a}{\partial t} &=& - \frac{C_d \vert \vec{v}_a \vert \vec{v}_a}{z_a} \label{sfu} \\
-\frac{\partial T_a}{\partial t} &=& \frac{C_H \vert \vec{v}_a \vert (T_s-T_a)}{z_a} \label{sft} \\
-\frac{\partial q_a}{\partial t} &=& \frac{C_E \vert \vec{v}_a \vert(q_{sat,s}-q_a)}{z_a}. \label{sfq}
-\end{eqnarray}
-We note that the wind at the surface is taken to be zero and therefore does not appear explicitly in (\ref{sfu}).
-In these equations $T_s$  denotes the prescribed sea surface temperature (SST) and $q_{sat,s}$ is the saturation specific humidity defined by the Clausius-Clapeyron equation
-\begin{eqnarray} \label{qsatcalc}
-q_{sat}(p) \approx \varepsilon\frac{e_s(T_s)}{p} \approx \frac{\varepsilon}{p}e^{*}_{0}e^{-(L/R_\nu)[(1/T_s)-(1/T_0)]},
-\end{eqnarray}
-where $e^{*}_{0}$ ($=610.78$ Pa) is the saturation vapor pressure at $T_0=273.16$ K.
-
-The final form of the surface fluxes will vary for models with other choices of prognostic variables. For example, if potential temperature $\Theta_a$ is used (\ref{sft}) takes the form
-\begin{eqnarray}
-\frac{\partial \Theta_a}{\partial t} &=& \frac{C_H \vert \vec{v}_a \vert (T_s-T_a)}{z_a} \left (\frac{p_0}{p_a} \right )^{R_d/c_p} \label{sftheta}
-\end{eqnarray}
-where $p_0 = 1000$ hPa is a reference pressure.
-This conversion uses the assumption that the pressure is time-invariant when individual physics parameterizations are applied.
-For other choices of prognostic variables like $(\rho u)_a$, $(\rho v)_a$, $(\rho \Theta)_a$ and $(\rho q)_a$ the right-hand-side of (\ref{sfu}), (\ref{sftheta}) and (\ref{sfq}) would need to be multiplied by the density of the air $\rho$.
-
-In order to ensure numerical stability, each of the aforementioned surface fluxes are applied via a semi-implicit operator.
-We demonstrate this procedure on the temperature evolution equation (\ref{sft}).
-First, the time derivative is expanded using a backward Euler operator,
-\begin{eqnarray}
-\frac{T_a^{n+1}-T_a^n}{\Delta t} &=& \frac{C_H \vert \vec{v}_a^n \vert (T_s-T_a^{n+1})}{z_a}.
-\end{eqnarray}
-The superscripts $n$ and $n+1$ represent the current time step (after the update from the large-scale condensation scheme) and the future time step, respectively.
-Note, that on the right-hand-side of the equation the only variable taken implicitly is $T_a$.
-$\vert \vec{v}_a^n \vert$ is evaluated at the current time step and $C_H$ is constant.
-The equation can now be solved for $T_a^{n+1}$
-\begin{eqnarray} \label{Tsurfflux}
-T_a^{n+1} &=& \frac{T_a^n + C_H \vert \vec{v}_a^n \vert T_s \frac{\Delta t}{z_a}}{1 + C_H \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}}.
-\end{eqnarray}
-Similar equations for $\vec{v}_a$ and $q_a$ can be calculated
-\begin{eqnarray}
-\vec{v}_a^{n+1} &=& \frac{\vec{v}_a^n}{1 + C_d^n \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}} \label{qusurfflux}\\
-q_a^{n+1} &=& \frac{q_a^n + C_E \vert \vec{v}_a^n \vert q_{sat,s}^n \frac{\Delta t}{z_a}}{1 + C_E \vert \vec{v}_a^n \vert\frac{\Delta t}{z_a}}, \label{qsurfflux}
-\end{eqnarray} 
-with the time-level dependent coefficient $C_d^n$.
-Notice that the second term in the numerator of (\ref{qusurfflux}) is absent in the case of the zonal and merdional wind.
-This is because the wind is set to zero at the surface.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Simplified Mixing in the Planetary Boundary Layer} \label{sec:PlanetaryBoundaryLayer}
-
-The forcing by the planetary boundary layer is described in \cite{reed2012idealized} and is partly reproduced here.
-To parameterize the surface fluxes that impact the zonal velocity $u$, the meridional velocity $v$ and moisture $q$ we start with the time rate of change equations
-\begin{eqnarray}
-\frac{\partial u}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'u'}}{\partial z} \label{eqturb}  \\
-\frac{\partial v}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'v'}}{\partial z}  \label{eqturb_v} \\
-\frac{\partial q}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'q'}}{\partial z}.  \label{eqturb_last}
-\end{eqnarray}  Potential temperature, as opposed to temperature, is used in the boundary layer parameterization because the vertical profile of the potential temperature is a suitable indicator of static stability.
-This adds the time rate of change equation
-\begin{eqnarray}
-\frac{\partial \Theta}{\partial t} &=& - \frac{1}{\rho} \frac{\partial \rho \ \overline{w'\Theta'}}{\partial z}.
-\end{eqnarray}  Here $u'$, $v'$, $w'$, $\Theta'$ and $q'$ symbolize the deviations of the zonal velocity, meridional velocity, vertical velocity, potential temperature and specific humidity from their averages, respectively.
-The average is indicated by an overbar.
-Note, assuming pressure is held constant (which is a common assumption in physical parameterizations), the potential temperature time tendency can be converted back to a temperature tendency of the following form
-\begin{eqnarray}
-\frac{\partial T}{\partial t} &=& - \frac{1}{\rho} \left (\frac{p}{p_0} \right )^{\kappa} \frac{\partial \rho \ \overline{w'\Theta'}}{\partial z}.
-\end{eqnarray}
-with the reference pressure $p_0 = 1000$ hPa.
-
-The turbulent mixing is characterized by a constant vertical eddy diffusivity to represent Ekman-like profiles of boundary layers
-\begin{eqnarray}\label{ktheory}
-\overline{w'u'} = -K_m \frac{\partial u}{\partial z} \\
-\overline{w'v'} = -K_m \frac{\partial v}{\partial z} \\
-\overline{w'\Theta'} = -K_E \frac{\partial \Theta}{\partial z} \\
-\overline{w'q'} = -K_E \frac{\partial q}{\partial z}. 
-\end{eqnarray}
-Here, $K_m$ is the eddy diffusivity coefficient for momentum and $K_E$ is the eddy diffusivity coefficient for energy and set equal to that for water vapor.
-In order to calculate the eddy diffusivity coefficients, the eddy diffusivity is matched to that for the surface flux calculated in Appendix~\ref{sec:OceanSurfaceFluxes} at the lowermost model level.
-To allow for a smooth transition above the boundary layer ($p_{top} = 850$ hPa)  the diffusivity coefficients for momentum taper to zero as
-\begin{equation} \label{Kmtaper}
-\begin{array}{ll}
-K_m = C_d \vert \vec{v}_{a} \vert z_a & \mbox{for} \; p > p_{top} \\
-K_m = C_d \vert \vec{v}_{a} \vert z_a \exp \left ( -\left [ \frac{p_{top} - p}{p_{strato}} \right ]^2 \right ) & \mbox{for} \; p \leq p_{top}.
-\end{array}
-\end{equation}
-Here the constant $p_{strato}$ determines the rate of decrease and is set to 100 hPa. $K_E$ is defined by
-\begin{equation}
-\begin{array}{ll}\label{KEtaper}
-K_E = C_E \vert \vec{v}_{a} \vert z_a & \mbox{for} \; p > p_{top} \\
-K_E = C_E \vert \vec{v}_{a} \vert z_a \exp \left ( -\left [ \frac{p_{top} - p}{p_{strato}} \right ]^2 \right ) & \mbox{for} \; p \leq p_{top}.
-\end{array}
-\end{equation}
-We suggest implementing the boundary layer scheme with an implicit temporal discretization to avoid numerical instabilities.
-The details of this discretization are somewhat complicated, and so we refer to implementation details in Appendix D of \cite{reed2012idealized}.
-In addition, we supply the DCMIP modeling groups with the complete ``simple-physics'' package as used in the model CAM which can serve as a template routine.}
-
 %\appendix

### 2017-04-27 DCMIP2016-Part1-Models.tex
- Added some more small modifications from James Kent

@@
 
-\Author[11]{David A.}{Hall}
+\Author[11]{David M.}{Hall}
 
@@
 $z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
-$u$ & Zonal wind \\
-$v$ & Meridional wind \\
-$w$ & Vertical velocity \\
+$u$ & Zonal wind velocity \\
+$v$ & Meridional wind velocity \\
+$w$ & Vertical wind velocity \\
 $\vb{u}$ & 3D wind vector \\
@@
 NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
-OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
 Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
@@
 MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
-NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
+NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
 OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
@@
 DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.
-It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature,\citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
+It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
 It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
@@
 Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
-Height is the vertical coordinate, and a Lorenz vertical grid stagger is used.
+Height is the vertical coordinate, and a Lorenz vertical grid staggering is used.
 A unique feature of OLAM is that grid levels are horizontal and intersect topography.
@@
 Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
-This grid is employed by the UK Met Office \citep{davies2005new}.
+This grid is employed by the UK Met Office \citep{davies2005new, wood2014inherently}.
 
@@
   \centering
-  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
+  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
  \end{subfigure}%
@@
   \centering
-  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
+  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
  \end{subfigure}
@@
 \includegraphics[height=6cm]{gem_yinyang.png}
-\caption{ Yin-Yang grid }
+\caption{Th Yin-Yang grid, a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
 \label{fig:gem_yinyang}
@@
 
-ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
+ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}), and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
 The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
@@
 
-OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:ContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and
+OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:EulerianContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and
 (\ref{eq:PotentialTemperatureFluxForm}).
@@
 Advection of all quantities, including all 3 velocity components that are diagnostically reconstructed at scalar cell centers, and advancement in time from the current to the future time level is based on the time- and space-centered Lax-Wendroff formulation.
-This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
+This scheme is horizontally explicit, but a trapezoidal- implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
+A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection.
+The vertical momentum equation is solved first so that the time-centered vertical momentum is available for computing transport of horizontal momentum and all scalar quantities.
+A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.
+
+%This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
 A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection.
@@
 GEM differs from the approaches above by using a semi-Lagrangian advection.
-Any model equations, prognostic or diagnostic, is written in the form
+Any model equations, prognostic or diagnostic, are written in the form
 \begin{align}
@@
 Consequently, the models discussed in this paper only represent a sample of the many dynamical cores that have been developed for general circulation modeling.
-Some of the models that have not been discussed include \cite{fox1997finite, davies2005new, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global} and \cite{doyle2014next}.
+Some of the models that have not been discussed include \cite{fox1997finite, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global, wood2014inherently} and \cite{doyle2014next}.
 
@@
 \pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
-\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis  that would arise from the momentum transport term $\vb{u} \cdot \nabla \vb{u}$ in (\ref{eq:PrognosticVelocity1}) (note that these derivatives sometimes appear as Christoffel symbols).
-In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
+\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis  that would arise from the momentum transport term $\vb{u} \cdot \nabla \vb{u}$ in (\ref{eq:PrognosticVelocity1}).
+In conjunction with (\ref{eq:ContinuityEquation}), (\ref{eq:PrognosticVelocity1}) also gives rise to the flux-form momentum equations,
 \begin{align} \label{eq:PrognosticVelocityFluxForm}
@@
 This work was made possible with support from our student and postdoctoral participants: Sabina Abba Omar, Scott Bachman, Amanda Back, Tobias Bauer, Vinicius Capistrano, Spencer Clark, Ross Dixon, Christopher Eldred, Robert Fajber, Jared Ferguson, Emily Foshee, Ariane Frassoni, Alexander Goldstein, Jorge Guerra, Chasity Henson, Adam Herrington, Tsung-Lin Hsieh, Dave Lee, Theodore Letcher, Weiwei Li, Laura Mazzaro, Maximo Menchaca, Jonathan Meyer, Farshid Nazari, John O'Brien, Bjarke Tobias Olsen, Hossein Parishani, Charles Pelletier, Thomas Rackow, Kabir Rasouli, Cameron Rencurrel, Koichi Sakaguchi, G\"okhan Sever, James Shaw, Konrad Simon, Abhishekh Srivastava, Nicholas Szapiro, Kazushi Takemura, Pushp Raj Tiwari, Chii-Yun Tsai, Richard Urata, Karin van der Wiel, Lei Wang, Eric Wolf, Zheng Wu, Haiyang Yu, Sungduk Yu and Jiawei Zhuang.
-We would also like to thank Rich Loft, Cecilia Banner, Kathryn Peczkowicz and Rory Kelly (NCAR), Carmen Ho, Perla Dinger, and Gina Skyberg (UC Davis) and Kristi Hansen (University of Michigan) for administrative support during the workshop and summer school.
+We would also like to thank Rich Loft, Cecilia Banner, Kathryn Peczkowicz and Rory Kelly (NCAR), Perla Dinger, Carmen Ho, and Gina Skyberg (UC Davis) and Kristi Hansen (University of Michigan) for administrative support during the workshop and summer school.
 \end{acknowledgements}

### 2017-04-28 DCMIP2016-Part1-Models.tex
update text formulations for FVM

@@
 
-The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
-ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-are available by means of simple switches. The formulation incorporates 
-a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
-A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
-Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-Conjugate Residual approach. The dynamics time step is adapted 
-at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-than 1). The physics time step is identical to the dynamics time step.
-
-The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
+ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
+compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
+A centered two-time-level semi-implicit 
+integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
+\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
+preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
+non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
 advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-The horizontal spatial discretization is fully unstructured finite-volume using the 
-median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-In both the horizontal and the vertical discretization, all prognostic variables are 
-co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
-reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.
-
-No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
-the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-the form of a Rayleigh-type forcing to the prognostic variables.
+The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
+combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
+the vertical discretization all variables are co-located. 
+The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
+Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
+grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
+interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
+broad classes of meshes including adaptivity, e.g.~\cite{szmelter2010,kuehnleinJCP2012}.
+%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.
+
+%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
+%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
+%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
+%are available by means of simple switches. The formulation incorporates 
+%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
+%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
+%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
+%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
+%Conjugate Residual approach. The dynamics time step is adapted 
+%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
+%than 1). The physics time step is identical to the dynamics time step. 
+% 
+%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
+%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
+%The horizontal spatial discretization is fully unstructured finite-volume using the 
+%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
+%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
+%In both the horizontal and the vertical discretization, all prognostic variables are 
+%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
+%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
+%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
+%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
+%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
+%
+%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
+%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
+%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
+%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
+%the form of a Rayleigh-type forcing to the prognostic variables.
 
@@
 is suitable for transformations involving spherical harmonics, and has been introduced for global 
-medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
+numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
 Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
@@
 \caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
-connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
-24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
+connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse octahedral grid with only
+24 latitudes between pole and equator 'O24' is used for illustration. The dual mesh resolution of the octahedral
 reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
@@
 
-FVM solves conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, and dry 
-entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
+The FVM formulation is based on conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, 
+and dry entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
 which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
-\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Underlying the governing equations 
-in FVM is a perturbational form with respect to a balanced ambient state, and a generalized curvilinear 
-coordinate formulation in a geospherical framework. These governing equations can 
-concisely be written as
-\citep{smolarkiewiczetalJCP2017}
+\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Moreover, underlying the conservation laws 
+in FVM is a perturbational form with respect to a balanced ambient state and a generalized curvilinear 
+coordinate formulation in a geospherical framework. Following \cite{smolarkiewiczetalJCP2017}, the
+governing equations can concisely be written as
 \begin{subequations} \label{fvm:compressible}
@@
 \label{fvm:mass}  \\
-& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left[\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
+& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
 \mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
@@
 \eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
-subscript `a' (see also \cite{prusa2008eulag,smolarkiewiczJCP2014}).   As far as geometric aspects are concerned, the nabla operator $\overline{\nabla}$ represents
+subscript `a', see \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. As far as geometric aspects 
+are concerned, the nabla operator $\overline{\nabla}$ represents
 the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
 a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
-the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see also \cite{prusa2003all} 
-and \cite{kuehnleinJCP2012}.
-The symbol $\mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
-subsumes the spherical Christoffel symbols \citep[][]{smolarkiewiczetalJCP2017}.
+the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$ where a contribution from
+optional time-dependency of the curvilinear coordinates is neglected for simplicity \citep{kuehnleinJCP2017}.
+The symbol $\mathcal{M}' = \mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
+subsumes the metric forces in the spherical domain, see \cite{smolarkiewiczetalJCP2017}.
 
@@
 of the fast buoyant and acoustic modes, and the slow rotational modes. Therefore, the model time step is 
-identical for all processes and selected according to the stability of the advective transport scheme---i.e.~%
+identical for all processes and typically selected with regard to the stability of the advective transport scheme---i.e.~%
 the time step is continuously adapted according to a given maximum advective Courant number 
@@
 A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016,kuehnleinJCP2017} for dry dynamics, 
-and in \cite{kurowskiJAS2014,smolarkiewiczetalJCP2017} for moist-precipitating dynamics. Here,
-we provide an outline of the basic integration scheme for the compressible Euler equations
-\eqref{fvm:compressible}. The scheme employs the two-time-level second-order-accurate template 
-algorithm given as
+and in \cite{kurowskiJAS2014,smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
+Here, we provide a short outline of the solution procedure for the compressible Euler 
+equations \eqref{fvm:compressible}. It employs the two-time-level second-order-accurate template algorithm given as
 \begin{equation}
@@
 where $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
-$\mathcal{A}$ symbolises the advective transport operator---the non-oscillatory 
-finite-volume MPDATA bespoke for the compressible dynamics on the hybrid horizontally-unstructured
-vertically-structured mesh \citep{kuehnleinJCP2017}---and the vector index $\mathbf{i} \equiv (k,i)$ denotes the 
-spatial position on the computational mesh.
-
-The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
-mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
-and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
-momentum \eqref{fvm:momentum}, and moisture equations enter
-\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
-which is generally depending on all prognostic variables. The high degree of implicitness in the representation of the rhs forcings
+$\mathcal{A}$ symbolises the advective transport operator given by the non-oscillatory finite-volume 
+MPDATA  scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}, and the spatial mesh vector 
+index $\mathbf{i} \equiv (k,i)$ denotes the position on the hybrid horizontally-unstructured 
+vertically-structured computational mesh.
+
+The solution procedure of the system \eqref{fvm:compressible} can then basically be divided into three steps. First, the homogenous
+mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
+and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, given already updated moisture variables 
+\citep{smolarkiewiczetalJCP2017}, the thermodynamic \eqref{fvm:thermodynamic} and momentum \eqref{fvm:momentum} equations enter
+\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
+which is generally depending on all these prognostic variables. The high degree of implicitness in the representation of the rhs forcings
 is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
-velocity updates---the procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
+velocity updates---this procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
 Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term.
@@
 The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
-Residual approach; see \cite{smolarkiewiczAG2011} and references therein. Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
+Residual approach; see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
+Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
 solution-dependent coefficients of the Helmholtz equation are lagged behind and executed in an outer iteration.

### 2017-04-28 DCMIP2016-Part1-Models.tex
V4 update

@@
 
-\title{DCMIP2016:  A Review of Nonhydrostatic Dynamical Core Design and Intercomparison of Participating Models}
+\title{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models}
 
@@
 
-\runningtitle{DCMIP2016: A Review of Nonhydrostatic Dynamical Core Design and Intercomparison}
+\runningtitle{DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison}
 
@@
 $\theta_{il}$ & Ice-liquid potential temperature \\
+$\theta_\rho$ & Density potential temperature \\
 %$P_{ls}$ & Large-scale precipitation rate \\
-$q$ & Water vapor mixing ratio \\
+$q$ & Specific humidity \\
+$q_v$ & Water vapor mixing ratio \\
 $q_c$ & Cloud water mixing ratio \\
@@
 
-\section{Dynamical Cores} \label{sec:DynamicalCores}
+\section{Dynamical cores} \label{sec:DynamicalCores}
 
@@
 GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
-HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
+ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories / University of Colorado, Boulder \\
 ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
@@
 GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
-HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
-ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
+ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
 MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
@@
 GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
-HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
+ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 ICON & Lorenz & fixed height \\
@@
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
-HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
+ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
 ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
@@
 
-\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}
+\subsection{Global Environmental Multiscale (GEM) model} \label{sec:GEM_core}
 
@@
 
-\subsection{High-Order Method Modeling Environment (HOMME)}
-
-The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.
+\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}
+
+The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}.
+ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model.
+Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
+ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
+The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
+Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
+All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
+Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
+
+%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.
 As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present.
@@
 
-%A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$.
-The PDEs are discretized using a split approximation, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.
-
 \subsection{Icosahedral Non-hydrostatic model (ICON)}
 
-The ICON model \citep{zangl2014icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
+The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
 Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
@@
 
-\section{Model Grids} \label{sec:ModelGrids}
+\section{Model grids} \label{sec:ModelGrids}
 
@@
 \includegraphics[height=6cm]{gem_yinyang.png}
-\caption{Th Yin-Yang grid, a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
+\caption{The Yin-Yang grid is a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
 \label{fig:gem_yinyang}
@@
 
-\subsection{Vertical Coordinates} \label{sec:VerticalCoordinates}
+\subsection{Vertical coordinates} \label{sec:VerticalCoordinates}
 
@@
 
-The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
-\begin{align} \label{eq:SurfacePressureEquation}
-\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
-\end{align} where the integrals are taken over the column.
+%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
+%\begin{align} \label{eq:SurfacePressureEquation}
+%{\color{red}\pdiff{p_s}{t} = - \int g \pdiff{\rho}{t} \left( \pdiff{z}{s} \right) ds = - \int \nabla_s \cdot (\rho_s \vb{u}_h) \left( \pdiff{z}{s} \right) ds,}
+%\end{align} where the integrals are taken over the column.
 
@@
 
-\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and HOMME)} \label{sec:FloatingLagrangian}
+\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and ACME--A)} \label{sec:FloatingLagrangian}
 
@@
 
-\subsection{Prognostic equations} \label{sec:PrognosticEquations}
+\subsection{Prognostic equations and treatment of moisture} \label{sec:PrognosticEquations}
 
@@
 The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.
-The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models.
+The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models, including any special treatment of terms related to moist physics.
 
@@
 \mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
-- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
+- \mathbf{k} \frac{g_c}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
 - 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
@@
 \frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
-\frac{d w}{dt}- g \mu &= 0, \\
+\frac{d w}{dt}- g_c \mu &= 0, \label{eq:GEMVerticalVelocity} \\
 \frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
@@
 \frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
-\frac{d \Phi}{dt}-  g w & = 0.
+\frac{d \Phi}{dt}-  g_c w & = 0.
 \end{align}
@@
 
-\subsubsection{HOMME}
-
-HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics.
-The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure (\ref{eq:SurfacePressureEquation}).
-
-In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model.
-This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}.
-In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical.
-Prognostic variables include the 3d wind velocity components $u,v,w$ ((\ref{eq:PrognosticWEquationK2Sb}) and (\ref{eq:PrognosticUhEquationK2Sb})), the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).
-
-%Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering.
-
-% The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios.
-Tracer advection is discretized in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}.
-The dynamics are also discretized using one of several explicit Runge-Kutta times schemes.
-Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.
-
-%This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability.
-Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
+\subsubsection{ACME--A}
+
+ACME-A hosts an experimental compressible shallow-atmosphere model in hybrid terrain-following pressure vertical coordinates $\eta$, similar to the model of \cite{laprise1992euler}.
+The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), pseudodensity (\ref{eq:PrognosticPseudoDensity}), and geopotential (\ref{eq:PrognosticShallowGeopotential}).
+The vertical velocity equation is formulated analogous to that of GEM (\ref{eq:GEMVerticalVelocity}).
 
@@
 
-ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}), and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.
-The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
-\begin{align}
-\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
-\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.
-The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangential velocity components $v_t$.
+ICON solves a non-hydrostatic equation set based on \cite{gassmann2008} 
+using terrain-following z-coordinates. The governing equations describe the mixture 
+of a two-component system of dry air and water, where water is allowed to occur in 
+all three phases, including precipitating constituents. Following \cite{wacker2006}, 
+the barycentric velocity $\vb{u}_{bc}=\sum_{k} \rho_{k} \vb{u}_{k}/\sum_{k}\rho_{k}$ -- 
+that is the mass-weighted sum of all constituent-specific velocites (including sedimenting ones) --
+is used as a prognostic variable. In contrast to \cite{gassmann2008}, 
+a vector invariant form is only used for the horizontal velocity equation 
+(\ref{eq:HorizontalVelocityOrthogonal}), whereas the vertical velocity equation is solved in advective form. 
+The pressure gradient force is formulated according to (\ref{eq:PressureGradientForceAlt}).
+
+%Note that in contrast to (\ref{eq:VirtualTemperatureExact}) the water loading term is included in the definition of $T_{v}$ and $\theta_{v}$.
+
+Additional prognostic variables include total air density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), 
+and mass fractions $q_{k}=\rho_{k}/\rho$ of all constituents (except for dry air) for which the 
+prognostic equation reads
+\begin{align}
+\frac{\partial \rho q_{k}}{\partial t} + \nabla\cdot\left(\rho q_{k} \vb{u}_{bc}\right) = 
+-\nabla\cdot \vb{J}_{k}  + \sigma_{k}\,,
+\end{align}
+with $\sigma_{k}$ describing sources/sinks due to phase changes, and  
+$\vb{J}_{k}=\rho q_{k}\left(\vb{u}_{k} - \vb{u}_{bc}\right)$ denoting diffusion fluxes, 
+which account for the motion of constituents relative to the frame of reference set by $\vb{u}_{bc}$.
+
+The specific heat capacities and ideal gas constant are approximated to be equal to their dry values 
+$R^{\ast}\approx R_{d}$, $c_{p}^{\ast}\approx c_{pd}$, $c_{v}^{\ast}\approx c_{vd}$. The model also 
+uses a prognostic equation for Exner pressure to simplify the treatment of vertical sound wave propagation, given by 
+\begin{align}
+\frac{\partial \pi}{\partial t} + \frac{R_{d}}{c_{vd}}\frac{\pi}{\rho \theta_{v}} \nabla\cdot\left(\vb{u}_{bc}\rho\theta_{v}\right) = \hat{Q}\,,
+\end{align}
+where $\hat{Q}$ is an appropriately formulated diabatic heat term. The horizontal uses a Arakawa C-grid formulation on the 
+triangular grid to prognose horizontal velocities normal to triangle edges $v_{n}$, making use of reconstructed 
+tangential velocity components $v_{t}$.
+
+In the current implementation, the following simplifcations are made with regards to the treatment of moisture: 
+The atmospheric mass loss/gain due to precipitation/evaporation is neglected in the total mass continuity 
+equation (\ref{eq:EulerianContinuityEquation}), by setting the vertical component of $\vb{u}_{bc}$ to zero at the lower boundary:  
+$w_{bc}|_{sfc} = 0$. In addition, only the vertical diffusion fluxes $\vb{J}_{k}$ of sedimenting 
+constituents and the surface evaporation flux $\vb{J}_{v}|_{sfc}$ are taken into account. The counter-flux 
+of non-sedimenting constituents is discarded. Since in the given framework the continuity equation (\ref{eq:EulerianContinuityEquation}) 
+is only valid if the constraint $\sum_{k} \vb{J}_{k}=0$ holds, we (implicitly) assume a fictitious 
+counter-flux of dry air to compensate for the considered vertical diffusion fluxes. As a consequence, 
+ICON currently conserves the global integral of total air mass rather than dry air mass.
 
@@
 
-\section{Diffusion, Stabilization, Filters and Fixers} \label{sec:DiffusionStabilization}
+\section{Diffusion, stabilization, filters and fixers} \label{sec:DiffusionStabilization}
 
@@
 
-\subsection{HOMME / Tempest}
-
-In both HOMME and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
+\subsection{ACME--A / Tempest}
+
+In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
 Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
@@
 The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2014icon}, and are summarized here.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2015icon}, and are summarized here.
 For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the (relatively) computationally  expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions,
@@
 
-\section{Temporal Discretizations} \label{sec:TemporalDiscretizations}
+\section{Temporal discretizations} \label{sec:TemporalDiscretizations}
 
@@
 
-\subsection{Mixed Implicit-Explicit, Forward-Backward, Semi-Implicit and Additive Runge-Kutta Schemes}
+\subsection{Mixed implicit-explicit, forward-backward, semi-implicit and additive Runge-Kutta schemes}
 
@@
 ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
-No time splitting is used with respect to sound waves, because the ratio of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
+No time splitting is used with respect to sound waves, because the ratio between the speed of sound and the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
 Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
 Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
-Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014icon}.
+Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2015icon}.
 
@@
 
-\conclusions[Summary and Conclusions] \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
+\conclusions[Summary and conclusions] \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
 
@@
 \diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
-\end{align} where $\vb{u}$ denotes the 3D vector velocity.  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
+\end{align} where $\vb{u}$ denotes the 3D vector velocity.
+Note that tracer variables $q_i$, including specific humidity $q$, in the absence of sources and sinks satisfy the simple Lagrangian relationship
 \begin{align} \label{eq:TracerTransport}
@@
 Common ratios $\kappa = R^\ast / c_p^\ast$, $\varepsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
+Note that as additional water species are added (cloud water, rain water, etc.) additional independent variables are needed to capture the thermodynamic effects of these species, and the ``virtual'' quantities modified accordingly, for instance through the adoption of density potential temperature $\theta_\rho$.
 
@@
 \end{align} which arises from the exact relationship
-\begin{align}
+\begin{align} \label{eq:VirtualTemperatureExact}
 T_v = \frac{T}{1 - \frac{e}{p} (1 - \varepsilon)},
@@
 Note that it is common to rewrite the pressure gradient force using the relationship
-\begin{align}
+\begin{align} \label{eq:PressureGradientForceAlt}
 - \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
-\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
+\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).
+Note that when condensate loading is included, the virtual potential temperautre in this equation is replaced with $\theta_\rho$.
+A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
 \begin{align}
@@
 \begin{align} \label{eq:DeepAtmosphereGeopotential}
-\Phi = g a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \varphi + \vb{j} \cos \varphi),
+\Phi = g_c a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \varphi + \vb{j} \cos \varphi),
 \end{align} where $g$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\varphi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
@@
 \begin{align}
-\diff{\Phi}{t} = \frac{a^2 g w}{(a + z)^2}.
+\diff{\Phi}{t} = \frac{a^2 g_c w}{(a + z)^2}.
 \end{align}  For \textit{shallow atmosphere} models, the geopotential takes the simpler form
 \begin{align} \label{eq:ShallowAtmosphereGeopotential}
-\Phi = g z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \varphi \vb{k},
+\Phi = g_c z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \varphi \vb{k},
 \end{align} where $z$ is the altitude above the surface.
@@
 \begin{align} \label{eq:PrognosticShallowGeopotential}
-\diff{\Phi}{t} = g w.
+\diff{\Phi}{t} = g_c w.
 \end{align}
@@
 
-\authorcontribution{TEXT}
+\authorcontribution{Text in this manuscript describing individual models was provided by the respective modeling teams.  Final composition and development of Appendix A was performed by P.A. Ullrich.}
 
@@
 1629819), the National Aeronautics and Space Administration (award no.
-{\color{red}??}), the National Oceanic and Atmospheric Administration (award no.
-{\color{red}??}), the Office of Naval Research and CU Boulder Research Computing.
+NNX16AK51G), the National Oceanic and Atmospheric Administration (award no.
+NA12OAR4320071), the Office of Naval Research and CU Boulder Research Computing.
 This work was made possible with support from our student and postdoctoral participants: Sabina Abba Omar, Scott Bachman, Amanda Back, Tobias Bauer, Vinicius Capistrano, Spencer Clark, Ross Dixon, Christopher Eldred, Robert Fajber, Jared Ferguson, Emily Foshee, Ariane Frassoni, Alexander Goldstein, Jorge Guerra, Chasity Henson, Adam Herrington, Tsung-Lin Hsieh, Dave Lee, Theodore Letcher, Weiwei Li, Laura Mazzaro, Maximo Menchaca, Jonathan Meyer, Farshid Nazari, John O'Brien, Bjarke Tobias Olsen, Hossein Parishani, Charles Pelletier, Thomas Rackow, Kabir Rasouli, Cameron Rencurrel, Koichi Sakaguchi, G\"okhan Sever, James Shaw, Konrad Simon, Abhishekh Srivastava, Nicholas Szapiro, Kazushi Takemura, Pushp Raj Tiwari, Chii-Yun Tsai, Richard Urata, Karin van der Wiel, Lei Wang, Eric Wolf, Zheng Wu, Haiyang Yu, Sungduk Yu and Jiawei Zhuang.

### 2017-05-01 DCMIP2016-Part1-Models.tex
more cosmetics regarding FVM parts of text

@@
 FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
-FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
+FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_v$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
 GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
@@
 interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
-broad classes of meshes including adaptivity, e.g.~\cite{szmelter2010,kuehnleinJCP2012}.
+broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
 %No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.
@@
 latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
-is suitable for transformations involving spherical harmonics, and has been introduced for global 
-numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
+is suitable for transformations involving spherical harmonics, and has been introduced for operational 
+weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
 Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
@@
 24 latitudes between pole and equator 'O24' is used for illustration. The dual mesh resolution of the octahedral
-reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
+reduced Gaussian grid is about a factor 2 finer at the poles than the equator, see \cite{smolarkiewiczetalJCP2016}.}
 \label{fig:octahedralgrid}
@@
 coordinate formulation in a geospherical framework. Following \cite{smolarkiewiczetalJCP2017}, the
-governing equations can concisely be written as
+FVM governing equations can concisely be written as
 \begin{subequations} \label{fvm:compressible}
@@
 \mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
-- \mathbf{k} \frac{g_c}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
+%- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
+%- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right)
+- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\varepsilon_b\,q'_v - q_c - q_r \right)
 - 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
@@
 \eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
-subscript `a', see \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. As far as geometric aspects 
+subscript `a', see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014} for discussions.
+The symbol $g$ in \eqref{fvm:momentum} denotes the gravitational acceleration 
+and $\varepsilon_b = 1/\varepsilon-1$. As far as geometric aspects 
 are concerned, the nabla operator $\overline{\nabla}$ represents
@@
 The symbol $\mathcal{M}' = \mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
-subsumes the metric forces in the spherical domain, see \cite{smolarkiewiczetalJCP2017}.
+subsumes the metric forces in the spherical domain \citep{smolarkiewiczetalJCP2017}.
 
@@
 permitted by the MPDATA scheme.
-A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016,kuehnleinJCP2017} for dry dynamics, 
-and in \cite{kurowskiJAS2014,smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
+A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016}
+and \cite{kuehnleinJCP2017} for dry dynamics, 
+whereas in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
 Here, we provide a short outline of the solution procedure for the compressible Euler 
@@
 The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
-Residual approach; see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
+Residual approach, see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
 Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 

### 2017-05-01 DCMIP2016-Part1-Models.tex
correct g -> g_c

@@
 \Phi = g_c a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \varphi + \vb{j} \cos \varphi),
-\end{align} where $g$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\varphi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
+\end{align} where $g_c$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\varphi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
 For models that don't utilize a height-based vertical coordinate, the geopotential is generally treated as a prognostic variable, with an evolution equation that emerges from the definition $w = dz/dt$,

### 2017-05-03 DCMIP2016-Part1-Models.tex
- Some small restructuring to keep horizontal stuff and vertical stuff separate

@@
 Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.
-The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
+The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  section \ref{sec:HorizontalDiscretization} describes aspects of the horizontal discretization, including model grids and horizontal placement of prognostic variables; section \ref{sec:VerticalDiscretization} describes the vertical placement of model variables and choice of vertical coordinates; section \ref{sec:PrognosticEquations} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
 The summary and conclusions then follow in section \ref{sec:Conclusions}.
@@
 \begin{table}[p]
-\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
+\caption{Participating modeling centers and associated dynamical cores that have submitted a model description and/or simulation results.} \label{tab:Models}
 \begin{center}
 \begin{tabular}{cll}
-\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
+\hline Short Name & Long Name & Modeling Center or Group \\ \hline 
+ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories and \\
+& & University of Colorado, Boulder, USA \\
 CSU & Colorado State University Model & Colorado State University, USA \\
@@
 GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
-ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories / University of Colorado, Boulder \\
 ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
@@
 OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
-Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
+TEMPEST & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
 \hline 
@@
 \begin{tabular}{cccccc}
-\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
+\hline Short Name & Equation Set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
 & & & & method & staggering \\ \hline
-CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
-DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
-FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_v$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
-GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
-ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
-ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
-MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
-NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
-OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
+CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
+DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
+FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_i$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
+GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q_i$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
+ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q_i$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
+MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q_i$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
+NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
+OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
+TEMPEST & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
 \hline 
@@
 \hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
+ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 CSU & Lorenz & fixed height \\
@@
 GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
-ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
 ICON & Lorenz & fixed height \\
@@
 OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
-Tempest & Lorenz & fixed height \\
+TEMPEST & Lorenz & fixed height \\
 \hline 
@@
 & stabilization, filters, or fixers & \\ \hline 
+ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
 CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
@@
 GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
-ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
 ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
@@
 OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
-Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
+TEMPEST & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
 \hline 
@@
 Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
+
+\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}
+
+The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}.
+ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model.
+Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
+ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
+The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
+Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
+All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
+Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
+
+%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.
+As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present.
+The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
+The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
+These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.
+The model employs a hybrid terrain-following pressure coordinate in the vertical.
+The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
+Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
+Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.
 
@@
 
-\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}
-
-The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}.
-ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model.
-Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
-The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
-Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
-All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
-
-%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.
-As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present.
-The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.
-The model employs a hybrid terrain-following pressure coordinate in the vertical.
-The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.
-
 \subsection{Icosahedral Non-hydrostatic model (ICON)}
@@
 
-\subsection{Tempest}
+\subsection{TEMPEST}
 
@@
 
-\section{Model grids} \label{sec:ModelGrids}
-
-Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.
-In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
+\section{Horizontal discretization and model grids} \label{sec:HorizontalDiscretization}
+
+The horizontal discretization determines how the atmosphere, which consists of a set of approximately continuous fields, is mapped into a very limited and discrete computational space.
+The horizontal discretization essentially consists of two major choices:  the model grid, which determines the density and connectivity of discrete regions \citep{staniforth2012horizontal}, and the arrangement of prognostic and diagnostic variables around each grid region \citep{arakawa1977computational}.
+In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have explored a number of options for model grids.
 The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.
-The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.
-Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.
+The choice of grid may be further decided by the numerical method -- for instance, finite element models that use tensor products to define basis functions require grids consisting entirely of quadrilaterals.
+Inevitably a choice must be made, and the pros and cons of that choice will impact other decisions related to the model.
 To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.
+Then, in section \ref{sec:HorizontalStaggering}, we discuss the ``staggering'' of model variables, referring to the distribution of variables within and around each grid cell.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Variable placement and prognosis} \label{sec:VariablePlacement}
-
-Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.
-Hence, in this section we turn our attention to the placement of prognostic variables on the grid, the choice of vertical coordinate, and the choice of corresponding prognostic equations.
-
 \subsection{Horizontal staggering} \label{sec:HorizontalStaggering}
 
-The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.
+The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, dispersion properties, and the maximum stable timestep size \citep{randall1994geostrophic, ullrich2014understanding}.
 The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.
 Later, other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.
-New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.
-Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).
-
-There are arguments to be made for and against each of these grids:  the A-grid is relatively simple to implement and tends to support large time step sizes, but also supports a $2 \Delta x$ computational mode (even in the context of finite element methods); the C-grid has excellent dispersion properties, and does not support extraneous computational modes (as long as the number of edges is equal to twice the number of volumes), but also requires a timestep that is half that of the A-grid; the D-grid provides a better representation of vorticity, but suffers to similar computational modes as the A-grid that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
+By interpreting ``staggerings'' to be analogous to a choice of finite element basis, new staggerings are under development in the context of mixed finite element methods \citep{cotter2012mixed}.
+Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves simple co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).
+
+Arguments in favor or against particular staggerings have generally emerged from linear analyses, and typically in the absence of either implicit or explicit diffusion.
+In this context, the A-grid tends to support large time step sizes, but produces unphysical phase speeds and negative group velocities at high wavenumbers, including a stationary $2 \Delta x$ wavelength mode (even in the context of finite element methods); the C-grid better represents short wave modes, does not support extraneous computational modes (as long as the number of edges is equal to twice the number of volumes), but also requires a timestep that is half that of the A-grid (in the 2nd order case), or 40\% smaller (in the 4th order case) \citep{ullrich2014understanding}; the D-grid provides a better representation of vorticity, but produces unphysical effects analogous to those on the A-grid at high wavenumbers that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
 
@@
 More details on this approach can be found in \cite{LR1997QJR}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Vertical discretization} \label{sec:VerticalDiscretization}
+
+%Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.
+Hence, in this section we turn our attention to the placement of prognostic variables on the grid, the choice of vertical coordinate, and the choice of corresponding prognostic equations.
+
+Because of the vast differences between horizontal and vertical scales, most atmospheric models use dimension splitting in order to separate the horizontal discretization from the vertical discretization.
+In this section, design considerations related to the vertical column are discussed, including the staggering of prognostic and diagnostic variables, and the choice of vertical coordinate.
 
@@
 
-Most atmospheric models use dimension splitting in order to separate the evolution equations for momentum into horizontal and vertical directions.
-Whereas ``horizontal'' in this context either refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g.
+In the context of dimension splitting, ``horizontal'' typically refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g.
 terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.
-In this context, the use of dimension splitting allows us to choose an arbitrary function $s(t,\vb{x},z)$ that is monotonic in $z$, so that model interfaces are equally spaced with respect to $s$.
+Vertical position is typically labelled using an arbitrary function $s(t,\vb{x},z)$ that is monotonic in $z$, so that model interfaces are equally spaced with respect to $s$.
 Typically $s$ is chosen so that the Earth's surface (the bottom boundary of the atmosphere) is a coordinate surface, allowing for easy specification of boundary conditions for the prognostic equations -- this leads to the so-called ``terrain-following'' family of vertical coordinates.
@@
 Here  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
-and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-
-\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and ACME--A)} \label{sec:FloatingLagrangian}
+$s_{top}$ is the coordinate value at the uppermost interface, and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
+
+\subsubsection{Floating Lagrangian coordinates (ACME--A, DYNAMICO and FV$^3$)} \label{sec:FloatingLagrangian}
 
@@
 
-\subsection{Prognostic equations and treatment of moisture} \label{sec:PrognosticEquations}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Prognostic equations and treatment of moisture} \label{sec:PrognosticEquations}
 
@@
 A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.
-The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.
+The particular prognostic equations used by the model can impact the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.
 The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models, including any special treatment of terms related to moist physics.
 
-\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
+\subsection{DYNAMICO} \label{sec:DYNAMICOEquations}
 
@@
 
-\subsubsection{FV$^3$} \label{sec:FV3Equations}
+\subsection{FV$^3$} \label{sec:FV3Equations}
 
@@
 
-\subsubsection{FVM} \label{sec:FVMEquations}
+\subsection{FVM} \label{sec:FVMEquations}
 
@@
 
-\subsubsection{GEM} \label{sec:GEM_equations}
+\subsection{GEM} \label{sec:GEM_equations}
 
@@
 \frac{d \Phi}{dt}-  g_c w & = 0.
-\end{align}
+\end{align}  Here $\nabla_\zeta$ denotes the horizontal gradient along $\zeta$ surfaces.
 
@@
 
-\subsubsection{ACME--A}
+\subsection{ACME--A}
 
@@
 
-\subsubsection{ICON} \label{sec:ICONEquations}
+\subsection{ICON} \label{sec:ICONEquations}
 
@@
 all three phases, including precipitating constituents. Following \cite{wacker2006}, 
-the barycentric velocity $\vb{u}_{bc}=\sum_{k} \rho_{k} \vb{u}_{k}/\sum_{k}\rho_{k}$ -- 
+the barycentric (bc) velocity $\vb{u}_{bc}=\sum_{k} \rho_{k} \vb{u}_{k}/\sum_{k}\rho_{k}$ -- 
 that is the mass-weighted sum of all constituent-specific velocites (including sedimenting ones) --
@@
 
-\subsubsection{OLAM} \label{sec:OLAMEquations}
+\subsection{OLAM} \label{sec:OLAMEquations}
 
@@
 
-\subsubsection{Tempest} \label{sec:TempestEquations}
+\subsection{TEMPEST} \label{sec:TempestEquations}
 
@@
 
-\subsection{ACME--A / Tempest}
+\subsection{ACME--A / TEMPEST}
 
@@
 \begin{align} \label{smagn2}
-F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
-& F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
-\end{align}
+F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,, & F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,,
+\end{align} where $a_c$ and $a_e$ denotes the area associated with the cell and edge under consideration, respectively.
 An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
@@
 The damping rate is zero at the bottom of the damping layer and increases upward, usually linearly.
-
 Throughout the model domain, vertical vorticity is filtered horizontally at the smallest resolvable scale in order to control a spurious computational mode.
@@
 The vorticity filter is constructed so as to have zero impact on divergence at any scale.
-
 Upwinding in the Lax-Wendroff formulation of the advection operator provides sufficient damping that no other type of filtering is required.
@@
 The vast diversity of the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.
-Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.
+Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the kinetic energy spectrum \citep{skamarock2004evaluating}.
 The repercussions that emerge from these choices can then be explored in the context of idealized test cases, such as the ones that have been proposed as part of DCMIP.
@@
 $c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{pv}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
 $c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
+$c_{vv}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
 $R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
+$R_v$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
 %$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
-$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
+$\varepsilon$ & Ratio of $R_d$ to $R_v$ & $0.622$ \\
 $M_v$ & Constant for virtual temperature conversion & $0.608$ \\
@@
 \begin{align}
-R^\ast =& R_d + (R_w - R_d) q, & c_p^\ast =& c_{pd} + (c_{pw} - c_{pd}) q, & c_v^\ast =& c_{vd} + (c_{vw} - c_{vd}) q.
+R^\ast =& R_d + (R_v - R_d) q, & c_p^\ast =& c_{pd} + (c_{pv} - c_{pd}) q, & c_v^\ast =& c_{vd} + (c_{vv} - c_{vd}) q.
 \end{align}  Note that in many models, $R^\ast$, $c_p^\ast$ and $c_v^\ast$ are approximated by $R_d$, $c_{pd}$ and $c_{vd}$, respectively.
@@
 Key thermodynamic variables include dry air density $\rho_d$, moist density $\rho$, pressure $p$, vapor pressure $e$, temperature $T$, virtual temperature $T_v$, Exner pressure $\pi$, potential temperature $\theta$, and virtual potential temperature $\theta_v$.
-Common ratios $\kappa = R^\ast / c_p^\ast$, $\varepsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
+Common ratios $\kappa = R^\ast / c_p^\ast$, $\varepsilon = R_d / R_v$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
 Note that as additional water species are added (cloud water, rain water, etc.) additional independent variables are needed to capture the thermodynamic effects of these species, and the ``virtual'' quantities modified accordingly, for instance through the adoption of density potential temperature $\theta_\rho$.
@@
 \vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
-\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
+\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
 This gives rise to the 3D vector-invariant form,
@@
 
-Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
+Due to its association with hydrostatic models, it is common to use the 2D kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
 Decomposing the momentum transport term into horizontal and vertical components gives

### 2017-05-04 DCMIP2016-Part1-Models.tex
- Final draft - Some textual modifications and reorganization

@@
 
-\Author[6]{Don}{Dazlich}
-\Author[6]{Ross}{Heikes}
-\Author[6]{Celal}{Konor}
-\Author[6]{David}{Randall}
-
-\Author[7]{Thomas}{Dubos}
-\Author[7]{Yann}{Meurdesoif}
-
-\Author[8]{Xi}{Chen}
-\Author[8]{Lucas}{Harris}
-
-\Author[9]{Christian}{K\"uhnlein}
-
-\Author[10]{Vivian}{Lee}
-\Author[10]{Abdessamad}{Qaddouri}
-\Author[10]{Claude}{Girard}
-
-\Author[11]{David M.}{Hall}
+\Author[6]{David M.}{Hall}
+
+\Author[7]{Don}{Dazlich}
+\Author[7]{Ross}{Heikes}
+\Author[7]{Celal}{Konor}
+\Author[7]{David}{Randall}
+
+\Author[8]{Thomas}{Dubos}
+\Author[8]{Yann}{Meurdesoif}
+
+\Author[9]{Xi}{Chen}
+\Author[9]{Lucas}{Harris}
+
+\Author[10]{Christian}{K\"uhnlein}
+
+\Author[11]{Vivian}{Lee}
+\Author[11]{Abdessamad}{Qaddouri}
+\Author[11]{Claude}{Girard}
 
@@
 \affil[5]{Stony Brook University}
-\affil[6]{Colorado State University}
-\affil[7]{Institut Pierre-Simon Laplace (IPSL)}
-\affil[8]{Geophysical Fluid Dynamics Laboratory (GFDL)}
-\affil[9]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
-\affil[10]{Environment and Climate Change Canada}
-\affil[11]{University of Colorado, Boulder}
+\affil[6]{University of Colorado, Boulder}
+\affil[7]{Colorado State University}
+\affil[8]{Institut Pierre-Simon Laplace (IPSL)}
+\affil[9]{Geophysical Fluid Dynamics Laboratory (GFDL)}
+\affil[10]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
+\affil[11]{Environment and Climate Change Canada}
 \affil[12]{Max Planck Institute for Meteorology}
@@
 To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
-This review includes choice of model grid, variable placement, vertical coordinate choices, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
+This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
 \end{abstract}
@@
 The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.
-Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.
-These factors include the choice of model grid, variable placement, prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.
+Although this component's role is simply to solve the equations of fluid motion governing atmospheric dynamics (the Navi\'er-Stokes equations), there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.
+These factors include the choice of model grid, variable placement, vertical coordinate, prognostic equations, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.
 
@@
 The 2016 workshop and summer school followed from earlier DCMIP and dynamical core workshops (held in 2012 and 2008, respectively), and other model intercomparison efforts.
-Its goals were to provide a forum for discussing outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
+Its goals were to provide an international forum for discussing outstanding issues in global atmospheric models, and provide a unique training experience for the future generation of climate scientists.
 Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling, and variable-resolution global modeling.
@@
 The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.
-The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.
+The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by allowing modeling groups to assess their models in the context of the global dynamical core ecosystem.
 Past and present intercomparison efforts have been leveraged by modeling groups to improve their own models, in turn leading to a positive impact on the quality of weather and climate simulations.
@@
 This paper is the first in a series of papers documenting the results of this workshop.
-Its purposes is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.
-For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
+Its purpose is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.
+For ease of reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
 Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.
@@
 Further details on these choices can be found in subsequent sections.
-Simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).
-Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
-The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.
+In total, simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).
 The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
@@
 Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
-The participant models are briefly described through the remainder of this section.
+A brief description of each participant model follows, focused on the unique features and decisions underlying the model design.
+
+%Although all models solve the Navi\'er-Stokes equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
+The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.
 
@@
 Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
-All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
+All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.
+
+%Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
 
@@
 Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid as described by \citet{LR1997QJR} (see section \ref{sec:CDGrid}).
+The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}).
 Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
@@
 To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
-via applications of the Laplacian operator.
-Viscosity operations are applied after the completion of the physics time step.
+via applications of the Laplacian operator, applied after the completion of the physics time step.
 
@@
 Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels.
-The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
+The discretization employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
 Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
@@
 
-For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
+The icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
 These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
@@
 An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
-For a discussion of the effective resolution see \cite{dipankar2015large}.
+A discussion of the effective resolution of such grids is given in \cite{dipankar2015large}.
 The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.
@@
 
-The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.
-Their cells are hexagonal, except for twelve pentagonal cells.
-Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.
+The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is most directly obtained by taking the dual to the icosahedral (triangular grid) -- that is, by replacing grid nodes with spherical polygons.
+The resulting grid's cells are hexagonal, except for twelve pentagonal cells.
+Given an icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are then defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO (see also section \ref{sec:grid_ccvt}), or a barycentric mesh, used by NICAM.
 A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.
@@
 
-It is useful to optimize icosahedral-hexagonal grids as well.
-DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh.
+It is often useful to optimize icosahedral-hexagonal grids as well.
+DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following by replacing the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generating the icosahedral-hexagonal mesh.
 This improves the homogeneity of the grid (e.g.
-ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
+ratio of largest cell area to smallest cell area) but several thousand iterations can be required for a significant improvement.
 
@@
 Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
-the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
+the edges of the primary mesh as applied in the context of the finite-volume discretization of FVM (Section~\ref{sec:FVM}).
 
@@
 
-\subsubsection{C-D-Grid in the FV$^3$ model} \label{sec:CDGrid}
-
-Although velocities in the FV$^3$ model are stored in accordance with the D-grid arrangement, at the intermediate stages of the forward-backwards timestepping scheme, velocities are actually prognosed on the C-grid.
+%\subsubsection{C-D-Grid in the FV$^3$ model} \label{sec:CDGrid}
+
+Other specialized staggerings have been developed that couple horizontal staggering with the formulation of the time integrator.
+In the FV$^3$ model, although velocities are stored in accordance with the D-grid arrangement, at the intermediate stages of the forward-backwards timestepping scheme, velocities are actually prognosed on the C-grid.
 The intermediate velocities then act as a simplified Riemann solver:  the intermediate stage velocities are time-centered and can be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
@@
 
-Because of the vast differences between horizontal and vertical scales, most atmospheric models use dimension splitting in order to separate the horizontal discretization from the vertical discretization.
+Because of the vast differences between horizontal and vertical scales in global simulations, most atmospheric models use dimension splitting in order to separate the horizontal discretization from the vertical discretization.
 In this section, design considerations related to the vertical column are discussed, including the staggering of prognostic and diagnostic variables, and the choice of vertical coordinate.
@@
 
-In the context of dimension splitting, ``horizontal'' typically refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g.
-terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.
+In the context of dimension splitting, the ``horizontal'' typically refers to either the contravariant basis, which is perpendicular to the vertical, or the covariant basis, which is directed along coordinate (e.g.
+terrain-following) surfaces.
+In contrast, the vertical dimension is strictly aligned with the radial vector pointing from the center of the Earth.
 Vertical position is typically labelled using an arbitrary function $s(t,\vb{x},z)$ that is monotonic in $z$, so that model interfaces are equally spaced with respect to $s$.
@@
 
-Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
+Mass-based coordinates \citep{laprise1992euler} are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
 \begin{align} \label{eq:s_mass-based}
@@
 
+\subsection{ACME--A}
+
+ACME-A hosts an experimental compressible shallow-atmosphere model in hybrid terrain-following pressure vertical coordinates $\eta$, similar to the model of \cite{laprise1992euler}.
+The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), pseudodensity (\ref{eq:PrognosticPseudoDensity}), and geopotential (\ref{eq:PrognosticShallowGeopotential}).
+The vertical velocity equation is formulated analogous to that of GEM,
+\begin{align}
+\frac{d w}{dt} &= - g_c \left( 1 - \pdiff{p}{s} \right).
+\end{align}
+
 \subsection{DYNAMICO} \label{sec:DYNAMICOEquations}
@@
 The prognostic equations employed by DYNAMICO are based on a Hamiltonian formulation \citep{Dubos2014Equations}.
-The specific prognostic variables employed pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+The specific prognostic variables employed pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, horizontal covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
 Prognostic equations are in flux-form for mass (\ref{eq:PrognosticPseudoDensity}) and $W$ (\ref{eq:PrognosticVelocityFluxForm}), in advective form for $\Phi$ (\ref{eq:PrognosticShallowGeopotential}), and in vector-invariant form for covariant horizontal momentum (\ref{eq:CovariantUpdateEquation}).
@@
 
-\subsection{ACME--A}
-
-ACME-A hosts an experimental compressible shallow-atmosphere model in hybrid terrain-following pressure vertical coordinates $\eta$, similar to the model of \cite{laprise1992euler}.
-The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), pseudodensity (\ref{eq:PrognosticPseudoDensity}), and geopotential (\ref{eq:PrognosticShallowGeopotential}).
-The vertical velocity equation is formulated analogous to that of GEM (\ref{eq:GEMVerticalVelocity}).
-
 \subsection{ICON} \label{sec:ICONEquations}
@@
 For DCMIP, the latter are limited to cloud and rain specific bulk densities.
-Water vapor density $\rho q$ is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$.
+Water vapor density is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$.
 The latter is in turn diagnosed from $\theta_{il}$ and from the latent heat required to convert any hydrometeors present to the vapor phase.
@@
 Momentum is C-staggered in the horizontal and vertical (Lorenz vertical staggering is used), meaning that prognosed components live on the grid cell faces and are each normal to the respective face and the pressure gradient force is evaluated and applied at those locations.
-However, evaluation of advective and turbulent momentum transport (as well as the Coriolis force) involves a diagnostic reconstruction of the total momentum vector at the centers of scalar grid cells \citep{perot2000conservation}, and cell-to-cell flux of momentum is computed from that reconstruction using the same A- grid control volumes as for scalars.
+However, evaluation of advective and turbulent momentum transport (as well as the Coriolis force) involves a diagnostic reconstruction of the total momentum vector at the centers of scalar grid cells \citep{perot2000conservation}, and cell-to-cell flux of momentum is computed from that reconstruction using the same A-grid control volumes as for scalars.
 This arrangement is particularly convenient for the cut-cell formulation at the topographic surface where reduced cell face areas and volumes regulate momentum and scalar fluxes in an identical manner.
@@
 Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-
-\subsection{ICON}
-
-The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2015icon}, and are summarized here.
-For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the (relatively) computationally  expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions,
-\begin{align}
-F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left\{\nabla \left[\tilde{\nabla} \cdot v + 
-\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right]\right\} \,.
-\end{align}
-$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
-
-ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
-The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
-
-The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
-\begin{align} \label{smagn2}
-F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,, & F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,,
-\end{align} where $a_c$ and $a_e$ denotes the area associated with the cell and edge under consideration, respectively.
-An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
-
-A fourth-order computational diffusion is also available for vertical wind speed $w$.
-This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
-This term appears as
-\begin{align}
-F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)).
-\end{align}
 
@@
 
+\subsection{ICON}
+
+The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
+The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2015icon}, and are summarized here.
+For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the (relatively) computationally  expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions,
+\begin{align}
+F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left\{\nabla \left[\tilde{\nabla} \cdot v + 
+\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right]\right\} \,.
+\end{align}
+$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
+
+ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
+The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
+
+The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
+\begin{align} \label{smagn2}
+F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,, & F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,,
+\end{align} where $a_c$ and $a_e$ denotes the area associated with the cell and edge under consideration, respectively.
+An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
+
+A fourth-order computational diffusion is also available for vertical wind speed $w$.
+This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
+This term appears as
+\begin{align}
+F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)).
+\end{align}
+
 \subsection{OLAM}
@@
 Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
+A number of different ARK schemes have compared and contrasted in this framework, with significant implications for model performance and stability (Gardner et al., under review).
 
@@
 \end{align}
-\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
+\noindent where $d/dt$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
 The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
@@
 These tests can also be used for future dynamical core development to identify where a new dynamical core diverges from a suite of modern models.
+
+\subsection*{Code availability}
 
@@
 1629819), the National Aeronautics and Space Administration (award no.
-NNX16AK51G), the National Oceanic and Atmospheric Administration (award no.
+NNX16AK51G), the National Oceanic and Atmospheric Administration Great Lakes Environmental Research Laboratory (award no.
 NA12OAR4320071), the Office of Naval Research and CU Boulder Research Computing.

### 2017-08-13 DCMIP2016-Part1-Abstract.tex
Updates based on returned reviews

--- 2017-05-01 DCMIP2016-Part1-Models.tex
@@
-%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
-%% ---------------------------------
-%% This template should be used for copernicus.cls
-%% The class file and some style files are bundled in the Copernicus Latex Package which can be downloaded from the different journal webpages.
-%% For further assistance please contact the Copernicus Publications at: publications@copernicus.org
-%% http://publications.copernicus.org
-
-%% Please use the following documentclass and Journal Abbreviations for Discussion Papers and Final Revised Papers.
-
-%% 2-Column Papers and Discussion Papers
-\documentclass[gmd, manuscript]{copernicus}
-
-%% Journal Abbreviations (Please use the same for Discussion Papers and Final Revised Papers)
-
-% Archives Animal Breeding (aab)
-% Atmospheric Chemistry and Physics (acp)
-% Advances in Geosciences (adgeo)
-% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
-% Annales Geophysicae (angeo)
-% ASTRA Proceedings (ap)
-% Atmospheric Measurement Techniques (amt)
-% Advances in Radio Science (ars)
-% Advances in Science and Research (asr)
-% Biogeosciences (bg)
-% Climate of the Past (cp)
-% Drinking Water Engineering and Science (dwes)
-% Earth System Dynamics (esd)
-% Earth Surface Dynamics (esurf)
-% Earth System Science Data (essd)
-% Fossil Record (fr)
-% Geographica Helvetica (gh)
-% Geoscientific Instrumentation, Methods and Data Systems (gi)
-% Geoscientific Model Development (gmd)
-% Geothermal Energy Science (gtes)
-% Hydrology and Earth System Sciences (hess)
-% History of Geo- and Space Sciences (hgss)
-% Journal of Sensors and Sensor Systems (jsss)
-% Mechanical Sciences (ms)
-% Natural Hazards and Earth System Sciences (nhess)
-% Nonlinear Processes in Geophysics (npg)
-% Ocean Science (os)
-% Proceedings of the International Association of Hydrological Sciences (piahs)
-% Primate Biology (pb)
-% Scientific Drilling (sd)
-% SOIL (soil)
-% Solid Earth (se)
-% The Cryosphere (tc)
-% Web Ecology (we)
-% Wind Energy Science (wes)
-
-%% \usepackage commands included in the copernicus.cls:
-%\usepackage[german, english]{babel}
-%\usepackage{tabularx}
-%\usepackage{cancel}
-%\usepackage{multirow}
-%\usepackage{supertabular}
-%\usepackage{algorithmic}
-%\usepackage{algorithm}
-%\usepackage{amsthm}
-%\usepackage{float}
-%\usepackage{subfig}
-%\usepackage{rotating}
-
-\usepackage{subcaption}
-
-% Custom commands
-\newcommand{\vb}{\mathbf}
-\newcommand{\vg}{\boldsymbol}
-\newcommand{\mat}{\mathsf}
-\newcommand{\diff}[2]{\frac{d #1}{d #2}}
-\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
-\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
-\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
-
-% User-defined mathematical symbols
-\renewcommand{\emph}[1]{{\color{red}\textbf{#1}}}
-\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
-\newcommand{\p}[1]{\partial #1}
-%Then, writing a partial is as simple as $\pd{J}{x}$.
-\newcommand{\ol}[1]{{\overline #1}}
-\newcommand{\mbf}[1]{{\mathbf #1}}
-\newcommand{\mbfol}[1]{\overline{{\mathbf #1}}}
-\newcommand{\wdh}[1]{{\widehat #1}}
-\newcommand{\wdt}[1]{{\widetilde #1}}
-\newcommand{\sss}[1]{{\scriptscriptstyle #1}}
-\newcommand{\scs}[1]{{\scriptstyle #1}}
+\documentclass[11pt]{article}
 
@@
 
-\title{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models}
+\noindent \textbf{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models} \\
 
-% \Author[affil]{given_name}{surname}
+\noindent {Paul A.
+Ullrich, Christiane Jablonowski, James Kent, Peter H.
+Lauritzen, Ramachandran Nair, Kevin A.
+Reed, Colin M.
+Zarzycki, David M.
+Hall, Don Dazlich, Ross Heikes, Celal Konor, David Randall, Thomas Dubos, Yann Meurdesoif, Xi Chen, Lucas Harris, Christian K\"uhnlein, Vivian Lee, Abdessamad Qaddouri, Claude Girard, Marco Giorgetta, Daniel Reinert, Joseph Klemp, Sang-Hun Park, William Skamarock, Hiroaki Miura, Tomoki Ohno,
+Ryuji Yoshida, Robert Walko, Alex Reinecke, and Kevin Viner} \\
 
-\Author[1]{Paul A.}{Ullrich}
-\Author[2]{Christiane}{Jablonowski}
-\Author[3]{James}{Kent}
-\Author[4]{Peter H.}{Lauritzen}
-\Author[4]{Ramachandran}{Nair}
-\Author[5]{Kevin A.}{Reed}
-\Author[4]{Colin M.}{Zarzycki}
-
-\Author[6]{Don}{Dazlich}
-\Author[6]{Ross}{Heikes}
-\Author[6]{Celal}{Konor}
-\Author[6]{David}{Randall}
-
-\Author[7]{Thomas}{Dubos}
-\Author[7]{Yann}{Meurdesoif}
-
-\Author[8]{Xi}{Chen}
-\Author[8]{Lucas}{Harris}
-
-\Author[9]{Christian}{K\"uhnlein}
-
-\Author[10]{Vivian}{Lee}
-\Author[10]{Abdessamad}{Qaddouri}
-\Author[10]{Claude}{Girard}
-
-\Author[11]{David M.}{Hall}
-
-\Author[12]{Marco}{Giorgetta}
-\Author[13]{Daniel}{Reinert}
-
-\Author[4]{Joseph}{Klemp}
-\Author[4]{Sang-Hun}{Park}
-\Author[4]{William}{Skamarock}
-
-\Author[14]{Hiroaki}{Miura}
-\Author[14]{Tomoki}{Ohno}
-\Author[15]{Ryuji}{Yoshida}
-
-\Author[16]{Robert}{Walko}
-
-\Author[17]{Alex}{Reinecke}
-\Author[17]{Kevin}{Viner}
-
-%\Author[18]{Elijah}{Goodfriend}
-%\Author[18]{Hans}{Johansen}
-
-\affil[1]{University of California, Davis}
-\affil[2]{University of Michigan}
-\affil[3]{University of South Wales}
-\affil[4]{National Center for Atmospheric Research}
-\affil[5]{Stony Brook University}
-\affil[6]{Colorado State University}
-\affil[7]{Institut Pierre-Simon Laplace (IPSL)}
-\affil[8]{Geophysical Fluid Dynamics Laboratory (GFDL)}
-\affil[9]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
-\affil[10]{Environment and Climate Change Canada}
-\affil[11]{University of Colorado, Boulder}
-\affil[12]{Max Planck Institute for Meteorology}
-\affil[13]{Deutscher Wetterdienst (DWD)}
-\affil[14]{University of Tokyo}
-\affil[15]{RIKEN}
-\affil[16]{University of Miami}
-\affil[17]{Naval Research Laboratory}
-
-%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
-
-\runningtitle{DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison}
-
-\runningauthor{Ullrich, et al.}
-
-\correspondence{Paul A. Ullrich (paullrich@ucdavis.edu)}
-
-\received{}
-\pubdiscuss{} %% only important for two-stage journals
-\revised{}
-\accepted{}
-\published{}
-
-%% These dates will be inserted by Copernicus Publications during the typesetting process.
-
-\firstpage{1}
-
-\maketitle
-
-\begin{abstract}
-Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
+\noindent Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
 These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.
@@
 To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
-This review includes choice of model grid, variable placement, vertical coordinate choices, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
-\end{abstract}
-
-\introduction  %% \introduction[modified heading if necessary]
-
-The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.
-Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.
-These factors include the choice of model grid, variable placement, prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.
-
-To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment.
-The 2016 workshop and summer school followed from earlier DCMIP and dynamical core workshops (held in 2012 and 2008, respectively), and other model intercomparison efforts.
-Its goals were to provide a forum for discussing outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists.
-Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling, and variable-resolution global modeling.
-The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.
-Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.
-
-The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.
-The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.
-Past and present intercomparison efforts have been leveraged by modeling groups to improve their own models, in turn leading to a positive impact on the quality of weather and climate simulations.
-The workshop component of DCMIP has also advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.
-Notably, the use of idealized test cases to isolate specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.
-Another important outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.
-The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.
-
-This paper is the first in a series of papers documenting the results of this workshop.
-Its purposes is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.
-For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.
-Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.
-The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.
-The summary and conclusions then follow in section \ref{sec:Conclusions}.
-Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.
-
-\begin{table}[p]
-\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
-\begin{center}
-\begin{tabular}{cl}
-\hline Symbol & Description \\ \hline 
-$\lambda$ & Longitude (in radians) \\
-$\varphi$ & Latitude (in radians) \\
-$z$ & Height with respect to mean sea level (set to zero) \\
-$s$ & Vertical model coordinate \\
-$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
-$\Phi_s$ & Surface geopotential \\
-$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
-$u$ & Zonal wind velocity \\
-$v$ & Meridional wind velocity \\
-$w$ & Vertical wind velocity \\
-$\vb{u}$ & 3D wind vector \\
-$\vb{u}_h$ & Horizontal wind vector \\
-$\vb{v}_h$ & Horizontal wind vector with covariant components \\
-$\omega$ & Vertical pressure velocity  \\
-$D$ & Divergence of the horizontal wind vector\\
-$\zeta$ & Vertical component of relative vorticity\\
-$p$ & Pressure (pressure of moist air if $q>0$) \\
-$e$ & Internal energy \\
-$\rho$ & Total air density \\
-$\rho_d$ & Dry air density \\
-$\rho_s$ & Pseudo-density \\
-$T$ &Temperature \\
-$T_v$ & Virtual temperature \\
-$\theta$ & Potential temperature \\
-$\theta_v$ & Virtual potential temperature \\
-$\theta_{il}$ & Ice-liquid potential temperature \\
-$\theta_\rho$ & Density potential temperature \\
-%$P_{ls}$ & Large-scale precipitation rate \\
-$q$ & Specific humidity \\
-$q_v$ & Water vapor mixing ratio \\
-$q_c$ & Cloud water mixing ratio \\
-$q_r$ & Rain water mixing ratio \\
-$q_i$ & General tracer mixing ratio \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Dynamical cores} \label{sec:DynamicalCores}
-
-This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.
-Further details on these choices can be found in subsequent sections.
-Simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).
-Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.
-The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.
-The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.
-The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.
-Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
-The participant models are briefly described through the remainder of this section.
-
-\begin{table}[p]
-\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
-\begin{center}
-\begin{tabular}{cll}
-\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
-CSU & Colorado State University Model & Colorado State University, USA \\
-DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
-FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
-FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
-GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
-ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories / University of Colorado, Boulder \\
-ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
-MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
-NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
-OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
-Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-\begin{table}[p]
-\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.
-Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).
-Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).
-More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
-\begin{center}
-\begin{tabular}{cccccc}
-\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
-& & & & method & staggering \\ \hline
-CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
-DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
-FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_v$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
-GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
-ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
-ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
-MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
-NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
-OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-\begin{table}[p]
-\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
-\begin{center}
-\begin{tabular}{cccccc}
-\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
-CSU & Lorenz & fixed height \\
-DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
-FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
-FVM & Co-located & fixed height \\
-GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
-ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
-ICON & Lorenz & fixed height \\
-MPAS & Lorenz & fixed height \\
-NICAM & Lorenz & fixed height \\
-OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
-Tempest & Lorenz & fixed height \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-\begin{table}[p]
-\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
-\begin{center}
-\begin{tabular}{cccccc}
-\hline Acronym & Standard options for diffusion, & Temporal discretization \\
-& stabilization, filters, or fixers & \\ \hline 
-CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
-DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
-FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
-FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
-GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
-ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
-ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
-MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
-NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
-OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
-Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
-\hline 
-\end{tabular}
-\end{center}
-\end{table}
-
-%Do not include information on the physical parameterizations used by the modeling system.
-Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}
-
-\subsection{Colorado State University Model (CSU)}
-
-The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate.
-The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
-The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
-Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.
-The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
-\subsection{DYNAMICO}
-
-DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
-DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.
-It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
-It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
-In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
-
-\subsection{FV Cubed (FV$^3$)}
-
-The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
-The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
-Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid as described by \citet{LR1997QJR} (see section \ref{sec:CDGrid}).
-Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
-Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
-Explicit viscosity is applied every acoustic timestep.
-
-The prognostic horizontal winds are stored in the native Gnomonic local coordinate.
-The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
-All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.
-
-\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
-\label{sec:FVM}
-
-The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
-ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
-compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
-A centered two-time-level semi-implicit 
-integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
-\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
-preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
-non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
-combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
-the vertical discretization all variables are co-located. 
-The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
-Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
-grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
-interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
-broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
-%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.
-
-%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
-%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-%are available by means of simple switches. The formulation incorporates 
-%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
-%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
-%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-%Conjugate Residual approach. The dynamics time step is adapted 
-%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-%than 1). The physics time step is identical to the dynamics time step. 
-% 
-%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-%The horizontal spatial discretization is fully unstructured finite-volume using the 
-%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-%In both the horizontal and the vertical discretization, all prognostic variables are 
-%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
-%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
-%
-%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
-%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-%the form of a Rayleigh-type forcing to the prognostic variables.
-
-%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
-%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
-
-%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
-%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
-%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
-%efficiency and accuracy.
-
-%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
-%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
-%framework developed at ECMWF.
-
-\subsection{Global Environmental Multiscale (GEM) model} \label{sec:GEM_core}
-
-The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.
-The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
-The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
-and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
-A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
-It gives rise to an {\it{iterative}} process where each step requires the solution of a
-linear system of equations that is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
-The dynamics and physics are time split.
-To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
-via applications of the Laplacian operator.
-Viscosity operations are applied after the completion of the physics time step.
-
-\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}
-
-The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}.
-ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model.
-Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
-The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
-Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
-All fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.
-
-%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.
-As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present.
-The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements.
-These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.
-The model employs a hybrid terrain-following pressure coordinate in the vertical.
-The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical.
-Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions.
-Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.
-
-\subsection{Icosahedral Non-hydrostatic model (ICON)}
-
-The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
-Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
-Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels.
-The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
-For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
-Further divergence damping is applied to the normal wind at every sub-step.
-Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
-
-\subsection{Model for Prediction Across Scales (MPAS)}
-
-The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
-Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
-In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
-The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
-Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
-Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
-
-\subsection{Non-hydrostatic ICosahedral Atmospheric Model (NICAM)}
-
-NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}.
-The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}.
-Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}.
-All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids.
-The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}.
-The Arakawa C-grid staggering is used vertically.
-A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended.
-The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}.
-The implicit time discretization is adopted for the vertically propagating one.
-A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.
-
-\subsection{Ocean-Land-Atmosphere Model (OLAM)}
-
-OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.
-Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting.
-Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
-Height is the vertical coordinate, and a Lorenz vertical grid staggering is used.
-A unique feature of OLAM is that grid levels are horizontal and intersect topography.
-This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms.
-Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground.
-The OLAM cut-cell formulation conserves mass and momentum.
-
-\subsection{Tempest}
-
-The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
-The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Model grids} \label{sec:ModelGrids}
-
-Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe \citep{staniforth2012horizontal}.
-In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.
-The choice of model grid can be motivated by simplicity, as in the case of the latitude-longitude grid, by a desire to maintain a local Cartesian structure, as with the cubed-sphere grid, or to support grid isotropy and homogeneity, as with many of the hexagonal or Voronoi grids that have been employed.
-The choice of grid may be further decided by the numerical method -- for instance, most finite element models rely on grids consisting entirely of quadrilaterals.
-Inevitably a choice must be made, and the pros and cons of that choice will have repercussions throughout the remainder of the model.
-To better understand the options that are available to dynamical core developers, we begin by reviewing many of the model grids that have been employed in global dynamical cores around the world.
-
-\subsection{Latitude-longitude grid} \label{sec:grid_rll}
-
-The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
-Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
-This grid is employed by the UK Met Office \citep{davies2005new, wood2014inherently}.
-
-\subsection{Cubed-sphere grid} \label{sec:grid_cs}
-
-The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed, PL2007JCP} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.
-More information on this choice of grid can be found in \cite{ullrich2014global}.
-On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.
-The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
-The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.
-Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.
-This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
-\end{center}
-\caption{(Left) A cubed-sphere grid.
-(Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.
-(Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
-\end{figure}
-
-%FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}.
-The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm.
-The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.
-The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.
-
-\subsection{Icosahedral (triangular) grid} \label{sec:grid_ico_tri}
-
-For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.
-These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained.
-For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle.
-By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
-
-Several methods are available for subdividing the triangular regions.
-One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement.
-Typical choices are $n$ = 2, 3 or 5.
-All additional $m$ refinement steps use $n$ = 2, i.e.
-are bisection steps.
-A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices.
-The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001shallow}.
-An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).
-For a discussion of the effective resolution see \cite{dipankar2015large}.
-The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.
-The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model.
-The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.
-
-\subsection{Icosahedral (hexagonal) grid / geodesic grid} \label{sec:grid_geo}
-
-The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.
-Their cells are hexagonal, except for twelve pentagonal cells.
-Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.
-A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.
-The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).
-Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
-
-It is useful to optimize icosahedral-hexagonal grids as well.
-DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh.
-This improves the homogeneity of the grid (e.g.
-ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.
-
-OLAM optimizes by applying the spring dynamics method of \cite{tomita2001shallow} to the dual triangular mesh prior to its mapping to the Voronoi grid.
-When local mesh refinement is applied, which OLAM achieves in a series of one or more resolution-doubling steps, each spanning a transition zone that is three grid rows wide (Figure \ref{fig:OLAMRefinement}), the equilibrium spring length is scaled to the target grid cell size in each refinement level and is varied incrementally across the transition zone.
-Spring dynamics is further modified by forcing angles on the dual triangular mesh in the transition zone in order to move the triangle edges closer to the centers of the hexagon edges they intersect.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.5in]{OLAM_Grid_Refinement.png}
-\end{center}
-\caption{Detail of one step of local mesh refinement used by OLAM Voronoi mesh.
-The transition zone is constructed by explicit topological reconnection of the grid lines, which produces pairings of heptagons (red dots) and pentagons (blue dots) along the refinement perimeter.} \label{fig:OLAMRefinement}
-\end{figure}
-
-\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids} \label{sec:grid_ccvt}
-
-Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.
-For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.
-However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.
-This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.
-Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e.
-for ocean basins or ice sheets) \citep{ringler2008multiresolution}.
-Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.
-CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:grid_geo}, since both typically contain a large number of hexagonal elements.
-However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
-\end{center}
-\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
-\end{figure}
-
-\subsection{Octahedral reduced Gaussian grid} \label{sec:grid_oct}
-\label{subsection:octahedralreducedgaussiangrid}
-
-As with the classical reduced Gaussian grid of \cite{hortal1991use}, the octahedral 
-reduced Gaussian grid \citep{malardel2016new,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
-roots of the Legendre polynomials. 
-The two grids differ in the arrangement of the points along the latitudes, which follows a simple
-rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
-are added with every latitude towards the equator, whereby the spacing between points along the 
-latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
-is suitable for transformations involving spherical harmonics, and has been introduced for operational 
-weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
-Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
-the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).
-
-\begin{figure}
-\centering
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
- \end{subfigure}%
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
- \end{subfigure}
-\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
-connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse octahedral grid with only
-24 latitudes between pole and equator 'O24' is used for illustration. The dual mesh resolution of the octahedral
-reduced Gaussian grid is about a factor 2 finer at the poles than the equator, see \cite{smolarkiewiczetalJCP2016}.}
-\label{fig:octahedralgrid}
-\end{figure}
-
-\subsection{Yin-Yang grid} \label{sec:grid_yinyang}
-
-The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).
-These components are combined to cover a spherical surface with partial overlap along their borders.
-The Yin component covers the latitude-longitude region
-\begin{equation}\label{eq:gem_yinyang}
-(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
-\frac{\pi}{4}+\delta_{\theta})  \cap
-(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
-\frac{3\pi}{4}+\delta_{\lambda}),
-\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.
-For instance, a common configuration employed by the GEM model for DCMIP fixes
-$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.
-The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
-This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
-
-\begin{figure}[]
-\includegraphics[height=6cm]{gem_yinyang.png}
-\caption{The Yin-Yang grid is a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
-\label{fig:gem_yinyang}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Variable placement and prognosis} \label{sec:VariablePlacement}
-
-Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.
-Hence, in this section we turn our attention to the placement of prognostic variables on the grid, the choice of vertical coordinate, and the choice of corresponding prognostic equations.
-
-\subsection{Horizontal staggering} \label{sec:HorizontalStaggering}
-
-The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.
-The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.
-Later, other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.
-New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.
-Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).
-
-There are arguments to be made for and against each of these grids:  the A-grid is relatively simple to implement and tends to support large time step sizes, but also supports a $2 \Delta x$ computational mode (even in the context of finite element methods); the C-grid has excellent dispersion properties, and does not support extraneous computational modes (as long as the number of edges is equal to twice the number of volumes), but also requires a timestep that is half that of the A-grid; the D-grid provides a better representation of vorticity, but suffers to similar computational modes as the A-grid that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
-
-%----------------------------------------------------
-\begin{figure}[]
-\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
-\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.  Here $\eta$ denotes the buoyancy variable.}
-\label{fig:horizontalstaggering}
-\end{figure}
-%----------------------------------------------------
-
-\subsubsection{C-D-Grid in the FV$^3$ model} \label{sec:CDGrid}
-
-Although velocities in the FV$^3$ model are stored in accordance with the D-grid arrangement, at the intermediate stages of the forward-backwards timestepping scheme, velocities are actually prognosed on the C-grid.
-The intermediate velocities then act as a simplified Riemann solver:  the intermediate stage velocities are time-centered and can be used to compute the fluxes and advance the flux terms by a full acoustic timestep.
-More details on this approach can be found in \cite{LR1997QJR}.
-
-\subsection{Vertical staggering} \label{sec:VerticalStaggering}
-
-Along with the choice of prognostic variables, the vertical discretization of the equations of motion also allows for the staggered placement of prognostic variables.
-As with hydrostatic models, certain discretizations give rise to spurious computational modes that can contaminate the solution \citep{tokioka1978some,arakawa1988baroclinic}.
-The choice of vertical staggering may also impact many physically-relevant properties of the model near the grid scale, such as the phase speed of Rossby waves \citep{thuburn2005vertical}.
-Finally, the choice of vertical staggering can have impacts on the physics-dynamics coupling \cite{holdaway2013comparison1,holdaway2013comparison2}.
-Taken altogether, these issues suggest care should be taken when selecting the discretization.
-Since co-located discretizations of the non-hydrostatic equations generally require some additional effort to control spurious computational modes, it is more common to employ either: (a) a Lorenz-type staggering \citep{lorenz1960energy}, which places horizontal velocity, buoyancy, and thermodynamic variables on model levels, and vertical velocity on model interfaces; or (b) a Charney-Phillips-type staggering \citep{charney1953numerical}, which places horizontal velocity and buoyancy variables on model levels and vertical velocity and thermodynamic variables on model interfaces (see Figure \ref{fig:verticalstaggering}).
-These approaches can be further augmented as needed, for instance by shifting the vertical velocity and thermodynamic variables from the bottom boundary to an intermediate level, as in the GEM model.
-Note that, in general, tracer variables are co-located with the buoyancy variable.
-
-%----------------------------------------------------
-\begin{figure}[]
-\includegraphics[width=5in]{VerticalStaggering_v2.pdf}
-\caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
-\label{fig:verticalstaggering}
-\end{figure}
-%----------------------------------------------------
-
-\subsection{Vertical coordinates} \label{sec:VerticalCoordinates}
-
-Most atmospheric models use dimension splitting in order to separate the evolution equations for momentum into horizontal and vertical directions.
-Whereas ``horizontal'' in this context either refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g.
-terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.
-In this context, the use of dimension splitting allows us to choose an arbitrary function $s(t,\vb{x},z)$ that is monotonic in $z$, so that model interfaces are equally spaced with respect to $s$.
-Typically $s$ is chosen so that the Earth's surface (the bottom boundary of the atmosphere) is a coordinate surface, allowing for easy specification of boundary conditions for the prognostic equations -- this leads to the so-called ``terrain-following'' family of vertical coordinates.
-Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which is in terms of the altitude $z$ and takes the form
-\begin{align}
-s(\vb{x},z) = z_{\footnotesize \mbox{top}} \left[ \frac{z - z_s(\vb{x})}{z_{\footnotesize \mbox{top}} - z_s(\vb{x})} \right],
-\end{align} where $\vb{x}$ denotes the horizontal position, $z_s(\vb{x})$ is the height of the topography at that position, and $z_{\footnotesize \mbox{top}}$ denotes the height of the model top (typically independent of position).
-Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.
-Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.
-All models in this paper except for OLAM use some variant of terrain-following coordinates, although work on developing modern cut-cell, embedded boundary and immersed boundary representations is ongoing (e.g.
-\cite{lock2012demonstration}).
-Note that time-dependent vertical coordinates are allowed and are typically referred to as ``floating'' coordinates.
-Several examples of vertical coordinates are now given.
-
-\subsubsection{Mass-based coordinates}
-
-Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
-\begin{align} \label{eq:s_mass-based}
-s = \int^{\infty}_{z} \rho g dz.
-\end{align}  Under this definition,
-\begin{align}
-\pdiff{s}{z} = - \rho g.
-\end{align}
-
-%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
-%\begin{align} \label{eq:SurfacePressureEquation}
-%{\color{red}\pdiff{p_s}{t} = - \int g \pdiff{\rho}{t} \left( \pdiff{z}{s} \right) ds = - \int \nabla_s \cdot (\rho_s \vb{u}_h) \left( \pdiff{z}{s} \right) ds,}
-%\end{align} where the integrals are taken over the column.
-
-\subsubsection{GEM zeta coordinate} \label{sec:GEM_zeta}
-
-The vertical coordinate in the GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
-Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
-\begin{align}\label{eq:gem_zeta}
-\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
-\end{align}
-\noindent with
-\begin{align}
-A(\zeta) = \zeta, \quad \mbox{and} \quad B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r.
-\end{align}
-Here  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
-and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
-
-\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and ACME--A)} \label{sec:FloatingLagrangian}
-
-In the floating Lagrangian formulation \citep{starr1945quasi, lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.
-The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
-\begin{align}
-s = \int^{\infty}_{z} \rho_0 g dz,
-\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
-\begin{align}
-\dot{s} = \frac{ds}{dt} = 0,
-\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.
-Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.
-
-\subsubsection{Cut-cells in OLAM} \label{sec:CutCells}
-
-A pure z coordinate with horizontal grid levels is used in OLAM \citep{walko2008ocean2} in order to completely avoid topographic imprinting on the model grid levels (Figure \ref{fig:OLAM_CutCell}).
-This implies that grid levels intersect the topographic surface, leading to some grid cells being partially above and partly below the surface.
-The face areas of these so-called cut cells are reduced accordingly, which in turn regulates cell-to-cell flux transport in accordance with the kinematic constraint imposed by the topography.
-Cut cell volumes are also reduced, and volumes and surface areas of all cells appear explicitly in the finite-volume formulation of the mass and momentum conservation equations.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2in]{OLAM_CutCellA.png} \quad \includegraphics[width=2in]{OLAM_CutCellB.png}
-\end{center}
-\caption{(Left) A terrain-following coordinate passing over rough topography.  (Right) A cut-cell coordinate used for representing the same topography.} \label{fig:OLAM_CutCell}
-\end{figure}
-
-\subsection{Prognostic equations and treatment of moisture} \label{sec:PrognosticEquations}
-
-The Navi\'er-Stokes equations that govern atmospheric motion can take on many forms, depending on the choice of prognostic variables and coordinate system.
-A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.
-The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.
-The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models, including any special treatment of terms related to moist physics.
-
-\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}
-
-The prognostic equations employed by DYNAMICO are based on a Hamiltonian formulation \citep{Dubos2014Equations}.
-The specific prognostic variables employed pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
-Prognostic equations are in flux-form for mass (\ref{eq:PrognosticPseudoDensity}) and $W$ (\ref{eq:PrognosticVelocityFluxForm}), in advective form for $\Phi$ (\ref{eq:PrognosticShallowGeopotential}), and in vector-invariant form for covariant horizontal momentum (\ref{eq:CovariantUpdateEquation}).
-
-\subsubsection{FV$^3$} \label{sec:FV3Equations}
-
-The hydrostatic FV$^3$ model uses a mass-based floating Lagrangian coordinate along with the shallow-atmosphere approximation \citep{lin2004vertically}.
-Prognostic equations include horizontal velocity in 2D vector-invariant form (\ref{eq:PrognosticUhEquationK2}), pseudo-density (\ref{eq:PrognosticPseudoDensity}), and virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}).
-The non-hydrostatic model further incorporates prognostic geopotential (\ref{eq:PrognosticShallowGeopotential}) and vertical momentum (\ref{eq:PrognosticWEquationK2}).
-
-\subsubsection{FVM} \label{sec:FVMEquations}
-
-The FVM formulation is based on conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, 
-and dry entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
-which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
-\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Moreover, underlying the conservation laws 
-in FVM is a perturbational form with respect to a balanced ambient state and a generalized curvilinear 
-coordinate formulation in a geospherical framework. Following \cite{smolarkiewiczetalJCP2017}, the
-FVM governing equations can concisely be written as
-\begin{subequations} \label{fvm:compressible}
-\begin{align}
- & \pdiff{\mathcal{G}\rho_d}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
-\label{fvm:mass}  \\
-& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
-\mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
-%- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a \left(\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right) \right)
-%- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r \right)
-- \mathbf{k}\,g \left(\frac{\theta'}{\theta_a}+\varepsilon_b\,q'_v - q_c - q_r \right)
-- 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
-+ \mathcal{M}' \right]~,
-%+ \mathbf{M} \right)~,
-\label{fvm:momentum} \\
-%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
-& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \overline{\nabla} \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
-- \mathcal{G}\rho_d\,\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \overline{\nabla} \theta_{a} ~,
-%- \frac{L}{c_{pd}\,\pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
-%+  \mathcal{H} \right)~,
-\label{fvm:thermodynamic} \\
-& \phi' = c_{pd} \left[\left( \frac{R_{d}}{p_{0}} \rho_d\,\theta\,(1+q_v/\varepsilon) \right)^{R_{d}/c_{vd}} - \pi_{a} \right]~.
-\label{fvm:gaslaw}
-\end{align}
-\end{subequations}
-Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, 3D
-physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and a modified
-Exner pressure perturbation $\phi'$, 
-with the thermodynamic variables related by the gas law 
-\eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
-subscript `a', see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014} for discussions.
-The symbol $g$ in \eqref{fvm:momentum} denotes the gravitational acceleration 
-and $\varepsilon_b = 1/\varepsilon-1$. As far as geometric aspects 
-are concerned, the nabla operator $\overline{\nabla}$ represents
-the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
-a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
-the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$ where a contribution from
-optional time-dependency of the curvilinear coordinates is neglected for simplicity \citep{kuehnleinJCP2017}.
-The symbol $\mathcal{M}' = \mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
-subsumes the metric forces in the spherical domain \citep{smolarkiewiczetalJCP2017}.
-
-\subsubsection{GEM} \label{sec:GEM_equations}
-
-%GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$, described in section \ref{sec:GEM_zeta}.
-In GEM the non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
-\begin{align}
-\mu = \pdiff{p}{s} - 1,
-\end{align}
-\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based}).  In this case the equations of GEM model \citep{Girard2014} are concisely given by
-\begin{align}
-\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
-\frac{d w}{dt}- g_c \mu &= 0, \label{eq:GEMVerticalVelocity} \\
-\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
-\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
-\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
-\frac{d \Phi}{dt}-  g_c w & = 0.
-\end{align}
-
-%\noindent with horizontal velocity ${\bf V}_h$, 
-%generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
-%vertical velocity $w$,
-%geopotential $\phi$, 
-%virtual temperature $T_v$, 
-%pressure $p$ and
-%hydrostatic pressure $\pi$. The variable 
-%$\mu$ is a measure of departure from hydrostatic balance, 
-%$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
-
-\subsubsection{ACME--A}
-
-ACME-A hosts an experimental compressible shallow-atmosphere model in hybrid terrain-following pressure vertical coordinates $\eta$, similar to the model of \cite{laprise1992euler}.
-The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), pseudodensity (\ref{eq:PrognosticPseudoDensity}), and geopotential (\ref{eq:PrognosticShallowGeopotential}).
-The vertical velocity equation is formulated analogous to that of GEM (\ref{eq:GEMVerticalVelocity}).
-
-\subsubsection{ICON} \label{sec:ICONEquations}
-
-ICON solves a non-hydrostatic equation set based on \cite{gassmann2008} 
-using terrain-following z-coordinates. The governing equations describe the mixture 
-of a two-component system of dry air and water, where water is allowed to occur in 
-all three phases, including precipitating constituents. Following \cite{wacker2006}, 
-the barycentric velocity $\vb{u}_{bc}=\sum_{k} \rho_{k} \vb{u}_{k}/\sum_{k}\rho_{k}$ -- 
-that is the mass-weighted sum of all constituent-specific velocites (including sedimenting ones) --
-is used as a prognostic variable. In contrast to \cite{gassmann2008}, 
-a vector invariant form is only used for the horizontal velocity equation 
-(\ref{eq:HorizontalVelocityOrthogonal}), whereas the vertical velocity equation is solved in advective form. 
-The pressure gradient force is formulated according to (\ref{eq:PressureGradientForceAlt}).
-
-%Note that in contrast to (\ref{eq:VirtualTemperatureExact}) the water loading term is included in the definition of $T_{v}$ and $\theta_{v}$.
-
-Additional prognostic variables include total air density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), 
-and mass fractions $q_{k}=\rho_{k}/\rho$ of all constituents (except for dry air) for which the 
-prognostic equation reads
-\begin{align}
-\frac{\partial \rho q_{k}}{\partial t} + \nabla\cdot\left(\rho q_{k} \vb{u}_{bc}\right) = 
--\nabla\cdot \vb{J}_{k}  + \sigma_{k}\,,
-\end{align}
-with $\sigma_{k}$ describing sources/sinks due to phase changes, and  
-$\vb{J}_{k}=\rho q_{k}\left(\vb{u}_{k} - \vb{u}_{bc}\right)$ denoting diffusion fluxes, 
-which account for the motion of constituents relative to the frame of reference set by $\vb{u}_{bc}$.
-
-The specific heat capacities and ideal gas constant are approximated to be equal to their dry values 
-$R^{\ast}\approx R_{d}$, $c_{p}^{\ast}\approx c_{pd}$, $c_{v}^{\ast}\approx c_{vd}$. The model also 
-uses a prognostic equation for Exner pressure to simplify the treatment of vertical sound wave propagation, given by 
-\begin{align}
-\frac{\partial \pi}{\partial t} + \frac{R_{d}}{c_{vd}}\frac{\pi}{\rho \theta_{v}} \nabla\cdot\left(\vb{u}_{bc}\rho\theta_{v}\right) = \hat{Q}\,,
-\end{align}
-where $\hat{Q}$ is an appropriately formulated diabatic heat term. The horizontal uses a Arakawa C-grid formulation on the 
-triangular grid to prognose horizontal velocities normal to triangle edges $v_{n}$, making use of reconstructed 
-tangential velocity components $v_{t}$.
-
-In the current implementation, the following simplifcations are made with regards to the treatment of moisture: 
-The atmospheric mass loss/gain due to precipitation/evaporation is neglected in the total mass continuity 
-equation (\ref{eq:EulerianContinuityEquation}), by setting the vertical component of $\vb{u}_{bc}$ to zero at the lower boundary:  
-$w_{bc}|_{sfc} = 0$. In addition, only the vertical diffusion fluxes $\vb{J}_{k}$ of sedimenting 
-constituents and the surface evaporation flux $\vb{J}_{v}|_{sfc}$ are taken into account. The counter-flux 
-of non-sedimenting constituents is discarded. Since in the given framework the continuity equation (\ref{eq:EulerianContinuityEquation}) 
-is only valid if the constraint $\sum_{k} \vb{J}_{k}=0$ holds, we (implicitly) assume a fictitious 
-counter-flux of dry air to compensate for the considered vertical diffusion fluxes. As a consequence, 
-ICON currently conserves the global integral of total air mass rather than dry air mass.
-
-\subsubsection{OLAM} \label{sec:OLAMEquations}
-
-OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:EulerianContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and
-(\ref{eq:PotentialTemperatureFluxForm}).
-Prognostic variables are the 3 components of momentum, ice-liquid potential temperature $\theta_{il}$ \citep{walko2000efficient}, total density $\rho$, specific density of total water, and specific bulk density and/or bulk number concentration of various scalar quantities including liquid and ice hydrometeors, aerosols, and trace gases.
-For DCMIP, the latter are limited to cloud and rain specific bulk densities.
-Water vapor density $\rho q$ is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$.
-The latter is in turn diagnosed from $\theta_{il}$ and from the latent heat required to convert any hydrometeors present to the vapor phase.
-Velocity components are diagnosed by dividing momentum components by total density.
-
-Momentum is C-staggered in the horizontal and vertical (Lorenz vertical staggering is used), meaning that prognosed components live on the grid cell faces and are each normal to the respective face and the pressure gradient force is evaluated and applied at those locations.
-However, evaluation of advective and turbulent momentum transport (as well as the Coriolis force) involves a diagnostic reconstruction of the total momentum vector at the centers of scalar grid cells \citep{perot2000conservation}, and cell-to-cell flux of momentum is computed from that reconstruction using the same A- grid control volumes as for scalars.
-This arrangement is particularly convenient for the cut-cell formulation at the topographic surface where reduced cell face areas and volumes regulate momentum and scalar fluxes in an identical manner.
-
-\subsubsection{Tempest} \label{sec:TempestEquations}
-
-Tempest is a shallow-atmosphere Eulerian model with terrain-following $z$-coordinates with prognostic density (\ref{eq:PrognosticPseudoDensity}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for covariant horizontal velocity (\ref{eq:CovariantUpdateEquation}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Diffusion, stabilization, filters and fixers} \label{sec:DiffusionStabilization}
-
-Most dynamical cores implement specialized techniques for diffusion or stabilization (see Table \ref{tab:ModelsTemporalDiffusion}).
-Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
-Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
-Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.
-Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.
-A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.
-In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.
-
-\subsection{ACME--A / Tempest}
-
-In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
-Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
-\begin{align}
-\nabla^2 \vb{u}_h = \nabla \nabla \cdot \vb{u}_h + \nabla \times \nabla \times \vb{u}_h.
-\end{align}  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.
-Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
-
-\subsection{ICON}
-
-The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure.
-The details of this scheme appear in sections 2.4 and 2.5 of \cite{zangl2015icon}, and are summarized here.
-For damping, in the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the (relatively) computationally  expensive diffusion operator (see below) at the physics time steps without incurring numerical stability problems under extreme conditions,
-\begin{align}
-F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left\{\nabla \left[\tilde{\nabla} \cdot v + 
-\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right]\right\} \,.
-\end{align}
-$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.
-
-ICON also includes Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top.
-The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.
-
-The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$, defined via
-\begin{align} \label{smagn2}
-F_{D2}(v_n) &= 4 K_h \tilde{\nabla}^2(v_n), & F_{D2}(\theta) &= a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
-& F_{D4}(v_n) &= - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
-\end{align}
-An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.
-
-A fourth-order computational diffusion is also available for vertical wind speed $w$.
-This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures.
-This term appears as
-\begin{align}
-F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)).
-\end{align}
-
-\subsection{FV$^3$}
-
-Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations.
-The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.
-FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion.
-The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
-If additional damping is desired, or if the non-monotonic advection is used, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
-The hyperdiffusion is of the same order as but much smaller than the divergence damping.
-Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep.
-
-Wave absorption at the model top is also provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
-An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
-
-FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment.
-This is only done before the physics is called and is not used in idealized simulations.
-
-\subsection{GEM}
-
-An explicit hyperviscosity in GEM is handled via applications of the Laplacian operator
-for both wind components and tracers.
-A vertical sponge layer, which uses a Laplacian operator,
-is employed on wind components and $T_v$ with a vertical modulation on the topmost levels.
-For stabilization purpose, the temporal discretization of GEM also uses an off-centering parameter.
-The quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}
-is used operationally to ensure tracer monotonicity
-for specific humidity and different hydrometeors.
-Other options are now available
-in GEM including a mass conserving monotonic scheme \citep{Sorenson2013} and
-a global mass fixer \citep{Bermejo2002MASS}. Those approaches have been evaluated using
-chemical constituents such as ozone \citep{deGrandpre2016OZONE}.
-
-\subsection{OLAM}
-
-OLAM requires two types of artificial damping.
-In the upper layers of the model, vertical velocity and small-scale horizontal divergence are damped in order to attenuate gravity waves and thereby mitigate their reflection off the rigid top boundary of the domain.
-The damping layer is commonly applied in the uppermost 10 km of the domain, where the model top is 35 or 40 km above sea level.
-The damping rate is zero at the bottom of the damping layer and increases upward, usually linearly.
-
-Throughout the model domain, vertical vorticity is filtered horizontally at the smallest resolvable scale in order to control a spurious computational mode.
-This vertical vorticity mode is inherent in C-staggered momentum formulations on hexagonal meshes because horizontal velocities are more numerous than twice the number of scalar (mass) values and are thus under-constrained.
-The vorticity filter is constructed so as to have zero impact on divergence at any scale.
-
-Upwinding in the Lax-Wendroff formulation of the advection operator provides sufficient damping that no other type of filtering is required.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Temporal discretizations} \label{sec:TemporalDiscretizations}
-
-Temporal discretizations are important for capturing the discrete dynamical evolution of the global atmosphere.
-In the past two decades, a variety of new temporal discretizations have been developed, leaving behind the days when the leapfrog scheme was ubiquitous across models.
-This diversity is in part because of the demands of non-hydrostatic models: unlike their hydrostatic counterparts, non-hydrostatic atmospheric models that do not use the anelastic \citep{ogura1962scale}, quasi-hydrostatic \citep{orlanski1981quasi}, pseudo-incompressible \citep{durran1989improving}, or unified approximation \citep{arakawa2009unification} must include a mechanism in the temporal discretization for dealing with vertically propagating sound waves.
-These waves are meteorologically insignificant, but with a vertical grid spacing of 100 meters, a purely explicit temporal discretization would require a time step size on the order of one second or less.
-In this section we discuss the timestepping schemes that have been employed across the DCMIP suite of models.
-
-\subsection{Mixed implicit-explicit, forward-backward, semi-implicit and additive Runge-Kutta schemes}
-
-Implicit-explicit schemes are a broad category of time integration schemes that divide the terms of the prognostic equations into a set of explicitly integrated terms and implicitly integrated terms.
-At the very least, terms associated with vertically propagating sound waves are included among the implicit terms.
-For the remaining terms, there is some freedom in choosing how to integrate terms associated with vertical advection and horizontally propagating sound waves.
-Semi-implicit schemes are one such class of schemes that typically incorporate horizontally propagating sound waves into the implicit solve, and so rely on a global Helmholtz-type solve.
-Additive Runge-Kutta schemes are another mechanism to ensure high-order temporal accuracy, and many such schemes have been described throughout the literature (see, for example, \cite{weller2013runge, ullrich2012operator}).
-Several examples of these schemes can be found among the DCMIP models:
-
-FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
-With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}.
-The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the non-hydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver.
-This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps.
-Physics tendencies are applied impulsively at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep.
-
-DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit.
-A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation).
-As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels.
-The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each.
-The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure.
-Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.
-
-ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves.
-No time splitting is used with respect to sound waves, because the ratio between the speed of sound and the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one.
-Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand.
-Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied.
-Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2015icon}.
-
-MPAS uses a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
-At the beginning of each Runge-Kutta sub-cycle, tendencies are computed for each of the prognostic variables and stored for the duration of the sub-cycle.
-Several iterations of an acoustic loop are then performed with a time-step much smaller than require for the Runge-Kutta sub-cycle.
-Within the acoustic loop, an implicit solve for vertically-integrated sound waves is performed to avoid timestep constraints that may arise from vertically-propagating sound waves.
-
-OLAM uses a unique temporal discretization that combines elements of the Adams-Bashforth (AB2) scheme and a Lax- Wendroff formulation for advected quantities.
-However, instead of extrapolating all prognostic tendencies forward to the half-future time level as in AB2, the horizontal momentum components alone (not their tendencies) are extrapolated in time at the cell boundaries where they reside.
-The extrapolated momentum provides the time-centered cell-to-cell total mass flux across the grid cell faces that is responsible for advective transport.
-Advection of all quantities, including all 3 velocity components that are diagnostically reconstructed at scalar cell centers, and advancement in time from the current to the future time level is based on the time- and space-centered Lax-Wendroff formulation.
-This scheme is horizontally explicit, but a trapezoidal- implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
-A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection.
-The vertical momentum equation is solved first so that the time-centered vertical momentum is available for computing transport of horizontal momentum and all scalar quantities.
-A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.
-
-%This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves.
-A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection.
-Scalar advection is performed afterward and uses the same total mass fluxes as momentum advection.
-A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.
-
-Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
-
-\subsection{The FVM Semi-Implicit method} \label{sec:FVMSemiImplicit}
-
-A characteristic feature of the FVM (Section~\ref{sec:FVM}) time-stepping scheme is the 3D implicit treatment 
-of the fast buoyant and acoustic modes, and the slow rotational modes. Therefore, the model time step is 
-identical for all processes and typically selected with regard to the stability of the advective transport scheme---i.e.~%
-the time step is continuously adapted according to a given maximum advective Courant number 
-permitted by the MPDATA scheme.
-A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016}
-and \cite{kuehnleinJCP2017} for dry dynamics, 
-whereas in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalJCP2017} for extension to moist-precipitating dynamics.  
-Here, we provide a short outline of the solution procedure for the compressible Euler 
-equations \eqref{fvm:compressible}. It employs the two-time-level second-order-accurate template algorithm given as
-\begin{equation}
-\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
-+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
-\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
-\label{fvm:solscalar}
-\end{equation}
-where $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
-$\mathcal{A}$ symbolises the advective transport operator given by the non-oscillatory finite-volume 
-MPDATA  scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}, and the spatial mesh vector 
-index $\mathbf{i} \equiv (k,i)$ denotes the position on the hybrid horizontally-unstructured 
-vertically-structured computational mesh.
-
-The solution procedure of the system \eqref{fvm:compressible} can then basically be divided into three steps. First, the homogenous
-mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
-and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, given already updated moisture variables 
-\citep{smolarkiewiczetalJCP2017}, the thermodynamic \eqref{fvm:thermodynamic} and momentum \eqref{fvm:momentum} equations enter
-\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
-which is generally depending on all these prognostic variables. The high degree of implicitness in the representation of the rhs forcings
-is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
-velocity updates---this procedure is facilitated by the co-located arrangement of all variables on the computational mesh.
-Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term.
-The subsequent third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
-an advective form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
-equation with a Euler backward scheme, in the spirit of \eqref{fvm:solscalar}, leads a Helmholtz equation
-\citep{smolarkiewiczJCP2014}.
-The associated 3D elliptic boundary value problem is solved iteratively using a preconditioned Generalised Conjugate 
-Residual approach, see \cite{smolarkiewiczAG2011} for a recent overview and comprehensive list of references.
-Nonlinear terms in $R^{\psi}|^{\scs n+1}$ and the 
-solution-dependent coefficients of the Helmholtz equation are lagged behind and executed in an outer iteration.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection {A semi-Lagrangian implicit time discretization in the GEM model} \label{sec:GEM_temporal}
-
-GEM differs from the approaches above by using a semi-Lagrangian advection.
-Any model equations, prognostic or diagnostic, are written in the form
-\begin{align}
-\frac {dF}{dt}+G=0,
-\label{eq:gem_temporal_O}
-\end{align}
-\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
-The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
-\begin{align}
-\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
-\label{eq:gem_temporal_D}
-\end{align}
-where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
-and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
-due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
-having occurred during the timestep $\Delta t$.
-$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
-The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
-\begin{align}
-\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
-\label{eq:gem_temporal_d_O}
-\end{align}
-discretized in the same way (trapezoidal method) though without off-centering:
-\begin{align}
-\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
-\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
-\label{eq:gem_temporal_d_D}
-\end{align}
-
-\noindent The process is of course a {\it{multi-step iterative}}
-one
-since both positions and velocities at departure positions (past time $t-\Delta t$)
-are unknown as well as, of course, the velocities at arrival positions (time $t$).
-Once a first estimate of the departure positions is obtained,
-the model equations are solved to obtain a first estimate of the velocities at time $t$.
-The model equations must be solved simultaneously and this is only possible for the linear part $L$
-which becomes a {\it {matrix inversion problem}}.
-Hence a suitable linearization is considered. The unknown (arrival) linear $L$ and non-linear $N$
-parts are then separated from the known (first departure estimate) remaining $R$ part.
-Thus, first separating space-times, (\ref{eq:gem_temporal_D}) is rewritten as follows
-\begin{align}
-\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D} \equiv R^{D},
-\label{eq:gem_temporal_AD}
-\end{align}
-where $\tau = \left ( 1/2 + \epsilon\right)\Delta t$ and $\beta=\left ( 1/2 - \epsilon\right)/\left ( 1/2 + \epsilon\right)$.
-Second separating linear from non-linear parts, we get:
-\begin{align}
-L^A + N^A = R^D,
-\label{eq:gem_temporal_LN}
-\end{align}
-\noindent
-with
-\begin{align}
-L^A=\left[\frac{F^A}{\tau} + G^A \right]_{linear}, \quad and \quad N^A \equiv \frac{F^A}{\tau} + G^A  - \left[\frac{F^A}{\tau} + G^A \right]_{linear}.
-\end{align}
-Note that both $F$ and $G$ may require linearization. $L^A$ may then be obtained if $N^A$ is first guessed:
-Once $L^A$ is found, an estimate of $N^A$ is obtained and $L^A$ is recalculated.
-This is called the {\it{non-linear iteration process}}
-(one iteration is usually sufficient).
-The overall process is then repeated once starting from a new estimate of the departure positions.
-
-There are two intensive calculation sections in this process:
-the so-called semi-Lagrangian calculations
-(twice estimating departure positions, twice interpolating right-hand sides $R$ on departure positions),
-and solution of the linear system (four times).
-Each time, the linear system is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid
-\citep{Qaddouri2008schwarz}.
-The composite variable solution is then used to update the prognostic variables (back substitution).
-At the end of the time-step, the static halo region of both panels of the Yin-Yang grid is updated \citep{Qaddouri2011operational}.
-All required interpolations throughout the semi-Lagrangian process and between Yin and Yang grids are cubic interpolations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\conclusions[Summary and conclusions] \label{sec:Conclusions}  %% \conclusions[modified heading if necessary]
-
-As discussed earlier, this paper represents the first in a series of papers documenting the results from the 2016 Dynamical Core Model Intercomparison Project workshop and summer school.
-In this paper we have provided a description of the differences and similarities between participating models, including the choice of computational grid, horizontal staggering, vertical staggering, vertical coordinates, prognostic equations, choice of diffusion, stabilization, filters and fixers, and temporal discretization.
-The literature on dynamical core development is vast, with origins that go back over half a century.
-Consequently, the models discussed in this paper only represent a sample of the many dynamical cores that have been developed for general circulation modeling.
-Some of the models that have not been discussed include \cite{fox1997finite, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global, wood2014inherently} and \cite{doyle2014next}.
-
-The vast diversity of the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.
-Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.
-The repercussions that emerge from these choices can then be explored in the context of idealized test cases, such as the ones that have been proposed as part of DCMIP.
-The remaining papers in this series investigate how the models described in this paper are able to simulate three idealized test cases, which each incorporate simplified model physics:  a moist baroclinic wave, an idealized tropical cyclone, and a splitting supercell storm on a small planet.
-Where appropriate, metrics have been included that may be indicative of model performance.
-These tests can also be used for future dynamical core development to identify where a new dynamical core diverges from a suite of modern models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}
-
-%This section provides an overview of the shared notation adopted by DCMIP for describing the equations and discretizations that have been adopted by the dynamical cores involved in this project.
-Table \ref{tab:symbols} lists the symbols used in this paper.
-%Great circle distance is used throughout the project and is computed via
-%\begin{equation}
-%R_c(\lambda_1, \varphi_1; \lambda_2, \varphi_2) = a \arccos \left( \sin \varphi_1 \sin \varphi_2 + \cos \varphi_1 \cos \varphi_2 \cos (\lambda_1 - \lambda_2) \right).
-%\end{equation}
-
-In this appendix we provide a detailed derivation of the fluid equations utilized by non-hydrostatic models.
-The physical constants which are used throughout this document is given in Table \ref{tab:PhysicalConstants}.
-The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
-\begin{align}
-\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla,
-\end{align} where $\vb{u}$ denotes the 3D vector velocity.
-Note that tracer variables $q_i$, including specific humidity $q$, in the absence of sources and sinks satisfy the simple Lagrangian relationship
-\begin{align} \label{eq:TracerTransport}
-\diff{q_i}{t} = 0.
-\end{align}
-
-\begin{table}[h]
-\caption{A list of physical constants used in this document.} \label{tab:PhysicalConstants}
-%\ \\
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lll}
-\hline Constant & Description & Value \\
-\hline $a_{\tiny \mbox{ref}}$ & Radius of the Earth & $6.37122 \times 10^{6}\ \mbox{m}$ \\
-$\Omega_{\tiny \mbox{ref}}$ & Rotational speed of the Earth & $7.292\ \times 10^{-5}\ \mbox{s}^{-1}$ \\
-%$a$ & Scaled radius of the Earth & $a_{\tiny \mbox{ref}} / X$ \\
-%$\Omega$ & Scaled rotational speed of the Earth & $\Omega_{\tiny \mbox{ref}} \cdot X$ \\
-$g_c$ & Gravitational acceleration & $9.80616\ \mbox{m}\ \mbox{s}^{-2}$ \\
-$p_0$ & Reference pressure & $1000\ \mbox{hPa}$ \\
-$c_{pd}$ & Specific heat capacity of dry air at constant pressure & $1004.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_{pw}$ & Specific heat capacity of water vapor at constant pressure & $1930.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_{vd}$ & Specific heat capacity of dry air at constant volume & $717.5\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$c_{vw}$ & Specific heat capacity of water vapor at constant volume & $1460.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$R_d$ & Gas constant for dry air & $287.0\ \mbox{J}\ \mbox{kg}^{-1}\ \mbox{K}^{-1}$ \\
-$R_w$ & Gas constant for water vapor & $461.5$ J kg$^{-1}$ K$^{-1}$ \\
-%$\kappa$ & Ratio of $R_d$ to $c_p$ & $2/7$ \\
-$\varepsilon$ & Ratio of $R_d$ to $R_w$ & $0.622$ \\
-$M_v$ & Constant for virtual temperature conversion & $0.608$ \\
-$\rho_{water}$ & Reference density of water & 1000 kg m$^{-3}$ \\
-\hline 
-\end{tabular*}
-\end{table}
-
-\subsection{Diagnostic relationships}
-
-The atmospheric fluid is assumed to be an ideal gas.
-For moist air, the ideal gas constant $R^\ast$, specific heat capacity at constant pressure $c_p^\ast$ and specific heat capacity at constant volume $c_v^\ast$ are given by
-\begin{align}
-R^\ast =& R_d + (R_w - R_d) q, & c_p^\ast =& c_{pd} + (c_{pw} - c_{pd}) q, & c_v^\ast =& c_{vd} + (c_{vw} - c_{vd}) q.
-\end{align}  Note that in many models, $R^\ast$, $c_p^\ast$ and $c_v^\ast$ are approximated by $R_d$, $c_{pd}$ and $c_{vd}$, respectively.
-Dry air, water vapor and moist air quantities all satisfy the linear relationship $R = c_p - c_v$.
-For a two-fluid system (dry air plus water vapor), two independent variables plus the specific humidity $q$ are needed to describe the thermodynamic state of the system.
-Key thermodynamic variables include dry air density $\rho_d$, moist density $\rho$, pressure $p$, vapor pressure $e$, temperature $T$, virtual temperature $T_v$, Exner pressure $\pi$, potential temperature $\theta$, and virtual potential temperature $\theta_v$.
-Common ratios $\kappa = R^\ast / c_p^\ast$, $\varepsilon = R_d / R_w$, and $\gamma = c_p^\ast / c_v^\ast$ are adopted here.
-Note that as additional water species are added (cloud water, rain water, etc.) additional independent variables are needed to capture the thermodynamic effects of these species, and the ``virtual'' quantities modified accordingly, for instance through the adoption of density potential temperature $\theta_\rho$.
-
-Relationships between key thermodynamic variables arise from the ideal gas law, along with definitions of Exner pressure, potential temperature and virtual potential temperature
-\begin{align} \label{eq:DiagnosticRelationships1}
-p =& \rho R_d T_v, & \pi =& \left( \frac{p}{p_0} \right)^{\kappa}, & \theta =& T \left( \frac{p_0}{p} \right)^{\kappa}, & \theta_v =& T_v \left( \frac{p_0}{p} \right)^{\kappa},
-\end{align}  which further give rise to
-\begin{align}
-p =& \left( \frac{\rho R_d \theta_v}{p_0^\kappa} \right)^{\gamma}, & \pi =& \left( \frac{\rho R_d \theta_v}{p_0} \right)^{R^\ast / c_v^\ast}, & \theta =& \frac{T}{\pi}, & \theta_v =& \frac{T_v}{\pi}.
-\end{align}
-
-Note that virtual temperature is typically approximated by
-\begin{align}
-T_v \approx& T \left(1 + \frac{(1-\varepsilon)}{\varepsilon} q \right),
-\end{align} which arises from the exact relationship
-\begin{align} \label{eq:VirtualTemperatureExact}
-T_v = \frac{T}{1 - \frac{e}{p} (1 - \varepsilon)},
-\end{align} upon applying $e/p = q / \varepsilon$ and using a Taylor expansion around $q = 0$.
-
-\subsection{Prognostic equations for thermodynamic variables}
-
-Note that, as a consequence of (\ref{eq:TracerTransport}), the following simplifications can be applied:
-\begin{align}
-\frac{1}{T_v} \diff{T_v}{t} =& \frac{1}{T} \diff{T}{t}, & \diff{R^\ast}{t} =& 0, & \diff{c_p^\ast}{t} =& 0, & \diff{c_v^\ast}{t} =& 0.
-\end{align}
-
-Mass conservation is typically represented through the continuity equation, which can be written in the Lagrangian frame as
-\begin{align} \label{eq:ContinuityEquation}
-\diff{\rho}{t} =& - \rho \nabla \cdot \vb{u},
-\end{align} or equivalently in the Eulerian frame,
-\begin{align} \label{eq:EulerianContinuityEquation}
-\pdiff{\rho}{t} =& - \nabla \cdot (\rho \vb{u}).
-\end{align}
-
-Further prognostic relationships can be derived from the thermodynamic equation, including the diabatic heating rate $J$,
-\begin{align}
-\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
-\end{align} which can be alternatively written as
-\begin{align} \label{eq:PotentialTemperaturePrognostic}
-\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
-\end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
-\begin{align} \label{eq:PotentialTemperatureFluxForm}
-\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
-\end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
-\begin{align}
-\frac{1}{p} \diff{p}{t} = \frac{1}{\rho} \diff{\rho}{t} + \frac{1}{T_v} \diff{T_v}{t},
-\end{align} the thermodynamic equation can be written in the form
-\begin{align}
-\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
-\end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
-\begin{align} \label{eq:VirtualTemperaturePrognosticEq}
-\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
-\end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
-\begin{align} \label{eq:PressurePrognosticEq}
-\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
-\end{align}  And similarly for Exner pressure,
-\begin{align}
-\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
-\end{align}
-
-\subsection{Momentum Equations}
-
-In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
-\begin{align} \label{eq:PrognosticVelocity1}
-\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
-\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.
-The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.
-In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis.
-Note that it is common to rewrite the pressure gradient force using the relationship
-\begin{align} \label{eq:PressureGradientForceAlt}
-- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
-\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).
-Note that when condensate loading is included, the virtual potential temperautre in this equation is replaced with $\theta_\rho$.
-A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
-\begin{align}
-\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
-\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
-This gives rise to the 3D vector-invariant form,
-\begin{align} \label{eq:PrognosticVelocity2}
-\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
-\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis  that would arise from the momentum transport term $\vb{u} \cdot \nabla \vb{u}$ in (\ref{eq:PrognosticVelocity1}).
-In conjunction with (\ref{eq:ContinuityEquation}), (\ref{eq:PrognosticVelocity1}) also gives rise to the flux-form momentum equations,
-\begin{align} \label{eq:PrognosticVelocityFluxForm}
-\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
-\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.
-
-%Finally, a prognostic equation for geopotential follows directly from the material derivative,
-%\begin{align} \label{eq:Geopotential}
-%\diff{\Phi}{t} = \vb{u} \cdot \nabla \Phi.
-%\end{align}
-
-The equations above still provide some flexibility with regards to the choice of $\Phi$ and $\vg{\Omega}$.  For \textit{deep atmosphere} models, one typically chooses
-\begin{align} \label{eq:DeepAtmosphereGeopotential}
-\Phi = g_c a^2 \left[\frac{1}{a} - \frac{1}{a+z}\right], \quad \mbox{and} \quad \vg{\Omega} = \Omega ( \vb{k} \sin \varphi + \vb{j} \cos \varphi),
-\end{align} where $g_c$ is gravitational acceleration at the surface, $a$ is the radius of the planet, $\Omega$ is the rotation rate (in s$^{-1}$), $\varphi$ is the latitude, $\vb{j}$ is the unit vector oriented in the meridional direction, and $\vb{k}$ is the unit vector oriented in the vertical direction.
-For models that don't utilize a height-based vertical coordinate, the geopotential is generally treated as a prognostic variable, with an evolution equation that emerges from the definition $w = dz/dt$,
-\begin{align}
-\diff{\Phi}{t} = \frac{a^2 g_c w}{(a + z)^2}.
-\end{align}  For \textit{shallow atmosphere} models, the geopotential takes the simpler form
-\begin{align} \label{eq:ShallowAtmosphereGeopotential}
-\Phi = g_c z, \quad \mbox{and} \quad \vg{\Omega} = \Omega \sin \varphi \vb{k},
-\end{align} where $z$ is the altitude above the surface.
-In this case we write $2 \vg{\Omega} = f \vb{k}$, where $f = 2 \Omega \sin \varphi$ is the Coriolis parameter.
-The evolution equation for the shallow atmosphere geopotential is then
-\begin{align} \label{eq:PrognosticShallowGeopotential}
-\diff{\Phi}{t} = g_c w.
-\end{align}
-
-\subsection{Orthogonal Formulation}
-
-Under the orthogonal formulation, projection of a vector field $\vb{b}$ onto its horizontal components is defined via
-\begin{align}
-[\vb{b}]_z = \vb{b} - (\vb{b} \cdot \vb{k}) \vb{k}.
-\end{align}  When applied to the velocity vector this gives rise to the decomposition
-\begin{align}
-\vb{u} = \vb{u}_h + w \vb{k},
-\end{align} where $\vb{k} = \nabla z$ is the unit vector in the vertical direction and $\vb{u}_h = [\vb{u}]_z$ ($\vb{u}_h$ is aligned with surfaces of constant $z$).
-In the orthogonal formulation, the material derivative expands as
-\begin{align} \label{eq:MaterialDerivative}
-\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla + w \vb{k} \cdot \nabla.
-\end{align}  For the special case of the material derivative applied to scalars, this equation can also be written as
-\begin{align} \label{eq:MaterialDerivativeScalar}
-\diff{}{t} = \pdiff{}{t} + \vb{u}_h \cdot \nabla_z + w \pdiff{}{z}.
-\end{align} where $\nabla_z b = [\nabla b]_z$ denotes the gradient along constant $z$ surfaces.
-From here, the vector-invariant form velocity equation obtained by taking (\ref{eq:PrognosticVelocity2}) $\cdot \vb{k}$ expands as
-\begin{align} \label{eq:VerticalVelocityOrthogonal}
-\pdiff{w}{t} =& - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{}{z} (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k},
-\end{align} which, from $\vb{u}_h = \vb{u} - w \vb{k}$, then gives rise to
-\begin{align} \label{eq:HorizontalVelocityOrthogonal}
-\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z.\end{align}
-
-%{\color{red}[Using $K = K_2 + w^2/2$ does this give rise to a simpler version of the $K_2$ evolution equations?]}
-
-Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.
-Decomposing the momentum transport term into horizontal and vertical components gives
-\begin{align}
-\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
-\end{align}  The first term in this expression admits the relationships
-\begin{align}
-\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \label{eq:UhDotNablaUhRelationship1} \\
-(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right] \label{eq:UhDotNablaUhRelationship2}
-\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.
-Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.
-
-Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1})$\cdot \vb{k}$, is
-\begin{align} \label{eq:PrognosticWEquationK2}
-\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{w}{z} - \vb{u}_h \cdot \nabla_z w - \pdiff{\Phi}{z} - \frac{1}{\rho} \pdiff{p}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
-\end{align}  Then subtracting (\ref{eq:PrognosticWEquationK2})$\cdot \vb{k}$ from (\ref{eq:PrognosticVelocity1}) gives
-\begin{align} \label{eq:PrognosticUhEquationK2}
-\pdiff{\vb{u}_h}{t} =& - w (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{\vb{u}_h}{z} - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
-\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.
-
-\subsection{Arbitrary vertical coordinates}
-
-The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e.
-following \cite{kasahara1974various} (hereafter K74).
-Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
-\begin{align} \label{eq:ChainRule}
-\pdiff{}{z} =& \pdiff{s}{z} \pdiff{}{s}, & \nabla_s =& \nabla_z + (\nabla_s z) (\vb{k} \cdot \nabla), & \left( \pdiff{}{t} \right)_s =& \pdiff{}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla),
-\end{align} which correspond to derivatives in the vertical, in the horizontal and in time.
-This final expression is used to describe the rate of change of a quantity on $s$ surfaces.
-These operators then yield the useful identities
-\begin{align} \label{eq:NablaZSIdentity}
-\pdiff{s}{z} =& \left( \pdiff{z}{s} \right)^{-1}, & \nabla_z s =& - \left( \pdiff{s}{z} \right) \nabla_s z, & \pdiff{s}{t} =& - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
-\end{align}  From here (\ref{eq:ChainRule}) also gives rise to
-\begin{align} \label{eq:HorizontalChainRule}
-\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s},
-\end{align} which can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.
-Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z A = 0$ and $\vb{k} \cdot \nabla_s A = 0$ for any scalar field $A$.
-
-From (\ref{eq:ChainRule}), it can be shown that the 2D divergence on $s$ surfaces (given by K74 eq. (3.17)) is
-\begin{align} \label{eq:DivergenceS}
-\nabla_s \cdot \vb{u}_h = \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \cdot \left( \pdiff{\vb{u}_h}{s} \right),
-\end{align} and that the 2D curl is given by
-\begin{align} \label{eq:CurlS}
-\nabla_s \times \vb{u}_h = \nabla_z \times \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \times \left( \pdiff{\vb{u}_h}{s} \right),
-\end{align} where $\nabla_z \times \vb{u}_h = \vb{k} (\vb{k} \cdot (\nabla \times \vb{u}_h))$.  Notably, these expressions are valid for both shallow- and deep-atmosphere formulations.
-
-The generalized velocity $\dot{s}$ following a fluid parcel is defined by
-\begin{align} \label{eq:DefinitionDotS}
-\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \vb{u}_h \cdot \nabla_z s + \left[ w - \left( \pdiff{z}{t} \right)_s \right] \pdiff{s}{z}.
-\end{align}  Then using (\ref{eq:ChainRule}) and (\ref{eq:DefinitionDotS}) to rewrite (\ref{eq:MaterialDerivativeScalar}) gives an expression for the material derivative for scalars on $s$ surfaces,
-\begin{align}
-\diff{A}{t} =& \left( \pdiff{A}{t} \right)_s - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s \pdiff{A}{s} + \vb{u}_h \cdot \left[ \nabla_s A + (\nabla_z s) \pdiff{A}{s} \right] + w \pdiff{s}{z} \pdiff{A}{s} \\
-=& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
-\end{align}  A similar expression arises for vectors, although in this case $\vb{u}_h \cdot \nabla \vb{a} \neq \vb{u}_h \cdot \nabla_z \vb{a}$ implies we cannot use the operator $\nabla_s$ in the form (\ref{eq:ChainRule}), and instead obtain
-\begin{align} \label{eq:MaterialDerivativeVectorS}
-\diff{\vb{a}}{t} =& \left( \pdiff{\vb{a}}{t} \right)_s + \left[ \vb{u}_h \cdot \nabla \vb{a} + (\vb{u}_h \cdot \nabla_s z) (\vb{k} \cdot \nabla \vb{a}) \right] + \dot{s} \pdiff{\vb{a}}{s}.
-\end{align}
-
-%{\color{red}[Is this needed?]} Note that
-%\begin{align}
-%\nabla A = \nabla_z A + \vb{k} \pdiff{A}{z} = \nabla_s A - \pdiff{s}{z} (\nabla_s z) \pdiff{A}{s} + \vb{k} \pdiff{s}{z} \pdiff{A}{s} = \nabla_s A + ( \vb{k} - \nabla_s z) \pdiff{s}{z} \pdiff{A}{s}
-%\end{align}
-
-\subsection{Conservation Laws in Arbitrary Vertical Coordinates}
-
-Using (\ref{eq:DivergenceS}), we observe that the 3D divergence on the sphere takes the form
-\begin{align} \label{eq:ThreeDDivergence}
-\nabla \cdot \vb{u} = \nabla_z \cdot \vb{u}_h + \frac{1}{\alpha} \pdiff{}{z} (\alpha w),
-\end{align} where $\alpha = 1$ for shallow-atmosphere models and $\alpha = r^2 = (a + z)^2$ for deep-atmosphere models.  Using $w = dz/dt$, this last term also takes the form
-\begin{align} \label{eq:GeneralVerticalTermX}
-\frac{1}{\alpha} \pdiff{}{z} (\alpha w) = \pdiff{w}{z} + \frac{w}{\alpha} \pdiff{\alpha}{z} = \pdiff{w}{z} + \frac{1}{\alpha} \diff{\alpha}{t}.
-\end{align}
-
-Using (\ref{eq:NablaZSIdentity}) to rewrite (\ref{eq:DefinitionDotS}) gives rise to
-\begin{align}
-w = \left( \pdiff{z}{t} \right)_s + \vb{u}_h \cdot \nabla_s z + \dot{s} \left( \pdiff{s}{z} \right)^{-1},
-\end{align} which is then differentiated to yield K74 eq. (3.16),
-\begin{align} \label{eq:WDerivativeS}
-\pdiff{w}{z} = \left( \pdiff{s}{z} \right) \left[ \diff{}{t} \left( \pdiff{s}{z} \right)^{-1} + \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \pdiff{\dot{s}}{s} = 0.
-\end{align}  Substituting this expression into the continuity equation (\ref{eq:ContinuityEquation}), and using (\ref{eq:ThreeDDivergence}), (\ref{eq:GeneralVerticalTermX}), and (\ref{eq:WDerivativeS}) then leads to
-\begin{align}
-\diff{}{t} \left[ \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \left[ \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) \left( \pdiff{\vb{u}_h}{s} \right) \cdot (\nabla_s z) \right] + \alpha \left( \pdiff{s}{z} \right)^{-1} \rho \pdiff{\dot{s}}{s} = 0.
-\end{align}  Defining the \textit{pseudodensity} as 
-\begin{align}
-\rho_s = \alpha \left( \pdiff{s}{z} \right)^{-1} \rho,
-\end{align} and using (\ref{eq:MaterialDerivativeS}) in the form
-\begin{align}
-\diff{\rho_s}{t} = \left( \pdiff{\rho_s}{t} \right)_s + \vb{u}_h \cdot \nabla_s \rho_s + \dot{s} \pdiff{\rho_s}{s},
-\end{align} along with (\ref{eq:DivergenceS}) leads to
-\begin{align} \label{eq:PrognosticPseudoDensity}
-\left( \pdiff{\rho_s}{t} \right)_s + \nabla_s \cdot (\rho_s \vb{u}_h) + \pdiff{}{s} (\rho_s \dot{s}) = 0.
-\end{align}
-
-Hence for any quantity that is conserved following a fluid parcel (\textit{i.e.}, $dq/dt = 0$),
-\begin{align}
-\left( \pdiff{\rho_s q}{t} \right)_s + \nabla_s \cdot (\rho_s q \vb{u}_h) + \pdiff{}{s} (\rho_s q \dot{s}) = 0.
-\end{align}  In particular, the prognostic equation for virtual potential temperature (or equivalently for potential temperature) reads
-\begin{align} \label{eq:PrognosticPseudoDensityTheta}
-\left( \pdiff{\rho_s \theta_v}{t} \right)_s + \nabla_s \cdot (\rho_s \theta_v \vb{u}_h) + \pdiff{}{s} (\rho_s \theta_v \dot{s}) = \frac{J \rho_s \theta_v}{c_p^\ast}.
-\end{align}
-
-\subsection{2D Vector Invariant Form}
-
-The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are derived by applying (\ref{eq:ChainRule}) to (\ref{eq:PrognosticWEquationK2}), yielding
-\begin{align} \label{eq:PrognosticWEquationK2Sa}
-\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
-\end{align}  Similarly, from (\ref{eq:PrognosticUhEquationK2}),
-\begin{align} \label{eq:PrognosticUhEquationK2Sa}
-\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) + \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \zeta_h \vb{k} \times \vb{u}_h - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z.
-\end{align}  Observe that both of these equations simplify when $w = (\partial z / \partial t)_s$, \textit{i.e.} model levels are advected with the vertical wind.
-
-An alternative form of these equation can similarly be obtained in terms of $\dot{s}$.  Substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticWEquationK2Sa}) then gives
-\begin{align} \label{eq:PrognosticWEquationK2Sb}
-\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla_s w - \dot{s} \pdiff{w}{s} + \left( \pdiff{s}{z} \right) \left[ - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right] - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
-\end{align}  Similarly, substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticUhEquationK2Sa}) and using the identity
-\begin{align}
-(\vb{u}_h \cdot \nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} = (\nabla_s \times \vb{u}_h) \times \vb{u}_h - \zeta_h \vb{k} \times \vb{u}_h + (\nabla_s z) \pdiff{K_2}{z}
-\end{align}
- then gives
-\begin{align} \label{eq:PrognosticUhEquationK2Sb}
-\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) - \nabla_s K_2 - \zeta_s \vb{k} \times \vb{u}_h - \dot{s} \pdiff{\vb{u}_h}{s} - \nabla_z \Phi - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z,
-\end{align} where
-\begin{align}
-\nabla_s \times \vb{u}_h = \vb{k} \zeta_s, \quad \mbox{and} \quad \zeta_s = \vb{k} \cdot (\nabla_s \times \vb{u}_h).
-\end{align}  In this case the vertical advection terms are removed when $\dot{s} = 0$, \textit{i.e.} the vertical coordinate is advected with the 3D wind $\vb{u}$.
-
-Note that under the shallow-atmosphere approximation, the first metric terms (those that include $(\vb{u}_h \cdot \nabla \vb{k})$) in (\ref{eq:PrognosticWEquationK2Sa})-(\ref{eq:PrognosticUhEquationK2Sb}) are typically dropped.
-
-\subsection{3D Vector Invariant Form}
-
-From (\ref{eq:PrognosticVelocity2}) and (\ref{eq:ChainRule}) the evolution equation for the 3D velocity vector takes the form
-%\begin{align}
-%\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}
-%\end{align} we have
-\begin{align}
-\left( \pdiff{\vb{u}}{t} \right)_s = \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}) - \nabla (K + \Phi) - \frac{1}{\rho} \nabla p - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}
-\end{align}  Then taking the dot product of this expression with $\vb{k}$ gives
-\begin{align} \label{eq:PrognosticWEquation}
-\left(\pdiff{w}{t} \right)_s =& \left( \pdiff{s}{z} \right) \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \pdiff{}{s} (K + \Phi) - \frac{1}{\rho} \pdiff{p}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}
-\end{align} where we have used $\vb{k} \cdot (\vb{k} \cdot \nabla \vb{u}) = \vb{k} \cdot \nabla w$.
- Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
-\begin{align} \label{eq:PrognosticUhEquationKS}
-\left( \pdiff{\vb{u}_h}{t} \right)_s =& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \nabla_z (K + \Phi) - \frac{1}{\rho} \nabla_z p - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z.
-\end{align}  Note that the vorticity term in this expression can be simplified further using
-\begin{align}
-[(\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_z = - (\zeta_h + \vb{k} \cdot 2 \vg{\Omega}) (\vb{u}_h \times \vb{k}) - w \vb{k} \times (\vg{\zeta} + 2 \vg{\Omega}),
-\end{align} and
-\begin{align}
-- \vb{k} \times \vg{\zeta} =& \vb{k} \cdot \nabla \vb{u} - \nabla (\vb{k} \cdot \vb{u}) + \vb{u} \cdot \nabla \vb{k} = \pdiff{\vb{u}_h}{z} - \nabla_z w + \vb{u}_h \cdot \nabla \vb{k}.
-\end{align}
-%so that (\ref{eq:PrognosticUhEquationKS}) takes the form
-%\begin{align}
-%\left( \pdiff{\vb{u}_h}{t} \right)_s =& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \nabla_z (K + \Phi) - \frac{1}{\rho} \nabla_z p - \left[ 2 \vg{\Omega} \times \vb{u} \right]_z - \zeta_h (\vb{u}_h \times \vb{k}) + w \left[ \pdiff{\vb{u}_h}{z} - \nabla_z w + \vb{u}_h \cdot \nabla \vb{k} \right]
-%\end{align}
-
-\subsection{Covariant Component Formulation}
-
-In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (in 2D vector invariant form as  (\ref{eq:PrognosticUhEquationK2Sa}) or (\ref{eq:PrognosticUhEquationK2Sb}), or in 3D vector invariant form as (\ref{eq:PrognosticUhEquationKS})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.
-This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.
-We define a horizontal covariance operator by
-\begin{align}
-[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
-\end{align} Applying this operator to the horizontal velocity gives
-\begin{align}
-\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
-\end{align}  For a time-dependent $s$ coordinate, we obtain the identity
-\begin{align}
-\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
-\end{align} and so can write
-\begin{align} \label{eq:CovariantExpansion}
-\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + (\nabla_s z) \left( \pdiff{w}{t} \right)_s + w \nabla_s \left( \pdiff{z}{t} \right)_s.
-\end{align}  Then using (\ref{eq:CovariantExpansion}), (\ref{eq:PrognosticWEquation}) and (\ref{eq:PrognosticUhEquationKS}) and identity
-\begin{align}
-& \left( \pdiff{z}{t} \right)_s \left( \pdiff{s}{z} \right) \left[ \pdiff{\vb{u}_h}{s} + (\nabla_s z) \pdiff{w}{s} \right] + w \nabla_s \left( \pdiff{z}{t} \right)_s \nonumber \\
-& \qquad \qquad  = \left( \pdiff{s}{z} \right) \left( \pdiff{z}{t} \right)_s \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\} + \nabla_s \left[ \left( \pdiff{z}{t} \right)_s w \right]
-\end{align} gives
-\begin{align}
-\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \nabla_s \left[ K - w \left( \pdiff{z}{t} \right)_s + \Phi \right] - \frac{1}{\rho} \nabla_s p - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s \\
-& \qquad + \left( \pdiff{s}{z} \right) \left( \pdiff{z}{t} \right)_s \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\}.
-\end{align}  Finally, we can expand the vorticity term and hence obtain
-\begin{align} \label{eq:CovariantUpdateEquation}
-\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \nabla_s \left[ K - w \left( \pdiff{z}{t} \right)_s + \Phi \right] - \frac{1}{\rho} \nabla_s p - \left[ 2 \vg{\Omega} \times \vb{u} \right]_s \nonumber \\
-& \qquad + \left[ \vb{k} \cdot \nabla_s \times \vb{v}_h \right] (\vb{u}_h \times \vb{k}) - \dot{s} \left\{ \pdiff{\vb{v}_h}{s} - \nabla_s \left[ \left( \pdiff{s}{z} \right)^{-1} w \right] \right\} - \left( \pdiff{z}{t} \right)_s \vb{u}_h \cdot \nabla \vb{k}.
-\end{align}
-
-\subsection{Vorticity-Divergence Form}
-
-The vorticity-divergence form of the dynamical equations in an arbitrary vertical coordinate predicts the absolute vorticity ($\zeta_h^\ast$) and velocity divergence ($D$) given by
-\begin{align} \label{eq:AbsoluteVorticityS}
-\zeta_h^\ast = (\nabla_s \times \vb{u}_h + 2 \vg{\Omega}) \cdot \vb{k},
-\end{align} and
-\begin{align} \label{eq:DivergenceSx}
-D \equiv \nabla_s \cdot \vb{u}_h,
-\end{align} respectively, instead of the horizontal velocity.  The horizontal velocity can be obtained from the streamfunction $\psi$ and the velocity potential $\chi$ following
-\begin{align} \label{eq:VelocityStreamfunctionPotentialRelationship}
-\vb{u}_h = \vb{k} \times \nabla_s \psi + \nabla_s \chi.
-\end{align}  By using (\ref{eq:VelocityStreamfunctionPotentialRelationship}) in (\ref{eq:AbsoluteVorticityS}) and (\ref{eq:DivergenceSx}), we obtain the elliptic equations that diagnose the streamfunction and velocity potential from the predicted velocity and divergence as
-\begin{align} \label{eq:LaplacianStreamfunctionPotential}
-\nabla_s^2 = \zeta_h^\ast - 2 \vg{\Omega} \cdot \vb{k}, \quad \mbox{and} \quad \nabla_s^2 \chi = D,
-\end{align} respectively.
-
-By taking the material derivative (\ref{eq:MaterialDerivativeS}) of (\ref{eq:AbsoluteVorticityS}) and using the horizontal momentum equation (\ref{eq:PrognosticUhEquationK2}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), the absolute vorticity prediction equation emerges,
-\begin{align}
-\left( \pdiff{\zeta_h^\ast}{t} \right)_s - J_s(\zeta_h^\ast, \psi) + \nabla_s \cdot (\zeta_h^\ast \nabla_s \chi) + \nabla_s \cdot \left( \dot{s} \pdiff{}{s} \nabla_s \psi \right) + \vb{k} \cdot \nabla_s \times \left( \dot{s} \pdiff{}{s} \nabla_s \chi \right) + J_s(\rho^{-1}, p) = 0,
-\end{align} where $J_s(a,b) = \vb{k} \cdot \nabla_s \times (a \nabla_s b)$ is the Jacobian operator.  It can also be shown that $\dot{s}$ relates to the vertical velocity $w$ through
-\begin{align}
-\dot{s} = \left( \pdiff{s}{z} \right) (w - w_c),
-\end{align} where
-\begin{align}
-w_c \equiv \left( \pdiff{z}{t} \right)_s + \left( \vb{k} \times \nabla_s \psi + \nabla_s \chi \right) \cdot (\nabla_s z).
-\end{align}
-
-By taking the material derivative of (\ref{eq:DivergenceSx}) and using (\ref{eq:PrognosticUhEquationK2}), (\ref{eq:VelocityStreamfunctionPotentialRelationship}) and (\ref{eq:LaplacianStreamfunctionPotential}), we can obtain the divergence prediction equation
-\begin{align}
-\left( \pdiff{D}{t} \right)_s - J_s(\zeta_h^\ast, \chi) - \nabla_s \cdot (\zeta_h^\ast \nabla_s \psi) + \nabla_s \cdot \left( \dot{s} \pdiff{}{s} \nabla_s \chi \right) + \left( \vb{k} \times \pdiff{}{s} \nabla_s \psi \right) \cdot \nabla_s \dot{s} & \\
-+ \nabla_s \cdot (\nabla_s K_2 + g \nabla_s z) + \nabla_s \cdot \left( \frac{1}{\rho} \nabla_s p \right) &= 0,
-\end{align}  where $K_2$ can be reformulated in terms of streamfunction and velocity potential as
-\begin{align}
-K_2 = \frac{1}{2} \left[ \nabla_s \cdot (\psi \nabla_s \psi) - \psi \nabla_s^2 \psi + \nabla_s \cdot (\chi \nabla_s \chi) - \chi \nabla_s^2 \chi \right] + J_s(\psi, \chi).
-\end{align}
-
-\subsection{Momentum Form}
-
-The momentum form of the prognostic equations emerges by combining the prognostic velocity equations with a continuity equation.
-Essentially any of the continuity equations can be chosen, as long as the mass field represented by the equation is everywhere non-zero.
-However, the most common options are moist pseudo-density \citep{ullrich2012mcore} or dry pseudo-density \citep{skamarock2012multiscale}.
-Here we denote our density variable by $\tilde{\rho}_s$, and assume no external sources or sinks of $\tilde{\rho}$.
-Multiplying (\ref{eq:PrognosticWEquationK2Sb}) through by $\tilde{\rho}_s$ and using (\ref{eq:PrognosticPseudoDensity}) gives
-\begin{align} \label{eq:PrognosticRhoWEquationK2Sb}
-\left( \pdiff{\tilde{\rho}_s w}{t} \right)_s =& \tilde{\rho}_s \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \nabla_s \cdot (\tilde{\rho}_s \vb{u}_h w) - \pdiff{}{s} (\tilde{\rho}_s \dot{s} w) + \tilde{\rho}_s \left( \pdiff{s}{z} \right) \left[ - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right] - (2 \vg{\Omega} \times \tilde{\rho}_s \vb{u}_h) \cdot \vb{k}.
-\end{align}  Similarly, from (\ref{eq:PrognosticUhEquationK2Sb}) we have
-\begin{align} \label{eq:PrognosticRhoUhEquationK2Sb}
-\left( \pdiff{\tilde{\rho}_s \vb{u}_h}{t} \right)_s =& - \tilde{\rho}_s w (\vb{u}_h \cdot \nabla \vb{k}) - \tilde{\rho}_s \nabla_s K_2 - \zeta_s \vb{k} \times \tilde{\rho}_s \vb{u}_h - \vb{u}_h \nabla_s \cdot (\tilde{\rho}_s \vb{u}_h) - \pdiff{}{s} (\tilde{\rho}_s \dot{s} \vb{u}_h) - \tilde{\rho}_s \left( \nabla_z \Phi + \frac{1}{\rho} \nabla_z p \right) - [2 \vg{\Omega} \times \tilde{\rho}_s \vb{u}]_z,
-\end{align}
-
-%\appendix
-%\section{}    %% Appendix A
-
-%\subsection{}                               %% Appendix A1, A2, etc.
-
-\authorcontribution{Text in this manuscript describing individual models was provided by the respective modeling teams.  Final composition and development of Appendix A was performed by P.A. Ullrich.}
-
-\begin{acknowledgements}
-DCMIP2016 is sponsored by the National Center for Atmospheric Research Computational Information Systems Laboratory, the Department of Energy Office of Science (award no.
-DE-SC0016015), the National Science Foundation (award no.
-1629819), the National Aeronautics and Space Administration (award no.
-NNX16AK51G), the National Oceanic and Atmospheric Administration (award no.
-NA12OAR4320071), the Office of Naval Research and CU Boulder Research Computing.
-This work was made possible with support from our student and postdoctoral participants: Sabina Abba Omar, Scott Bachman, Amanda Back, Tobias Bauer, Vinicius Capistrano, Spencer Clark, Ross Dixon, Christopher Eldred, Robert Fajber, Jared Ferguson, Emily Foshee, Ariane Frassoni, Alexander Goldstein, Jorge Guerra, Chasity Henson, Adam Herrington, Tsung-Lin Hsieh, Dave Lee, Theodore Letcher, Weiwei Li, Laura Mazzaro, Maximo Menchaca, Jonathan Meyer, Farshid Nazari, John O'Brien, Bjarke Tobias Olsen, Hossein Parishani, Charles Pelletier, Thomas Rackow, Kabir Rasouli, Cameron Rencurrel, Koichi Sakaguchi, G\"okhan Sever, James Shaw, Konrad Simon, Abhishekh Srivastava, Nicholas Szapiro, Kazushi Takemura, Pushp Raj Tiwari, Chii-Yun Tsai, Richard Urata, Karin van der Wiel, Lei Wang, Eric Wolf, Zheng Wu, Haiyang Yu, Sungduk Yu and Jiawei Zhuang.
-We would also like to thank Rich Loft, Cecilia Banner, Kathryn Peczkowicz and Rory Kelly (NCAR), Perla Dinger, Carmen Ho, and Gina Skyberg (UC Davis) and Kristi Hansen (University of Michigan) for administrative support during the workshop and summer school.
-\end{acknowledgements}
-
-%% REFERENCES
-
-%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
-%% authors experienced with BibTeX only have to include the following two lines:
-%%
-\bibliographystyle{copernicus}
-\bibliography{DCMIP2016-Part1.bib}
-%%
-%% URLs and DOIs can be entered in your BibTeX file as:
-%%
-%% URL = {http://www.xyz.org/~jones/idx_g.htm}
-%% DOI = {10.5194/xyz}
-
-%% LITERATURE CITATIONS
-%%
-%% command                        & example result
-%% \citet{jones90}|               & Jones et al. (1990)
-%% \citep{jones90}|               & (Jones et al., 1990)
-%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
-%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
-%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
-%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
-%% \citeauthor{jones90}|          & Jones et al.
-%% \citeyear{jones90}|            & 1990
-
-%% FIGURES
-
-%% ONE-COLUMN FIGURES
-
-%%f
-%\begin{figure}[t]
-%\includegraphics[width=8.3cm]{FILE NAME}
-%\caption{TEXT}
-%\end{figure}
-%
-%%% TWO-COLUMN FIGURES
-%
-%%f
-%\begin{figure*}[t]
-%\includegraphics[width=12cm]{FILE NAME}
-%\caption{TEXT}
-%\end{figure*}
-%
-%
-%%% TABLES
-%%%
-%%% The different columns must be seperated with a & command and should
-%%% end with \\ to identify the column brake.
-%
-%%% ONE-COLUMN TABLE
-%
-%%t
-%\begin{table}[t]
-%\caption{TEXT}
-%\begin{tabular}{column = lcr}
-%\tophline
-%
-%\middlehline
-%
-%\bottomhline
-%\end{tabular}
-%\belowtable{} % Table Footnotes
-%\end{table}
-%
-%%% TWO-COLUMN TABLE
-%
-%%t
-%\begin{table*}[t]
-%\caption{TEXT}
-%\begin{tabular}{column = lcr}
-%\tophline
-%
-%\middlehline
-%
-%\bottomhline
-%\end{tabular}
-%\belowtable{} % Table Footnotes
-%\end{table*}
-%
-%
-%%% NUMBERING OF FIGURES AND TABLES
-%%%
-%%% If figures and tables must be numbered 1a, 1b, etc. the following command
-%%% should be inserted before the begin{} command.
-%
-%\addtocounter{figure}{-1}\renewcommand{\thefigure}{\arabic{figure}a}
-%
-%
-%%% MATHEMATICAL EXPRESSIONS
-%
-%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
-%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
-%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
-%%%
-%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
-%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
-%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
-%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
-%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
-%%% Vectors are identified in bold italic font using \vec{x}
-%%% Matrices are identified in bold roman font
-%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
-%%% The character * should not be applied as mutliplication sign
-%
-%
-%%% EQUATIONS
-%
-%%% Single-row equation
-%
-%\begin{equation}
-%
-%\end{equation}
-%
-%%% Multiline equation
-%
-%\begin{align}
-%& 3 + 5 = 8\\
-%& 3 + 5 = 8\\
-%& 3 + 5 = 8
-%\end{align}
-%
-%
-%%% MATRICES
-%
-%\begin{matrix}
-%x & y & z\\
-%x & y & z\\
-%x & y & z\\
-%\end{matrix}
-%
-%
-%%% ALGORITHM
-%
-%\begin{algorithm}
-%\caption{…}
-%\label{a1}
-%\begin{algorithmic}
-%…
-%\end{algorithmic}
-%\end{algorithm}
-%
-%
-%%% CHEMICAL FORMULAS AND REACTIONS
-%
-%%% For formulas embedded in the text, please use \chem{}
-%
-%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
-%
-%\begin{reaction}
-%%% \rightarrow should be used for normal (one-way) chemical reactions
-%%% \rightleftharpoons should be used for equilibria
-%%% \leftrightarrow should be used for resonance structures
-%\end{reaction}
-%
-%
-%%% PHYSICAL UNITS
-%%%
-%%% Please use \unit{} and apply the exponential notation
+This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
 

### 2017-08-13 DCMIP2016-Part1-EditorResponse.tex
Updates based on returned reviews

--- 2017-08-13 DCMIP2016-Part1-Abstract.tex
@@
 
-\noindent Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.
-These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.
-In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.
-To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
-This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
+\noindent \textbf{Response to Editor} \\
+
+\noindent A throughly revised code availability table has been included with the latest version of the paper.
+Six of the modeling groups have uploaded model source code to Zenodo.
+Licensing agreements prevent the remainder from directly posting model code publicly, however nearly all provide an avenue for public access to the codebase.
+Only two of the model groups (ACME-A and FVM) cannot allow public access at this time, but plan for a public release of the code in the future.
+In the mean time, these groups will allow the topical editor (Julia Hargreaves) access the code.
+The following page contains the updated code availability table.
 

### 2017-08-13 DCMIP2016-Part1-Models-20170604.tex
Updates based on returned reviews

--- 2017-05-04 DCMIP2016-Part1-Models.tex
@@
 \documentclass[gmd, manuscript]{copernicus}
+
+ \usepackage{xcolor} \usepackage{ulem} \usepackage{changebar}
+ \newcommand\TLSins[1]{\cbstart{}\textcolor{ins}{\uline{#1}}\cbend{}}
+ \newcommand\TLSdel[1]{\cbdelete{}\textcolor{del}{\sout{#1}}}
 
@@
 \end{align}  Here $\nabla_\zeta$ denotes the horizontal gradient along $\zeta$ surfaces.
+With respect to the treatment of moisture in GEM, the cloud water and all non-gases are embedded in the total air density $\rho$,
+affecting the virtual temperature defined in (\ref{eq:VirtualTemperatureExact}). Also, specific mass is used in GEM (not mixing ratio).
 
@@
 \subsection*{Code availability}
+
+Information on the availability of source code for the models featured in this paper is tabulated below.
+
+\begin{center}
+\begin{tabular}{cp{5in}}
+\hline Short Name & Code availability \\ \hline 
+ACME--A & ACME, including ACME--A, is under active development funded by the U.S.
+Department of Energy.
+ACME version 1.0 is scheduled to be publicly released under an open source license in 2018, but is not available at present.$^\dagger$ \\
+CSU & CSU model source code is available under the BSD 3-clause license.  The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.580099}}). \\
+DYNAMICO & DYNAMICO is open source and available online from IPSL Forge (\mbox{\url{http://forge.ipsl.jussieu.fr/dynamico}}) or directly by request to Thomas Dubos (\mbox{dubos@lmd.polytechnique.fr}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583718}}).
+\\
+FV$^3$ & FV$^3$ model source code is available through the GFDL Virtual Lab (\url{https://vlab.ncep.noaa.gov/web/fv3gfs}).  Access requires users to create a Virtual Lab account. \\
+FVM & Model codes developed at ECMWF, including the IFS and FVM, are intellectual 
+property of ECMWF and its member states.
+Although the FVM code is not publicly available at present, it is expected that FVM will be available in the near future under the OpenIFS license (\url{http://www.ecmwf.int/en/research/projects/openifs}).
+The repo tag for the version of FVM that applies for DCMIP is ``v0.1''.$^\dagger$ \\
+GEM & Due to licensing requirements, GEM model code is only available by request to Abdessamad Qaddouri (Abdessamad.Qaddouri@canada.ca) or Vivian Lee (Vivian.Lee2@canada.ca). \\
+ICON & ICON is freely available to the scientific community for non-commercial research under an institutional license issued by project partners DWD+MPI-M.
+Because of the restrictions of this license, access to the code is only available by request to G\"unther Z\"{a}ngl (\mbox{Guenther.zaengl@dwd.de}) or Marco Giorgetta (\mbox{marco.giorgetta@mpimet.mpg.de}).
+\\
+\hline \multicolumn{2}{p{6in}}{$\dagger$ \footnotesize In compliance with the GMD editorial requirements, this code has been made available to the topical editor in charge of the manuscript.}
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{cp{5in}}
+\hline Short Name & Code availability (cont'd) \\ \hline 
+MPAS & Open-source via GitHub (\url{https://github.com/MPAS-Dev/MPAS-Release}).  The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583316}}) \\
+NICAM & NICAM source code is available under the BSD 2-clause license via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.580128}}).
+Further information on collaborating with the NICAM team can be found at \mbox{\url{http://nicam.jp/hiki/?Research+Collaborations}}.
+\\
+OLAM & OLAM is open source and available online via SourceForge (\url{https://sourceforge.net/projects/olam-model/}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://doi.org/10.5281/zenodo.582308}}).
+\\
+TEMPEST & Tempest source code is available under the Lesser GNU Public License on GitHub (\mbox{\url{https://github.com/paullric/tempestmodel}}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.579649}}).
+\\
+\hline 
+\end{tabular}
+\end{center}
+
+\clearpage
 

### 2017-10-26 DCMIP2016-Part1-ReviewerResponse1.tex
- Near final production update

\documentclass{article}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{color}
\usepackage{comment}

\oddsidemargin 0cm
\evensidemargin 0cm

\textwidth 16.5cm
\topmargin -2.0cm
\parindent 0cm
\textheight 24cm
\parskip 0.5cm

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\fancyhead[L]{AOSS Reference Sheet}
%\fancyhead[CH]{test}
\fancyfoot[C]{Page \thepage}

\newcommand{\vb}{\mathbf}
\newcommand{\vg}{\boldsymbol}
\newcommand{\mat}{\mathsf}
\newcommand{\diff}[2]{\frac{d #1}{d #2}}
\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
\newcommand{\topic}{\textbf}
\newcommand{\arccot}{\mathrm{arccot}}
\newcommand{\arcsinh}{\mathrm{arcsinh}}
\newcommand{\arccosh}{\mathrm{arccosh}}
\newcommand{\arctanh}{\mathrm{arctanh}}

\begin{document}

\textbf{{This is an exceptionally well written paper leading to the most minor review that I have
every written. The paper describes the 11 models which took part in DCMIP 2016 with
clarity, consistency and the kind of insight that very few people have. This will be an
extremely useful resource for those seeking to understand how any of the models work
and how they compare to other models. I have a very few minor comments.}}

We would like to thank the reviewer for this very kind comment.  We are very hopeful that this manuscript will be a valuable resource for future model developers and students of model development.

\textbf{{1. OLAM uses cut cells to represent smooth topography - how is the small cell problem solved.}}

To address this question, the following text has been added to section 4.2.4:

{\color{blue}One or more methods are used to avoid the so-called small cell problem where volume to area ratios of cut cells are much less than for full cells and therefore can lead to instability. The smallest cells are eliminated by adjusting topography slightly, which is usually justified by noting that local topographic sampling is approximate. In larger cut cells, volumes can be increased (without changing surface areas) which stabilizes the cell at the expense of slowing its response to advected transients. When either of the above adjustments is unacceptable for a particular application, a flux-balance method based partly on Berger and Helzel (2012) is used to stabilize small cut cells.}

\textbf{{2. Perhaps mention the advantages of the lat-lon grid.}}

Agreed.  We have added the following text to the description of the latitude-longitude grid:

{\color{blue}The latitude-longitude grid has the benefits of being globally rectilinear, which simplifies data access and subdivision of computation across processors, and yields a vector basis that is locally orthogonal nearly everywhere.  This structure accurately maintains purely zonal flows and simplifies data post-processing for visualization.}

\textbf{{3. In section 6.5, give a citation describing the hexagonal C-grid computational mode
and either give a citation or describe the filter.}}

Agreed.  In the revised manuscript we have included citations to Weller, H.: Controlling the computational modes of the arbitrarily structured C grid, Mon. Weather Rev., 140, 3220--3234, 2012 and Weller, H., Thuburn, J., and Cotter, C. J.: Computational modes and grid imprinting on five quasi-uniform spherical C grids, Mon. Weather Rev., 140, 2734--2755, 2012.

\textbf{{4. In section 7, you say that fully compressible non-hydrostatic models need a temporal
discretisation for dealing with vertically propagating sound waves. You give the
impression that using some form of approximation that filters sound waves implies that
the problem goes away. It doesn't, it makes the problem elliptic rather than hyperbolic
and so requires the solution of a Poisson equation rather than a Helmholtz equation.
It is a common misconception that these approximations somehow make solving the
equations easier. Please help to dispel this misconception.}}

Thank you for pointing this out.  The text has been modified to read:

{\color{blue}This diversity is in part because of the demands of non-hydrostatic models: unlike their hydrostatic counterparts, non-hydrostatic atmospheric models must include a mechanism for dealing with vertically propagating sound waves.  These waves are meteorologically insignificant, but with a vertical grid spacing of 100 meters, a purely explicit temporal discretization of the unmodified fluid equations would require a time step size on the order of one second or less.  Consequently, sound waves are either filtered explicitly through the use of an alternative equation set, or artificially slowed through the use of implicit temporal discretizations.  Some commonly employed alternative equation sets include the anelastic (Ogura and Phillips, 1962), quasi-hydrostatic (Orlanski, 1981), pseudo-incompressible (Durran, 1989), or unified approximation (Arakawa and Konor, 2009).  These filtered equation sets generally require that a global elliptic solve be performed as prognostic variables are updated.}

\textbf{{5. Line 29 of page 28. Remove the word ``basically''.}}

Removed.

\end{document}

### 2017-10-26 DCMIP2016-Part1-ReviewerResponse2.tex
- Near final production update

\documentclass{article}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{color}
\usepackage{comment}

\oddsidemargin 0cm
\evensidemargin 0cm

\textwidth 16.5cm
\topmargin -2.0cm
\parindent 0cm
\textheight 24cm
\parskip 0.5cm

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\fancyhead[L]{AOSS Reference Sheet}
%\fancyhead[CH]{test}
\fancyfoot[C]{Page \thepage}

\newcommand{\vb}{\mathbf}
\newcommand{\vg}{\boldsymbol}
\newcommand{\mat}{\mathsf}
\newcommand{\diff}[2]{\frac{d #1}{d #2}}
\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
\newcommand{\topic}{\textbf}
\newcommand{\arccot}{\mathrm{arccot}}
\newcommand{\arcsinh}{\mathrm{arcsinh}}
\newcommand{\arccosh}{\mathrm{arccosh}}
\newcommand{\arctanh}{\mathrm{arctanh}}

\begin{document}

\textbf{{This paper gives an overview of the models that participated in the DCMIP2016 workshop
on dynamical core intercomparisons. By itself the paper provides a useful reference
on the current state of the art in regards to dynamical core development, highlighting
the wide range of choices that have been made by modeling groups across
the globe as well as highlighting some of the choices, such as equation sets, used in
dynamical core design. As noted by the authors this paper is the first of an envisioned
sequence detailing the models and their performance on a number of idealized test
cases and it will be interesting to observe how the sequence develops and what can
be learned from the intercomparison.}}

\textbf{{It is some achievement to condense the wealth of information needed to describe a
range of modeling issues into one concise paper and the authors should be applauded
for succeeding in such a difficult task. The paper is well written and provides a useful
source of information to model developers and is therefore suitable for publication after
a number of minor issues are dealt with.}}

We thank the reviewer for his positive comments on the manuscript.  We agree that, in the interests of completeness, each section should feature descriptions from all models being assessed and so have added text to this effect to each of sections 3-7.  Further, the model descriptions in section 2 have been rewritten to maintain a more consistent structure.

\textbf{{Main comments:}}

\textbf{{1. The main issue that should be corrected in this paper is that not all of the models
are covered in each of sections 3-7 describing aspects of the model formulations.
This could be due a desire not to replicate information (if models share the same
governing equations etc) but I think it would be useful if the reader could find the
appropriate model description in each of sections 3-7. In detail I suggest:}}

We agree with your assessment and have added the missing sections, as requested.

\textbf{{Section 5 lists the equation sets used by each model but is missing the
CSU, MPAS and NICAM models. It would be useful for the reader to add
brief sections for these models, or if they are the same as some of the other
models to combine them into the appropriate subsection.}}

Text for CSU, MPAS and NICAM has been added to this section:

\textbf{CSU:}  {\color{blue}The CSU model uses the vorticity-divergence form of the equations of motion, as described in section A10, discretized on the geodesic mesh with absolute vorticity and velocity divergence scalars stored at cell-centers.  The unified approximation of the equations of motion (Arakawa and Konor, 2009) is employed to avoid vertically propagating sound waves.}

\textbf{MPAS:}  {\color{blue}The evolution equations used by MPAS are fully described in Skamarock et al. (2012), based on the formulation of Dutton (1986). The MPAS model uses the momentum form of the update equations, as described in section A11, with dry mass utilized for the density variable $\tilde{\rho}_s$. MPAS further evolves dry mass using a continuity equation of the form (A10) and moist potential temperature following (A13).}

\textbf{NICAM:}  {\color{blue}NICAM prognoses horizontal and vertical momentum analogous to the approach described in section A11. It further evolves the density perturbation from the background reference state using (A10) and sensible heat part of internal energy. A detailed explanation of the evolution equations can be found in Satoh et al. (2008).}

\textbf{{Section 6 describes the diffusion mechanisms in each model but omits the
CSU, DYNAMICO, FVM, MPAS and NICAM models. Since table 5 indicates
these models to have explicit diffusion mechanisms then it would be good
to add subsections for the missing models, or where appropriate combine
them, e.g. CSU and DYNAMICO both use 4th order hyperviscosity which
is covered in subsection 6.1 on ACME-A and so these models could be
combined into a single section.}}

The following text has been added in section 6.

\textbf{CSU:}  {\color{blue}The CSU model utilizes an explicit diffusion scheme that consists of fourth-order hyperdiffusion ($\nabla^4$) applied to the vorticity, divergence, and potential temperature. The model does not include any explicit diffusion in the vertical column. However, for the idealized DCMIP test cases explicit diffusion was disabled.}

\textbf{DYNAMICO:}  {\color{blue}In DYNAMICO, (hyper-)diffusive filters are used to eliminate spurious noise due to the energy-conservative centered discretization. Filters are applied every $N_{\mbox{\textit{diff}}}$ Runge-Kutta time steps in a forward-Euler manner, with $N_{\mbox{\textit{diff}}}$ as large as allowed by stability. The scalar Laplacian is computed as div grad and the vector Laplacian is decomposed into its divergent (grad div) and rotational (curl curl) parts. The strength of filtering is controlled by dissipation time scales $\tau$:  Given $\tau$ the hyperviscous coefficient that multiplies operator $D^p$ is  $\delta^{2p} \tau^-1$ where $\delta^-2$ is the largest eigenvalue of operator $D$. For DCMIP, DYNAMICO uses $p=2$ (fourth-order hyperviscosity) for all filters.}

\textbf{FVM:}  {\color{blue}Within the dynamical core, FVM does not apply any explicit diffusion.  For the DCMIP test cases, the implicit diffusion of the monotonic MPDATA provides the right amount of diffusion/dissipation needed to maintain model stability and remove excess energy from the finest scales.  An absorbing layer is also available in the model for damping vertically propagating waves near the model top.}

\textbf{MPAS:}  {\color{blue}The MPAS model applies fourth-order hyperdiffusion and Smagorinsky diffusion (Smagorinsky, 1963), as described in Skamarock et al. (2012).  When applied to the momentum, the Laplacian is evaluated as
\begin{align}
\nabla^2 u_i = \pdiff{}{x_i} \nabla_s \cdot \vb{v} - \pdiff{\eta}{x_j},
\end{align} where $u_i$ is the edge-normal velocity defined on cell edge $i$, $\eta$ is the vertical component of the relative vorticity, computed on vertices, and $\nabla_s \cdot \vb{v}$ is the horizontal divergence on $s$ surfaces, computed on edges.  The evaluation of divergence and vorticity in this expression is described in \cite{ringler2010unified}.  The fourth-order hyperdiffusion operator is then computed by twice applying the above Laplacian operator to the momentum.

Smagorinsky diffusion, which is often applied in atmospheric models to parameterize turbulent processes, uses a second-order Laplacian and a physically-motivated eddy viscosity $K_h$, defined in terms of Cartesian velocities $(u,v)$,
\begin{align}
K_h = c_s^2 \ell^2 \sqrt{(u_x - v_y)^2 + (u_y + v_x)^2},
\end{align} where $c_s$ is a constant parameter and $\ell$ is the grid scale.  The diffusion operator then takes the form $\nabla \cdot (K_h \nabla \psi)$ for a scalar field $\psi$.}

\textbf{NICAM:}  {\color{blue}NICAM implements three types of diffusion: 3D divergence damping, fourth-order horizontal hyperdiffusion, and sixth-order vertical hyperdiffusion as described in Tomita and Satoh (2004). Specifically, the divergence damping term (Skamarock and Klemp, 1992) aims to suppress instabilities that arise due to the time splitting scheme, and is applied to both horizontal and vertical velocities. The hyperdiffusion operators are applied to all prognostic variables. For tracer advection, upwinding is used to remove spurious oscillations, as described in Miura (2007); Niwa et al. (2011).}

\textbf{{Related to the previous point the methods of diffusion \& stabilization in table
5 and section 6 are somewhat different, for example some model subsections describe
using sponge layers (FV3, ICON) but these are not listed in
table 5 and the same applies to monotonic limiting for some models. Is it
the case that table 5 only lists the principle methods of stabilization and diffusion?
In which case I suggest adding words to this effect in the caption.
I appreciate that it is beyond the scope of this paper to list in detail all the
methods of stabilization and diffusion applied in all the models, maybe some
words in the introduction to section 6 indicating that this section only covers
the principle diffusion methods used?}}

The header of Table 5 has been changed to read ``Principal options for diffusion, stabilization, filters or fixers.''  Further, the following sentence has been added to the introduction of section 6:

{\color{blue}Diffusion also includes mechanisms for damping vertically propagating internal gravity waves, such as model-top Rayleigh layers, which are fairly ubiquitous across models and hence not discussed in detail here.}

\textbf{{Section 6 lists the temporal discretization methods used but omits the methods
used by the ACME-A, CSU and NICAM models, it would be useful it the
methods used by these models was indicated in this section.}}

ACME-A uses the same method as Tempest, so these have been combined in the text.  NICAM uses the same method as MPAS, so these have also been combined in the text.

The following text has been added for the CSU model:

{\color{blue}CSU uses a semi-implicit time integration scheme with third-order Adams-Bashforth scheme for explicit integration of the continuity equation, potential temperature equation, and terms related to advection.  Since potential temperature is updated prior to the computation of the pressure-gradient force, this term can be thought of as implicit in time.  The horizontal wind field is then predicted through integration of the vorticity and divergence of the horizontal wind and a multigrid method applied to solve a pair of two-dimensional Poisson equations for the stream function and velocity potential, which are then differentiated to obtain the velocity field.  Horizontal diffusion is then applied forward in time.}

\textbf{{2. Section 2 gives a brief description of each model and as noted in the author contributions
these are provided by the modeling teams themselves, however this
has led to a rather uneven section where the model descriptions provide differing
levels of detail. I think this section could use some editorial input to unify the
descriptions. Based upon the sections for the rest of the paper I would like to be
able to ascertain the following properties for each model from this section: equation
set, horizontal grid and discretization, vertical grid and discretization, temporal
discretization, principal diffusion and stabilization mechanisms and transport
scheme. Only a couple of words to a sentence are needed and much of this
information can be found in tables 2-5 but i think it would help readability to unify
this description section.}}

Thank you for this suggestion.  Section 2 has been modified heavily in order to better align the wording for each model.  The changes are sufficiently extensive that they are not reproduced here.

\textbf{{3. The paper does a very good job of describing the key features of a wide range of
models, however I would have been interested in seeing a specific section detailing
the transport schemes used by the models in a similar fashion to Sections 3-7
(and including the information in table 2 if possible). However in order to avoid
over lengthening the paper I suggest this could be covered to some extent by
the descriptions in section 2. This would require details of the transport schemes
used by ACME-A, OLAM and TEMPEST to be mentioned in the appropriate subsections
of section 2.}}

The following updates to the text have been made:

\textbf{ACME-A:}  {\color{blue}By default, tracer transport is sub-cycled relative to the hydrodynamics, but also performed using the spectral element method using the tracer mass as a prognostic variable.}

\textbf{OLAM:}  {\color{blue}Tracer transport is second-order in space and time using the scheme of Miura (2007), with consistent fluxes obtained by time averaging over the acoustic time steps.}

\textbf{TEMPEST:}  {\color{blue}Tracer transport is performed using the spectral element method with the same timestep as the hydrodynamics and using the tracer mass density as a prognostic variable.  As with the hydrodynamics, tracer transport is updated explicitly in the horizontal and implicitly in the vertical.}

\textbf{{Minor comments:}}

\textbf{{Table 1: $\Phi$, $\delta \Phi$, $\dot{\zeta}$ and $\theta^\prime$ are missing entries but listed in the prognostic variables of table 3.}}

Thank you for catching this.  Entries have been added for $\Phi$ and $\dot{\zeta}$ in Table 1.  In Table 3, $\delta \Phi$ has been changed to $\Phi$ and $\theta^\prime$ has been changed to $\theta$.

\textbf{{The DCMIP2016 website lists HOMME, UZIM and NEPTUNE (NEP) as models
taking part, I assume that HOMME is ACME-A, UZIM is CSU, if I'm mistaken then
could these models be added? Is there a reason NEPTUNE is not included in
this paper?}}

HOMME was used at DCMIP2016 using a hydrostatic formulation.  This model has since been updated to include a nonhydrostatic formulation that was subsequently included in ACME-A.  Since it would be odd to include a model that was purely hydrostatic in a paper that focused on nonhydrostatic models, it was decided that HOMME would be substituted for ACME-A.

You are correct that UZIM and CSU are the same model (although a new curl-curl form model is now under development at CSU).

NEPTUNE participated in the DCMIP2016 workshop but did not submit text for this paper describing their model, and so are not included.  Since they did invest the considerable effort in participating in the workshop, we believed it would be appropriate to retain them as co-authors even though the model description was absent.

\textbf{{Section 2.3: If Dubos and Dubery has been submitted this reference could be
updated.}}

Although there is not much remaining to be done, this paper has not been submitted as of the time of writing.

\textbf{{Section 2.7 ``Icosahedral'' should be ``ICOsahedral'' in the subsection title to match
the format of other model names.}}

Corrected.

\textbf{{Section 3.5, last line: Is it possible that the CCVT method produces polygons
with less than 5 sides? If so this should be mentioned.}}

Yes, although it's very unlikely.  The text has been updated as follows:

{\color{blue}Although hexagons are, by far, the most common polygon on CCVT grids, CCVT grids on the sphere will also include at least 12 pentagons and sometimes other polygons with more than six sides.  Quadrilateral elements are theoretically possible, but are never found in practice on the final grid due to being a locally unstable solution of the underlying CCVT system of equations.}

\textbf{{Section 3.7 last line. I don?t think it is entirely correct that GEM uses two regional
climate models on the patches of the YinYang grid. Qaddouri 2011 States that the
numerics come from the original GEM latlong model which is used for medium
range weather forecasts. I suggest changing this to ``utilizing a pair of local area
models based with the numerics from the GEM latitude-longitude model''. If the
GEM modeling team feel the current description is accurate then I am happy for
it to be left as is.}}

We agree with the suggested change.  The modified text has been incorporated in the manuscript.

\textbf{{Page 16, Line 8. The A-grid collocates all scalar and velocity components. To
avoid confusion with the B- and E-grid (which only collocate velocity components)
I suggest changing ``co-location of all velocity components'' to ``co-location of all
velocity components and scalar fields''.}}

Agreed and changed accordingly.

\textbf{{Page 16 Line 9: To be consistent with the descriptions of the other grids I would
add ``which co-locates the vorticity, divergence and buoyancy variable.'' after ``and
the Z-grid''.}}

Agreed and changed accordingly.

\textbf{{Page 16 Line 14: There is a mix up of dimensionality of the mesh objects here,
for a 3D mesh the C-grid stores velocities on faces not edges. I would suggest
saying ``as long as the number of horizontal faces is twice the number of volumes''.}}

Agreed and changed accordingly.

\textbf{{Page 16 Lines 1 and 14-15: The maximum stable timestep size (if it exists) is
given by a combination of factors such as the time scheme, horizontal and vertical
discretization, grid staggering and waves supported in the equation set. The
comments in this section give the reader the impression that staggering is the
most important (or only) factor. I suggest that ``for explicit timestepping schemes''
is added after ``timestep size'' on line 1 and that the text on lines 14-15 from ``but
also'' to the semi-colon is removed since I believe this statement is only true for
a given choice of horizontal discretization (2nd order fd?) and defined explicit
timestepping schemes.}}

Agreed.  Line 1 has been changed accordingly.  The text on lines 14-15 has been changed to

{\color{blue}...the C-grid better represents short wave modes, does not support extraneous computational modes (as long as the number of horizontal faces is equal to twice the number of volumes), but typically has a more restrictive timestep with explicit timestepping schemes than the A-grid...}

\textbf{{Page 16 Line 18: In general I think it is a Poisson problem that needs to be solved
for the z-grid rather than the more general Helmholtz problem.}}

Agreed.  ``Helmholtz'' has been replaced with ``Poisson''.

\textbf{{Page 28 Line 14: Could a citation (at least title and authors) be given for this
paper if it is under review.}}

A citation has been added.

\textbf{{Typos}}

Thank you for catching these.  All have been corrected

\end{document}

### 2017-10-26 DCMIP2016-Part1-ReviewerResponse3.tex
- Near final production update

\documentclass{article}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{color}
\usepackage{comment}

\oddsidemargin 0cm
\evensidemargin 0cm

\textwidth 16.5cm
\topmargin -2.0cm
\parindent 0cm
\textheight 24cm
\parskip 0.5cm

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\fancyhead[L]{AOSS Reference Sheet}
%\fancyhead[CH]{test}
\fancyfoot[C]{Page \thepage}

\newcommand{\vb}{\mathbf}
\newcommand{\vg}{\boldsymbol}
\newcommand{\mat}{\mathsf}
\newcommand{\diff}[2]{\frac{d #1}{d #2}}
\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
\newcommand{\topic}{\textbf}
\newcommand{\arccot}{\mathrm{arccot}}
\newcommand{\arcsinh}{\mathrm{arcsinh}}
\newcommand{\arccosh}{\mathrm{arccosh}}
\newcommand{\arctanh}{\mathrm{arctanh}}

\begin{document}

%\textbf{{I reopened the online peer review a few weeks ago as I had not yet gained access to
%all of the three not-publicly-accessible codes (ACME, GEM and FVM), and I was also
%hoping for input from the third reviewer who had agreed to look at the manuscript. I
%have now been given access to all three codes. This required some effort on the part
%of the authors, for which I am grateful. This manuscript is not in the model description
%paper type, and is much closer to being a review article, and I do not see reason to
%be critical over anything related to the code - I will just say that I enjoyed seeing how
%the different developers have approached the design of their modelling infrastructure.}}

\textbf{{Before submitting a revision, I would appreciate it if the lead author of the paper would
test the veracity of the links to (or other ways of accessing) the other models included
in the code accessibility table. In my experience there are often small issues that need
to be resolved to make sure the information is accurate and that the code is really
accessible to all.}}

As requested, we have ensured the veracity of the links to the model codes.

%\textbf{{I have received some comments from a third reviewer, who did not like the paper at
%all, but declined to give a full review. I will pass on a paraphrased version of their
%comments as, particularly in relation to the organisation of the paper, I do see where
%they are coming from. Please consider their complaints seriously, respond to each
%point, and revise the structure of the paper to make it more logically organised and
%readable. It is possible that this reviewer did not take the time to consider the nature
%of a GMD paper. I am encouraged in this thinking by the fact that one of the other
%reviewers is a former GMD editor, and they gave a very encouraging review. In that
%context I am still considering this to be a minor revision despite this negative review.}}

\textbf{{Paraphrased comments from anonymous 3rd reviewer.
The reviewer commented to the effect that. . .
The purpose of the paper is not clear.
There is nothing of interest or importance to the modelling community in the paper.
The paper is too long, has too many equations and all of it is already published in the
literature.}}

We disagree strongly with the reviewer regarding the utility of the manuscript.  The manuscript itself should be viewed in the light of reviewing existing material from the wealth of (often unavailable) non-hydrostatic modeling technical reports and peer-reviewed publications, rather than introducing substantially new concepts.  The structure of this paper provides us with a mechanism to compare the different design decisions that have been made in these models, and potentially as a resource for future students of non-hydrostatic modeling systems to learn about the types of decisions that need to be made.  In the view of the authors, the content of our manuscript is simply not easily accessible in the existing literature.

Regarding the length of the paper, we certainly acknowledge that this material could be the basis for a textbook on the subject, but believe that an open-access and peer-reviewed compilation provides broader access to this content than a textbook.

\textbf{{The material is presented in a highly confusing manner, ``for instance, Section 2 is subdivided
by model (with 11 sections, 1 per model), but Section 3 only has 8 subsections
(why not 11?), and Section 4 only 2 subsections, the second of which only has 4 subsubsection
(yet, when you look at Table 4, last column, I only see 3 different methods,
so I have no idea where there are 4 sub-subsections for 3 methods); then Section 5 is
again subsectioned by model (why this back and forth), but instead of 11 I only see 8 subsections: why are 3 models missing?''}}

\textbf{{[editor's note: GMD papers are allowed to be long and may contain many equations,
but because of this, an easy to navigate logical and readable structure is even more
important than it is for regular science papers.]}}

The issues with the structure have also been highlighted by the second reviewer.  In response to his specific criticisms, the revised manuscript now fills in all gaps within sections that are structured around the different options available to each model.  We note that each section is meant to highlight one of the potential decisions that need to be made in building a non-hydrostatic model.  If the potential options for that decision were limited, it made more sense to the authors to break up the section by the choices available.  If essentially all modeling groups pursued a different strategy, then it made more sense to break up each section by the model.  We hope the revisions have been sufficient to address this concern.

\end{document}

### 2017-10-26 DCMIP2016-Part1-Abstract.tex
- Near final production update

@@
 In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.
-To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
+To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven non-hydrostatic dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
 This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.

### 2017-10-26 DCMIP2016-Part1-Models-20171008.tex
- Near final production update

--- 2017-08-13 DCMIP2016-Part1-Models-20170604.tex
@@
 \documentclass[gmd, manuscript]{copernicus}
-
- \usepackage{xcolor} \usepackage{ulem} \usepackage{changebar}
- \newcommand\TLSins[1]{\cbstart{}\textcolor{ins}{\uline{#1}}\cbend{}}
- \newcommand\TLSdel[1]{\cbdelete{}\textcolor{del}{\sout{#1}}}
 
@@
 \Author[4]{Joseph}{Klemp}
-\Author[4]{Sang-Hun}{Park}
+\Author[14]{Sang-Hun}{Park}
 \Author[4]{William}{Skamarock}
 
-\Author[14]{Hiroaki}{Miura}
-\Author[14]{Tomoki}{Ohno}
-\Author[15]{Ryuji}{Yoshida}
-
-\Author[16]{Robert}{Walko}
-
-\Author[17]{Alex}{Reinecke}
-\Author[17]{Kevin}{Viner}
+\Author[15]{Hiroaki}{Miura}
+\Author[15]{Tomoki}{Ohno}
+\Author[16]{Ryuji}{Yoshida}
+
+\Author[17]{Robert}{Walko}
+
+\Author[18]{Alex}{Reinecke}
+\Author[18]{Kevin}{Viner}
 
@@
 \affil[13]{Deutscher Wetterdienst (DWD)}
-\affil[14]{University of Tokyo}
-\affil[15]{RIKEN}
-\affil[16]{University of Miami}
-\affil[17]{Naval Research Laboratory}
+\affil[14]{Yonsei University}
+\affil[15]{University of Tokyo}
+\affil[16]{RIKEN AICS / Kobe University}
+\affil[17]{University of Miami}
+\affil[18]{Naval Research Laboratory}
 
@@
 $p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
+$\Phi$ & Geopotential \\
 $\Phi_s$ & Surface geopotential \\
@@
 $w$ & Vertical wind velocity \\
+$\dot{\zeta}$ & GEM vertical coordinate velocity \\
 $\vb{u}$ & 3D wind vector \\
@@
 
-This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.
+This section provides a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.
 Further details on these choices can be found in subsequent sections.
@@
 The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.
-Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
+Principal options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.
 A brief description of each participant model follows, focused on the unique features and decisions underlying the model design.
@@
 CSU & Colorado State University Model & Colorado State University, USA \\
-DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
+DYNAMICO & DYNAMical core on the ICOsahedron & Institut Pierre Simon Laplace (IPSL), France \\
 FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
@@
 GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
-ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
+ICON & ICOsahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
 MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
@@
 DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
-FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
+FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
 FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_i$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
@@
 \begin{table}[p]
-\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
+\caption{Principal options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
 \begin{center}
 \begin{tabular}{cccccc}
-\hline Acronym & Standard options for diffusion, & Temporal discretization \\
+\hline Acronym & Principal options for diffusion, & Temporal discretization \\
 & stabilization, filters, or fixers & \\ \hline 
@@
 Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
-The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
+ACME-A is built upon an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
+The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates.
 Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
 All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.
+Tracer transport is sub-cycled relative to the hydrodynamics, using the spectral element method, with tracer mass as the prognostic variable.
 
@@
 
-The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate.
+The CSU model is a finite-volume model using an optimized geodesic grid \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate.
 The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
 The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
-Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.
+Horizontal diffusion is included in the form of a fourth-order hyperviscosity operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.
 The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
@@
 
-DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
-DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.
+DYNAMICO is a mimetic finite-difference / finite-volume model using a geodesic grid and a floating vertical mass coordinate.
+Although originally a hydrostatic model, it has been recently extended to solve the shallow-atmosphere non-hydrostatic Euler equations.
+DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}.
 It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
 It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
+Horizontal diffusion is included via a fourth-order hyperviscosity operator.
 In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
@@
 
-The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
+The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with a floating Lagrangian vertical coordinate.
 The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
+The non-hydrostatic formulation extends the hydrostatic model described in  \cite{lin2004vertically} by adding a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}), although the prognostic horizontal winds are stored in the native Gnomonic local coordinate.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.
 Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}).
-Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+Since divergence is effectively invisible to the solver, a 2D divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
 Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
@@
 
-The prognostic horizontal winds are stored in the native Gnomonic local coordinate.
-The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
-All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.
-
 \subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
@@
 The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
-ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
-compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
-A centered two-time-level semi-implicit 
+ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}.
+FVM solves the non-hydrostatic Euler equations on an octahedral reduced Gaussian grid with a height-based terrain-following vertical coordinate \citep{szmelter2010,smolarkiewiczetalJCP2016}.
+The horizontal spatial discretization uses the median-dual finite-volume approach, combined with a structured-grid finite-difference method in the vertical.
+In both the horizontal and
+the vertical discretization all variables are co-located. A centered two-time-level semi-implicit 
 integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
@@
 preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
-non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
-combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
-the vertical discretization all variables are co-located. 
-The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
-Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
+non-oscillatory finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  The non-oscillatory (i.e.~monotonic) MPDATA 
+also provides sufficient dissipation/diffusion to stabilize the model, so no other explicit filtering mechanism is required. 
+Note that the octahedral reduced Gaussian 
 grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
 interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
-broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
-%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.
-
-%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
-%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-%are available by means of simple switches. The formulation incorporates 
-%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
-%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
-%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-%Conjugate Residual approach. The dynamics time step is adapted 
-%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-%than 1). The physics time step is identical to the dynamics time step. 
-% 
-%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-%The horizontal spatial discretization is fully unstructured finite-volume using the 
-%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-%In both the horizontal and the vertical discretization, all prognostic variables are 
-%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
-%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
-%
-%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
-%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-%the form of a Rayleigh-type forcing to the prognostic variables.
-
-%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
-%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
-
-%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
-%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
-%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
-%efficiency and accuracy.
-
-%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
-%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
-%framework developed at ECMWF.
+broad classes of meshes \citep[][]{szmelter2010,kuehnleinJCP2012,deconinck2017atlas}.
 
@@
 The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.
-The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
+GEM solves the non-hydrostatic Euler equations on the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering of prognostic variables.
 The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
@@
 linear system of equations that is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
-The dynamics and physics are time split.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.  Semi-Lagrangian advection is also used for tracer transport.
 To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
@@
 
-\subsection{Icosahedral Non-hydrostatic model (ICON)}
-
-The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
-Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
-Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels.
-The discretization employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+\subsection{ICOsahedral Non-hydrostatic model (ICON)}
+
+The ICON model \citep{zangl2015icon} is a finite-volume model that solves the non-hydrostatic Euler equations in 2D vector invariant form on an icosahedral (triangular) Arakawa C-grid, utilizing a smoothed terrain-following height-based Lorenz vertical discretization.
+Prognostic horizontal velocities are stored as normal wind components at the edge mid-points of full levels.
+Prognostic vertical velocity is stored at the circumcenters of the triangles on half levels.
+The discretization employs a two time-level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
 For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
@@
 Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
+The horizontal diffusion, which is applied at full model time steps, combines a flow-dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator.
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which leads to local mass conservation and consistency with the air motion.
+Specifically, the average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+These numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
+
+%Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
 
@@
 
-The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
-Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
+The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} is a finite-volume model that solves the non-hydrostatic Euler equations using an Arakawa C-grid staggering on a centroidal Voronoi tessellation mesh (see section \ref{sec:grid_ccvt}), and the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
 In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
+Advection is nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
 The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
 Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
-Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
+Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, fourth-order hyperdiffusion, and 2D and 3D divergence damping operators.
 
@@
 
-NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}.
-The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}.
-Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}.
-All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids.
+NICAM is a finite-volume model that solves the non-hydrostatic Euler equations using an optimized geodesic grid optimized by the spring dynamics using the method proposed by \cite{tomita2002optimization}.
+A terrain-following height coordinate system is used in the vertical \citep{tomita2004new} with Lorenz staggering.
+Instead of temperature or potential temperature, total energy is prognosed following the method of \citep{satoh2002conservative, satoh2003conservative}.
+All prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell-integrated quantities to point values at cell centroids.
 The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}.
-The Arakawa C-grid staggering is used vertically.
-A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended.
-The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}.
-The implicit time discretization is adopted for the vertically propagating one.
+For integration in time, a two-stage Runge-Kutta scheme is usually employed because of low computational cost, although a three-stage Runge-Kutta scheme \citep{wicker2002time} is available and recommended.
+The split-explicit time discretization is used for the horizontally propagating sound waves with the 3D divergence damping term \citep{skamarock1992stability}.
+An implicit time discretization is adopted for the vertically propagating wave modes.
 A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.
@@
 
-OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.
-Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting.
-Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
+OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a finite-volume model that solves the deep-atmosphere non-hydrostatic Euler equations in momentum conservation form on a hexagonal Voronoi mesh with Arakawa C-grid staggering.
+The model supports optional local mesh refinement, which introduces some pentagons and heptagons to the grid.
 Height is the vertical coordinate, and a Lorenz vertical grid staggering is used.
@@
 The OLAM cut-cell formulation conserves mass and momentum.
+Acoustic modes are solved explicitly in the horizontal using time splitting and a second-order Lax Wendroff method, and implicitly in the vertical.
+Tracer transport is second-order in space and time using the scheme of \cite{miura2007upwind}, with consistent fluxes obtained by time averaging over the acoustic time steps.
+
+%Acoustic modes are solved explicitly in the horizontal using a 2nd-order Adams-Bashforth method and implicitly in the vertical by means of time splitting.
+Tracer transport is 2nd-order in space and time, using the scheme of \cite{crowley1968numerical} for obtaining consistent numerical fluxes, and Lax-Wendroff for time integration.
+
+%, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.
+Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
 
@@
 
-The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
+%Based upon the sections for the rest of the paper I would like to be
+%able to ascertain the following properties for each model from this section: equation
+%set, horizontal grid and discretization, vertical grid and discretization, temporal
+%discretization, principal diffusion and stabilization mechanisms and transport
+%scheme.
+
+%The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with a floating Lagrangian vertical coordinate.
+
+The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that solves the non-hydrostatic Euler equations on a cubed-sphere grid using a horizontally co-located spectral element discretization.
+In the vertical Tempest uses an Eulerian finite-volume discretization with Lorenz staggering and terrain-following height coordinate.
 The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.
+Fourth-order hyperviscosity is used in the horizontal to prevent a buildup of energy at the grid scale.
+The model further provides an optional upwind-biased transport scheme in the vertical column.
+Tracer transport is performed using the spectral element method with the same timestep as the hydrodynamics and using the tracer mass density as a prognostic variable.
+As with the hydrodynamics, tracer transport is performed explicitly in the horizontal and implicitly in the vertical.
 
@@
 
-The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
+The classic latitude-longitude grid is produced by subdividing the sphere along lines of constant latitude and longitude.
+The latitude-longitude grid has the benefits of being globally rectilinear, which simplifies data access and subdivision of computation across processors, and yields a vector basis that is locally orthogonal nearly everywhere.
+This structure accurately maintains purely zonal flows and simplifies data post-processing for visualization.
 Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
@@
 
-OLAM optimizes by applying the spring dynamics method of \cite{tomita2001shallow} to the dual triangular mesh prior to its mapping to the Voronoi grid.
+OLAM optimizes by applying the spring dynamics method of \cite{tomita2001shallow} to the dual triangular mesh prior to its mapping to the Voronoi mesh.
 When local mesh refinement is applied, which OLAM achieves in a series of one or more resolution-doubling steps, each spanning a transition zone that is three grid rows wide (Figure \ref{fig:OLAMRefinement}), the equilibrium spring length is scaled to the target grid cell size in each refinement level and is varied incrementally across the transition zone.
@@
 
-\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids} \label{sec:grid_ccvt}
+\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) meshes} \label{sec:grid_ccvt}
 
@@
 Figure \ref{fig:CCVT} depicts one such CCVT grid that is compatible with the MPAS model.
-CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:grid_geo}, since both typically contain a large number of hexagonal elements.
-However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.
+CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:grid_geo}, since both typically contain a large number of hexagonal elements; however, CCVT grids are fundamentally constructed using a very different technique.
+Although hexagons are, by far, the most common polygon on CCVT grids, CCVT grids on the sphere will also include at least 12 pentagons and sometimes other polygons with more than six sides.
+Quadrilateral elements are theoretically possible, but are never found in practice on the final grid due to being a locally unstable solution of the underlying CCVT system of equations.
 
@@
 \end{center}
-\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
+\caption{A constrained centroidal Voronoi tessellation mesh with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
 \end{figure}
@@
 The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
-This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
+This grid is employed by the GEM model, utilizing a pair of local area models based with the numerics from the GEM latitude-longitude model.
 
@@
 
-The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, dispersion properties, and the maximum stable timestep size \citep{randall1994geostrophic, ullrich2014understanding}.
+The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, dispersion properties, and the maximum stable timestep size for explicit timestepping schemes \citep{randall1994geostrophic, ullrich2014understanding}.
 The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.
@@
 By interpreting ``staggerings'' to be analogous to a choice of finite element basis, new staggerings are under development in the context of mixed finite element methods \citep{cotter2012mixed}.
-Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves simple co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).
+Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves simple co-location of all velocity components and scalar fields; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid, which co-locates the vorticity, divergence and buoyancy variables (see Figure \ref{fig:horizontalstaggering}).
 
 Arguments in favor or against particular staggerings have generally emerged from linear analyses, and typically in the absence of either implicit or explicit diffusion.
-In this context, the A-grid tends to support large time step sizes, but produces unphysical phase speeds and negative group velocities at high wavenumbers, including a stationary $2 \Delta x$ wavelength mode (even in the context of finite element methods); the C-grid better represents short wave modes, does not support extraneous computational modes (as long as the number of edges is equal to twice the number of volumes), but also requires a timestep that is half that of the A-grid (in the 2nd order case), or 40\% smaller (in the 4th order case) \citep{ullrich2014understanding}; the D-grid provides a better representation of vorticity, but produces unphysical effects analogous to those on the A-grid at high wavenumbers that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Helmholtz problem at each timestep to extract the velocity field from the divergence and vorticity.
+In this context, the A-grid tends to support large time step sizes, but produces unphysical phase speeds and negative group velocities at high wavenumbers, including a stationary $2 \Delta x$ wavelength mode (even in the context of finite element methods); the C-grid better represents short wave modes, does not support extraneous computational modes (as long as the number of horizontal faces is equal to twice the number of volumes), but typically has a more restrictive timestep with explicit timestepping schemes than the A-grid; the D-grid provides a better representation of vorticity, but produces unphysical effects analogous to those on the A-grid at high wavenumbers that must be controlled with divergence damping; finally, the Z-grid yields optimal dispersion properties, but requires the inversion of a Poisson problem at each timestep to extract the velocity field from the divergence and vorticity.
 
@@
 
-\subsubsection{Floating Lagrangian coordinates (ACME--A, DYNAMICO and FV$^3$)} \label{sec:FloatingLagrangian}
+\subsubsection{Floating Lagrangian coordinates (ACME--A and FV$^3$)} \label{sec:FloatingLagrangian}
 
@@
 Cut cell volumes are also reduced, and volumes and surface areas of all cells appear explicitly in the finite-volume formulation of the mass and momentum conservation equations.
+One or more methods are used to avoid the so-called small cell problem where volume to area ratios of cut cells are much less than for full cells and therefore can lead to instability.
+The smallest cells are eliminated by adjusting topography slightly, which is usually justified by noting that local topographic sampling is approximate.
+In larger cut cells, volumes can be increased (without changing surface areas) which stabilizes the cell at the expense of slowing its response to advected transients.
+When either of the above adjustments is unacceptable for a particular application, a flux-balance method based partly on \cite{berger2012simplified} is used to stabilize small cut cells.
 
@@
 
+\subsection{CSU} \label{sec:CSUEquations}
+
+The CSU model uses the vorticity-divergence form of the equations of motion, as described in section \ref{sec:VorticityDivergenceForm}, discretized on the geodesic mesh with absolute vorticity and velocity divergence scalars stored at cell-centers.
+The unified approximation of the equations of motion \citep{arakawa2009unification} is employed to avoid vertically propagating sound waves.
+
 \subsection{DYNAMICO} \label{sec:DYNAMICOEquations}
@@
 The prognostic equations employed by DYNAMICO are based on a Hamiltonian formulation \citep{Dubos2014Equations}.
-The specific prognostic variables employed pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, horizontal covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
+The specific prognostic variables employed are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, horizontal covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$.
 Prognostic equations are in flux-form for mass (\ref{eq:PrognosticPseudoDensity}) and $W$ (\ref{eq:PrognosticVelocityFluxForm}), in advective form for $\Phi$ (\ref{eq:PrognosticShallowGeopotential}), and in vector-invariant form for covariant horizontal momentum (\ref{eq:CovariantUpdateEquation}).
@@
 of non-sedimenting constituents is discarded. Since in the given framework the continuity equation (\ref{eq:EulerianContinuityEquation}) 
-is only valid if the constraint $\sum_{k} \vb{J}_{k}=0$ holds, we (implicitly) assume a fictitious 
+is only valid if the constraint $\sum_{k} \vb{J}_{k}=0$ holds, it is (implicitly) assumed that a fictitious 
 counter-flux of dry air to compensate for the considered vertical diffusion fluxes. As a consequence, 
 ICON currently conserves the global integral of total air mass rather than dry air mass.
+
+\subsection{MPAS} \label{sec:MPASEquations}
+
+The evolution equations used by MPAS are fully described in \cite{skamarock2012multiscale}, based on the formulation of \cite{dutton1986ceaseless}.
+The MPAS model uses the momentum form of the update equations, as described in section \ref{sec:MomentumForm}, with dry mass utilized for the density variable $\tilde{\rho}_s$.
+MPAS further evolves dry mass using a continuity equation of the form (\ref{eq:EulerianContinuityEquation}) and moist potential temperature following (\ref{eq:PotentialTemperatureFluxForm}).
+
+\subsection{NICAM} \label{sec:NICAMEquations}
+
+NICAM prognoses horizontal and vertical momentum analogous to the approach described in section \ref{sec:MomentumForm}.
+It further evolves the density perturbation from the background reference state using (\ref{eq:EulerianContinuityEquation}) and sensible heat part of internal energy.
+A detailed explanation of the evolution equations can be found in \cite{satoh2008nonhydrostatic}.
 
@@
 Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.
+Diffusion also includes mechanisms for damping vertically propagating internal gravity waves, such as model-top Rayleigh layers, which are fairly ubiquitous across models and hence not discussed in detail here.
 Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.
@@
 
-In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
+In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{dennis2012cam, ullrich2014global}.
 Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
@@
 Several limiter options are available for tracer transport including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}.
+
+\subsection{CSU}
+
+The CSU model utilizes an explicit diffusion scheme that consists of fourth-order hyperdiffusion ($\nabla^4$) applied to the vorticity, divergence, and potential temperature.
+The model does not include any explicit diffusion in the vertical column.
+However, for the idealized DCMIP test cases explicit diffusion was disabled.
+
+\subsection{DYNAMICO}
+
+In DYNAMICO, (hyper-)diffusive filters are used to eliminate spurious noise due to the energy-conservative centered discretization.
+Filters are applied every $N_{\footnotesize \mbox{\textit{diff}}}$ Runge-Kutta time steps in a forward-Euler manner, with $N_{\footnotesize \mbox{\textit{diff}}}$ as large as allowed by stability.
+The scalar Laplacian is computed as the divergence of the gradient and the vector Laplacian is decomposed into its divergent (grad div) and rotational (curl curl) parts.
+The strength of filtering is controlled by dissipation time scales $\tau$:  Given $\tau$ the hyperviscous coefficient that multiplies operator $D^p$ is  $\delta^{2p} \tau^-1$ where $\delta^-2$ is the largest eigenvalue of operator $D$.
+For DCMIP, DYNAMICO uses $p=2$ (fourth-order hyperviscosity) for all filters.
 
@@
 The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow.
-If additional damping is desired, or if the non-monotonic advection is used, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
+The model also supports an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion.
 The hyperdiffusion is of the same order as but much smaller than the divergence damping.
@@
 
-Wave absorption at the model top is also provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
+FV$^3$ is constructed with a flexible-lid (constant-pressure) upper boundary that is effective at damping internal gravity wave modes; however, FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain.
 An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.
@@
 This is only done before the physics is called and is not used in idealized simulations.
+
+\subsection{FVM}
+
+Within the dynamical core, FVM does not apply any explicit dissipation/diffusion.  For the DCMIP test cases, the implicit 
+regularization of the monotonic MPDATA provides sufficient dissipation/diffusion needed to remove excess 
+energy from the finest scales and maintain model stability.  An absorbing layer is also available for damping vertically 
+propagating waves near the model top.
 
@@
 
+\subsection{MPAS}
+
+The MPAS model applies fourth-order hyperdiffusion and Smagorinsky diffusion \citep{smagorinsky1963general}, as described in \cite{skamarock2012multiscale}.
+When applied to the momentum, the Laplacian is evaluated as
+\begin{align}
+\nabla^2 u_i = \pdiff{}{x_i} \nabla_s \cdot \vb{v} - \pdiff{\eta}{x_j},
+\end{align} where $u_i$ is the edge-normal velocity defined on cell edge $i$, $\eta$ is the vertical component of the relative vorticity, computed on vertices, and $\nabla_s \cdot \vb{v}$ is the horizontal divergence on $s$ surfaces, computed on edges.
+The evaluation of divergence and vorticity in this expression is described in \cite{ringler2010unified}.
+The fourth-order hyperdiffusion operator is then computed by twice applying the above Laplacian operator to the momentum.
+
+Smagorinsky diffusion, which is often applied in atmospheric models to parameterize turbulent processes, uses a second-order Laplacian and a physically-motivated eddy viscosity $K_h$, defined in terms of Cartesian velocities $(u,v)$,
+\begin{align}
+K_h = c_s^2 \ell^2 \sqrt{(u_x - v_y)^2 + (u_y + v_x)^2},
+\end{align} where $c_s$ is a constant parameter and $\ell$ is the grid scale.  The diffusion operator then takes the form $\nabla \cdot (K_h \nabla \psi)$ for a scalar field $\psi$.
+
+%{\color{red}The evaluation of the eddy viscosity Kh is accomplished by projecting the velocities onto a tangent plane, integrating the squared terms involving the velocities in (17) over the cell, and applying Stokes theorem to transform the cell integrals to discrete line integrals around the cell edge.
+The evaluation is inexpensive because the prognostic normal and diagnostic tangential velocities exist, and the time-independent coefficients for the line integrals are precomputed and stored before the time integration begins.}
+
+\subsection{NICAM}
+
+NICAM implements three types of diffusion: 3D divergence damping, fourth-order horizontal hyperdiffusion, and sixth-order vertical hyperdiffusion as described in \cite{tomita2004new}.
+Specifically, the divergence damping term \citep{skamarock1992stability} aims to suppress instabilities that arise due to the time splitting scheme, and is applied to both horizontal and vertical velocities.
+The hyperdiffusion operators are applied to all prognostic variables.
+For tracer advection, upwinding is used to remove spurious oscillations, as described in \cite{miura2007upwind,niwa2011three}.
+
 \subsection{OLAM}
@@
 Throughout the model domain, vertical vorticity is filtered horizontally at the smallest resolvable scale in order to control a spurious computational mode.
-This vertical vorticity mode is inherent in C-staggered momentum formulations on hexagonal meshes because horizontal velocities are more numerous than twice the number of scalar (mass) values and are thus under-constrained.
+This vertical vorticity mode is inherent in C-staggered momentum formulations on hexagonal meshes because horizontal velocities are more numerous than twice the number of scalar (mass) values and are thus under-constrained \citep{weller2012computational, weller2012controlling}.
 The vorticity filter is constructed so as to have zero impact on divergence at any scale.
@@
 In the past two decades, a variety of new temporal discretizations have been developed, leaving behind the days when the leapfrog scheme was ubiquitous across models.
-This diversity is in part because of the demands of non-hydrostatic models: unlike their hydrostatic counterparts, non-hydrostatic atmospheric models that do not use the anelastic \citep{ogura1962scale}, quasi-hydrostatic \citep{orlanski1981quasi}, pseudo-incompressible \citep{durran1989improving}, or unified approximation \citep{arakawa2009unification} must include a mechanism in the temporal discretization for dealing with vertically propagating sound waves.
-These waves are meteorologically insignificant, but with a vertical grid spacing of 100 meters, a purely explicit temporal discretization would require a time step size on the order of one second or less.
+This diversity is in part because of the demands of non-hydrostatic models: unlike their hydrostatic counterparts, non-hydrostatic atmospheric models must include a mechanism for dealing with vertically propagating sound waves.
+These waves are meteorologically insignificant, but with a vertical grid spacing of 100 meters, a purely explicit temporal discretization of the unmodified fluid equations would require a time step size on the order of one second or less.
+Consequently, sound waves are either filtered explicitly through the use of an alternative equation set, or artificially slowed through the use of implicit temporal discretizations.
+Some commonly employed alternative equation sets include the anelastic \citep{ogura1962scale}, quasi-hydrostatic \citep{orlanski1981quasi}, pseudo-incompressible \citep{durran1989improving}, or unified approximation \citep{arakawa2009unification}.
+These filtered equation sets generally require that a global elliptic solve be performed as prognostic variables are updated.
 In this section we discuss the timestepping schemes that have been employed across the DCMIP suite of models.
@@
 
+ACME-A and Tempest both use the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
+A number of different ARK schemes have been compared and contrasted in this framework, with significant implications for model performance and stability \citep{gardner2017implicit}.
+
+CSU uses a semi-implicit time integration scheme with third-order Adams-Bashforth scheme for explicit integration of the continuity equation, potential temperature equation, and terms related to advection.
+Since potential temperature is updated prior to the computation of the pressure-gradient force, this term can be thought of as implicit in time.
+The horizontal wind field is then predicted through integration of the vorticity and divergence of the horizontal wind and a multigrid method applied to solve a pair of two-dimensional Poisson equations for the stream function and velocity potential, which are then differentiated to obtain the velocity field.
+Horizontal diffusion is then applied forward in time.
+
 FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics.
@@
 
-MPAS uses a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
+MPAS and NICAM use a split-explicit formulation \citep{klemp2007conservative} consisting of an outer Runge-Kutta loop (typically RK3) and inner acoustic loop.
 At the beginning of each Runge-Kutta sub-cycle, tendencies are computed for each of the prognostic variables and stored for the duration of the sub-cycle.
@@
 A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.
-
-Tempest uses the ARS(2,3,2) scheme described in \cite{ascher1997implicit}, with all horizontal and vertical advection terms treated explicitly and the remaining vertical terms, associated with sound wave propagation, treated implicitly.
-A number of different ARK schemes have compared and contrasted in this framework, with significant implications for model performance and stability (Gardner et al., under review).
 
@@
 \begin{equation}
-\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
+\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, G^{n}, G^{n+1})
 +0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
@@
 
-The solution procedure of the system \eqref{fvm:compressible} can then basically be divided into three steps. First, the homogenous
-mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
+The solution procedure of the system \eqref{fvm:compressible} can then be divided into three steps. First, the homogenous
+mass continuity equation \eqref{fvm:mass} is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $, $G \equiv \mathcal{G}$,
 and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, given already updated moisture variables 
 \citep{smolarkiewiczetalJCP2017}, the thermodynamic \eqref{fvm:thermodynamic} and momentum \eqref{fvm:momentum} equations enter
-\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
+\eqref{fvm:solscalar} with $\psi = u, v, w, \theta'$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, $G \equiv \mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
 which is generally depending on all these prognostic variables. The high degree of implicitness in the representation of the rhs forcings
@@
 
-The vast diversity of the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.
+The vast diversity within the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.
 Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the kinetic energy spectrum \citep{skamarock2004evaluating}.
@@
 \hline Short Name & Code availability (cont'd) \\ \hline 
-MPAS & Open-source via GitHub (\url{https://github.com/MPAS-Dev/MPAS-Release}).  The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583316}}) \\
+MPAS & MPAS is an open-source model available under the BSD 3-clause license via GitHub (\url{https://github.com/MPAS-Dev/MPAS-Release}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583316}}) \\
 NICAM & NICAM source code is available under the BSD 2-clause license via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.580128}}).
@@
 
-Note that virtual temperature is typically approximated by
-\begin{align}
-T_v \approx& T \left(1 + \frac{(1-\varepsilon)}{\varepsilon} q \right),
-\end{align} which arises from the exact relationship
+Note that virtual temperature is typically written as
+\begin{align}
+T_v =& T \left(1 + \frac{(1-\varepsilon)}{\varepsilon} q \right),
+\end{align} which arises from the relationship
 \begin{align} \label{eq:VirtualTemperatureExact}
 T_v = \frac{T}{1 - \frac{e}{p} (1 - \varepsilon)},
-\end{align} upon applying $e/p = q / \varepsilon$ and using a Taylor expansion around $q = 0$.
+\end{align} upon applying $e/p = q / \varepsilon$.
 
@@
 \begin{align}
-\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{c_p^\ast},
+\frac{1}{T} \diff{T}{t} - \frac{\kappa}{p} \diff{p}{t} =& \frac{J}{T c_p^\ast},
 \end{align} which can be alternatively written as
 \begin{align} \label{eq:PotentialTemperaturePrognostic}
-\diff{\theta}{t} = \frac{J \theta}{c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{c_p^\ast}.
+\diff{\theta}{t} = \frac{J \theta}{T c_p^\ast}, \quad \mbox{or} \quad \diff{\theta_v}{t} = \frac{J \theta_v}{T c_p^\ast}.
 \end{align}  These equations can then be combined with (\ref{eq:ContinuityEquation}) to obtain
 \begin{align} \label{eq:PotentialTemperatureFluxForm}
-\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{c_p^\ast},
+\pdiff{}{t} (\rho \theta_v) + \nabla \cdot (\rho \theta_v \vb{u}) = \frac{J \rho \theta_v}{T c_p^\ast},
 \end{align} or similarly for $\theta$.  In conjunction with the material derivative of the ideal gas law,  
@@
 \begin{align}
-\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{R^\ast}.
+\frac{c_v^\ast}{R^\ast T_v} \diff{T_v}{t} - \frac{1}{\rho} \diff{\rho}{t} = \frac{J}{T R^\ast}.
 \end{align}  Then substituting (\ref{eq:ContinuityEquation}) gives a prognostic equation for virtual temperature,
 \begin{align} \label{eq:VirtualTemperaturePrognosticEq}
-\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{R^\ast}.
+\frac{c_v^\ast}{R^\ast} \diff{T_v}{t} + T_v \nabla \cdot \vb{u} = \frac{J T_v}{T R^\ast}.
 \end{align}  The prognostic equation for temperature is identical except with $T$ substituted for $T_v$.  An analogous equations for pressure can be obtained through a similar procedure,
 \begin{align} \label{eq:PressurePrognosticEq}
-\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{c_p^\ast}.
+\frac{c_v^\ast}{c_p^\ast} \diff{p}{t} + p \nabla \cdot \vb{u} = \frac{J p}{T c_p^\ast}.
 \end{align}  And similarly for Exner pressure,
 \begin{align}
-\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{c_p^\ast}.
+\frac{c_v^\ast}{R^\ast} \diff{\pi}{t} + \pi \nabla \cdot \vb{u} = \frac{J \pi}{T c_p^\ast}.
 \end{align}
@@
 \begin{align} \label{eq:PressureGradientForceAlt}
-- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
+- \frac{1}{\rho} \nabla p = - c_p^\ast \theta \left[ \nabla \pi - \pi \ln \left( \frac{p}{p_0} \right) \nabla \kappa \right],
 \end{align} which follows from (\ref{eq:DiagnosticRelationships1}).
-Note that when condensate loading is included, the virtual potential temperautre in this equation is replaced with $\theta_\rho$.
+Note that often in non-hydrostatic models, $\kappa$ is assumed constant and the $\nabla \kappa$ term neglected.
 A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
@@
 \vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
-\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
+\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D specific kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.
 This gives rise to the 3D vector-invariant form,
@@
 
+%\subsection{Momentum Form}
+
+%The equations of motion can similarly be formulated in terms of momentum, which we denote by
+%\begin{align}
+%\vb{U} = \rho \vb{u}, \qquad \vb{U}_h = \rho \vb{u}_h, \qquad \mbox{and} \qquad W = \rho w,
+%\end{align} for total, horizontal, and vertical velocities, respectively.  Following (\ref{eq:DefinitionDotS}), these definitions gives rise to the generalized vertical momentum,
+%\begin{align}
+%\dot{S} = W - \vb{U}_h \cdot \nabla_s z
+%\end{align}  Then multiplying (\ref{eq:PrognosticUhEquationK2Sb}) through by $\rho$ and using (\ref{eq:ContinuityEquation}) gives
+%\begin{align}
+%\pdiff{\vb{U}_h}{t} =& - W (\vb{u}_h \cdot \nabla \vb{k}) - \rho \nabla_s K_2 - \zeta_s \vb{k} \times \vb{U}_h - \pdiff{}{s} \left( \dot{S} \vb{u}_h \right) - \rho \nabla_z \Phi - \nabla_z p - [2 \vg{\Omega} \times \rho \vb{u}]_z
+%\end{align}
+%\begin{align}
+%\pdiff{\vb{U}_h}{t} =& - \nabla_z p - \eta \vb{k} \times \vb{U}_h - (\nabla_s \cdot \vb{U}_h) \vb{u}_h - \pdiff{}{z} \left( \dot{S} \vb{u}_h \right) - \rho \nabla_s K_2
+%\end{align}
+
+%\begin{align}
+%\pdiff{W}{t} =&\left( \pdiff{s}{z} \right) \left[ - \rho \pdiff{\Phi}{s} - \pdiff{p}{s} \right] - (\nabla \cdot \vb{U} w) - (2 \vg{\Omega} \times \vb{U}_h) \cdot \vb{k}
+%\end{align}
+
 \subsection{Covariant Component Formulation}
@@
 
-\subsection{Vorticity-Divergence Form}
+\subsection{Vorticity-Divergence Form} \label{sec:VorticityDivergenceForm}
 
@@
 
-\subsection{Momentum Form}
+\subsection{Momentum Form} \label{sec:MomentumForm}
 

### 2017-10-26 DCMIP2016-Part1-Models.tex
- Near final production update

@@
 Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts.
-ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed.
-The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
+ACME-A is built upon an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration (\S \ref{sec:grid_cs}), although unstructured regionally-refined meshes with conforming edges may also be employed.
+The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates (\S \ref{sec:FloatingLagrangian}).
 Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}.
 All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.
+Tracer transport is sub-cycled relative to the hydrodynamics, using the spectral element method, with tracer mass as the prognostic variable.
 
@@
 
-The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate.
+The CSU model is a finite-volume model using an optimized geodesic grid \citep{heikes1995numerical,heikes2013optimized} (\S \ref{sec:grid_geo}), with height as the vertical coordinate.
 The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state.
 The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential.
-Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.
+Horizontal diffusion is included in the form of a fourth-order hyperviscosity operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer (\S \ref{sec:diffusion_csu}).
 The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
@@
 
-DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations.
-DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.
+DYNAMICO is a mimetic finite-difference / finite-volume model using a geodesic grid (\S \ref{sec:grid_geo}) and a floating vertical mass coordinate (\S \ref{sec:FloatingLagrangian}).
+Although originally a hydrostatic model, it has been recently extended to solve the shallow-atmosphere non-hydrostatic Euler equations.
+DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}.
 It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.
 It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).
+Horizontal diffusion is included via a fourth-order hyperviscosity operator (\S \ref{sec:diffusion_dynamico}).
 In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.
@@
 
-The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}.
+The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid (\S \ref{sec:grid_cs}) with a floating Lagrangian vertical coordinate.
 The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below).
+The non-hydrostatic formulation extends the hydrostatic model described in  \cite{lin2004vertically} by adding a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.
+The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}), although the prognostic horizontal winds are stored in the native Gnomonic local coordinate.
+All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.
 Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport.
-The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}).
-Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.
+Since divergence is effectively invisible to the solver, a 2D divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale (\S \ref{sec:diffusion_fv3}).
 Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise.
@@
 The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
-ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
-compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
-A centered two-time-level semi-implicit 
+ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}.
+FVM solves the non-hydrostatic Euler equations on an octahedral reduced Gaussian grid (\S \ref{sec:grid_oct}) with a height-based terrain-following vertical coordinate \citep{szmelter2010,smolarkiewiczetalJCP2016}.
+The horizontal spatial discretization uses the median-dual finite-volume approach, combined with a structured-grid finite-difference method in the vertical.
+In both the horizontal and
+the vertical discretization all variables are co-located. A centered two-time-level semi-implicit 
 integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
-\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
+\citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}). The associated 3D Helmholtz problem is solved iteratively using a bespoke 
 preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
-non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
-advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
-combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
-the vertical discretization all variables are co-located. 
-The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
-Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
+non-oscillatory finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
+advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  The non-oscillatory (i.e.~monotonic) MPDATA 
+also provides sufficient dissipation/diffusion to stabilize the model, so no other explicit filtering mechanism is required (\S \ref{sec:diffusion_fvm}). 
+Note that the octahedral reduced Gaussian 
 grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
 interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
-broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
-%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.
-
-%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
-%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
-%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
-%are available by means of simple switches. The formulation incorporates 
-%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
-%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
-%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
-%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
-%Conjugate Residual approach. The dynamics time step is adapted 
-%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
-%than 1). The physics time step is identical to the dynamics time step. 
-% 
-%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
-%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
-%The horizontal spatial discretization is fully unstructured finite-volume using the 
-%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
-%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
-%In both the horizontal and the vertical discretization, all prognostic variables are 
-%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
-%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
-%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
-%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
-%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
-%
-%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
-%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
-%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
-%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
-%the form of a Rayleigh-type forcing to the prognostic variables.
-
-%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
-%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.
-
-%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
-%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
-%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
-%efficiency and accuracy.
-
-%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
-%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
-%framework developed at ECMWF.
+broad classes of meshes \citep[][]{szmelter2010,kuehnleinJCP2012,deconinck2017atlas}.
 
@@
 The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.
-The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
-The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
-and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
+GEM solves the non-hydrostatic Euler equations on the Yin-Yang grid \citep{kageama2004yinyang} (\S \ref{sec:grid_yinyang}) with Arakawa C-grid staggering of prognostic variables.
+The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type (\S \ref{sec:GEM_zeta})
+and the vertical discretization is based on the Charney-Phillips grid (\S \ref{sec:VerticalStaggering}).
 A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
@@
 linear system of equations that is reduced to a Helmholtz problem for one composite variable.
-For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
-The dynamics and physics are time split.
+For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.  Semi-Lagrangian advection is also used for tracer transport.
 To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
-via applications of the Laplacian operator, applied after the completion of the physics time step.
-
-\subsection{Icosahedral Non-hydrostatic model (ICON)}
-
-The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.
-Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.
-Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels.
-The discretization employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
-Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
-One complete time step typically includes 5 dynamical sub-steps.
-The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+via applications of the Laplacian operator, applied after the completion of the physics time step (\S \ref{sec:diffusion_gem}).
+
+\subsection{ICOsahedral Non-hydrostatic model (ICON)}
+
+The ICON model \citep{zangl2015icon} is a finite-volume model that solves the non-hydrostatic Euler equations in 2D vector invariant form on an icosahedral (triangular) grid (\S \ref{sec:grid_ico_tri}) with Arakawa C-grid staggering, and further utilizing a smoothed terrain-following height-based Lorenz vertical discretization.
+Prognostic horizontal velocities are stored as normal wind components at the edge mid-points of full levels.
+Prognostic vertical velocity is stored at the circumcenters of the triangles on half levels.
+The discretization employs a two time-level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.
 For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells.
@@
 Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves.
-The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator.
-For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion.
-The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
+The horizontal diffusion, which is applied at full model time steps, combines a flow-dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator (\S \ref{sec:diffusion_icon}).
+For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which leads to local mass conservation and consistency with the air motion.
+Specifically, the average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport.
+These numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.
+
+%Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics.
+One complete time step typically includes 5 dynamical sub-steps.
 
@@
 
-The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
-Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
+The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} is a finite-volume model that solves the non-hydrostatic Euler equations using an Arakawa C-grid staggering on a centroidal Voronoi tessellation mesh (\S \ref{sec:grid_ccvt}), and the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.
 In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.
+Advection is nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.
 The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.
 Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.
-Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.
+Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, fourth-order hyperdiffusion, and 2D and 3D divergence damping operators (\S \ref{sec:diffusion_mpas}).
 
@@
 
-NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}.
-The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}.
-Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}.
-All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids.
+NICAM is a finite-volume model that solves the non-hydrostatic Euler equations using a geodesic grid (\S \ref{sec:grid_geo}) optimized with spring dynamics using the method of \cite{tomita2002optimization}.
+A terrain-following height coordinate system is used in the vertical \citep{tomita2004new} with Lorenz staggering.
+Instead of temperature or potential temperature, total energy is prognosed following the method of \citep{satoh2002conservative, satoh2003conservative}.
+All prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell-integrated quantities to point values at cell centroids.
 The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}.
-The Arakawa C-grid staggering is used vertically.
-A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended.
-The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}.
-The implicit time discretization is adopted for the vertically propagating one.
+For integration in time, a two-stage Runge-Kutta scheme is usually employed because of low computational cost, although a three-stage Runge-Kutta scheme \citep{wicker2002time} is available and recommended.
+The split-explicit time discretization is used for the horizontally propagating sound waves with the 3D divergence damping term \citep{skamarock1992stability} (\S \ref{sec:diffusion_nicam}).
+An implicit time discretization is adopted for the vertically propagating wave modes.
 A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.
@@
 
-OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.
-Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting.
-Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).
+OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a finite-volume model that solves the deep-atmosphere non-hydrostatic Euler equations in momentum conservation form on a hexagonal Voronoi mesh (\S \ref{sec:grid_geo}) with Arakawa C-grid staggering.
+The model supports optional local mesh refinement, which introduces some pentagons and heptagons to the grid.
 Height is the vertical coordinate, and a Lorenz vertical grid staggering is used.
-A unique feature of OLAM is that grid levels are horizontal and intersect topography.
+A unique feature of OLAM is that grid levels are horizontal and intersect topography (\S \ref{sec:CutCells}).
 This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms.
@@
 
-The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.
+%Based upon the sections for the rest of the paper I would like to be
+%able to ascertain the following properties for each model from this section: equation
+%set, horizontal grid and discretization, vertical grid and discretization, temporal
+%discretization, principal diffusion and stabilization mechanisms and transport
+%scheme.
+
+%The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with a floating Lagrangian vertical coordinate.
+
+The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that solves the non-hydrostatic Euler equations on a cubed-sphere grid (\S \ref{sec:grid_cs}) using a horizontally co-located spectral element discretization.
+In the vertical Tempest uses an Eulerian finite-volume discretization with Lorenz staggering and terrain-following height coordinate.
 The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.
+Fourth-order hyperviscosity is used in the horizontal to prevent a buildup of energy at the grid scale (\S \ref{sec:diffusion_acme}).
+The model further provides an optional upwind-biased transport scheme in the vertical column.
+Tracer transport is performed using the spectral element method with the same timestep as the hydrodynamics and using the tracer mass density as a prognostic variable.
+As with the hydrodynamics, tracer transport is performed explicitly in the horizontal and implicitly in the vertical.
 
@@
 
-The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.
+The classic latitude-longitude grid is produced by subdividing the sphere along lines of constant latitude and longitude.
+The latitude-longitude grid has the benefits of being globally rectilinear, which simplifies data access and subdivision of computation across processors, and yields a vector basis that is locally orthogonal nearly everywhere.
+This structure accurately maintains purely zonal flows and simplifies data post-processing for visualization.
 Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.
-This grid is employed by the UK Met Office \citep{davies2005new, wood2014inherently}.
+This grid is presently employed in many global models, including the UK Met Office New Dynamics and ENDGame dynamical cores \citep{davies2005new, wood2014inherently}.
+The latitude-longitude grid is also an option in the GEM model.
 
@@
 This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).
+Among the DCMIP2016 models, the cubed-sphere grid is employed by the ACME, FV$^3$, and TEMPEST dynamical cores.
 
@@
 \begin{center}
-\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
+\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{Figure1a.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{Figure1b.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{Figure1c.pdf}
 \end{center}
@@
 By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.
+Among the DCMIP2016 models, the icosahedral (triangular) grid is employed operationally in the ICON dynamical core.
 
@@
 Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).
+Among the DCMIP2016 models, the geodesic grid is employed by the CSU, DYNAMICO, NICAM, and OLAM dynamical cores.
 
@@
 \begin{center}
-\includegraphics[width=2.5in]{OLAM_Grid_Refinement.png}
+\includegraphics[width=2.5in]{Figure2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
+\includegraphics[width=2in]{Figure3.pdf}
 \end{center}
-\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
+\caption{A constrained centroidal Voronoi tessellation mesh with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
 \end{figure}
@@
 Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
-the edges of the primary mesh as applied in the context of the finite-volume discretization of FVM (Section~\ref{sec:FVM}).
+the edges of the primary mesh as applied in the context of the finite-volume discretization of FVM (\S \ref{sec:FVM}).
 
@@
   \centering
-  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
+  \includegraphics[width=6cm]{Figure4a.pdf}
  \end{subfigure}%
@@
   \centering
-  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
+  \includegraphics[width=5.4cm]{Figure4b.pdf}
  \end{subfigure}
@@
 The Yang component covers an analogous area, but is rotated perpendicularly so as to cover the region of the sphere outside of the Yin grid.
-This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.
+This grid is employed by the GEM model, utilizing a pair of local area models based with the numerics from the GEM latitude-longitude model.
 
 \begin{figure}[]
-\includegraphics[height=6cm]{gem_yinyang.png}
+\includegraphics[height=6cm, clip, trim=0cm 3.0cm 0cm 0cm]{Figure5.pdf}
 \caption{The Yin-Yang grid is a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
@@
 \begin{figure}[]
-\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
+\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{Figure6.pdf}
 \caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.  Here $\eta$ denotes the buoyancy variable.}
@@
 \begin{figure}[]
-\includegraphics[width=5in]{VerticalStaggering_v2.pdf}
+\includegraphics[width=5in]{Figure7.pdf}
 \caption{(a) A Lorenz-type variable staggering for a model utilizing height coordinates, (b) a Charney-Phillips-type variable staggering for a model utilizing height coordinates, (c) a modified Charney-Phillips-type staggering used in the GEM model that introduces a new near-surface level for vertical velocity and temperature.}
@@
 
-\subsubsection{Mass-based coordinates}
+\subsubsection{Mass-based coordinates} \label{sec:mass_coords}
 
@@
 \begin{center}
-\includegraphics[width=2in]{OLAM_CutCellA.png} \quad \includegraphics[width=2in]{OLAM_CutCellB.png}
+\includegraphics[width=5in,trim={0cm 5cm 0cm 15.5cm},clip]{Figure8.pdf}
 \end{center}
@@
 
-\subsection{ACME--A / TEMPEST}
-
-In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{ullrich2014global}.
+\subsection{ACME--A / TEMPEST} \label{sec:diffusion_acme}
+
+In both ACME--A and Tempest, scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of a scalar Laplacian \citep{dennis2012cam, ullrich2014global}.
 Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping and vorticity damping terms via the vector identity
@@
 
-\subsection{FV$^3$}
+\subsection{CSU} \label{sec:diffusion_csu}
+
+The CSU model utilizes an explicit diffusion scheme that consists of fourth-order hyperdiffusion ($\nabla^4$) applied to the vorticity, divergence, and potential temperature.
+The model does not include any explicit diffusion in the vertical column.
+However, for the idealized DCMIP test cases explicit diffusion was disabled.
+
+\subsection{DYNAMICO} \label{sec:diffusion_dynamico}
+
+In DYNAMICO, (hyper-)diffusive filters are used to eliminate spurious noise due to the energy-conservative centered discretization.
+Filters are applied every $N_{\footnotesize \mbox{\textit{diff}}}$ Runge-Kutta time steps in a forward-Euler manner, with $N_{\footnotesize \mbox{\textit{diff}}}$ as large as allowed by stability.
+The scalar Laplacian is computed as the divergence of the gradient and the vector Laplacian is decomposed into its divergent (grad div) and rotational (curl curl) parts.
+The strength of filtering is controlled by dissipation time scales $\tau$:  Given $\tau$ the hyperviscous coefficient that multiplies operator $D^p$ is  $\delta^{2p} \tau^-1$ where $\delta^-2$ is the largest eigenvalue of operator $D$.
+For DCMIP, DYNAMICO uses $p=2$ (fourth-order hyperviscosity) for all filters.
+
+\subsection{FV$^3$} \label{sec:diffusion_fv3}
 
@@
 
-\subsection{GEM}
+\subsection{FVM} \label{sec:diffusion_fvm}
+
+Within the dynamical core, FVM does not apply any explicit dissipation/diffusion.  For the DCMIP test cases, the implicit 
+regularization of the monotonic MPDATA provides sufficient dissipation/diffusion needed to remove excess 
+energy from the finest scales and maintain model stability.  An absorbing layer is also available for damping vertically 
+propagating waves near the model top.
+
+\subsection{GEM} \label{sec:diffusion_gem}
 
@@
 
-\subsection{ICON}
+\subsection{ICON} \label{sec:diffusion_icon}
 
@@
 
-\subsection{OLAM}
+\subsection{MPAS} \label{sec:diffusion_mpas}
+
+The MPAS model applies fourth-order hyperdiffusion and Smagorinsky diffusion \citep{smagorinsky1963general}, as described in \cite{skamarock2012multiscale}.
+When applied to the momentum, the Laplacian is evaluated as
+\begin{align}
+\nabla^2 u_i = \pdiff{}{x_i} \nabla_s \cdot \vb{v} - \pdiff{\eta}{x_j},
+\end{align} where $u_i$ is the edge-normal velocity defined on cell edge $i$, $\eta$ is the vertical component of the relative vorticity, computed on vertices, and $\nabla_s \cdot \vb{v}$ is the horizontal divergence on $s$ surfaces, computed on edges.
+The evaluation of divergence and vorticity in this expression is described in \cite{ringler2010unified}.
+The fourth-order hyperdiffusion operator is then computed by twice applying the above Laplacian operator to the momentum.
+
+Smagorinsky diffusion, which is often applied in atmospheric models to parameterize turbulent processes, uses a second-order Laplacian and a physically-motivated eddy viscosity $K_h$, defined in terms of Cartesian velocities $(u,v)$,
+\begin{align}
+K_h = c_s^2 \ell^2 \sqrt{(u_x - v_y)^2 + (u_y + v_x)^2},
+\end{align} where $c_s$ is a constant parameter and $\ell$ is the grid scale.  The diffusion operator then takes the form $\nabla \cdot (K_h \nabla \psi)$ for a scalar field $\psi$.
+
+%{\color{red}The evaluation of the eddy viscosity Kh is accomplished by projecting the velocities onto a tangent plane, integrating the squared terms involving the velocities in (17) over the cell, and applying Stokes theorem to transform the cell integrals to discrete line integrals around the cell edge.
+The evaluation is inexpensive because the prognostic normal and diagnostic tangential velocities exist, and the time-independent coefficients for the line integrals are precomputed and stored before the time integration begins.}
+
+\subsection{NICAM} \label{sec:diffusion_nicam}
+
+NICAM implements three types of diffusion: 3D divergence damping, fourth-order horizontal hyperdiffusion, and sixth-order vertical hyperdiffusion as described in \cite{tomita2004new}.
+Specifically, the divergence damping term \citep{skamarock1992stability} aims to suppress instabilities that arise due to the time splitting scheme, and is applied to both horizontal and vertical velocities.
+The hyperdiffusion operators are applied to all prognostic variables.
+For tracer advection, upwinding is used to remove spurious oscillations, as described in \cite{miura2007upwind,niwa2011three}.
+
+\subsection{OLAM} \label{sec:diffusion_olam}
 
@@
 \subsection*{Code availability}
+
+Information on the availability of source code for the models featured in this paper is tabulated below.
+
+\begin{center}
+\begin{tabular}{cp{5in}}
+\hline Short Name & Code availability \\ \hline 
+ACME--A & ACME, including ACME--A, is under active development funded by the U.S.
+Department of Energy.
+ACME version 1.0 is scheduled to be publicly released under an open source license in 2018, but is not available at present.$^\dagger$ \\
+CSU & CSU model source code is available under the BSD 3-clause license.  The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.580099}}). \\
+DYNAMICO & DYNAMICO is open source and available online from IPSL Forge (\mbox{\url{http://forge.ipsl.jussieu.fr/dynamico}}) or directly by request to Thomas Dubos (\mbox{dubos@lmd.polytechnique.fr}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583718}}).
+\\
+FV$^3$ & FV$^3$ model source code is available through the GFDL Virtual Lab (\url{https://vlab.ncep.noaa.gov/web/fv3gfs}).  Access requires users to create a Virtual Lab account. \\
+FVM & Model codes developed at ECMWF, including the IFS and FVM, are intellectual 
+property of ECMWF and its member states.
+Although the FVM code is not publicly available at present, it is expected that FVM will be available in the near future under the OpenIFS license (\url{http://www.ecmwf.int/en/research/projects/openifs}).
+The repo tag for the version of FVM that applies for DCMIP is ``v0.1''.$^\dagger$ \\
+GEM & Due to licensing requirements, GEM model code is only available by request to Abdessamad Qaddouri (Abdessamad.Qaddouri@canada.ca) or Vivian Lee (Vivian.Lee2@canada.ca). \\
+ICON & ICON is freely available to the scientific community for non-commercial research under an institutional license issued by project partners DWD+MPI-M.
+Because of the restrictions of this license, access to the code is only available by request to G\"unther Z\"{a}ngl (\mbox{Guenther.zaengl@dwd.de}) or Marco Giorgetta (\mbox{marco.giorgetta@mpimet.mpg.de}).
+\\
+\hline \multicolumn{2}{p{6in}}{$\dagger$ \footnotesize In compliance with the GMD editorial requirements, this code has been made available to the topical editor in charge of the manuscript.}
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{cp{5in}}
+\hline Short Name & Code availability (cont'd) \\ \hline 
+MPAS & MPAS is an open-source model available under the BSD 3-clause license via GitHub (\url{https://github.com/MPAS-Dev/MPAS-Release}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.583316}}) \\
+NICAM & NICAM source code is available under the BSD 2-clause license via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.580128}}).
+Further information on collaborating with the NICAM team can be found at \mbox{\url{http://nicam.jp/hiki/?Research+Collaborations}}.
+\\
+OLAM & OLAM is open source and available online via SourceForge (\url{https://sourceforge.net/projects/olam-model/}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://doi.org/10.5281/zenodo.582308}}).
+\\
+TEMPEST & Tempest source code is available under the Lesser GNU Public License on GitHub (\mbox{\url{https://github.com/paullric/tempestmodel}}).
+The release used for DCMIP2016 can be found via Zenodo (\mbox{\url{http://dx.doi.org/10.5281/zenodo.579649}}).
+\\
+\hline 
+\end{tabular}
+\end{center}
+
+\clearpage
 

### 2017-10-26 YinYangFigure/YinYangFigure.m

@@ -1,0 +1,70 @@
+function YinYangFigure()
+
+pad = 0.05;
+
+% Figure window
+gcf = figure(1);
+set(gcf, 'Position', [100, 100, 1400, 600]);
+
+% Resolution of grid (# of points in latitudinal direction)
+np = 16;
+halo = 1;
+
+% Generate points on Yin grid
+dlat = pi/2/np;
+yinlatp = dlat*[0:np+2*halo]'-pi/4-dlat*halo;
+yinlonp = dlat*[0:3*np+2*halo]'-3*pi/4-dlat*halo;
+
+nlat = length(yinlatp);
+nlon = length(yinlonp);
+
+[yinlon, yinlat] = meshgrid(yinlonp, yinlatp);
+
+yinx = cos(yinlat) .* cos(yinlon);
+yiny = cos(yinlat) .* sin(yinlon);
+yinz = sin(yinlat);
+
+ax1 = axes('Position',[pad/3 pad-0.02 1/3-2/3*pad 1-2*pad], 'Visible', 'off');
+annotation('textbox', [0.14 0.97 0 0], 'string', 'Yin', 'FontSize', 36);
+draw3_cartesian_mesh(ax1, yinx, yiny, yinz, 'r');
+
+% Generate points on Yang grid
+yangx = -yinx;
+yangy = yinz;
+yangz = yiny;
+
+ax2 = axes('Position',[0.3+pad/3 pad-0.02 1/3-2/3*pad 1-2*pad], 'Visible', 'off');
+annotation('textbox', [0.44 0.97 0 0], 'string', 'Yang', 'FontSize', 36);
+draw3_cartesian_mesh(ax2, yangx, yangy, yangz, 'b');
+
+ax3 = axes('Position',[2/3+pad/3 pad-0.02 1/3-2/3*pad 1-2*pad], 'Visible', 'off');
+annotation('textbox', [0.78 0.97 0 0], 'string', 'Yin-Yang', 'FontSize', 36)
+draw3_cartesian_mesh(ax3, yinx, yiny, yinz, 'r');
+draw3_cartesian_mesh(ax3, yangx, yangy, yangz, 'b');
+
+end
+
+% Draw a Cartesian mesh in 3D using the given x,y,z coordinates of mesh
+% vertices.
+function draw3_cartesian_mesh(ax,x,y,z,colorspec)
+
[…]
